---
title: "Week 3, Lecture 2: Bestiary of Distributions"
subtitle: "Instructor notes -- Part 1: prep reading, Part 2: podium cheat sheet"
fontsize: 12pt
output:
  pdf_document: default
header-includes:
  - \usepackage{enumitem}
  - \setlist[itemize]{itemsep=4pt, parsep=0pt, topsep=4pt}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 5, fig.height = 3.5)
```

This single file has two parts, both instructor-only (not a student
handout). **Part 1** is read beforehand -- full worked examples and the
reasoning behind the flow. **Part 2** (after the page break, large
print) is the podium copy.

Source material: Bolker 4.4-4.6 (p. 155-188): analyzing probability
distributions (moments), the bestiary of distributions, and extending
simple distributions. Also mined: `LECTURE2.Rmd` (legacy version of
this lecture pair -- its Poisson/binomial/beta R demos are reused
below with fresh parameter values), and Monday's lecture notes.

# Where this sits in the week

- Monday: probability foundations, Bayes' rule.
- Tuesday (Lab 3): Act 1 practiced Bayes' rule; Acts 2-3 got hands-on
  with `dbinom()`/`dpois()`/`dnorm()`/`pnorm()`/`qnorm()` on a
  camera-trap monitoring scenario, *before* today's lecture formally
  names and organizes these distributions.
- **Today:** name what the lab already touched, add moments and two
  distributions the lab didn't cover (a skewed continuous one, and a
  quick look at what happens when simple distributions don't fit), and
  give students a framework for choosing a distribution from the shape
  of their own data.

# Rationale for the flow

Bolker's own advice for section 4.5: skim it as a map, not a
memorization list. Students already have `dbinom()` and `dpois()` in
their hands from yesterday's lab -- today isn't a first introduction,
it's organizing what they already did and adding vocabulary (moments,
support, the data-generating-process framing) plus 1-2 new shapes.
Keep the survey shallow and wide, with one deeper worked example
(moments) and one genuinely new distribution (a bounded continuous one
-- beta -- since nothing in the lab needed it).

# Part A: Moments (~8-10 min)

**Moments** describe the shape of a distribution. The $n$th moment is
$E[X^n]$; the $n$th *central* moment replaces $X$ with $X-\mu$.

- 0th central moment: always 1 (axiom 2, restated)
- 1st central moment: always 0 (that's what makes $\mu$ the mean)
- 2nd central moment: the **variance**
- 3rd, standardized: **skewness**; 4th, standardized: **kurtosis**

**Worked derivation** (Calc-II algebra only, no new calculus): show
that $\text{Var}(X) = E[X^2] - \mu^2$, a formula students will use
constantly but rarely see derived.

$$\text{Var}(X) = E[(X-\mu)^2] = E[X^2 - 2\mu X + \mu^2] = E[X^2] - 2\mu E[X] + \mu^2$$

Since $E[X]=\mu$:
$$= E[X^2] - 2\mu^2 + \mu^2 = E[X^2] - \mu^2$$

Worth pointing out explicitly: this is just expanding a square and
using linearity of expectation -- nothing here is harder than what
Week 1's algorithms lecture already assumed they could do.

**Parameters vs. moments:** parameters are the numbers that go into a
distribution's formula (e.g. $\lambda$ for Poisson); moments are
properties you can compute *from* the distribution once you know its
parameters. Sometimes they're the same thing (Poisson's $\lambda$ is
simultaneously its mean *and* its variance) and sometimes they're not
(the normal's $\sigma^2$ is its variance directly, but the gamma's
variance is a function of two separate parameters).

# Part B: Discrete distributions -- counts (~8 min)

Fresh example, not yesterday's camera-trap detections: seed counts per
quadrat. Suppose the average number of a given plant's seeds found per
$1\text{m}^2$ quadrat is $\lambda = 4$.

```{r poisson-example}
lambda <- 4
xvals <- 0:14
probs <- dpois(xvals, lambda)
barplot(probs, names.arg = xvals, ylab = "Probability",
        xlab = "Seeds per quadrat", main = "Poisson(4)")
sum(probs)     # doesn't quite hit 1 -- tail beyond x=14 is still out there
```

**Binomial recap (already met in Lab 3 Act 2):** counts of successes
out of a fixed number of independent trials, each with the same
success probability $p$. **Poisson recap (also met in lab):** counts of
independent events in a fixed window of time or space, with rate
$\lambda$ -- the limiting case of the binomial as $N\to\infty$,
$p\to0$, with $Np=\lambda$ held fixed (this is exactly what Act 2's
$N=100$, $p=0.024$ comparison demonstrated numerically).

**Sticking point to flag:** Poisson's single parameter $\lambda$ is
*simultaneously* the mean and the variance (mean = variance is called
**equidispersion**). Real count data is very often *more* variable than
this (**overdispersion**) -- worth a one-sentence preview that the
negative binomial exists specifically to relax this constraint (not
covered today).

# Part C: Continuous distributions -- measurements (~8-10 min)

**Normal recap (met in Lab 3 Act 3):** symmetric, unbounded, fully
determined by its first two moments ($\mu$, $\sigma^2$) -- central to
the whole course because of the central limit theorem.

**New today -- the beta distribution**, for data that's a *proportion*
or *probability* itself (bounded between 0 and 1). Example: fraction of
a quadrat's area covered by an invasive species.

```{r beta-example}
a <- 2; b <- 5   # shape parameters
curve(dbeta(x, a, b), from = 0, to = 1, lwd = 2,
      xlab = "Proportion cover", ylab = "Density", main = "Beta(2, 5)")
integrate(dbeta, lower = 0, upper = 1, shape1 = a, shape2 = b)   # check: integrates to 1
```

Point out: beta's two shape parameters ($\alpha$, $\beta$) don't map
onto mean/variance as directly as the normal's do, but the mean works
out to $\alpha/(\alpha+\beta)$ -- worth showing since it's exactly the
"fraction of total" intuition the shape parameters suggest. This is
also the natural distribution for *a probability itself* -- worth a
one-sentence flag forward to the Bayesian inference weeks, where a
prior on a probability parameter is very often a beta.

**Sticking point to flag (recap from Lab 3 Act 3, worth restating with
a new example):** the height of `dbeta()`'s curve is a *density*, not a
probability -- it can exceed 1 (try `dbeta(0.2, 2, 5)`), and only an
*area* under the curve is a probability.

# Part D: Choosing a distribution (~6-8 min)

The organizing question, straight from Bolker: **what process generated
this data?**, not "what shape does the histogram look like?" A table
worth putting on the board:

| Data looks like...                          | Consider...          |
|:---------------------------------------------|:----------------------|
| successes out of a fixed number of trials    | binomial               |
| counts of independent events, no fixed upper bound | Poisson (or negative binomial, if overdispersed) |
| a bounded proportion or probability          | beta                   |
| a continuous, roughly symmetric measurement  | normal                 |
| a continuous, right-skewed, positive measurement | gamma or lognormal (not covered in depth today) |

The two "not covered in depth today" mentions (negative binomial,
gamma/lognormal) are deliberate -- flag them as *named* so students
recognize them later in the semester, without deriving them now.

# Take-home message

- Moments describe shape; $\text{Var}(X)=E[X^2]-\mu^2$ is worth
  remembering, not just looking up.
- Distributions are chosen by the *mechanism* that generated the data,
  not by eyeballing a histogram shape.
- Density $\neq$ probability (only area under a density curve is a
  probability); Poisson's mean = variance is a real, checkable
  assumption, not just a mathematical curiosity.

# Roadmap tie-in

- The shapes surveyed today (Weeks 1-2's deterministic functions) will
  soon get noise wrapped around them -- Week 5 (synthetic data
  simulation) combines a deterministic mean function with one of
  today's distributions to build a full stochastic model.
- Overdispersion (Poisson's broken equidispersion assumption) comes
  back explicitly in Week 12 (modeling variance).

# Discussion seed (if time remains)

Ask: *looking back at yesterday's lab -- which distribution from
today's tour best describes the camera-trap detection counts (Act 2)?
The body mass measurements (Act 3)? Can you imagine a way reality might
violate the distribution you picked (e.g., detections clustering in
time rather than being independent)?*

\newpage
\large

# PART 2 -- PODIUM CHEAT SHEET

\normalsize
*Podium copy -- terse on purpose. Full derivations, rationale, and
code: see Part 1 above.*
\large

## 0. Open (~2 min)

- Yesterday's lab already used `dbinom()`, `dpois()`, `dnorm()` -- today
  organizes and names what you touched, adds moments + one new shape

## 1. Moments (~8-10 min)

- 0th central moment = 1, 1st = 0, 2nd = **variance**, 3rd = skewness,
  4th = kurtosis
- **Derive on board:** $\text{Var}(X)=E[X^2]-2\mu E[X]+\mu^2 = E[X^2]-\mu^2$
  (just expand the square + linearity of $E[\cdot]$)
- parameters (go into the formula) vs. moments (computed from it) --
  Poisson's $\lambda$ is *both* at once

## 2. Discrete: counts (~8 min)

```{r podium-pois, echo=FALSE, fig.width=3.3, fig.height=2.3}
par(mar = c(3.5, 3, 1, 0.5))
barplot(dpois(0:14, 4), names.arg = 0:14, cex.names = 0.6,
        ylab = "Prob.", xlab = "Seeds/quadrat")
```

- **Binomial** (Lab 3 Act 2): fixed $N$ trials, constant $p$
- **Poisson** (Lab 3 Act 2): rate $\lambda$, limit of binomial as
  $N\to\infty,p\to0,Np=\lambda$
- **Sticking point:** Poisson mean = variance (**equidispersion**) --
  real counts are often **overdispersed** (negative binomial handles
  this, not today)

## 3. Continuous: measurements (~8-10 min)

```{r podium-beta, echo=FALSE, fig.width=3.3, fig.height=2.3}
par(mar = c(3.5, 3, 1, 0.5))
curve(dbeta(x, 2, 5), from = 0, to = 1, lwd = 2, xlab = "Proportion", ylab = "Density")
```

- **Normal** (Lab 3 Act 3): symmetric, unbounded, $\mu,\sigma^2$ fully
  determine it (CLT)
- **New -- Beta**: bounded $[0,1]$, for proportions/probabilities;
  mean $=\alpha/(\alpha+\beta)$; natural prior for a probability
  parameter (preview: Bayesian weeks)
- **Sticking point (again):** density $\neq$ probability -- can exceed
  1, only *area* is a probability

## 4. Choosing a distribution (~6-8 min)

- Fixed trials, successes $\to$ **binomial**
- Counts, no fixed upper bound $\to$ **Poisson** (or neg. binomial if
  overdispersed)
- Bounded proportion $\to$ **beta**
- Symmetric continuous $\to$ **normal**
- Skewed positive continuous $\to$ gamma/lognormal (named only)
- **Match the mechanism, not the histogram shape**

## Discussion (if time)

- *Which distribution fits yesterday's detection counts (Act 2)? Body
  mass (Act 3)? How might reality violate that choice?*
