---
title: "Week 8, Lecture 1: Priors and Posteriors"
subtitle: "Instructor notes. Part 1: prep reading. Part 2: podium cheat sheet"
fontsize: 12pt
output:
  pdf_document: default
header-includes:
  - \usepackage{enumitem}
  - \setlist[itemize]{itemsep=4pt, parsep=0pt, topsep=4pt}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 5, fig.height = 3.3)
```

# How to prepare (for students)

- **Big picture.** On Wednesday we built Bayes' rule for parameter
  values and worked out why a Beta prior and a binomial likelihood give
  a Beta posterior. Today we use that result on a real example with two
  different priors, learn how to summarize a posterior, and talk about
  common misunderstandings and why Bayesian methods are so widely used.
  You'll also get a short take-home WOD on Newton's method, due
  Wednesday, to get ready for Wednesday's lecture on optimization.
- **Read:** Bolker 6.2.2 (printed p. 237-243) and 6.4.2 (p. 258-260),
  if you haven't yet. For the take-home and Wednesday: Bolker 7.1-7.2
  (p. 294-310).
- **Review:** your notes on Beta-binomial conjugacy from Wednesday, the
  Beta distribution (Bolker 4.5.2.5, p. 176), and the log-odds,
  $\log(p/(1-p))$, from Wednesday's logistic regression opening.

# Overview (instructor only)

This file has two parts, both instructor-only. **Part 1** is read
beforehand. **Part 2** (after the page break, large print) is the
podium copy.

| Beat | Time |
|---|---|
| WOD: "How much did the prior matter?" | ~8 min |
| Lecture | ~32 min |
| Discussion | ~8 min |

**Where this came from.** Sections 2-5 and the discussion were planned
for Week 7 Wednesday, which instead covered logistic regression, Bayes'
rule for parameters, and the conjugacy math (see
`NOTES_Week7_Lecture2_BayesianInference.Rmd`). Start with the
germination example.

**Pacing.** Both Week 7 lectures covered about half of what was
planned. Core: sections 1-2, confusions (a), (e) from section 4 (with
a quick callback for (b)), two reasons from section 5, and section 6.
Section 3 and the rest of 4-5 are "if time." The optimization preview
planned earlier is gone: Wednesday covers it, and the take-home WOD
(below) does the setup.

**WODs.**

- **In class:** `wod/WOD_Week8_Lecture1_HowMuchDidThePriorMatter.Rmd`
  (key in `wod/answer_keys/`). Three tortoise-survival panels, prior
  and posterior only; students rank prior influence, sketch the
  missing likelihood, and decide which study had the most data. The
  trap is Team B: narrowest posterior, least data (it's all prior).
  Debrief straight into section 2: the germination supplier prior is
  worth 200 seeds against 40 of data, between Teams B and C.
- **Take-home, due Wednesday:** `wod/WOD_Week8_Lecture1_NewtonByHand.Rmd`
  (key in `wod/answer_keys/`). Pika NLL derivatives, finite differences,
  and Newton steps by hand, ending on an overshoot from $\lambda = 7$.
  Hand it out at the end of class with a one-line pitch: "Wednesday we
  open up `optim()`; this is the first piece."
- `wod/WOD_Week7_Lecture2_CurvatureByHand.Rmd` stays shelved; the
  take-home absorbs its pika part.

Source material: Bolker 4.3.3 (priors, and why flat priors aren't
neutral), 6.2.2.1 (binomial/Beta conjugate example; the numbers below
are his, recast as a germination trial), 6.4.2 (credible intervals);
legacy `LECTURE6.Rmd` ("Bayesian Analysis #1: Concepts": mean vs.
mode, credible intervals).

**Example choice.** The conjugate example is a seed-germination trial
rather than Bolker's tadpoles. The numbers (30 of 40, and a prior worth
120 of 200) are Bolker's from p. 237-238, so students who did the
reading can check them, but the setting is one the agricultural and
restoration students will recognize.

# Where this sits in the week

- **Last week:** likelihood intervals (Monday), the quadratic
  approximation and profiles (Lab 6), then logistic regression and the
  start of Bayesian inference (Wednesday).
- **Today:** finish the one-parameter Bayesian material. Hand out the
  Newton's method take-home.
- **Tuesday:** group meetings on final project proposals (no lab).
- **Wednesday:** optimization (Bolker 7.1-7.2), opening with the
  take-home.
- **Coming up:** MCMC and Lab 7 on Bayesian inference (Week 9).

# Part 1: Lecture content

## 1. Big picture and recap (~3 min)

Loop on the board:

$$\text{THEORY} \to \text{MODEL(S)} \to \text{DATA (fit)} \to \text{COMPARE} \to \text{repeat}$$

- Same model, same likelihood; Bayesian inference describes our
  uncertainty about a parameter with a probability distribution.
- Recap: posterior $\propto$ likelihood $\times$ prior. With a Beta
  prior and binomial data, the posterior is Beta($\alpha + k$,
  $\beta + n - k$).
- Today's question: how much does the prior matter, and what do we
  report?

## 2. A worked example: germination with two priors (~10 min)

**Bridge from the WOD.** The tortoise panels showed priors of
different strength by eye. Here we put numbers on it: how much is a
prior "worth," in data?

**Setting.** A restoration crew tests a native seed lot: 30 of 40 seeds
germinate. What is the lot's germination probability $p$?

**The likelihood** is binomial: $\mathcal{L}(p) \propto p^{30}(1-p)^{10}$.

**Recap from Wednesday.** With a Beta($\alpha$, $\beta$) prior, the
posterior is Beta($\alpha + k$, $\beta + n - k$): just add counts.

- Interpretation (Bolker p. 237): $\alpha - 1$ and $\beta - 1$ act
  like earlier successes and failures. Beta(1, 1) is flat: no prior
  seeds.
- **Flat prior:** posterior Beta(31, 11).
- **Informative prior:** the supplier's records say 120 of 200 seeds
  from earlier lots germinated. Prior Beta(121, 81); posterior
  Beta(151, 91).

```{r germ-setup}
k <- 30; n <- 40
post_flat <- c(1 + k, 1 + n - k)
post_info <- c(121 + k, 81 + n - k)
beta_mean <- function(ab) ab[1] / sum(ab)
beta_mode <- function(ab) (ab[1] - 1) / (sum(ab) - 2)
summ <- rbind(
  flat = c(mean = beta_mean(post_flat), mode = beta_mode(post_flat),
           qbeta(c(0.025, 0.975), post_flat[1], post_flat[2])),
  informative = c(mean = beta_mean(post_info), mode = beta_mode(post_info),
                  qbeta(c(0.025, 0.975), post_info[1], post_info[2]))
)
colnames(summ)[3:4] <- c("lower95", "upper95")
round(summ, 3)
f2 <- function(v) formatC(v, format = "f", digits = 2)
```

```{r germ-fig, fig.width=6.5, fig.height=2.8}
p_seq <- seq(0.001, 0.999, length.out = 500)
lik_scaled <- dbeta(p_seq, k + 1, n - k + 1)   # likelihood rescaled to area 1
plot_panel <- function(prior_ab, post_ab, title) {
  plot(p_seq, dbeta(p_seq, post_ab[1], post_ab[2]), type = "l", lwd = 2,
       ylim = c(0, 14), xlab = "Germination probability, p", ylab = "Density",
       main = title, cex.main = 0.9)
  lines(p_seq, dbeta(p_seq, prior_ab[1], prior_ab[2]), lwd = 2, col = "grey60")
  lines(p_seq, lik_scaled, lwd = 2, lty = 2, col = "darkorange")
}
par(mfrow = c(1, 2), mar = c(4, 4, 1.5, 0.5))
plot_panel(c(1, 1), post_flat, "Flat prior, Beta(1, 1)")
legend("topleft", c("prior", "likelihood (rescaled)", "posterior"),
       lwd = 2, lty = c(1, 2, 1), col = c("grey60", "darkorange", "black"),
       bty = "n", cex = 0.7)
plot_panel(c(121, 81), post_info, "Supplier prior, Beta(121, 81)")
```

- **Flat prior:** the posterior *is* the rescaled likelihood. Its mode
  is 0.75, the MLE; its mean is `r round(beta_mean(post_flat), 3)`,
  slightly pulled toward 0.5.
- **Supplier prior:** the posterior sits between prior and data, closer
  to the prior (mean `r round(beta_mean(post_info), 3)`), because the
  prior is "worth" 200 seeds and the data only 40.
- **More data wins.** If the crew had tested 400 seeds with 300
  germinating, the same supplier prior gives a posterior mean of
  `r round(beta_mean(c(121 + 300, 81 + 100)), 3)`: the data now
  outweigh the prior two to one.
- **Updating in sequence.** Today's posterior is tomorrow's prior.
  Testing 40 seeds and then 40 more gives the same answer as testing 80
  at once (Bolker p. 237).

## 3. Summarizing a posterior (~4 min)

- **Point estimates.** Posterior mean (most common), median, or mode.
  With a flat prior, the mode equals the MLE. For skewed posteriors the
  three differ; Bolker prefers the mean because it uses the whole
  distribution.
- **Credible interval.** The central 95% of the posterior (2.5% and
  97.5% quantiles), or the highest posterior density (HPD) interval,
  the shortest interval holding 95% (Bolker p. 258-259). For
  symmetric posteriors they coincide.
- **Interpretation.** "Given the model, the prior, and the data, there
  is a 95% probability that $p$ is between
  `r f2(summ["flat", "lower95"])` and
  `r f2(summ["flat", "upper95"])`." That is the sentence most
  people *want* to say about a confidence interval (Week 7 Monday's
  discussion question 1), and a Bayesian is allowed to say it.

## 4. Where new Bayesians get stuck (~8 min)

Core: (a), (e), and a 30-second callback for (b), which was covered
well in Week 7 Wednesday's lecture. The rest if time; each takes a
minute or two.

**(a) "The likelihood is the posterior."** Only after multiplying by a
prior and rescaling to area 1. The likelihood alone doesn't integrate
to 1 (Week 5 WOD) and isn't a distribution for $p$.

**(b) "A flat prior means no information."** (Callback only; covered
Wed., Oct. 7.) Flat on one scale isn't flat on another (Bolker 4.3.3,
the bird-mass example). A uniform prior on $p$ implies a prior on the
log-odds, $\log\frac{p}{1-p}$, that is peaked at 0:

```{r flat-not-flat, fig.width=6.5, fig.height=2.4}
set.seed(746)
p_draws <- runif(1e5)
par(mfrow = c(1, 2), mar = c(4, 4, 1.5, 0.5))
hist(p_draws, breaks = 40, freq = FALSE, col = "grey80", border = "white",
     main = "Flat prior on p", xlab = "p", cex.main = 0.9)
hist(qlogis(p_draws), breaks = 60, freq = FALSE, col = "grey80", border = "white",
     main = "Same prior, log-odds scale", xlab = "log(p / (1 - p))",
     xlim = c(-8, 8), cex.main = 0.9)
curve(dlogis(x), add = TRUE, lwd = 2)
```

So "uninformative" always means "uninformative on some scale." With
plenty of data it rarely matters; with little data it can.

**(c) "Priors let you get any answer you want."** In principle yes
(Bolker p. 150: a prior of 0 on a hypothesis means no data can ever
support it). In practice: priors must be reported, reviewers can
object, and the standard move is a **prior sensitivity analysis**: rerun
with a few reasonable priors and see whether the conclusion changes.
If it does, the data are weak and that's worth knowing.

**(d) "Bayesian means MCMC."** MCMC is a computing method for
approximating posteriors when the denominator can't be computed. The
germination example needed none. Conversely, MCMC can be used for
non-Bayesian problems too.

**(e) "The credible interval and the confidence interval are the same
thing."** With a flat prior and plenty of data the numbers are often
close (Week 7 pikas: likelihood ratio [1.82, 4.61]; with a flat prior
the posterior is Gamma(19, 6), credible interval [1.91, 4.74]), but the
statements differ: a credible interval is a
probability statement about the parameter; a confidence interval is a
statement about the procedure.

**(f) "A high posterior probability proves the model."** Every
posterior is conditional on the model being right. A Bayesian model
still has to be confronted with data (does it predict data that look
like what we saw?) and compared with rivals. The loop on the board
doesn't change.

## 5. Why Bayesian methods are so widely used (~5 min)

**Interpretation.**

- Direct probability statements about parameters and hypotheses: "95%
  probability the germination rate is between
  `r f2(summ["flat", "lower95"])` and `r f2(summ["flat", "upper95"])`."
- Natural accumulation of evidence: last year's posterior is this
  year's prior.

**Practical.**

- **Uncertainty in derived quantities is easy to get.** Callback to
  Lab 6: for the LD50, a Bayesian draws many $(a, b)$ pairs from the
  posterior and computes $-a/b$ for each; the spread of those values is
  the posterior for the LD50. No profile, no reparameterization.
- **Hierarchical models.** Plots within sites within regions, animals
  measured repeatedly, many species sharing information: these models
  have many parameters, some describing variation among groups. They
  are awkward to fit by maximum likelihood and natural in a Bayesian
  framework (later in the semester).
- **Hidden quantities.** True occupancy when detection is imperfect
  (the frog from Lab 5), true infection status behind a noisy test, the
  true population size behind a count. These are just more unknowns in
  a Bayesian model.
- **Small samples.** No reliance on large-sample approximations like
  the Hessian interval, which goes negative for the Week 7 chytrid data
  (2 of 20 positive).
- **Prior information.** Previous studies, expert knowledge, or
  physical limits can be built in openly.

**Costs.** Priors must be chosen and defended; computation
can be slow; MCMC output has to be checked for convergence.

## 6. Return to the big picture (~1 min)

- Same model, same likelihood. Bayesian inference multiplies by a prior
  and reads uncertainty off a posterior distribution.
- With one parameter and a conjugate prior, it's just adding counts.
  With many parameters, we need MCMC (Week 9).
- Hand out the Newton's method take-home (due Wednesday). Every MLE so
  far came from calculus, a grid, or `optim()` as a black box;
  Wednesday opens the box.

# Discussion (~8 min)

Pick one or two. (The earlier "flat on $p$ vs. flat on the log-odds"
question is dropped; Week 7 Wednesday covered it.)

1. With the supplier prior, the posterior mean germination rate is
   `r f2(beta_mean(post_info))`; with a flat prior it's
   `r f2(beta_mean(post_flat))`. A restoration manager says, "Keep the
   supplier's numbers out of my estimate." Is that reasonable? What
   would you report to them?
2. In Lab 6 you built a profile interval for the LD50. How would a
   Bayesian get an interval for the LD50, and which approach would you
   rather use for a model with ten parameters?
3. The Week 7 likelihood ratio interval (95%) for the pikas was about
   [1.8, 4.6]. A land manager asks, "So there's a 95% chance the true
   density is between 1.8 and 4.6?" Can you say yes? What could a
   Bayesian say instead?

Question 1 surfaces the "priors are cheating" worry. The manager has a
point (the prior is worth 200 seeds to the data's 40, and it comes from
other lots and an interested party), but the earlier lots aren't
worthless either. Where to land: report both posteriors (a prior
sensitivity analysis), perhaps with a down-weighted prior in between
(the supplier's 60% scaled down to 12 of 20 seeds, Beta(13, 9)). Whether past
lots are relevant to this one is an ecological question, not a
statistical one. Question 2 connects today to Lab 6 and sets up MCMC.
Question 3 lands on the credible vs. confidence distinction from 4(e).

\newpage
\large

# PART 2: PODIUM CHEAT SHEET

\normalsize
\textit{Terse on purpose. Full rationale and code: Part 1.}
\large

**WOD ~8 min. Lecture ~32 min. Discussion ~8 min.**

## WOD debrief: tortoise panels

- Prior influence: B > C > A
- C's likelihood peaks ~0.5, opposite side of posterior from prior
- Most data: A (60); fewest: B (5). **Narrow posterior $\neq$ lots of data**
- Bridge: supplier prior worth 200 seeds vs. 40 of data

## 1. Big picture and recap (~3 min)

- Loop on board: THEORY $\to$ MODEL $\to$ DATA (fit) $\to$ COMPARE $\to$ repeat
- Posterior $\propto$ likelihood $\times$ prior
- Beta($\alpha$, $\beta$) + $k$ of $n$ $\to$ Beta($\alpha + k$, $\beta + n - k$)
- Today: how much does the prior matter?

## 2. Germination, two priors (~10 min)

Seeds: **30 of 40** germinate. Supplier: 120 of 200.

- Beta prior $\times$ binomial $\to$ Beta($\alpha + k$, $\beta + n - k$)
- Conjugate = add counts; $\alpha-1$, $\beta-1$ = prior successes, failures
- Flat: Beta(31, 11); mean `r round(beta_mean(post_flat), 3)`, mode 0.75 = MLE
- Supplier: Beta(151, 91); mean `r round(beta_mean(post_info), 3)`
- 300 of 400 with supplier prior: mean `r round(beta_mean(c(421, 181)), 3)`
- Sequential updating = same answer

```{r podium-germ, echo=FALSE, fig.width=6, fig.height=2.5}
par(mfrow = c(1, 2), mar = c(4, 4, 1.5, 0.5))
plot_panel(c(1, 1), post_flat, "Flat")
plot_panel(c(121, 81), post_info, "Supplier")
```

\newpage

## 3. Summaries (~4 min)

- Mean (usual), median, mode (= MLE if flat)
- Credible interval: quantiles or HPD
- Flat: [`r f2(summ["flat", "lower95"])`, `r f2(summ["flat", "upper95"])`]; supplier: [`r f2(summ["informative", "lower95"])`, `r f2(summ["informative", "upper95"])`]
- "95% probability $p$ is in here" is allowed

## 4. Where people get stuck (~8 min)

- (a) Likelihood $\neq$ posterior (Week 5 WOD)
- (b) Flat isn't flat on every scale (callback only; covered Oct. 7)
- (c) Priors: report them; sensitivity analysis
- (d) Bayes $\neq$ MCMC
- (e) Credible $\neq$ confidence (pikas: [1.82, 4.61] vs. [1.91, 4.74])
- (f) Posterior is conditional on the model; still check and compare

\newpage

## 5. Why so popular (~5 min)

- Direct probability statements
- Evidence accumulates: posterior $\to$ next prior
- Derived quantities: LD50 = $-a/b$ for each posterior draw (Lab 6)
- Hierarchical models (later)
- Hidden quantities: occupancy, true infection status
- Small samples: no Hessian approximation (chytrid)
- Costs: priors, computation, convergence

## 6. Wrap (~1 min)

- Conjugate = add counts; many parameters = MCMC (Week 9)
- **Hand out Newton take-home (due Wed.)**

## Discussion (~8 min)

1. Supplier prior: mean `r f2(beta_mean(post_info))`; flat: `r f2(beta_mean(post_flat))`. Manager: "keep the supplier's numbers out." Reasonable? What to report?
2. Bayesian interval for the LD50? Which approach for 10 parameters?
3. Week 7 pikas [1.8, 4.6]: "95% chance"? Can you say yes?
