---
title: "Lab 3: Probability and Distributions"
subtitle: "Instructor notes -- Part 1: prep reading, Part 2: podium/circulating cheat sheet"
fontsize: 12pt
output:
  pdf_document: default
header-includes:
  - \usepackage{enumitem}
  - \setlist[itemize]{itemsep=4pt, parsep=0pt, topsep=4pt}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 5, fig.height = 3.5)
```

This single file has two parts, both instructor-only (not a student
handout). **Part 1** is read beforehand -- an answer key and the
reasoning behind each Act. **Part 2** (after the page break, large
print) is the copy to have open while circulating during lab.

**Schedule note:** Labor Day cancels this week's Monday meeting, so
unlike a normal week, this lab now runs with **no prior lecture at
all** -- Wednesday's probability/Bayes' rule lecture comes *after* this
lab, not before, and the bestiary of distributions has moved to next
Monday (Week 4). That puts this lab in a similar position to
`NOTES_Lab1_MathRefresher.Rmd` (content ahead of any formal
instruction), except here it's a new concept (Bayes' rule) rather than
rusty-but-familiar calculus, so groups may need more up-front framing
than usual -- consider a short (5 min) live walkthrough of conditional
probability notation before releasing Act 1, since the Act itself
doesn't re-derive Bayes' rule from scratch. This doc exists mainly as a
quick-reference answer key for spot-checking groups, kept brief.

# Act 1: camera-trap classifier PPV

Same false-positive/PPV technique as the medical-testing example
Wednesday's lecture will formalize (Bolker 4.3.1), different numbers
($P(\text{species})=0.0001$, sensitivity $=0.98$, false positive rate
$=0.01$) -- since the lecture hasn't happened yet, Act 1 is students'
first encounter with this idea, not a repeat.

```{r act1-key}
p_S <- 1e-4; sens <- 0.98; fpr <- 0.01
p_notS <- 1 - p_S
p_F_and_S    <- sens * p_S
p_F_and_notS <- fpr * p_notS
p_F  <- p_F_and_S + p_F_and_notS
PPV  <- p_F_and_S / p_F
c(P_F_and_S = p_F_and_S, P_F_and_notS = p_F_and_notS, P_F = p_F, PPV = PPV)
```

**Answer key:** $P(F\cap S)=0.000098$, $P(F\cap\lnot S)=0.009999$,
$P(F)\approx0.0101$, $\text{PPV}\approx0.0097$ ($\approx1\%$). The
"negligible term" approximation ($P(F)\approx0.01$, dropping the
$S$-contribution) gives essentially the same answer -- a group that
uses the shortcut isn't wrong.

**Natural-frequency table (N = 100,000), rounded:**

| | Flagged | Not flagged | Total |
|:--|--:|--:|--:|
| Actually the species | $\approx10$ | $\approx0$ | 10 |
| Not the species | $\approx1{,}000$ | $\approx98{,}990$ | 99,990 |
| Total | $\approx1{,}010$ | $\approx98{,}990$ | 100,000 |

$\text{PPV} = 10/1010 \approx 0.0099$ -- matches Part 1.

**Watch for:** groups who compute $P(S\mid F)$ by dividing by $P(S)$
instead of $P(F)$ (the classic $P(A\mid B)$ vs. $P(B\mid A)$ mix-up --
this is exactly the base rate fallacy Wednesday's lecture will name
explicitly). If a group's PPV comes out $>0.5$, that's almost always
the tell.

# Act 2: binomial detections, and when Poisson is a good stand-in

$N=8$, $p=0.3$, $k=3$.

```{r act2-key}
N <- 8; p <- 0.3; k <- 3
choose(N, k) * p^k * (1 - p)^(N - k)
dbinom(k, N, p)
```

**Answer key:** $P(X=3) = 0.2541$. Full distribution mode is at $k=2$
(check with `which.max(dbinom(0:8,8,0.3))`).

**Poisson comparison, part A** ($N=8,p=0.3$ vs. $\lambda=2.4$):
binomial variance $Np(1-p)=1.68$; Poisson variance $\lambda=2.4$ --
mismatched, and the two barplots should look visibly different (Poisson
has a fatter right tail).

**Poisson comparison, part B** ($N=100,p=0.024$ vs. $\lambda=2.4$):
binomial variance $=100(0.024)(0.976)=2.3424$ -- close to $2.4$, and
the two barplots should look nearly identical. Rule of thumb (2f):
*Poisson is a good stand-in for binomial when $N$ is large and $p$ is
small* (so $(1-p)\approx1$ and $Np(1-p)\approx Np=\lambda$).

**Watch for:** groups who conclude Poisson is "just always fine" from
part B alone without noticing part A disagreed -- push them back to the
variance mismatch in part A as the reason it isn't universal.

# Act 3: continuous body mass ($\mu=4.5$, $\sigma=0.6$)

```{r act3-key}
mu <- 4.5; sigma <- 0.6
1 - pnorm(5.5, mu, sigma)
qnorm(0.9, mu, sigma)
```

**Answer key:** $P(\text{mass}>5.5) \approx 0.048$ ($\approx5\%$); the
heaviest-10% cutoff is $\approx5.27$ kg.

**Watch for:** 1b/1c (density vs. probability) is the conceptual crux
of this Act -- a group that answers "$P(\text{mass}=4.5)$ is very small
but not zero" hasn't fully separated density from probability yet.
Point them back at the shaded-area picture from 1a: a single point has
zero *width*, hence zero area, hence zero probability, regardless of
how tall the curve is there.

\newpage
\large

# PART 2 -- CHEAT SHEET (circulating)

\normalsize
*Copy for circulating during lab. Full reasoning and code: see Part 1
above.*
\large

## Act 1 -- classifier PPV

- $P(F\cap S)=0.000098$, $P(F\cap\lnot S)=0.009999$, $P(F)\approx0.0101$
- $\text{PPV}\approx0.0097$ ($\approx1\%$)
- Freq. table check (N=100k): $10/1010\approx1\%$ -- matches
- **Watch for:** PPV $>0.5$ usually means $P(A|B)$/$P(B|A)$ flipped

## Act 2 -- binomial vs. Poisson

- $N=8,p=0.3,k=3$: hand calc and `dbinom(3,8,0.3)` both $=0.2541$
- Mode of full distribution: $k=2$
- Var mismatch ($N=8,p=0.3$): binom $1.68$ vs. Pois $2.4$ -- **don't
  match**
- Var match ($N=100,p=0.024$): binom $2.3424$ vs. Pois $2.4$ -- **do
  match**
- Rule of thumb: Poisson $\approx$ binomial when $N$ large, $p$ small

## Act 3 -- continuous mass

- $P(\text{mass}>5.5)\approx0.048$; top-10% cutoff $\approx5.27$ kg
- **Watch for:** "small but nonzero" answers to $P(X=4.5)$ -- push back
  to the shaded-area picture (zero width $\to$ zero probability)
