---
title: "Week 2, Lecture 2: Bestiary of Functions"
subtitle: "Instructor notes -- Part 1: prep reading, Part 2: podium cheat sheet"
fontsize: 12pt
output:
  pdf_document: default
header-includes:
  - \usepackage{enumitem}
  - \setlist[itemize]{itemsep=4pt, parsep=0pt, topsep=4pt}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 5, fig.height = 3.5)
```

This single file has two parts, both instructor-only (not a student
handout). **Part 1** is read beforehand -- full worked examples and the
reasoning behind the flow. **Part 2** (after the page break, large
print) is the podium copy.

Source material: Bolker 3.5 ("Bestiary of functions"), leaning heavily
on Table 3.1 (p. 119) and section 3.5.2 (exponential-based functions,
p. 127-130) for the logistic. Also mined: this morning's WOD
(`wod/WOD_Week2_Lecture2_BestiaryLimits.Rmd`), Monday's lecture notes,
and Lab 2 Acts 1-2 (flexible Ricker, Holling IV).

# Where this sits in the week

- Monday: the toolkit -- limits (throw out lower-order terms),
  derivatives, critical points -- worked through Michaelis-Menten and
  Ricker.
- Monday's WOD: R syntax + derived/coded a seasonal sine function whose
  steepest point (day 91) was never named.
- Tuesday (Lab 2): Act 1 (grid search), Act 2 (flexible Ricker, Holling
  IV -- another rational function, one degree harder than
  Michaelis-Menten).
- Wednesday AM (WOD4, already done): limits practice on four functions
  (two familiar -- Michaelis-Menten, Ricker -- and two new -- logistic,
  Holling III), then a derivative of the logistic and a leading question
  about what kind of point $x=0$ is.
- **Wednesday lecture (this one):** reveal WOD4, name the new concept
  it was building toward (inflection points), and give the map of the
  bestiary that today's and Monday's examples are drawn from.

# Rationale for the flow

WOD4's Part 2 is a setup, not a self-contained exercise -- it walks
students to the edge of "the logistic's midpoint has a nonzero
derivative, so it's not a critical point... so what is it?" and leaves
the actual concept (inflection point, concavity, second derivative)
unnamed, with only a hint. That's today's centerpiece: give the reveal,
not a whole new derivation from scratch. The payoff is bigger than just
the logistic -- Monday's WOD sine function has the *exact same*
feature at day 91 (maximum slope, not zero slope), and nobody named it
then either. Connecting those two moments is worth building the lecture
around.

Bolker himself says to treat 3.5 as a map, not a memorization list:
*"You should skim through this section on the first reading to get an
idea of what functions are available... use the section for reference
as needed."* So the bestiary-survey portion of this lecture is
deliberately shallow and wide -- a table, not a derivation marathon --
with one deeper worked example (Holling III's limits, which
double-checks WOD4 and sets up a "staircase" of difficulty: MM's
numerator and denominator have different degrees; Holling III's have
the *same* degree; Lab 2's Holling IV has the same degree plus an extra
middle term. Today is the middle rung.

# Part A: reveal WOD4's limits (quick -- ~5 min)

**1a-1b (Michaelis-Menten, Ricker):** already fully worked Monday --
just flash the answers, don't re-derive.

- MM: $x\to0$ gives a line through the origin, slope $a/b$; $x\to\infty$
  gives the asymptote $a$.
- Ricker: $x\to0$ gives $0$ (both factors vanish); $x\to\infty$ gives
  $0$ (exponential beats the linear $x$ term).

**1c (Logistic), $f(x)=\dfrac{1}{1+e^{-bx}}$:** as $x\to\infty$,
$e^{-bx}\to0$, so $f(x)\to 1/(1+0)=1$. As $x\to-\infty$ (or, for the
WOD's $x\to0$ framing, just evaluate directly), $e^{-bx}\to\infty$ so
$f(x)\to 1/\infty = 0$. The logistic is bounded between 0 and 1 -- no
new dominance trick needed, just careful bookkeeping of what $e^{-bx}$
does at each end.

**1d (Holling III), $f(x)=\dfrac{ax^2}{b^2+x^2}$ -- the one worth
doing on the board:**

- $x\to0$: numerator $ax^2\to0$ (it's the higher-order term here,
  since there's no linear term to compete with); denominator
  $b^2+x^2\approx b^2$ (constant wins). So $f(x)\approx ax^2/b^2\to0$,
  and *quadratically* -- the curve leaves the origin flat (slope zero)
  and curves upward, unlike Michaelis-Menten's straight-line start.
- $x\to\infty$: **numerator and denominator are now the *same* degree**
  ($x^2$ on both sides) -- new wrinkle beyond Monday. Throw out the
  lower-order term in the denominator ($b^2 \ll x^2$), leaving
  $f(x)\approx ax^2/x^2 = a$. Same asymptote as Michaelis-Menten, same
  logic, just both leading terms survive and cancel instead of one side
  disappearing entirely.

**Sticking point to flag:** when numerator and denominator are the same
degree, students sometimes think the dominance trick "doesn't apply"
because nothing fully vanishes. It still applies -- you throw out the
lower-order term *within* each of the numerator and denominator
separately, then simplify what's left. This is exactly the skill Lab 2
Act 2 needs for Holling IV (same idea, with an extra $cx$ term to also
discard in the denominator).

# Part B: critical points vs. inflection points (the core new idea, ~12-14 min)

**Recap the vocabulary from Monday first:** a critical point is where
$f'(x)=0$ -- a candidate peak or trough. Ricker has one; Michaelis-Menten,
being monotonic, has none.

**Now work the logistic's derivative on the board** (this is WOD4
2a-2b, revealed): $f(x) = (1+e^{-bx})^{-1}$. Chain rule, twice:
$$f'(x) = -(1+e^{-bx})^{-2}\cdot(-b\,e^{-bx}) = \frac{b\,e^{-bx}}{(1+e^{-bx})^2}$$

```{r logistic-fprime}
b <- 1
f  <- function(x) 1 / (1 + exp(-b * x))
fp <- function(x) b * exp(-b * x) / (1 + exp(-b * x))^2
fp(0)
```

At $x=0$: $e^{0}=1$, so $f'(0) = b\cdot1/(1+1)^2 = b/4$ -- **not zero**
(for any $b\neq0$). This matches Bolker's own Figure 3.9, which labels
exactly this point "$b/4$" on the logistic panel.

**So $x=0$ is not a critical point -- what is it?** It's the point of
*steepest* slope on the whole curve: the logistic climbs slowly, then
fastest right around $x=0$, then slowly again as it flattens toward the
asymptote at 1. A point where the slope is at a local maximum (rather
than zero) is an **inflection point**: formally, where the *second*
derivative $f''(x)=0$ and changes sign -- where the curve's concavity
flips from curving upward (concave up) to curving downward (concave
down). Don't grind through $f''(x)$ symbolically today (that's more
than WOD4 asked for); the geometric argument is enough: just before
$x=0$ the curve bends upward (accelerating), just after it bends
downward (decelerating) -- concavity changes, so it's an inflection
point, not a critical point.

```{r logistic-plot, echo=FALSE, fig.height=3}
curve(f(x), from = -6, to = 6, lwd = 2, col = "steelblue",
      xlab = "x", ylab = "f(x)")
abline(h = c(0, 1), lty = 3, col = "grey70")
points(0, 0.5, pch = 19)
curve(0.5 + fp(0) * x, from = -1.5, to = 1.5, add = TRUE, col = "darkorange", lwd = 2)
text(2.5, 0.35, "slope = b/4\n(steepest point)", col = "darkorange", cex = 0.8)
```

**The payoff -- callback to Monday's WOD:** the seasonal sine function
students derived Monday (period 1 year, steepest at day 91) has exactly
this feature. Day 91 wasn't a critical point either -- the function
wasn't at a max or min there, it was climbing at its *fastest* rate.
Same concept, unnamed until right now. Worth saying explicitly: *every
sigmoid has exactly one inflection point (its steepest point); every
periodic function like the sine has one every half-cycle.* Two
completely different-looking functions, same calculus feature.

**Sticking point to flag (restated from Monday, now with a second
example):** critical point ($f'=0$, zero slope) vs. inflection point
(steepest slope, $f''=0$ with a sign change) are easy to blur together
because both involve "a special point found using a derivative." Keep
the plain-English distinction front and center: critical point =
*flattest* nearby; inflection point = *steepest* nearby (for a sigmoid;
more generally, just "concavity changes here").

# Part C: the bestiary map (~8-10 min)

Bolker's Table 3.1 organizes the whole bestiary into four families.
Don't derive every row -- this is a map, students already own several
of these landmarks:

- **Polynomial / piecewise** -- line, quadratic, cubic; threshold,
  hockey stick. *Not yet met.*
- **Rational (ratios of polynomials)** -- hyperbolic, Michaelis-Menten,
  Holling III, Holling IV. *Already met: MM (Monday), Holling IV (Lab
  2), Holling III (this morning's WOD).*
- **Exponential-based** -- negative exponential, monomolecular,
  **Ricker**, **logistic**. *Already met: Ricker (Monday, Lab 1-2),
  logistic (this morning's WOD).*
- **Power-based** -- power law, von Bertalanffy, Gompertz, Shepherd.
  *Not yet met.*

Point out explicitly: **you have already personally met roughly half
this table** without necessarily realizing it was "the bestiary" --
today just gives it a name and a map.

**One mechanism worth telling, since it's the most famous equation in
ecology:** the logistic function is the *solution* to the logistic
growth equation,
$$\frac{dn}{dt} = rn\left(1-\frac{n}{K}\right)$$
-- per-capita growth rate declines linearly from $r$ (when $n\ll K$) to
zero (when $n=K$, the carrying capacity). The equation is separable;
the resulting integral $\int \frac{dn}{n(1-n/K)}$ is solved via
*partial fraction decomposition* ($\frac{K}{n(K-n)} = \frac{1}{n} +
\frac{1}{K-n}$), not shown today, and produces exactly the sigmoid
curve just derived. This is a nice moment to connect the
purely-curve-fitting view of "logistic = an S-shaped function that
happens to fit data" with the mechanistic view every student already
has from intro ecology: population growth that decelerates as it
approaches a ceiling.

**Briefly mention, don't derive:** the Gompertz function is an
alternative sigmoid whose inflection point sits at $1/e\approx1/3$ of
the way up rather than halfway -- worth one sentence as a reminder that
"sigmoid" is a shape category with more than one member, the same way
"hump-shaped" wasn't just Ricker (flexible Ricker, Holling IV).

# Take-home message

- Two new limit moves today: same-degree rational functions (Holling
  III/IV) still use dominance, just applied separately to numerator and
  denominator before taking the ratio.
- Inflection point $\neq$ critical point: steepest slope, not zero
  slope; concavity change, not a peak/trough.
- The logistic's midpoint and the sine's day-91 point (Monday's WOD)
  are the same calculus feature wearing two different costumes.
- The bestiary is a map, not a memorization list -- four families
  (polynomial, rational, exponential-based, power-based), and students
  already own examples from three of the four.

# Roadmap tie-in

- Week 3: probability and distributions -- deterministic functions
  (this week) become the *mean* or *expected* part of a full stochastic
  model; today's shapes (saturating, hump-shaped, sigmoid) are about to
  get noise wrapped around them.
- Later: model selection weeks will ask "does the data support a
  Holling III (sigmoid) vs. a Michaelis-Menten (no inflection) shape?"
  -- that question only makes sense because of the vocabulary built
  today.

# Discussion seed (if time remains)

Ask: *besides population growth, where else in your own field might a
quantity have a single point of fastest change -- a real inflection
point -- rather than a maximum or minimum?* Good prompt for surfacing
that "steepest, not biggest" is a genuinely different (and under-used)
question to ask of data.

\newpage
\large

# PART 2 -- PODIUM CHEAT SHEET

\normalsize
*Podium copy -- terse on purpose. Full derivations, rationale, and
code: see Part 1 above.*
\large

## 0. Open (~2 min)

- Monday: toolkit (limits, critical points) via MM and Ricker
- This AM's WOD: limits on 2 new functions + a derivative with a hint
- Today: reveal the hint's answer, then the map

## 1. WOD4 limits, reveal (~5 min)

- MM, Ricker -- already done Monday, just flash it
- Logistic $\dfrac{1}{1+e^{-bx}}$: bounded 0 to 1, no new trick needed
- **Holling III** $\dfrac{ax^2}{b^2+x^2}$ -- do this one on the board:
  - $x\to0$: numerator wins the "who's higher order" question this
    time $\to \approx ax^2/b^2 \to 0$, quadratically
  - $x\to\infty$: **same degree top and bottom** -- new wrinkle!
    throw out $b^2$ vs. $x^2$ separately $\to ax^2/x^2 = a$
- **Sticking point:** same-degree fractions still use dominance --
  apply it inside num. and denom. separately, then simplify

\newpage

## 2. Critical points vs. inflection points (~12-14 min)

- Recap: critical point $= f'(x)=0$ (Ricker's peak, Monday)
- **Logistic derivative** (chain rule twice):
  $f'(x)=\dfrac{be^{-bx}}{(1+e^{-bx})^2}$
- $f'(0) = b/4$ -- **not zero!** (matches Bolker Fig. 3.9's own label)

```{r logistic-podium-fig, echo=FALSE, fig.width=3.3, fig.height=2.3}
par(mar = c(3.5, 3, 1, 0.5))
curve(f(x), from = -6, to = 6, lwd = 2, col = "steelblue",
      xlab = "x", ylab = "f(x)")
points(0, 0.5, pch = 19)
curve(0.5 + fp(0) * x, from = -1.5, to = 1.5, add = TRUE, col = "darkorange", lwd = 2)
```

- $x=0$ = **steepest** point, not flattest $\to$ **inflection point**
  ($f''=0$, concavity flips), not a critical point
- geometric argument only -- concave-up before, concave-down after; no
  need to grind through $f''$ symbolically today
- **Payoff:** Monday's sine WOD, day 91 = fastest increase = same
  feature, unnamed until now. Sigmoid's steepest point and a periodic
  function's steepest point are the *same calculus idea*.
- **Sticking point:** critical pt = flattest nearby; inflection pt =
  steepest nearby (for a sigmoid) -- easy to blur, keep separate

## 3. The bestiary map (~8-10 min)

- Table 3.1 (Bolker p.119) -- four families, don't derive every row:
  - polynomial/piecewise -- line, quadratic, hockey stick
  - **rational** -- MM, Holling III, Holling IV
  - **exponential-based** -- Ricker, **logistic**
  - power-based -- power law, von Bertalanffy, Gompertz
- **you've already met about half this table** -- today just names it
- **logistic's origin story:** solution to
  $dn/dt = rn(1-n/K)$ -- per-capita growth declines linearly to zero at
  carrying capacity $K$. Mechanistic view, not just curve-fitting.
- one-liner: Gompertz = another sigmoid, inflection at $1/e$ up not
  halfway -- "sigmoid" is a category, not one function

## 4. Roadmap (~2-3 min)

- Week 3: these shapes become the *mean* of a full stochastic model
- Later: model selection asks "sigmoid or not?" -- that question needs
  today's vocabulary

## If time: discussion seed

- *Where else in your field is there a single point of fastest change
  -- a real inflection point -- rather than a max/min?*
