---
title: "Week 8, Lecture 2: Optimization"
subtitle: "Instructor notes. Part 1: prep reading. Part 2: podium cheat sheet"
fontsize: 12pt
output:
  pdf_document: default
header-includes:
  - \usepackage{enumitem}
  - \setlist[itemize]{itemsep=4pt, parsep=0pt, topsep=4pt}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 5.5, fig.height = 3.8)
```

# How to prepare (for students)

- **Big picture.** Since Week 6 we've found maximum likelihood
  estimates by handing a negative log-likelihood to `optim()` and
  trusting what came back. Today we open the black box: how do
  optimizers search for the lowest point of an NLL surface, and why do
  they sometimes stop in the wrong place?
- **Bring:** your take-home WOD on Newton's method. We start by going
  over it.
- **Read:** Bolker 7.1-7.2 (printed p. 294-310). Skim 7.4.4 on
  multiple minima (p. 323-326).
- **Review:** derivatives and the Hessian from Lab 6 (the curvature of
  the NLL), and the Gamma distribution (Bolker 4.5.2.3, p. 172-174).
  Today's example fits a Gamma distribution to virus titers.

# Overview (instructor only)

This file has two parts, both instructor-only. **Part 1** is read
beforehand. **Part 2** (after the page break, large print) is the
podium copy.

| Beat | Time |
|---|---|
| Take-home WOD debrief (discussion) | ~8 min |
| Lecture, with three short discussion pauses | ~40 min |
| Separate discussion period | none |

**Pacing.** This is a full period by design. Core: sections 0-3 and
the start of section 4 (the rules and the acceptance probability).
Whatever of section 4 doesn't fit opens Week 9 Monday, where it leads
straight into MCMC (same acceptance rule). The "if time" section at the
end (reparameterizing the Gamma, a fitting checklist) can go to Week 9
or Lab 7. The three discussion pauses are short (~2 min each); if
running behind, ask the question, take one answer, and give the
landing point yourself.

**Source material.** Bolker 7.1-7.2 (direct search, Newton's method,
BFGS, Nelder-Mead, simulated annealing), 7.4.4 (multiple minima; the
Vonesh and Bolker story); legacy `LECTURE5.Rmd` ("Optimization!":
myxomatosis example, brute force, Newton with `pracma`, hand-coded
simplex and simulated annealing). The code below is rewritten from the
legacy version: no extra packages beyond `emdbook`, and one set of
helper functions shared across sections.

**Example choice.** The myxomatosis titers (grade 1 virus, 27 rabbits)
with a Gamma distribution, shape and scale parameters. This is Bolker's
running example in 7.2 (his Figures 7.5 and 7.6), so the reading and
the lecture line up. It was mentioned only in passing in Week 6. Two
features make it a good teaching surface: the NLL is undefined for
negative parameters, and shape and scale are strongly correlated, so
the minimum sits in a long, narrow, curved valley. The pikas carry
over from the take-home for the one-dimensional case.

# Where this sits in the week

- **Monday:** finished the one-parameter Bayesian material (germination
  example); handed out the Newton's method take-home.
- **Tuesday:** group meetings on final project proposals (no lab).
- **Today:** how optimizers work, and how they fail.
- **Coming up:** Week 9 Monday picks up simulated annealing and turns
  it into MCMC; Lab 7 on Bayesian inference.

# Part 1: Lecture content

```{r data-and-helpers}
library(emdbook)
titer <- subset(MyxoTiter_sum, grade == 1)$titer

# gamma NLL; Inf outside the allowed range so optimizers can't use it
nll_gamma <- function(p) {
  if (any(p <= 0)) return(Inf)
  -sum(dgamma(titer, shape = p[1], scale = p[2], log = TRUE))
}

# central finite differences: f'(x) ~ [f(x+h) - f(x-h)] / 2h, per parameter
grad_fd <- function(f, p, h = 1e-4 * abs(p)) {
  sapply(seq_along(p), function(i) {
    e <- replace(numeric(length(p)), i, h[i])
    (f(p + e) - f(p - e)) / (2 * h[i])
  })
}

# matrix of second derivatives (the Hessian), by finite differences
hess_fd <- function(f, p, h = 1e-3 * abs(p)) {
  k <- length(p)
  H <- matrix(0, k, k)
  for (i in seq_len(k)) {
    for (j in seq_len(k)) {
      ei <- replace(numeric(k), i, h[i])
      ej <- replace(numeric(k), j, h[j])
      H[i, j] <- (f(p + ei + ej) - f(p + ei - ej) - f(p - ei + ej) +
                    f(p - ei - ej)) / (4 * h[i] * h[j])
    }
  }
  H
}

shape_vec <- seq(10, 90, length.out = 120)
scale_vec <- seq(0.03, 0.32, length.out = 120)
surf <- outer(shape_vec, scale_vec, Vectorize(function(a, s) nll_gamma(c(a, s))))
nll_min <- min(surf)

# contour plot of NLL minus its minimum over the shape-scale grid
plot_surface <- function(...) {
  contour(shape_vec, scale_vec, surf - nll_min,
          levels = c(1, 2, 5, 10, 20, 50, 100, 200, 500, 1000),
          xlab = "shape", ylab = "scale", col = "grey60", labcex = 0.6, ...)
}
f2 <- function(v) formatC(v, format = "f", digits = 2)
f3 <- function(v) formatC(v, format = "f", digits = 3)
```

## 0. Take-home debrief (~8 min, discussion)

Run this as a discussion, not a solution walkthrough. Key in
`wod/answer_keys/WOD_Week8_Lecture1_NewtonByHand_Key.Rmd`.

- **3b (Newton from $\lambda = 2$).** Ask for the guesses: 2, 2.667,
  2.963, 3.000. Ask how fast the error shrank: 1, 0.33, 0.037, 0.0005.
  Each error is roughly the square of the last. That speed near the
  answer is the appeal of derivative-based methods.
- **3c (start at $\lambda = 7$).** Ask who got a negative number
  ($-2.33$) and why. Far from the MLE the NLL is nearly straight, so
  the matching parabola is very wide and its bottom is past zero.
  Collect fixes on the board: shorter steps (halve until the NLL
  improves), better starting values, bounds, fitting on the log
  scale. Newton on $\theta = \log\lambda$ from 7 goes 3.95, 3.11, 3.002:
  no overshoot.
- **2b and 3d (finite differences).** The second-difference formula is
  how `optim(..., hessian = TRUE)` got the Hessian in Lab 6. A Newton
  step built from finite-difference estimates (2.69) lands almost where
  the exact one does (2.67), so an optimizer doesn't need calculus.

Landing point: *every fix students proposed is built into a real
optimizer.* Today: what those optimizers do in more than one
dimension.

## 1. Big picture (~3 min)

Loop on the board:

$$\text{THEORY} \to \text{MODEL(S)} \to \text{DATA (fit)} \to \text{COMPARE} \to \text{repeat}$$

- Today is the "fit" arrow. Fitting means finding the lowest point of
  the NLL surface. Every method today is a strategy for searching that
  surface without evaluating it everywhere.
- **Today's data.** Virus titers (log$_{10}$ virus load) in 27 rabbits
  infected with the most virulent grade of myxomatosis (Bolker's
  example). Model: Gamma(shape, scale). Two parameters, so we can still
  draw the whole surface:

```{r myxo-surface, fig.width=5.5, fig.height=3.8}
par(mar = c(4, 4, 0.5, 0.5))
plot_surface()
fit_nm <- optim(c(20, 0.05), nll_gamma)
points(fit_nm$par[1], fit_nm$par[2], pch = 4, cex = 1.5, lwd = 2)
```

- **Brute force works here** (Week 6 grids): this picture took
  `r format(length(surf), big.mark = ",")` NLL evaluations. Bolker's coarser grid (shape 10 to
  100 by 1, scale 0.01 to 0.3 by 0.01) takes 2,730.
- **It doesn't scale.** At 30 values per parameter, a 10-parameter
  model needs $30^{10} \approx 6 \times 10^{14}$ evaluations: at 1 ms
  each, about 19,000 years.
- **The MLE.** Shape `r f2(fit_nm$par[1])`, scale
  `r f3(fit_nm$par[2])`, NLL `r f3(fit_nm$value)`.
- Note the shape of the valley: long, narrow, and curved. That shape
  causes trouble for every method today.

## 2. Newton's method in more than one dimension (~10 min)

**From the take-home.** One parameter: step $= -\text{NLL}'/\text{NLL}''$.
Two or more parameters:

- the slope becomes the **gradient**, a vector of slopes, one per
  parameter;
- the curvature becomes the **Hessian**, a matrix of second derivatives
  (Lab 6);
- "divide by the curvature" becomes "multiply by the inverse of the
  Hessian": $\theta_{\text{new}} = \theta - H^{-1}\,g$. In R,
  `solve(H, g)`.

```{r newton-2d}
newton <- function(f, start, n_steps = 8) {
  path <- matrix(start, nrow = 1)
  p <- start
  for (i in seq_len(n_steps)) {
    p <- p - solve(hess_fd(f, p), grad_fd(f, p))
    path <- rbind(path, p, deparse.level = 0)
  }
  path
}
path_good <- newton(nll_gamma, c(40, 0.2))
path_bad <- newton(nll_gamma, c(20, 0.05))
round(cbind(path_good, nll = apply(path_good, 1, nll_gamma)), 3)
round(cbind(path_bad, nll = apply(path_bad, 1, nll_gamma)), 3)
```

```{r newton-fig, fig.width=5.5, fig.height=3.8}
par(mar = c(4, 4, 0.5, 0.5))
plot_surface()
lines(path_good, type = "o", pch = 19, cex = 0.7, col = "steelblue", lwd = 2)
lines(path_bad, type = "o", pch = 19, cex = 0.7, col = "darkorange", lwd = 2)
legend("topright", c("start (40, 0.2)", "start (20, 0.05)"), lwd = 2,
       col = c("steelblue", "darkorange"), bty = "n", cex = 0.8)
```

- From (40, 0.2), Newton reaches the minimum in about six steps.
- From (20, 0.05), it lurches across the valley, the NLL goes *up* on
  one step, and after eight steps it still isn't done. Same lesson as
  3c on the take-home: the parabola is a good guide near the minimum
  and a bad one far from it.
- **Why not always use Newton?** It needs the full Hessian at every
  step: for $k$ parameters, about $k^2$ second derivatives, each costing
  several NLL evaluations by finite differences.
- **BFGS** (`method = "BFGS"`, a "quasi-Newton" method) computes only
  the gradient. It builds up an approximate Hessian from how the
  gradient changes from step to step. This is the workhorse for smooth
  surfaces (and `mle2()`'s default).
- **One parameter?** Use `optimize()`, which brackets the minimum and
  narrows in. `optim()`'s default warns if you give it one parameter.

**Pause for discussion (~2 min).**

```{r bfgs-compare}
fit_bfgs <- optim(c(20, 0.05), nll_gamma, method = "BFGS")
fit_bfgs_log <- optim(log(c(20, 0.05)), function(q) nll_gamma(exp(q)),
                      method = "BFGS")
cmp <- rbind(
  nelder_mead = c(fit_nm$par, fit_nm$value, fit_nm$counts[1], fit_nm$convergence),
  bfgs = c(fit_bfgs$par, fit_bfgs$value, fit_bfgs$counts[1], fit_bfgs$convergence),
  bfgs_log_scale = c(exp(fit_bfgs_log$par), fit_bfgs_log$value,
                     fit_bfgs_log$counts[1], fit_bfgs_log$convergence)
)
colnames(cmp) <- c("shape", "scale", "nll", "n_evals", "convergence")
round(cmp, 4)
```

> Same data, same starting point. Nelder-Mead (`optim()`'s default)
> returns NLL `r f3(fit_nm$value)` with `convergence = 0`. BFGS returns
> NLL `r f3(fit_bfgs$value)` with `convergence = `r fit_bfgs$convergence``,
> and estimates (`r f2(fit_bfgs$par[1])`, `r f3(fit_bfgs$par[2])`) that
> look close to Nelder-Mead's. Is the BFGS fit done? What would you do?

*Where to land:* no. `convergence = 1` means BFGS hit its iteration
limit (100) before it stopped improving. The NLL difference is tiny,
but nothing guarantees it would be tiny on another problem. Always
check `$convergence`. Fixes, from the take-home list: a better start,
restarting from where it stopped, or a better-behaved scale. Fitting
log(shape) and log(scale) instead converges in
`r fit_bfgs_log$counts[1]` evaluations (and can't wander into negative
values). Watch for students treating "the numbers look close" as
"converged."

## 3. Derivative-free: Nelder-Mead simplex (~12 min)

**The idea.** No slopes at all. Keep $k + 1$ points (a triangle for two
parameters), and keep moving the worst one. Sketch on the board:

- **Reflect** the worst point through the midpoint of the other two.
- If that's the best point yet, **expand**: go twice as far.
- If the reflection is still bad, **contract**: try halfway back.
- If nothing works, **shrink** the whole triangle toward the best point.

The triangle tumbles downhill, stretches out along valleys, and shrinks
around the minimum. `optim()` stops when the vertices' NLL values are
all within a small tolerance.

```{r nelder-mead}
# hand-coded Nelder-Mead for two parameters; stores every simplex
nelder_mead <- function(f, start, step, n_iter = 80) {
  simplex <- rbind(start, start + c(step[1], 0), start + c(0, step[2]))
  vals <- apply(simplex, 1, f)
  history <- vector("list", n_iter)
  moves <- character(n_iter)
  for (it in seq_len(n_iter)) {
    o <- order(vals)
    simplex <- simplex[o, ]
    vals <- vals[o]
    history[[it]] <- simplex
    centroid <- colMeans(simplex[1:2, ])
    refl <- centroid + (centroid - simplex[3, ])
    f_r <- f(refl)
    if (f_r < vals[1]) {
      expd <- centroid + 2 * (centroid - simplex[3, ])
      f_e <- f(expd)
      if (f_e < f_r) {
        simplex[3, ] <- expd; vals[3] <- f_e; moves[it] <- "expand"
      } else {
        simplex[3, ] <- refl; vals[3] <- f_r; moves[it] <- "reflect"
      }
    } else if (f_r < vals[2]) {
      simplex[3, ] <- refl; vals[3] <- f_r; moves[it] <- "reflect"
    } else {
      contr <- centroid + 0.5 * (simplex[3, ] - centroid)
      f_c <- f(contr)
      if (f_c < vals[3]) {
        simplex[3, ] <- contr; vals[3] <- f_c; moves[it] <- "contract"
      } else {
        simplex[2:3, ] <- (simplex[2:3, ] + rbind(simplex[1, ], simplex[1, ])) / 2
        vals[2:3] <- apply(simplex[2:3, ], 1, f)
        moves[it] <- "shrink"
      }
    }
  }
  list(best = simplex[which.min(vals), ], value = min(vals),
       history = history, moves = moves)
}
nm <- nelder_mead(nll_gamma, c(20, 0.05), step = c(5, 0.02))
round(c(shape = nm$best[1], scale = nm$best[2], nll = nm$value), 4)
table(nm$moves)
```

```{r nm-fig, fig.width=5.5, fig.height=3.8}
move_cols <- c(expand = "darkorange", reflect = "steelblue",
               contract = "firebrick", shrink = "purple")
par(mar = c(4, 4, 0.5, 0.5))
plot_surface()
for (i in 1:30) polygon(nm$history[[i]], border = move_cols[nm$moves[i]], lwd = 1.5)
legend("topright", names(move_cols), col = move_cols, lwd = 2, bty = "n", cex = 0.8)
```

- The first 30 triangles are shown. Big expanding steps across the
  contours, then contractions as it "turns the corner" into the valley,
  then reflections and contractions crawling along the valley floor
  (compare Bolker's Figure 7.5, p. 305).
- `optim()`'s version needed `r fit_nm$counts[1]` NLL evaluations. BFGS
  on the log scale needed `r fit_bfgs_log$counts[1]` (plus gradient
  evaluations). Nelder-Mead is slower but doesn't care whether the
  surface is smooth: bumps, kinks, and thresholds don't fool it the way
  they fool a slope-based method.

**Pause for discussion (~2 min).**

```{r ridge-corr}
vc <- solve(optimHess(fit_nm$par, nll_gamma))
corr_ss <- cov2cor(vc)[1, 2]
prod(fit_nm$par)   # shape x scale = Gamma mean
mean(titer)
```

> Every method today crawled along the same long, curved valley.
> Using the Hessian at the MLE (Lab 6), the estimated correlation
> between shape and scale is `r f3(corr_ss)`. What does that say about
> what these data can and can't tell us? (Hint: the mean of a Gamma is
> shape $\times$ scale.)

*Where to land:* along the valley floor, shape goes up and scale goes
down while their product, the mean titer
(`r f2(prod(fit_nm$par))`), barely changes. The data are very
informative about the mean and much less about how it splits between
shape and scale. That's what makes the valley long and narrow, and
it's why optimizers creep along it. Callback: in Lab 6 Act 1, centering
the dose at 1.8 removed the same kind of correlation between $a$ and
$b$. Reparameterizing (fit mean and shape instead) is the fix here;
it's in the "if time" section.

## 4. Stochastic global search: simulated annealing (start; ~8 min)

**The problem.** Everything so far goes downhill from the start. If
the surface has more than one valley, a downhill method finds whichever
valley it starts in. Sketch on the board: an NLL curve with a shallow
valley near the start and a deeper one farther away (Bolker's Figure
7.1). Bolker's own story (7.4.4, p. 323-325): in a published tadpole
predation analysis, Nelder-Mead found one minimum; later, BFGS from
the same start found a different, slightly better one.

**The idea (Bolker p. 306).** Take random steps, and *sometimes accept
a worse one*, so the search can climb out of a shallow valley. The name
comes from metalworking: cooling metal slowly lets it settle into a
low-energy state.

1. From the current point, propose a random nearby point.
2. If its NLL is lower, move there.
3. If it's higher by $\Delta$, move there anyway with probability
   $e^{-\Delta / T}$, where $T$ is the "temperature."
4. Repeat, and lower $T$ every so often.

```{r accept-table}
accept_tab <- outer(c(10, 1, 0.1), c(0.5, 2, 10),
                    function(temp, delta) exp(-delta / temp))
dimnames(accept_tab) <- list(paste("T =", c(10, 1, 0.1)),
                             paste("NLL worse by", c(0.5, 2, 10)))
accept_show <- ifelse(accept_tab < 0.001, "<0.001",
                      formatC(accept_tab, format = "f", digits = 3))
noquote(accept_show)
```

- Hot ($T = 10$): most steps are accepted; the search wanders
  widely. Cold ($T = 0.1$): it's almost pure downhill.
- Start hot to explore, cool down to settle.

```{r anneal}
# simulated annealing: accept worse steps w.p. exp(-delta / temp); cool temp over time
anneal <- function(f, start, step_sd, n_steps = 3000, temp0 = 10,
                   cool = 0.9, every = 100) {
  current <- start
  f_current <- f(current)
  best <- current
  f_best <- f_current
  temp <- temp0
  track <- matrix(NA, n_steps, 4,
                  dimnames = list(NULL, c("shape", "scale", "nll", "temp")))
  for (i in seq_len(n_steps)) {
    proposal <- current + rnorm(length(current), 0, step_sd)
    f_prop <- f(proposal)
    if (is.finite(f_prop) && runif(1) < exp(-(f_prop - f_current) / temp)) {
      current <- proposal
      f_current <- f_prop
    }
    if (f_current < f_best) {
      best <- current
      f_best <- f_current
    }
    track[i, ] <- c(current, f_current, temp)
    if (i %% every == 0) temp <- temp * cool
  }
  list(best = best, value = f_best, track = track)
}
set.seed(746)
sa <- anneal(nll_gamma, c(20, 0.05), step_sd = c(2, 0.005))
round(c(shape = sa$best[1], scale = sa$best[2], nll = sa$value), 4)
```

```{r anneal-fig, fig.width=6.5, fig.height=3}
par(mfrow = c(1, 2), mar = c(4, 4, 0.5, 0.5))
plot_surface()
lines(sa$track[, "shape"], sa$track[, "scale"], col = adjustcolor("darkorange", 0.5))
plot(sa$track[, "nll"], type = "l", log = "y", xlab = "Step", ylab = "Current NLL")
```

- Left: the search drops into the valley within a few dozen steps, then
  wanders up and down it (it keeps accepting slightly worse points).
  Right: as $T$ falls, the wandering settles near the minimum.
- Best point found: (`r f2(sa$best[1])`, `r f3(sa$best[2])`), NLL
  `r f3(sa$value)`, after 3,000 evaluations (Nelder-Mead:
  `r fit_nm$counts[1]`). Slow and imprecise on a one-valley surface; its
  value is on surfaces with many valleys.

**Pause for discussion (~2 min).**

```{r sann}
set.seed(746)
fit_sann <- optim(c(20, 0.05), nll_gamma, method = "SANN")
round(c(shape = fit_sann$par[1], scale = fit_sann$par[2], nll = fit_sann$value,
        evals = unname(fit_sann$counts[1]),
        convergence = fit_sann$convergence), 4)
```

> `optim()` has a built-in simulated annealer, `method = "SANN"`. With
> its default settings it returns NLL `r f2(fit_sann$value)` after
> `r format(fit_sann$counts[1], big.mark = ",")` evaluations, and
> `convergence = 0`. Nelder-Mead found `r f2(fit_nm$value)` in
> `r fit_nm$counts[1]`. Is SANN broken? Should you trust
> `convergence = 0`?

*Where to land:* not broken, but untuned. SANN runs a fixed number of
steps and *always* reports `convergence = 0` (see `?optim`), so that
code tells you nothing. Simulated annealing needs its temperature and
step size tuned to the problem (Bolker p. 310). The usual practice: use
a global method to find the right valley, then polish with Nelder-Mead
or BFGS started from its answer. And whatever the method, try several
starting points.

**Bridge to Week 9.** Keep the acceptance rule, fix $T = 1$, never
cool, and keep *every* point visited instead of only the best. The
collection of points then forms a sample from the posterior
distribution. That's MCMC (the Metropolis algorithm), Monday's topic.

## 5. Return to the big picture (~1 min)

- Fitting = searching the NLL surface. Grids are simple and hard to
  fool but don't scale.
- Derivative-based (Newton, BFGS): fast near the answer, can misbehave
  far from it. Derivative-free (Nelder-Mead): slower, robust to bumps.
  Stochastic (simulated annealing): slowest, can escape local valleys.
- Whatever you use: check `$convergence`, try several starts, and
  look at the surface (or a profile) when something seems off.

## If time, or Week 9 / Lab 7

**Reparameterize.** Fit the Gamma in terms of its mean and shape
(scale = mean/shape, Bolker's parameterization for Gamma regressions):

```{r reparam}
nll_mean_shape <- function(p) nll_gamma(c(p[2], p[1] / p[2]))
fit_ms <- optim(c(7, 20), nll_mean_shape)
vc_ms <- solve(optimHess(fit_ms$par, nll_mean_shape))
round(c(mean = fit_ms$par[1], shape = fit_ms$par[2], nll = fit_ms$value,
        evals = unname(fit_ms$counts[1]),
        corr = cov2cor(vc_ms)[1, 2]), 3)
```

Same minimum, correlation near zero, and a round valley instead of a
banana.

**Checklist when fitting (Bolker 7.4).** Test the NLL at the starting
values (finite?). Check `$convergence`. Refit from the answer, and from
a few other starts. Try a second method. Put parameters on comparable
scales (log, or `control = list(parscale = ...)`). Plot a slice or
profile around the answer.

\newpage
\large

# PART 2: PODIUM CHEAT SHEET

\normalsize
\textit{Terse on purpose. Full rationale and code: Part 1.}
\large

**Take-home debrief ~8 min. Lecture ~40 min (3 pauses). No separate discussion.**

## 0. Take-home debrief (~8 min)

- 3b: 2 $\to$ 2.667 $\to$ 2.963 $\to$ 3.000; error 1, 0.33, 0.037, 0.0005
- 3c: from 7 $\to$ $-2.33$. Why? Flat NLL $\to$ wide parabola
- Board: fixes (shorter steps, better start, bounds, log scale)
- Log scale from 7: 3.95, 3.11, 3.002
- 2b: $[f(x+h) - 2f(x) + f(x-h)]/h^2$ = Lab 6 Hessian
- **Every fix you proposed is built into real optimizers**

\newpage

## 1. Big picture (~3 min)

- Loop: THEORY $\to$ MODEL $\to$ **DATA (fit)** $\to$ COMPARE
- Fit = find the bottom of the NLL surface
- Myxo titers, 27 rabbits, Gamma(shape, scale)
- MLE (`r f2(fit_nm$par[1])`, `r f3(fit_nm$par[2])`), NLL `r f3(fit_nm$value)`
- Grid: 2,730 evals here; 10 params at 30 values $\approx$ 19,000 years

```{r podium-surface, echo=FALSE, fig.width=5, fig.height=3}
par(mar = c(4, 4, 0.5, 0.5))
plot_surface()
points(fit_nm$par[1], fit_nm$par[2], pch = 4, cex = 1.5, lwd = 2)
```

\newpage

## 2. Newton in 2D (~10 min)

- Slope $\to$ gradient $g$; curvature $\to$ Hessian $H$
- $\theta_{\text{new}} = \theta - H^{-1} g$ (`solve(H, g)`)
- Good start: ~6 steps. Bad start: lurches, NLL goes up

```{r podium-newton, echo=FALSE, fig.width=5, fig.height=3}
par(mar = c(4, 4, 0.5, 0.5))
plot_surface()
lines(path_good, type = "o", pch = 19, cex = 0.6, col = "steelblue", lwd = 2)
lines(path_bad, type = "o", pch = 19, cex = 0.6, col = "darkorange", lwd = 2)
```

- BFGS: gradient only, builds approximate $H$; `mle2()` default
- 1 parameter: `optimize()`

**Pause:** NM `r f3(fit_nm$value)` (conv 0) vs. BFGS `r f3(fit_bfgs$value)` (conv `r fit_bfgs$convergence`). Done?

- No: hit maxit. Check `$convergence`. Log scale: `r fit_bfgs_log$counts[1]` evals, conv 0

\newpage

## 3. Nelder-Mead (~12 min)

- Sketch triangle: reflect, expand, contract, shrink
- No derivatives; tumbles downhill, crawls along valleys
- `optim()` default; `r fit_nm$counts[1]` evals

```{r podium-nm, echo=FALSE, fig.width=5, fig.height=3}
par(mar = c(4, 4, 0.5, 0.5))
plot_surface()
for (i in 1:30) polygon(nm$history[[i]], border = move_cols[nm$moves[i]], lwd = 1.5)
```

**Pause:** corr(shape, scale) = `r f3(corr_ss)`. What can the data tell us?

- Mean = shape $\times$ scale = `r f2(prod(fit_nm$par))`: well determined; split isn't
- Lab 6 centering; reparameterize (if time)

\newpage

## 4. Simulated annealing (~8 min; rest to Week 9)

- Board: two valleys; downhill finds the nearest one
- Bolker's tadpoles: NM and BFGS found different minima
- Random step; worse by $\Delta$: accept w.p. $e^{-\Delta/T}$; cool $T$

```{r podium-accept, echo=FALSE}
noquote(accept_show)
```

```{r podium-anneal, echo=FALSE, fig.width=6, fig.height=2.6}
par(mfrow = c(1, 2), mar = c(4, 4, 0.5, 0.5))
plot_surface()
lines(sa$track[, "shape"], sa$track[, "scale"], col = adjustcolor("darkorange", 0.5))
plot(sa$track[, "nll"], type = "l", log = "y", xlab = "Step", ylab = "Current NLL")
```

**Pause:** SANN defaults: NLL `r f2(fit_sann$value)`, `r format(fit_sann$counts[1], big.mark = ",")` evals, conv 0. Broken?

- Untuned; always reports 0. Find the valley, then polish with NM/BFGS

## 5. Wrap (~1 min)

- Grid / Newton-BFGS / Nelder-Mead / annealing: speed vs. robustness
- Check convergence; several starts; look at the surface
- Monday: $T = 1$, keep every point $\to$ MCMC
