---
title: "Lab 3: Probability and Distributions"
subtitle: "Instructor notes -- Part 1: prep reading, Part 2: podium/circulating cheat sheet"
fontsize: 12pt
output:
  pdf_document: default
header-includes:
  - \usepackage{enumitem}
  - \setlist[itemize]{itemsep=4pt, parsep=0pt, topsep=4pt}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 5, fig.height = 3.5)
```

This single file has two parts, both instructor-only (not a student
handout). **Part 1** is read beforehand -- an answer key and the
reasoning behind each Act. **Part 2** (after the page break, large
print) is the copy to have open while circulating during lab.

**Schedule note:** Labor Day cancels this week's Monday meeting, so
this lab runs with no prior lecture on Bayes' rule -- Wednesday's
lecture comes after, and the bestiary of distributions has moved to
next Monday (Week 4). **Design note:** Act 1 is deliberately
discovery-based rather than "apply the formula you were just taught" --
students derive the PPV calculation themselves from plain counting
before ever seeing it written as $P(H\mid D)$. **No laptops this lab**
-- all three Acts are pen-and-paper (a calculator is enough); Acts 2
and 3 use pre-rendered figures and a printed z-table instead of
`dbinom()`/`dpois()`/`pnorm()`/`qnorm()`, since sampling/evaluating a
named distribution isn't, by itself, enough reason to open a laptop --
that's reserved for weeks with real algorithm-development content.

## Suggested opening (~5 min, before releasing Act 1)

Keep this short -- a full treatment of conditional probability and
Bayes' rule is Wednesday's job, not today's. Two rules are enough to
unlock Act 1:

- **Addition:** probabilities of mutually exclusive (disjoint) outcomes
  add up (e.g., "flagged" splits into "flagged AND really the species"
  plus "flagged AND not the species" -- no overlap, so these two pieces
  sum to the total flagged).
- **Multiplication / applying a rate:** a probability or rate applied
  to a count of events gives you the expected count in that subgroup
  (e.g., if 10 events are truly the species and the classifier catches
  98% of them, expect about $10\times0.98\approx10$ flagged).

That's it -- don't derive Bayes' rule or use conditional-probability
notation ($P(A\mid B)$) in this opening; Act 1 gets there on its own,
and Wednesday's lecture gives it a name.

# Act 1: camera-trap classifier PPV (discovery-based)

**Part 1 (counting).** Same numbers as before
($P(\text{species})=0.0001$, sensitivity $=0.98$, false positive rate
$=0.01$), $N=100{,}000$.

```{r act1-freq-key}
N <- 1e5; p_S <- 1e-4; sens <- 0.98; fpr <- 0.01
species     <- N * p_S
not_species <- N * (1 - p_S)
flag_S      <- species * sens
flag_notS   <- not_species * fpr
data.frame(row = c("Actually the species", "Not the species"),
           flagged = round(c(flag_S, flag_notS)),
           not_flagged = round(c(species - flag_S, not_species - flag_notS)),
           total = c(species, not_species))
```

**Answer key (1a-1b):** Flagged/true-species $\approx10$; flagged/not-species
$\approx1{,}000$; total flagged $\approx1{,}010$. PPV (1b) $=10/1010\approx0.0099\approx1\%$
-- pure counting, no formula.

**Answer key (1c):** open-ended, but the number should read as
"surprisingly low given how good the classifier sounded" -- a single
flag isn't enough to trust; the field team should look for corroborating
evidence (a second detection, a track, scat) before acting on one image.

**Part 2 (formalize).** Guided derivation:

```{r act1-symbolic-key}
# numerator: N * p_S * sens ; denominator: N*p_S*sens + N*(1-p_S)*fpr
# N cancels, leaving:
ppv_formula <- (p_S * sens) / (p_S * sens + (1 - p_S) * fpr)
ppv_formula
```

**Answer key (2a-2c):** numerator $=N\times\text{prevalence}\times\text{sensitivity}$;
denominator $=N\times\text{prevalence}\times\text{sensitivity} + N\times(1-\text{prevalence})\times\text{false positive rate}$.
$N$ cancels, leaving
$$\text{PPV} = \frac{\text{prevalence}\times\text{sensitivity}}{\text{prevalence}\times\text{sensitivity} + (1-\text{prevalence})\times\text{false positive rate}} \approx 0.0097$$
This is Bayes' rule / PPV, arrived at without ever naming it -- don't
name it for them either; that reveal is Wednesday's job. If a group
gets stuck on 2a, the hint is: "what did you multiply by 100,000 to get
each cell of your table in Part 1? Do that again, but write $N$ instead
of 100,000."

**Part 3 (transfer -- Site B).** Prevalence $=0.01$, same sensitivity
and false positive rate.

```{r act1-siteB-key}
p_S_B <- 0.01
ppv_B <- (p_S_B * sens) / (p_S_B * sens + (1 - p_S_B) * fpr)
ppv_B
```

**Answer key (3a):** PPV $\approx0.497\approx50\%$ -- a dramatic jump
from Site A's $\approx1\%$, same classifier. **(3b):** the intended
takeaway is that prevalence dominates PPV every bit as much as the
classifier's own accuracy -- a "good" classifier is only as useful as
the base rate of what it's looking for allows.

**Watch for:** groups who compute the fraction backwards ($P(S)$
instead of $P(F)$ in the denominator of 1b). A PPV $>0.5$ at Site A is
almost always the tell. In Part 2, watch for groups who try to cancel
something other than $N$, or who get stuck because they're trying to
recall "the formula" instead of generalizing their own Part 1 numbers.

# Act 2: binomial detections, and when Poisson is a good stand-in

$N=8$, $p=0.3$, $k=3$; all figures below are pre-rendered in the
handout (no code for students to run).

```{r act2-key}
N <- 8; p <- 0.3; k <- 3
dbinom(k, N, p)
which.max(dbinom(0:8, 8, 0.3)) - 1   # mode, 0-indexed
```

**Answer key:** $P(X=3)=0.2541$; mode $=2$.

**Poisson comparison, part A** ($N=8,p=0.3$ vs. $\lambda=2.4$): binomial
variance $Np(1-p)=1.68$; Poisson variance $\lambda=2.4$ -- mismatched,
and the pre-rendered barplots look visibly different (Poisson has a
fatter right tail).

**Poisson comparison, part B** ($N=100,p=0.024$ vs. $\lambda=2.4$):
binomial variance $=100(0.024)(0.976)=2.3424$ -- close to $2.4$, and the
barplots should look nearly identical. Rule of thumb (2d): *Poisson is
a good stand-in for binomial when $N$ is large and $p$ is small* (so
$(1-p)\approx1$ and $Np(1-p)\approx Np=\lambda$).

**Watch for:** groups who conclude Poisson is "just always fine" from
Scenario B alone without noticing Scenario A disagreed -- push them
back to the variance mismatch in Scenario A as the reason it isn't
universal.

# Act 3: continuous body mass ($\mu=4.5$, $\sigma=0.6$), via z-table

```{r act3-key}
mu <- 4.5; sigma <- 0.6
z <- (5.5 - mu) / sigma
z
1 - pnorm(round(z, 2))          # table-precision version of 1 - Phi(z)
1 - pnorm(5.5, mu, sigma)       # exact, for comparison
mu + 1.28 * sigma               # table-lookup version of the 90th percentile
qnorm(0.9, mu, sigma)           # exact, for comparison
```

**Answer key:** $z=(5.5-4.5)/0.6=1.67$ (table row 1.6, column .07);
$\Phi(1.67)\approx0.9525$, so $P(\text{mass}>5.5)\approx0.0475$
($\approx5\%$; the exact value via `pnorm` is $0.0478$ -- table
rounding accounts for the small difference). For 3c, the table's
closest entry to $0.90$ is $\Phi(1.28)=0.8997$, giving a cutoff of
$4.5+1.28(0.6)\approx5.27$ kg (exact value via `qnorm` is $5.27$ kg --
matches).

**Watch for:** 1b/1c (density vs. probability) is the conceptual crux
of Part 1 -- a group that answers "$P(\text{mass}=4.5)$ is very small
but not zero" hasn't fully separated density from probability yet.
Point them back at the shaded-area picture from 1a: a single point has
zero *width*, hence zero area, hence zero probability, regardless of
how tall the curve is there. In Part 3, watch for groups who forget to
convert the z-score back to a body mass in 3c (the table gives $z$, not
$x$).

\newpage
\large

# PART 2 -- CHEAT SHEET (circulating)

\normalsize
*Copy for circulating during lab. Full reasoning and code: see Part 1
above.*
\large

## Opening (~5 min)

- Two rules only: **addition** (disjoint outcomes sum) and
  **multiplication/rate-application** (rate $\times$ count = expected
  count in a subgroup)
- Don't derive Bayes' rule or use $P(A\mid B)$ notation -- Act 1 gets
  there on its own; Wednesday names it

## Act 1 -- classifier PPV (discovery)

- Part 1 (count, $N=100$k): flagged/true $\approx10$, flagged/false
  $\approx1{,}000$, total flagged $\approx1{,}010$ $\to$ PPV
  $\approx1\%$
- Part 2 hint if stuck: "what did you multiply by 100,000 in Part 1?
  Do that again with $N$ instead"
- Part 2 answer: PPV $=\dfrac{\text{prev}\times\text{sens}}{\text{prev}\times\text{sens}+(1-\text{prev})\times\text{fpr}}\approx0.0097$
- Part 3 (Site B, prevalence $=0.01$): PPV $\approx50\%$ -- prevalence
  dominates
- **Watch for:** PPV $>0.5$ at Site A usually means $P(S)$/$P(F)$
  flipped

## Act 2 -- binomial vs. Poisson

- $N=8,p=0.3,k=3$: hand calc $=0.2541$; mode $=2$
- Var mismatch ($N=8,p=0.3$): binom $1.68$ vs. Pois $2.4$ -- **don't
  match**
- Var match ($N=100,p=0.024$): binom $2.3424$ vs. Pois $2.4$ -- **do
  match**
- Rule of thumb: Poisson $\approx$ binomial when $N$ large, $p$ small

## Act 3 -- continuous mass, via z-table

- $z=(5.5-4.5)/0.6=1.67 \to \Phi(1.67)\approx0.9525 \to P(\text{mass}>5.5)\approx0.048$
- Table's closest entry to 0.90 is $z\approx1.28 \to$ cutoff
  $\approx4.5+1.28(0.6)=5.27$ kg
- **Watch for:** "small but nonzero" answers to $P(X=4.5)$ (push back
  to zero width $\to$ zero area); forgetting to convert $z$ back to $x$
  in 3c
