---
title: "Week 4, Lecture 2: Synthetic Data Simulation"
subtitle: "Instructor notes -- Part 1: prep reading, Part 2: podium cheat sheet"
fontsize: 12pt
output:
  pdf_document: default
header-includes:
  - \usepackage{enumitem}
  - \setlist[itemize]{itemsep=4pt, parsep=0pt, topsep=4pt}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 5, fig.height = 3.5)
```

This single file has two parts, both instructor-only (not a student
handout). **Part 1** is read beforehand -- full worked examples and the
reasoning behind the flow. **Part 2** (after the page break, large
print) is the podium copy.

Source material: Bolker Ch. 5 intro and 5.2 (stochastic simulation,
p. 197-201, the single-group linear/hyperbolic examples). Also mined:
legacy `LECTURE3.Rmd` ("Data Generating Models") -- its "why simulate
fake data" list, deterministic/stochastic decomposition, and the tree
volume goodness-of-fit example are well-tested and reused below with
modernized R code; its power-analysis monitoring-program example is
flagged for next week's Lab 4 rather than rebuilt here, since fully
coding it up is more than today's lecture has time for.

# Where this sits in the week

- Monday (this week): combinatorics, random variables, probability
  distributions (pmf vs. pdf), expectation and variance -- students now
  have a toolbox of named distributions and know how to compute their
  moments.
- Tuesday (this week): no lab Acts -- lab time is final project
  proposal pitches instead.
- **Today:** those distributions stop being static objects to describe
  and become the *stochastic* half of something you can build and run
  -- a full **data-generating model**. This is also direct setup for
  next week's Lab 4 ("Data generating models"), which builds a full
  power-analysis simulation; today stays conceptual and example-driven,
  leaving the deep hands-on coding for the lab.

# Rationale for the flow

Bolker's own framing for Chapter 5 is useful to state explicitly:
simulation is *forward* modeling (pick a model and parameters, predict
data), while the rest of the course is mostly about *inverse* modeling
(start from data, work backward to parameters). Students haven't done
any inverse modeling yet (that starts in Week 5, likelihood), so
today's forward-modeling framing is genuinely their first exposure to
"build a model and run it," not a contrast with something they already
know how to do -- keep the framing simple rather than assuming the
inverse-modeling contrast lands yet.

Keep today to one clean running example (the deterministic + stochastic
decomposition, worked twice -- once as pure demonstration, once as a
goodness-of-fit check against real-looking data) rather than touring
many models. The payoff is conceptual: students should leave able to
name *why* you'd build a synthetic data generator, not just how to call
`rnorm()`.

# Part A: Why simulate data? (~8 min)

Open with the Popperian framing from earlier in the semester:
$$\text{THEORY/QUESTION} \to \text{CANDIDATE MODEL(S)} \to \text{confront with DATA} \to \text{COMPARE} \to \text{repeat}$$
A synthetic data generator *is* a candidate model, made concrete enough
to run. If you can write the computer program that generates data under
your hypothesis, you've forced every assumption into the open -- there's
nowhere for a vague verbal hypothesis to hide.

Four reasons this matters (ask students to guess some before listing):

- **Validate model fitting.** Before trusting an estimation procedure on
  real data (where you never know the true answer), run it on data you
  generated yourself -- you *know* the true parameters, so you can check
  whether the method recovers them.
- **Power analysis.** Simulate data under a hypothesized effect and a
  proposed sampling design, then ask how often the analysis would
  actually detect the effect. Tells you whether a study is worth
  running before you run it.
- **Improved understanding.** Chaining a deterministic function to a
  stochastic layer and watching what comes out builds intuition for a
  model's behavior that staring at the equation alone doesn't -- you
  see the consequences of your assumptions, including ones you didn't
  realize you'd made.
- **Forecasting.** A validated data-generating model can be run forward
  under new conditions (new covariate values, future time points) to
  predict what hasn't been observed yet.

**Sticking point to flag:** "fake data" sounds like something you'd
never show a reviewer, but the point isn't to replace real data -- it's
a diagnostic tool you use *before and alongside* real data, the same
way you'd sanity-check any other model output.

# Part B: Anatomy of a data-generating model (~10-12 min)

Bolker's running distinction: a **deterministic component** (the
"signal," giving an expected value) plus a **stochastic component**
(the "noise," a probability distribution wrapped around that expected
value). Monday's lecture supplied exactly the vocabulary needed for the
second half of this.

**Simplest case -- linear regression's data-generating model:**
$$\hat{\mu}_i = a + b x_i \qquad Y_i \sim \text{Normal}(\hat{\mu}_i, \sigma)$$

```{r linreg-sim}
x <- 1:20                                    # predictor: e.g. nitrogen concentration
a <- 2; b <- 1
mu <- a + b * x                              # deterministic component

set.seed(1001)
y <- rnorm(length(x), mean = mu, sd = 2)     # stochastic component layered on top

plot(x, y, pch = 16, xlab = "x", ylab = "y")
lines(x, mu, lwd = 2)
legend("topleft", legend = "true deterministic mean", lwd = 2, bty = "n")
```

Point out explicitly: the normal's mean parameter *is* the deterministic
function's output, evaluated at each $x_i$ -- this is the same
$Y_i \sim \text{Normal}(\mu, \sigma^2)$ notation from Monday, just with
$\mu$ replaced by a function of $x$ instead of a constant.

**A case where the components don't just add together.** Not every
model can be written as "deterministic plus noise" -- sometimes the
deterministic function controls a parameter of a distribution that
isn't the mean of a symmetric error term. Example: a hyperbolic
decline in fecundity with population density, with Poisson-distributed
counts (rather than negative binomial, so this stays inside Monday's
distributions and doesn't require introducing overdispersion):

```{r hyperbolic-sim}
dens <- runif(30, min = 0, max = 5)          # population density, sampled from the field
a <- 20; b <- 1
mu_det <- a * b / (b + dens)                 # deterministic: hyperbolic decline

set.seed(1001)
fecundity <- rpois(30, lambda = mu_det)      # stochastic: Poisson counts around that mean

plot(dens, fecundity, pch = 16, xlab = "Density", ylab = "Offspring count")
curve(a * b / (b + x), add = TRUE, lwd = 2)
```

**Sticking point to flag:** the Poisson's rate parameter $\lambda$ is
*simultaneously* its mean and variance (Monday's equidispersion point)
-- so as the deterministic curve declines, the *spread* of the noise
shrinks right along with it. That's visible in the plot (points hug the
curve more tightly at low density) and is a real, checkable consequence
of choosing this distribution, not an artifact.

# Part C: Using synthetic data as a goodness-of-fit check (~10-12 min)

Worked example (adapted from legacy `LECTURE3.Rmd`, tree girth/volume):
imagine a small set of "real" measurements.

```{r tree-data}
tree_data <- data.frame(girth = seq(1, 10, length.out = 10),
                         volume = c(0.4, 2.7, 5.3, 13.3, 42.4,
                                    63.5, 45.8, 233.4, 213.7, 383.1))
plot(volume ~ girth, data = tree_data, pch = 16)
```

**The question a data-generating model can answer:** could a *specific,
fully-specified* hypothesis have plausibly produced this data? Propose
one: $\text{Volume}_i \sim \text{Normal}(a + b\cdot\text{Girth}_i,\ \sigma)$,
with $a=10$, $b=40$, $\sigma=31$. Simulate many datasets under that
exact hypothesis and see whether the real data falls inside the
plausible range:

```{r tree-goodness-of-fit}
n_sim <- 1000
n_obs <- nrow(tree_data)
sim_volume <- replicate(n_sim, rnorm(n_obs, mean = 10 + 40 * tree_data$girth, sd = 31))

boxplot(t(sim_volume), names = tree_data$girth,
        xlab = "Girth", ylab = "Volume (simulated)")
points(seq_len(n_obs), tree_data$volume, pch = 16, col = "red", cex = 1.5)
legend("topleft", legend = "real data", pch = 16, col = "red", bty = "n")
```

The real data (red) sits well inside the simulated spread -- this
particular hypothesis is at least *plausible*. Contrast with an
obviously wrong hypothesis (flat mean, no girth effect):

```{r tree-null}
sim_null <- replicate(n_sim, rnorm(n_obs, mean = 100, sd = 400))
boxplot(t(sim_null), names = tree_data$girth,
        xlab = "Girth", ylab = "Volume (simulated, null model)")
points(seq_len(n_obs), tree_data$volume, pch = 16, col = "red", cex = 1.5)
```

Both technically "contain" the red points somewhere in their range (the
null's variance is huge), but the pattern is obviously worse -- worth
asking students *why* a wide-enough null can always "fit" in this loose
visual sense, and what that implies about needing sharper comparison
tools later in the course (a preview of Weeks 6-7's likelihood-based
model comparison, which replaces "eyeball it" with a number).

**Sticking point to flag:** this is a *plausibility* check, not proof --
"the data are consistent with this model" is a much weaker (and more
honest) claim than "this model is correct." That distinction is the
same corroborate-vs-prove vocabulary from the semester's opening
Popperian framing.

# Take-home message

- A data-generating model = a deterministic function (the signal) plus
  a probability distribution (the noise) -- Monday's distributions
  supply the second half.
- Simulating data is useful before you ever touch real data: it
  validates fitting procedures, powers a power analysis, builds
  intuition, and lets you forecast.
- Comparing real data against simulated replicates is a visual
  goodness-of-fit / plausibility check -- a first, informal version of
  the model comparison tools coming later in the semester.

# Roadmap tie-in

- Next Tuesday (Lab 4, "Data generating models"): build a full
  power-analysis simulation from scratch (a monitoring-program example
  -- how large a sample size is needed to reliably detect a population
  decline), going well beyond today's conceptual introduction.
- Weeks 6-7 (likelihood, model comparison): replace today's "eyeball
  the boxplot" goodness-of-fit check with a quantitative one.

# Discussion seed (if time remains)

Ask: *if you were designing a monitoring study for a species you
study, what would your data-generating model need to include -- what's
the deterministic part (the ecological signal you actually care about)
and what's the stochastic part (sampling error, detection error,
natural variability)? Would more than one source of randomness need to
be chained together?*

\newpage
\large

# PART 2 -- PODIUM CHEAT SHEET

\normalsize
*Podium copy -- terse on purpose. Full derivations, rationale, and
code: see Part 1 above.*
\large

## 0. Open (~2 min)

- Recap Popperian loop: THEORY $\to$ MODEL $\to$ DATA $\to$ COMPARE
  $\to$ repeat -- a synthetic data generator makes a candidate model
  concrete enough to *run*

## 1. Why simulate? (~8 min)

- **Validate** model-fitting (you know the true answer, so you can
  check recovery)
- **Power analysis** (how often would this design detect a real
  effect?)
- **Understanding** (chaining functions reveals consequences of
  assumptions)
- **Forecasting** (run the validated model forward under new
  conditions)
- Not a replacement for real data -- a diagnostic used alongside it

## 2. Anatomy: deterministic + stochastic (~10-12 min)

```{r podium-linreg, echo=FALSE, fig.width=3.3, fig.height=2.3}
par(mar = c(4, 3, 1, 0.5))
x <- 1:20; mu <- 2 + 1 * x
set.seed(1001); y <- rnorm(length(x), mu, 2)
plot(x, y, pch = 16, cex = 0.7, xlab = "x", ylab = "y")
lines(x, mu, lwd = 2)
```

- $\hat\mu_i = a+bx_i$ (deterministic/"signal") $+$
  $Y_i\sim\text{Normal}(\hat\mu_i,\sigma)$ (stochastic/"noise")
- Distribution parameter = deterministic function's output (Monday's
  $\mu$, now a function of $x$)
- **Sticking point:** not always additive -- a hyperbolic $\times$
  Poisson example shrinks noise right along with the shrinking mean
  (equidispersion)

## 3. Synthetic data as a goodness-of-fit check (~10-12 min)

```{r podium-gof, echo=FALSE, fig.width=3.3, fig.height=2.6}
tree_data <- data.frame(girth = seq(1, 10, length.out = 10),
                         volume = c(0.4, 2.7, 5.3, 13.3, 42.4,
                                    63.5, 45.8, 233.4, 213.7, 383.1))
sim_volume <- replicate(500, rnorm(10, mean = 10 + 40 * tree_data$girth, sd = 31))
par(mar = c(4, 3, 1, 0.5))
boxplot(t(sim_volume), names = round(tree_data$girth), cex.axis = 0.6,
        xlab = "Girth", ylab = "Volume")
points(seq_len(10), tree_data$volume, pch = 16, col = "red")
```

- Propose a fully-specified hypothesis, simulate many datasets under
  it, overlay real data -- does it fall inside the plausible range?
- **Sticking point:** plausibility $\neq$ proof -- "consistent with"
  is weaker than "correct" (corroborate, don't prove)

## 4. Roadmap (~2 min)

- Next Tuesday (Lab 4): full power-analysis build, from scratch
- Weeks 6-7: quantitative model comparison replaces "eyeball the
  boxplot"

## Discussion (if time)

- *Designing a monitoring study for your own system: what's the
  deterministic part? What's the stochastic part? More than one source
  of randomness to chain together?*
