---
title: "Week 6, Lecture 2: Live Coding a Power Analysis"
subtitle: "Instructor notes. Part 1: prep reading. Part 2: podium cheat sheet"
fontsize: 12pt
output:
  pdf_document: default
header-includes:
  - \usepackage{enumitem}
  - \setlist[itemize]{itemsep=4pt, parsep=0pt, topsep=4pt}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 5.5, fig.height = 3.5)
```

\fbox{\parbox{0.95\textwidth}{\textbf{Before class:}
\begin{itemize}
\item Print the WOD (\texttt{WOD\_Week6\_Lecture2\_FindTheBugs}); bring its key.
\item Open a blank R script and an AI assistant in a browser tab, both
visible on the projector. Bump the font size in both.
\item Have the Lab 4, Act 2 handout open for the checks in each part.
\item Optional safety net: run the grid chunk below the night before and
save it (\texttt{saveRDS(grid, "grid.rds")}) in case the live run is slow.
\end{itemize}}}

This file has two parts, both instructor-only. **Part 1** is read
beforehand: the full working code with real output, the plan for which
pieces to type by hand and which to hand to an AI, and a set of good
and bad prompts to explore with the class. **Part 2** (after the page
break, large print) is the podium copy.

| Beat | Time |
|---|---|
| WOD: find the bugs | ~5 min |
| Live coding: Lab 4, Act 2 | ~40 min |
| Discussion | ~5 min |

Source material: `labs/Lab4_Act2.Rmd` and its key
(`labs/answer_keys/Lab4_Act2_Key.Rmd`). All code below matches the key,
so the numbers agree with what groups should have found.

# Where this sits in the week

- **Last week, Lab 4:** Act 1 validated the log-scale trend test. Act 2
  (the monitoring design problem) never started in lab.
- **Monday:** WOD debrief, course check-in, and an overview of maximum
  likelihood.
- **Tuesday, Lab 5:** likelihood by hand (Act 1), then a first
  `optim()` fit (Act 2). Act 2 introduced the new AI policy: each
  coding question is labeled "Recommended: no AI" or "AI optional/encouraged," students
  disclose their AI use after each one, and they turn in outputs, not
  code.
- **Today:** a step back to build Lab 4, Act 2 live. Two goals: get
  students' coding going again after two math-heavy days, and show
  in practice what "build and test one function at a time" means,
  with and without an AI assistant. The demo mirrors the new policy:
  the core functions by hand, the repetitive parts with AI.

# Rationale for the flow

- **Type the first four functions by hand.** They are short, and each
  one has a check from the Lab 4 handout. Typing them live lets you say
  each check out loud before running it. This is the habit to model.
  WOD bug 1a is in the first of these functions, and two more bugs
  that didn't make the WOD (listed below) fit naturally into the
  others.
- **Hand `estimate_power()` and the design grid to an AI.** This is
  where the code gets longer and more repetitive, which is where AI
  help is most useful. Try a bad prompt first, then a good one, and
  test whatever comes back against numbers you already know (the
  false-decline rate and the power at one design).
- **Don't aim to finish the cost table live.** If time is short, show
  the precomputed grid and the cheapest designs, and point to the key's
  recommendation. The point of today is process, not the final answer.

# The WOD (~5 min)

Three minutes alone, two minutes with a neighbor on Part 2, then
straight into the demo. The two fixes:

- 1a, `detect_prob()`: `1 - pobs^k` should be `1 - (1 - pobs)^k`.
  Check: `detect_prob(1)` should equal 0.02.
- 1b, `estimate_power()`: `detected <-` should be `detected[i] <-`.
  Check: the power should never come out exactly 0 or 1.

Both bugs run without an error and give plausible-looking wrong
answers. Say so as you move into the demo: "it runs" isn't a test, so
every function today gets a check like these before we move on.

Two more bugs that didn't make the WOD, to raise while typing the
matching function in part A:

- `simulate_counts()`: `rbinom(1, ...)` instead of
  `rbinom(length(t_survey), ...)` returns one count, not 25.
- `detect_decline()`: `||` instead of `&&` counts any negative slope
  (or any significant increase) as a decline.

# Live coding, part A: four functions by hand (~15 min)

Put subheadings in the script as you go (`# Detection probability ----`
and so on), and say why: it makes the script easy to navigate, and it
mirrors how the pieces were planned on paper.

**Detection probability.** A bout of $k$ person-days detects an animal
unless every day misses it: $q(k) = 1 - (1 - p_{obs})^k$.

```{r detect-prob}
detect_prob <- function(k, pobs = 0.02) {
  1 - (1 - pobs)^k
}
detect_prob(1)   # should equal pobs
detect_prob(3)   # handout check: about 0.0588
```

Two checks: the handout's value, and the edge case $k = 1$, where the
answer must equal `pobs`. Mention that the edge case would have caught
WOD bug 1a.

**True abundance.** $N_t = N_0\lambda^t$, with $\lambda^{25} = 0.75$.

```{r true-abundance}
true_abundance <- function(t, N0 = 1000, lambda = 0.75^(1/25)) {
  N0 * lambda^t
}
true_abundance(c(0, 25))   # should return 1000 and 750 together
```

Test it with a *vector* of years, not one at a time. That is how the
next function will call it.

**Simulated counts.** Each survey's count is
Binomial($N_t$, $q(k)$).

```{r simulate-counts}
simulate_counts <- function(t_survey, k, N0 = 1000, lambda = 0.75^(1/25),
                            pobs = 0.02) {
  N_t <- true_abundance(t_survey, N0, lambda)
  rbinom(length(t_survey), size = round(N_t), prob = detect_prob(k, pobs))
}
set.seed(11)
counts <- simulate_counts(1:25, k = 3)
length(counts)   # should be 25
```

```{r simulate-counts-plot, fig.height=3}
par(mar = c(4, 4.5, 0.5, 0.5))
plot(1:25, counts, pch = 16, xlab = "Year", ylab = "Count",
     ylim = c(0, max(counts) * 1.1))
```

Things to say out loud: `round()` because `size` must be a whole
number; `length(t_survey)` draws, not `1` (the extra bug above); counts sit far below
the true abundance (1000 falling to 750) because detection is under
6%. Plot it. A quick picture is a test too.

**Decision rule.** Negative slope **and** $p \le \alpha$.

```{r detect-decline}
detect_decline <- function(t_survey, counts, alpha = 0.05) {
  s <- summary(lm(log(counts) ~ t_survey))$coefficients[2, ]
  unname(s[1] < 0 && s[4] <= alpha)
}
detect_decline(1:25, counts)
```

Now a test with known truth: simulate a stable population
($\lambda = 1$) many times. The rule should fire only about 2.5% of the
time, half of 5%, since significant *increases* don't count.

```{r detect-decline-null}
set.seed(12)
null_hits <- replicate(1000, {
  detect_decline(1:25, simulate_counts(1:25, 3, lambda = 1))
})
mean(null_hits)
```

This is the check that would catch the `||` bug: with "or," it would
come out near 50%.

# Live coding, part B: bring in the AI (~12 min)

Switch to the AI assistant for `estimate_power()`. Run the bad prompt
first (prompt pair 1 or 2 below), ask the class what they expect to go
wrong, then run the good prompt. Whatever code comes back, test it
against two numbers you already have:

1. **False-decline rate** at $\lambda = 1$, $k = 3$, $s = 1$: about
   0.02 to 0.03 (just computed above).
2. **Power** at $k = 3$, $s = 1$: about 0.8 (from the Lab 4 key).

A reference version to compare against:

```{r estimate-power}
estimate_power <- function(k, s, R = 1000, lambda = 0.75^(1/25),
                           alpha = 0.05) {
  t_survey <- seq(1, 25, by = s)
  mean(replicate(R, {
    counts <- simulate_counts(t_survey, k, lambda = lambda)
    detect_decline(t_survey, counts, alpha)
  }))
}
set.seed(14)
estimate_power(3, 1, lambda = 1)   # false-decline rate
estimate_power(3, 1)               # power
run_time <- system.time(estimate_power(3, 1))["elapsed"]
run_time
```

Note the run time: about `r round(run_time, 1)` s per design with
`R = 1000` on the machine that knit these notes, so the full 75-design
grid takes roughly `r round(75 * run_time)` s here, and likely longer on
a classroom laptop. For the live demo, use `R = 200` and say why: fast
and rough first, then more replicates only where precision matters.

**Things to look for in AI output** (ask the class to spot them):

- Does it use the functions we already wrote, or rewrite everything
  from scratch (possibly with a different model)?
- Is the survey schedule right: `seq(1, 25, by = s)`, not
  `seq(1, 25, length.out = s)`?
- Does each replicate simulate *new* counts, or does it reuse one
  dataset?
- Does it store every replicate's result (WOD bug 1b)?
- Did it quietly add something we didn't ask for (a package, a
  different test, a `set.seed()` inside the function)?

# Live coding, part C: the design space (~13 min)

```{r grid, cache=TRUE}
set.seed(746)
grid <- expand.grid(k = 1:15, s = 1:5)
grid$power <- mapply(estimate_power, grid$k, grid$s,
                     MoreArgs = list(R = 1000))
```

```{r grid-plot, fig.height=3.5}
par(mar = c(4, 4.5, 0.5, 0.5))
plot(NA, xlim = c(1, 15), ylim = c(0, 1), xlab = "k (person-days per bout)",
     ylab = "Estimated power")
abline(h = 0.75, lty = 2, col = "grey40")
for (ss in 1:5) {
  lines(grid$k[grid$s == ss], grid$power[grid$s == ss], col = ss, lwd = 2)
}
legend("bottomright", legend = paste("s =", 1:5), col = 1:5, lwd = 2,
       bty = "n", cex = 0.8)
```

Cost is \$200 per person-day plus \$2,000 per expedition:

```{r cost}
cost_of <- function(k, s) length(seq(1, 25, by = s)) * (2000 + 200 * k)
grid$cost <- mapply(cost_of, grid$k, grid$s)
ok <- grid[grid$power >= 0.75, ]
head(ok[order(ok$cost), ], 5)
```

Points to make:

- Power rises with $k$ and falls with $s$, and the two trade off at
  similar rates across most of the grid (compare with question 1d in the Lab 4, Act 2 handout).
- The cheapest designs sit right at 0.75. With 1,000 replicates, the
  Monte Carlo standard error near 0.75 is about
  $\sqrt{0.75 \times 0.25 / 1000} \approx 0.014$, so a design at 0.76
  could easily be below the line. The key re-checks the top candidates
  with 5,000 replicates and recommends $k = 11$, $s = 4$ (\$29,400),
  which clears the bar with margin, over the cheapest $k = 10$, $s = 4$.
- If time allows, ask: would you trust the recommendation more if the
  AI had written the whole grid in one go? Why or why not?

# Good and bad AI prompts to explore together

Each pair targets the same task. Show the bad prompt, ask what the
class expects to go wrong, run it if time allows, then show the better
version and ask what changed. You don't need to run all five: pairs 1,
2, and 4 fit best into part B.

**Pair 1: the whole problem at once**

- *Bad:* "Write R code for a power analysis for a declining toad
  population."
- *Better:* "Write an R function `estimate_power(k, s, R)` that
  estimates the power of a monitoring design. It should call my
  existing functions `simulate_counts(t_survey, k)` and
  `detect_decline(t_survey, counts)`. Survey years are
  `seq(1, 25, by = s)`. For each of `R` replicates, simulate new
  counts and record whether a decline was detected. Return the
  fraction detected. Base R only."
- *Why:* the bad prompt leaves the model unstated, so the AI fills it
  in with its own defaults (a canned t-test, a `pwr` package call, a
  Poisson or normal model, a different decision rule). The code may
  run and look professional while answering a different question.

**Pair 2: the model in words vs. the model in equations**

- *Bad:* "Simulate survey counts for a population that declines 25%
  over 25 years with imperfect detection."
- *Better:* "Write `simulate_counts(t_survey, k, N0 = 1000, lambda =
  0.75^(1/25), pobs = 0.02)`. True abundance is
  $N_t = N_0 \lambda^t$. Detection probability for a bout of `k`
  person-days is $1 - (1 - p_{obs})^k$. Each count is
  Binomial($N_t$ rounded, detection probability). Return one count per
  element of `t_survey`."
- *Why:* "declines 25%" could mean linear or exponential decline;
  "imperfect detection" could be modeled many ways. If you've already
  worked out the equations on paper, give them to the AI.

**Pair 3: debugging**

- *Bad:* "Fix my code." (pasting the whole script)
- *Better:* "This function should return 25 counts, one per survey
  year, but `length(simulate_counts(1:25, 3))` returns 1. Here is the
  function. Explain why it returns one value before suggesting a fix."
- *Why:* the better prompt gives the symptom, the expected behavior,
  and a minimal example, and it asks for a diagnosis first. "Fix my
  code" invites the AI to rewrite everything, possibly introducing new
  bugs, and you learn nothing about what went wrong.

**Pair 4: testing**

- *Bad:* "Is this code correct?"
- *Better:* "Write three quick tests for `detect_decline()` with known
  answers: a clearly declining series should return `TRUE`, a flat
  series `FALSE`, and a clearly increasing series `FALSE`. Then tell me
  what fraction of `TRUE`s I should expect if I run it on many
  simulated datasets with no decline, and why."
- *Why:* "Is this correct?" usually gets a confident yes plus a
  paraphrase of the code. Asking for tests with specific expected
  values turns the question into something you can check. Also check
  the tests themselves: an AI-written test sometimes compares the
  function to itself, which always passes.

**Pair 5: leading questions**

- *Bad:* "Show that survey frequency matters more than effort per
  bout."
- *Better:* "Using `estimate_power()`, compute power for
  $k = 1, \ldots, 15$ and $s = 1, \ldots, 5$, and plot power against
  $k$ with one line per $s$. Don't interpret the results yet."
- *Why:* a leading prompt invites the AI to find support for the
  conclusion you handed it. Ask for the computation, then interpret it
  yourself.

# General guidance: AI for writing and testing code

These points are for the podium and for students. They restate what
the lab handouts already ask for (plan on paper, one function at a
time, record prompts) with reasons.

**Before prompting**

- **Do the thinking first.** Work out the model and the algorithm on
  paper (the pseudocode step in every lab). The AI is good at turning a
  clear plan into code, and poor at choosing the right model for your
  question.
- **Decide your tests before you see any code.** Know what numbers
  should come out (an edge case, a known-truth scenario, a value from
  the handout). Otherwise it's easy to accept whatever looks
  reasonable.

**While prompting**

- **One function per prompt.** Give its name, arguments, what it
  returns, and the equations it implements.
- **Give the context.** Existing functions it should call, the data's
  structure, allowed packages (base R here).
- **Ask for explanations.** "Explain each line" and "What assumptions
  did you make?" often surface a substitution you didn't ask for.
- **Don't paste sensitive data** (unpublished data, locations of
  sensitive species, anything covered by a data agreement). Use a small
  made-up example with the same structure.

**After the code comes back**

- **Read it before running it.** Could you explain every line to a
  classmate? If not, ask, or look it up.
- **Run your tests.** Check shapes (`length()`, `dim()`), edge cases,
  and known-truth scenarios, not just "it runs without errors."
- **Debug with evidence.** Give the error message, a minimal example,
  and what you expected, and ask for a diagnosis before a fix.
- **Disclose it.** Note where you used AI and what you changed (the
  AI-use lines in Lab 5, Act 2 ask for this). Using AI never costs
  points in this course; hiding it does.
- **You own the result.** If AI-written code gives a wrong power
  estimate and the agency buys the wrong design, "the AI wrote it" is
  not an answer.

# Discussion (~5 min)

Pick one:

- *Which of today's checks would you have skipped if you were in a
  hurry? What would that have cost you?*
- *Where did the AI help most today, and where would you not trust it
  without checking?*
- *Tuesday's Act 2 marked some coding tasks "Recommended: no AI." Which of today's
  tasks would you have labeled that way, and why?*

\newpage
\large

# PART 2: PODIUM CHEAT SHEET

\normalsize
\textit{Terse on purpose. Full code, output, and rationale: Part 1.}
\large

**Setup:** R script + AI tab on projector, big fonts. `grid.rds` saved
as backup.

## WOD: find the bugs (~5 min)

- 3 min alone, 2 min neighbor
- 1a `1 - pobs^k` $\to$ `1 - (1 - pobs)^k`; check `detect_prob(1)` = 0.02
- 1b `detected <-` $\to$ `detected[i] <-`; power never exactly 0 or 1
- Both silent: "it runs" isn't a test

\newpage

## A. Four functions by hand (~15 min)

| Function | Check (say it before running) |
|---|---|
| `detect_prob(k)` | `detect_prob(1)` = 0.02; `detect_prob(3)` $\approx$ 0.0588 |
| `true_abundance(t)` | `c(0, 25)` $\to$ 1000, 750 together |
| `simulate_counts(t, k)` | `length()` = 25; plot it; counts $\ll$ $N_t$ |
| `detect_decline(t, y)` | $\lambda = 1$, many reps $\to$ TRUE $\approx$ 2.5% |

- Subheadings as you go
- Extra bugs to raise: `rbinom(1, ...)` in `simulate_counts`; `||` in
  `detect_decline` ($\to$ ~50% under $\lambda = 1$)
- `round(N_t)`: `size` must be whole

\newpage

## B. Bring in the AI (~12 min)

- `estimate_power(k, s, R)`: bad prompt first, class predicts problems,
  then good prompt
- Test against known numbers:
    - $\lambda = 1$, $k = 3$, $s = 1$: $\approx$ 0.02-0.03
    - $k = 3$, $s = 1$: $\approx$ 0.8
- Look for: reuses our functions? `by = s`? new counts each rep? stores
  every rep? added anything unasked?
- `system.time()` $\to$ use `R = 200` live

**Prompt pairs** (use 1, 2, 4):

1. "Power analysis for a declining toad" vs. named function, calls our
   functions, `by = s`, base R
2. "Declines 25%, imperfect detection" vs. the equations
3. "Fix my code" vs. symptom + expected + minimal example + diagnose first
4. "Is this correct?" vs. three tests with known answers
5. "Show that $s$ matters more" vs. compute and plot, don't interpret

\newpage

## C. Design space (~13 min)

```{r podium-grid, echo=FALSE, fig.width=4.5, fig.height=2.8}
par(mar = c(4, 4.5, 0.5, 0.5))
plot(NA, xlim = c(1, 15), ylim = c(0, 1), xlab = "k", ylab = "Power")
abline(h = 0.75, lty = 2, col = "grey40")
for (ss in 1:5) lines(grid$k[grid$s == ss], grid$power[grid$s == ss], col = ss, lwd = 2)
legend("bottomright", legend = paste("s =", 1:5), col = 1:5, lwd = 2, bty = "n", cex = 0.7)
```

- Cost = bouts $\times$ (\$2,000 + \$200 $k$)
- $k$ and $s$ trade off at similar rates
- Cheapest designs sit on 0.75; MC SE $\approx$ 0.014 at R = 1000
- Key's pick: $k = 11$, $s = 4$, \$29,400 (margin over cheapest)

\newpage

## AI guidance (weave in throughout)

- Think first: model + pseudocode on paper
- Decide tests before seeing code
- One function per prompt: name, args, return, equations
- Give context: existing functions, data, base R
- Ask "explain each line" / "what did you assume?"
- No sensitive data; use a toy example
- Read before running; run your tests
- Debug with evidence; diagnose before fixing
- Disclose AI use; it never costs points, hiding it does
- You own the result

## Discussion (~5 min)

- Which check would you skip in a hurry? Cost?
- Where did AI help most? Where not trust it?
- Which of today's tasks would you label "Recommended: no AI"? Why?
