---
title: "Why Likelihood Intervals Work: A Graphical Guide"
subtitle: "Reference notes on the score, Fisher information, and the sampling distribution of the MLE"
fontsize: 12pt
output:
  pdf_document: default
  # word_document: default
header-includes:
  - \usepackage{enumitem}
  - \setlist[itemize]{itemsep=3pt, parsep=0pt, topsep=3pt}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 3.2, fig.align = "center")

lambda0 <- 3
lam <- seq(1.4, 6, length.out = 400)
nll <- function(l, y) -sum(dpois(y, l, log = TRUE))
score <- function(l, y) sum(y) / l - length(y)
```

These notes explain why the standard error of a maximum likelihood
estimate is $1/\sqrt{\text{curvature}}$, why the MLE is approximately
normal, and where the 1.92 cutoff for likelihood ratio intervals comes
from. The approach is graphical; the only math needed is a derivative
and the central limit theorem.

The running example is the pika survey from lecture: active haypiles
counted in $n$ talus plots, each count Poisson with mean $\lambda$. To
see how estimates behave across repeated surveys, we simulate surveys
from a known truth, $\lambda_0 = 3$.

# 1. Two views of the same curve

For one survey, the negative log-likelihood (NLL) is a valley whose
bottom is the MLE, $\hat\lambda$. Its slope, with the sign flipped so
that it points uphill on the log-likelihood, is called the **score**:
$$U(\lambda) = \frac{d}{d\lambda}\log\mathcal{L}(\lambda) = -\frac{d}{d\lambda}\text{NLL}(\lambda).$$
For the Poisson, $U(\lambda) = \sum y_i/\lambda - n$.

For the data 2, 5, 3, 0, 4, 4 (figure below):

- The bottom of the valley (top panel) is where the score crosses zero
  (bottom panel). Both mark the MLE, $\hat\lambda = 3$.
- The **curvature** of the valley is the **steepness of the score**.
  A sharp valley means a steep score line; a broad valley means a
  shallow one.
- Near the MLE, a **parabola-shaped valley** and a **straight-line
  score** are the same statement. Everything below depends on it.

The expected steepness of the score (the expected curvature of the NLL)
is the **Fisher information**, $I$. For the Poisson, $I = n/\lambda$.

```{r fig-two-views, fig.height=4.4}
set.seed(1)
y1 <- c(2, 5, 3, 0, 4, 4)
lam_hat1 <- mean(y1)
d_nll <- sapply(lam, nll, y = y1) - nll(lam_hat1, y1)
u1 <- sapply(lam, score, y = y1)

par(mfrow = c(2, 1), mar = c(2, 4.5, 1.5, 0.5), oma = c(2, 0, 0, 0))
plot(lam, d_nll, type = "l", lwd = 2, ylim = c(0, 5), xlab = "",
     ylab = "NLL minus its minimum", main = "Negative log-likelihood", cex.main = 0.9)
abline(v = lam_hat1, lty = 3)
plot(lam, u1, type = "l", lwd = 2, col = "steelblue", xlab = "",
     ylab = "Score U", main = "Score (slope of the log-likelihood)", cex.main = 0.9)
abline(h = 0)
abline(v = lam_hat1, lty = 3)
points(lam_hat1, 0, pch = 16, col = "red")
mtext("lambda", side = 1, line = 0.5, outer = TRUE)
```


# 2. Many surveys, many score curves

A new survey gives new counts, so a new score curve. The figure shows
the score from 15 simulated surveys at two sample sizes.

```{r fig-many, fig.height=3.4}
score_panel <- function(n, n_sims = 15, ylim = c(-12, 12)) {
  info <- n / lambda0
  plot(NA, xlim = c(1.6, 5.5), ylim = ylim, xlab = "lambda", ylab = "Score U",
       main = paste0(n, " plots (I = ", round(info, 1), ")"), cex.main = 0.9)
  abline(h = 0, lwd = 1.5)
  abline(v = lambda0, lty = 3, col = "grey40")
  for (i in seq_len(n_sims)) {
    y <- rpois(n, lambda0)
    lines(lam, sapply(lam, score, y = y), col = adjustcolor("steelblue", 0.6))
    points(mean(y), 0, pch = 16, col = "red", cex = 0.8)
    points(lambda0, score(lambda0, y), pch = 16, col = "steelblue4", cex = 0.6)
  }
  yy <- seq(ylim[1], ylim[2], length.out = 200)
  lines(lambda0 + 0.55 * dnorm(yy, 0, sqrt(info)) / dnorm(0, 0, sqrt(info)), yy,
        col = "steelblue4", lwd = 2)
  xx <- seq(1.6, 5.5, length.out = 300)
  dx <- dnorm(xx, lambda0, sqrt(1 / info))
  lines(xx, ylim[1] + 5 * dx / max(dx), col = "red", lwd = 2)
}
set.seed(746)
par(mfrow = c(1, 2), mar = c(4, 4, 2, 0.5))
score_panel(6)
score_panel(24)
```

Read it in four steps.

1. **Each curve crosses zero at that survey's MLE** (red dots). The
   crossings spread along the axis: that spread is the sampling
   distribution of $\hat\lambda$.
2. **At the true value, look up and down.** The blue dots are each
   curve's height at $\lambda_0 = 3$. They are centered on 0, because
   at the truth the log-likelihood is flat on average.
3. **Those heights are normal.** The score is a sum over plots,
   $U(\lambda_0) = \sum_i (y_i/\lambda_0 - 1)$, so the central limit
   theorem applies. Their variance turns out to be $I$ (section 4
   explains why), drawn as the blue bell.
4. **Each curve slopes down at about $-I$.** Section 3 turns this into
   the spread of the red dots.

# 3. Turning vertical spread into horizontal spread

Zoom in near the truth, where each score curve is close to a straight
line with slope $-I$.

```{r fig-shift, fig.height=3}
info6 <- 6 / lambda0
par(mfrow = c(1, 2), mar = c(4, 4, 2, 0.5))
plot(NA, xlim = c(2.4, 4.1), ylim = c(-1.5, 2.2), xlab = "lambda", ylab = "Score U",
     main = "Shift = height / slope", cex.main = 0.9)
abline(h = 0)
abline(v = lambda0, lty = 3, col = "grey40")
h <- 1.6
for (s in c(info6, 2.5 * info6)) {
  xs <- seq(2.4, 4.1, length.out = 50)
  lines(xs, h - s * (xs - lambda0), lwd = 2, col = if (s == info6) "steelblue" else "darkorange")
  points(lambda0 + h / s, 0, pch = 16, col = "red")
}
points(lambda0, h, pch = 16, col = "steelblue4")
legend("topright", c("shallow (small I)", "steep (large I)"), lwd = 2,
       col = c("steelblue", "darkorange"), bty = "n", cex = 0.75)

set.seed(746)
sims_n <- 24
hats <- replicate(4000, mean(rpois(sims_n, lambda0)))
hist(hats, breaks = seq(1.5, 4.6, by = 1 / sims_n), freq = FALSE, col = "grey85",
     border = "white", main = "MLE across 4,000 surveys (24 plots)", cex.main = 0.9,
     xlab = "MLE of lambda")
curve(dnorm(x, lambda0, sqrt(lambda0 / sims_n)), add = TRUE, lwd = 2, col = "red")
```

- **Left:** a curve that sits a height $h$ above zero at the truth has
  to travel $h/I$ to the right before it reaches zero. A steeper line
  (larger $I$) gets there sooner, so its MLE lands closer to the truth.
- The heights have variance $I$. Dividing by the constant slope $I$
  gives the MLEs a variance of
  $$\frac{\text{Var}(U)}{I^2} = \frac{I}{I^2} = \frac{1}{I}.$$
- **Right:** across 4,000 simulated surveys of 24 plots, the MLEs
  follow the predicted $\text{Normal}(\lambda_0,\ 1/I) =
  \text{Normal}(3,\ 3/24)$ (red curve).

**More data.** With 24 plots instead of 6, the heights at the truth
spread *more* (their variance $I$ grows with $n$), but the lines get
steeper just as fast (their slope is also $I$). Spread grows like
$\sqrt{n}$, slope like $n$, so the MLEs tighten like $1/\sqrt{n}$. This
also settles a common puzzle: more data lowers the maximum likelihood
itself, but what matters for the estimate is the steepness, not the
height.

**Check with the pikas.** Here the straight-line step is exact:
$$\hat\lambda - \lambda_0 = \frac{U(\lambda_0)}{n/\lambda_0}
 = \frac{\sum y_i/\lambda_0 - n}{n/\lambda_0} = \bar y - \lambda_0,
 \qquad \text{Var}(\hat\lambda) = \frac{\lambda_0}{n} = \frac{1}{I}.$$

# 4. Why the variance of the score equals its slope

Section 3 needed the variance of the score and its slope to be the same
number, $I$. This is the **information identity**, and the score
curves show why it has to hold: **the curves must move when the truth
moves.**

```{r fig-identity, fig.height=3.2}
n_id <- 6
delta <- 0.1
info_id <- n_id / lambda0
exp_score <- function(l, lam_true) n_id * lam_true / l - n_id
par(mar = c(4, 4, 1.5, 0.5))
plot(lam, exp_score(lam, lambda0), type = "l", lwd = 2, col = "steelblue",
     xlim = c(2.6, 3.5), ylim = c(-0.9, 0.9), xlab = "lambda",
     ylab = "Average score")
lines(lam, exp_score(lam, lambda0 + delta), lwd = 2, col = "darkorange")
abline(h = 0)
abline(v = c(lambda0, lambda0 + delta), lty = 3, col = "grey40")
gap <- exp_score(lambda0, lambda0 + delta)
arrows(lambda0 - 0.02, 0, lambda0 - 0.02, gap, length = 0.06, lwd = 2)
text(lambda0 - 0.03, gap / 2, "gap (route 2)", pos = 2, cex = 0.8)
l_seg <- seq(lambda0, lambda0 + delta, length.out = 20)
lines(l_seg, exp_score(l_seg, lambda0 + delta), lwd = 5, col = "darkorange3")
text(lambda0 + delta / 2, gap / 2 + 0.03, "route 1", pos = 4, cex = 0.8)
legend("topright", c("data from lambda = 3", "data from lambda = 3.1"),
       lwd = 2, col = c("steelblue", "darkorange"), bty = "n", cex = 0.8)
```

The blue curve is the average score when the data come from
$\lambda_0 = 3$; it crosses zero at 3. Now nudge the truth by
$\delta = 0.1$, to 3.1 (orange). The new curve must cross zero at 3.1,
or the MLE wouldn't track the truth. The **gap** is how far the
average score at $\lambda = 3$ rises. There are two ways to get it.

**Route 1: what the slope requires.** The curves fall at slope $-I$, so
to cross zero $\delta$ further right, the new curve must sit about
$\delta \cdot I$ above zero at 3. For six plots, $I = 6/3 = 2$, so the
gap should be $0.1 \times 2 = 0.2$.

**Route 2: what reweighting the surveys delivers.** Keep reading the
score at $\lambda = 3$. For six plots with total count $S$, that's
$U(3) = S/3 - 6$, a fixed formula applied to each survey. Changing the
truth doesn't change the formula; it changes **how often each survey
turns up**, and the average is a probability-weighted sum over surveys.

How much more likely does a survey get? The score is, by definition,
how fast a survey's log-probability rises as $\lambda$ increases. So a
small step $\delta$ multiplies its probability by about
$e^{\delta U} \approx 1 + \delta U$:

```{r reweight-table}
s_ex <- c(12, 18, 24)
u_ex <- s_ex / lambda0 - n_id
ratio_ex <- dpois(s_ex, n_id * (lambda0 + delta)) / dpois(s_ex, n_id * lambda0)
knitr::kable(data.frame(s_ex, u_ex, sprintf("%.3f", ratio_ex),
                        sprintf("%.1f", 1 + delta * u_ex)),
             col.names = c("Total count S", "Score at 3, U", "Exact ratio",
                           "1 + 0.1 U"),
             align = "c")
```

Surveys whose score points toward a larger $\lambda$ become more common,
and surveys pointing lower become rarer. The new average score at 3 is
then
$$\sum U \cdot P_3 \cdot (1 + \delta U)
  = \underbrace{E[U]}_{0} + \delta\, E[U^2] = \delta \cdot \text{Var}(U).$$
The variance appears because reweighting only moves the average when
surveys disagree. If every survey had a score of 0, tilting the weights
would change nothing; the more the scores spread, the more the average
moves. For six plots, $\text{Var}(U) = \text{Var}(S)/9 = 18/9 = 2$, so
the gap is again $0.1 \times 2 = 0.2$.

Both routes describe the same gap, so $\delta \cdot I = \delta \cdot
\text{Var}(U)$, which means $\text{Var}(U) = I$. In the figure, the gap
is `r round(gap, 2)`.

**The path in the figure.** Start where the blue curve crosses zero, at
3, and get to where the orange curve crosses zero, at 3.1, in two
moves: up the gap (switching the truth from 3 to 3.1 while still
reading the score at 3; route 2), then down along the orange curve
from 3 to 3.1 (moving the reading point with the truth fixed; route 1,
the thick segment). The path starts and ends on the zero line, so the
two moves cancel: $\delta \cdot \text{Var}(U) - \delta \cdot I = 0$.
The calculus version below is this path, made exact.

## The calculus version

Write $f(y; \lambda)$ for the probability of a whole survey $y$ (all
six counts) when the true mean is $\lambda$. Because the counts are
discrete, "averaging over surveys" means summing over every possible
survey, weighted by $f$. The score is
$$U(y; \lambda) = \frac{d}{d\lambda}\log f(y; \lambda) = \frac{f'}{f},
\qquad\text{so}\qquad f' = U f,$$
where a prime means a derivative with respect to $\lambda$.

**Step 1: the score averages to zero at the truth.** The claim is that
the expected score, a sum over every possible outcome weighted by its
probability, is zero when the score is read at the true $\lambda$:
$$E[U] = \sum_y U(y; \lambda)\, f(y; \lambda) = 0.$$

*One count first.* For a single plot, $f(y; \lambda) = \lambda^y
e^{-\lambda}/y!$ for $y = 0, 1, 2, \ldots$, and the log is $y \log\lambda
- \lambda - \log y!$, so the score for one count is $U = y/\lambda - 1$.
A count of 3 says nothing ($U = 0$ at $\lambda = 3$), a count of 6 says
"$\lambda$ is higher" ($U = 1$), and a count of 0 says "$\lambda$ is
lower" ($U = -1$). Weighting each possible count by its probability at
$\lambda = 3$:

```{r step1-table}
y_one <- 0:60
f_one <- dpois(y_one, lambda0)
u_one <- y_one / lambda0 - 1
uf_one <- u_one * f_one
show <- 1:7
tab_one <- data.frame(
  y = c(y_one[show], "7+"),
  f = c(sprintf("%.3f", f_one[show]), ""),
  u = c(sprintf("%.3f", u_one[show]), ""),
  uf = sprintf("%.3f", c(uf_one[show], sum(uf_one[-show])))
)
knitr::kable(tab_one, align = "c",
             col.names = c("Count y", "Probability f(y; 3)",
                           "Score U = y/3 - 1", "U times f"))
neg_part <- sum(uf_one[u_one < 0])
pos_part <- sum(uf_one[u_one > 0])
```

The negative terms add to `r sprintf("%.3f", neg_part)` and the
positive terms to `r sprintf("+%.3f", pos_part)`. Low counts point
down, high counts point up, and weighted by how often each turns up,
they cancel. (Directly: $E[U] = E[y]/3 - 1 = 3/3 - 1 = 0$, since a
Poisson's mean is $\lambda$.)

*Why it has to be zero, for any model.* Whatever $\lambda$ is, the
probabilities of all possible outcomes add to 1:
$\sum_y f(y; \lambda) = 1$. Something that stays at 1 has derivative 0,
so
$$\sum_y f' = \sum_y U f = 0.$$
Each $f'$ is how fast one outcome's probability changes as $\lambda$
increases. Some go up (high counts) and some go down (low counts), but
the total can't change, so the changes cancel. Since $f' = Uf$, the
last column of the table is $f'$: the change in each count's
probability per unit of $\lambda$. So "the score averages to zero" is
another way of saying "probabilities always add to 1." This is why the
blue dots in section 2 are centered on 0.

*The condition that matters: "at the truth."* Nothing in the algebra
above fixes $\lambda$: the substitution $f' = Uf$ is just the
definition of $U$, so $\sum_y U(y; \lambda)\, f(y; \lambda) = 0$ holds
for every $\lambda$. What it requires is that the **same** $\lambda$
appears in the score and in the weights. The truth enters when we read
the sum as an average over real surveys: the weights $f(y; \lambda)$
are how often each survey turns up only if $\lambda$ is the value that
generated the data. So the result says "if the data come from
$\lambda$, the score read at that same $\lambda$ averages to zero," and
for real data that means reading the score at the true value.

If the two don't match, the argument doesn't apply: $U(y; a)\, f(y; b)$
with $a \neq b$ isn't the derivative of anything that adds to 1, and
the average isn't zero. Read the score at 3 when the data come from
3.1 and the average is $3.1/3 - 1 =
`r sprintf("%.3f", 3.1 / 3 - 1)`$. That is the gap from route 2
for one count, $\delta \cdot \text{Var}(U) = 0.1 \times 1/3$ (for one
Poisson count, $\text{Var}(U) = I = 1/\lambda$). For the six-plot
survey every quantity is six times larger, which gives the gap of 0.2
and $I = 2$ in the figure. From here on, $y$ is the whole survey again.

**Step 2: separate the two jobs $\lambda$ is doing.** In
$\sum_y U(y; \lambda)\, f(y; \lambda)$, $\lambda$ appears twice: once as
the point where the score is read (the horizontal axis of the figure),
and once as the truth that generates the surveys (which curve we're
on). Give them separate names:
$$g(a, b) = \sum_y U(y; a)\, f(y; b)
 = \text{average score read at } a \text{, when the data come from } b.$$
The blue curve is $g(a, 3)$ and the orange curve is $g(a, 3.1)$, each
plotted against $a$. Step 1 says $g(\lambda, \lambda) = 0$ for every
$\lambda$: each curve crosses zero at its own truth.

**Step 3: move both together.** As $\lambda$ increases, $g(\lambda,
\lambda)$ stays at 0, so its rate of change is 0. A change in
$g(\lambda, \lambda)$ comes from two sources, the change in $a$ and the
change in $b$, and these add (the chain rule for a function of two
inputs, each changed with the other held fixed):
$$0 = \underbrace{\sum_y U'(y; \lambda)\, f(y; \lambda)}_{\text{change } a \text{, hold } b}
 \;+\; \underbrace{\sum_y U(y; \lambda)\, f'(y; \lambda)}_{\text{change } b \text{, hold } a}.$$

- **The first term is route 1.** With the truth held fixed, it's the
  rate at which the average score falls as the reading point moves:
  the slope of a curve in the figure. It equals $E[U']$, the average
  slope of the score, which is $-I$.
- **The second term is route 2.** With the reading point held fixed,
  it's how fast the average score rises as the truth moves: the gap per
  unit $\delta$. Substituting $f' = U f$ turns it into
  $\sum_y U \cdot U f = E[U^2] = \text{Var}(U)$ (the last step because
  the score averages to 0). The substitution is the reweighting from
  route 2: $f' = Uf$ says a step $\delta$ multiplies each survey's
  probability by about $1 + \delta U$.

So $0 = -I + \text{Var}(U)$, which is the identity. Many texts write
Step 3 in one line, differentiating $\sum U f = 0$ with the product
rule: $\sum (U' f + U f') = 0$. The product rule's two terms are the
two routes.

**Check with the pikas.** For six plots, the average score read at $a$
when the truth is $b$ is $g(a, b) = 6b/a - 6$ (the curves in the
figure). Holding $b = 3$, the slope in $a$ at $a = 3$ is $-6b/a^2 = -2$
(route 1, $-I$). Holding $a = 3$, the rate in $b$ is $6/a = 2$ (route 2,
$\text{Var}(U)$). They cancel.

# 5. The parabola, the interval, and the 1.92 cutoff

Sections 2 and 3 described many surveys. In practice we have one, and
we use the shape of its NLL near the MLE. The second-order Taylor
expansion around $\hat\lambda$ (the first derivative is zero there) is
$$\text{NLL}(\lambda) \approx \text{NLL}(\hat\lambda) +
 \tfrac{1}{2}\, I\, (\lambda - \hat\lambda)^2,$$
a parabola whose curvature is the slope of the score.

```{r fig-parabola, fig.height=3.1}
info_obs <- sum(y1) / lam_hat1^2
quad <- 0.5 * info_obs * (lam - lam_hat1)^2
cut <- qchisq(0.95, 1) / 2
f_cut <- function(l) nll(l, y1) - nll(lam_hat1, y1) - cut
lr_int <- c(uniroot(f_cut, c(0.5, 3))$root, uniroot(f_cut, c(3, 8))$root)
wald_int <- lam_hat1 + c(-1, 1) * qnorm(0.975) / sqrt(info_obs)

par(mfrow = c(1, 2), mar = c(4, 4, 2, 0.5))
plot(lam, d_nll, type = "l", lwd = 2, ylim = c(0, 5), xlab = "lambda",
     ylab = "NLL minus its minimum", main = "NLL and its parabola", cex.main = 0.9)
lines(lam, quad, lwd = 2, lty = 2, col = "blue")
abline(h = cut, col = "red")
abline(v = lr_int, lty = 3)
abline(v = wald_int, lty = 3, col = "blue")
plot(lam, u1, type = "l", lwd = 2, col = "steelblue", ylim = c(-4, 6), xlab = "lambda",
     ylab = "Score U", main = "Score and its straight line", cex.main = 0.9)
lines(lam, -info_obs * (lam - lam_hat1), lwd = 2, lty = 2, col = "blue")
abline(h = 0)
legend("topright", c("exact", "approximation"), lwd = 2, lty = c(1, 2),
       col = c("steelblue", "blue"), bty = "n", cex = 0.75)
```

- **The interval from the parabola.** Setting the rise equal to the
  cutoff, $\tfrac12 I d^2 = 1.92$, gives $d = 1.96/\sqrt{I}$: the
  Hessian (Wald) interval, $\hat\lambda \pm 1.96\,\text{SE}$, here
  [`r round(wald_int[1], 2)`, `r round(wald_int[2], 2)`].
- **Where 1.92 comes from.** Put the true value into the parabola:
  $$2\,[\text{NLL}(\lambda_0) - \text{NLL}(\hat\lambda)] \approx
   I\,(\hat\lambda - \lambda_0)^2 =
   \left(\frac{\hat\lambda - \lambda_0}{1/\sqrt{I}}\right)^2 = Z^2.$$
  Section 3 showed that the quantity in parentheses is a standard
  normal, $Z$, so twice the NLL difference is $\chi^2_1$ (Wilks'
  theorem). Its 95th percentile is 3.84, and half of that is 1.92.
- **The likelihood ratio interval** uses the 1.92 cutoff on the true
  curve instead of the parabola: [`r round(lr_int[1], 2)`,
  `r round(lr_int[2], 2)`]. It follows the curve's lopsidedness, which
  the parabola can't.

So the Hessian interval and the likelihood ratio interval come from the
same idea. They agree when the score is close to a straight line, and
they part ways when it bends, as it does here with only 6 plots.

```{r fig-wilks, fig.height=2.8}
set.seed(746)
dev_stat <- replicate(4000, {
  y <- rpois(24, lambda0)
  2 * (nll(lambda0, y) - nll(mean(y), y))
})
par(mar = c(4, 4, 1.5, 0.5))
hist(dev_stat, breaks = 60, freq = FALSE, col = "grey85", border = "white",
     xlim = c(0, 8), main = "", xlab = "2 x [NLL(true value) - NLL(MLE)], 24 plots")
curve(dchisq(x, 1), add = TRUE, lwd = 2, col = "red", from = 0.05)
abline(v = qchisq(0.95, 1), lty = 2)
```

Across 4,000 simulated surveys, twice the NLL difference at the true
value follows a $\chi^2_1$ distribution (red curve); about 95% of
surveys fall below 3.84 (dashed line). In this simulation the share is
`r round(mean(dev_stat <= qchisq(0.95, 1)), 3)`.

# 6. When the picture breaks

Every step above assumed the score curves are close to straight lines,
their heights are close to normal, and the model is right. Each can
fail.

```{r fig-breaks, fig.height=3.1}
set.seed(746)
par(mfrow = c(1, 2), mar = c(4, 4, 2, 0.5))
small_lam <- 0.5
hats_small <- replicate(4000, mean(rpois(6, small_lam)))
hist(hats_small, breaks = seq(-0.5 / 6, 3, by = 1 / 6), freq = FALSE, col = "grey85",
     border = "white", main = "Small lambda (0.5), 6 plots", cex.main = 0.9,
     xlab = "MLE of lambda", xlim = c(-0.5, 2))
curve(dnorm(x, small_lam, sqrt(small_lam / 6)), add = TRUE, lwd = 2, col = "red")
abline(v = 0, lty = 3)

heights_nb <- replicate(4000, score(lambda0, rnbinom(24, mu = lambda0, size = 2)))
hist(heights_nb, breaks = 60, freq = FALSE, col = "grey85", border = "white",
     main = "Overdispersed counts, 24 plots", cex.main = 0.9,
     xlab = "Score at the true value", xlim = c(-15, 20))
curve(dnorm(x, 0, sqrt(24 / lambda0)), add = TRUE, lwd = 2, col = "red")
```

- **Near a boundary or with little data (left).** With $\lambda_0 =
  0.5$ and 6 plots, the MLE is skewed, and the normal approximation
  (red) puts probability below zero, where $\lambda$ can't go. This is
  why a Hessian interval for a prevalence of 2 in 20 can have a
  negative lower limit. The likelihood ratio interval or a Bayesian
  interval handles this better.
- **Wrong model (right).** Here the counts are more variable than a
  Poisson allows (negative binomial), but we still fit a Poisson. The
  score heights spread far more than the predicted bell (red). Variance
  and slope no longer match, so the Hessian standard errors are too
  small. "Sandwich" (robust) standard errors estimate the two pieces
  separately for this reason.
- **Several parameters.** The same argument runs with vectors and
  matrices: $\hat\theta - \theta_0 \approx I^{-1} U$, so the
  variance-covariance matrix of the estimates is $I^{-1}$, the inverse
  of the Hessian. Its off-diagonal entries are the correlations between
  estimates, the tilt of the likelihood surface.

# Summary

| Idea | Graphical meaning | Formula |
|:-------------------|:-------------------------------------|:--------------------------|
| Score | slope of the log-likelihood | $U(\theta) = \frac{d}{d\theta}\log\mathcal{L}(\theta)$ |
| MLE | where the score crosses zero | $U(\hat\theta) = 0$ |
| Fisher information | steepness of the score; curvature of the NLL | $I = E[-U'(\theta_0)]$ |
| Score at the truth | heights of the curves at $\theta_0$; normal by the CLT | $U(\theta_0) \approx \text{Normal}(0,\, I)$ |
| Information identity | curves must move when the truth moves | $\text{Var}(U) = I$ |
| Sampling distribution of the MLE | crossing points = height / slope | $\hat\theta \approx \text{Normal}(\theta_0,\, 1/I)$ |
| Hessian (Wald) interval | width of the parabola at 1.92 | $\hat\theta \pm 1.96/\sqrt{I}$ |
| Likelihood ratio cutoff | squared standard normal | $2\,\Delta\text{NLL} \sim \chi^2_1$, cutoff 1.92 |

One sentence holds it together: **a parabola-shaped log-likelihood and a
straight-line score are the same assumption, and the normal sampling
distribution, the Hessian interval, and the 1.92 cutoff all rest on
it.**
