---
title: "Week 7, Lecture 1: How Sure Are We? Uncertainty from the Shape of the Likelihood"
subtitle: "Instructor notes. Part 1: prep reading. Part 2: podium cheat sheet"
fontsize: 12pt
output:
  pdf_document: default
header-includes:
  - \usepackage{enumitem}
  - \setlist[itemize]{itemsep=4pt, parsep=0pt, topsep=4pt}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 5, fig.height = 3.5)
```

# How to prepare (for students)

- **Big picture.** Last week we found the single best-supported
  parameter value (the MLE). A point estimate alone doesn't tell us
  much. Today's question: how uncertain is a parameter estimate, and
  how do we turn that uncertainty into a confidence interval? We'll also
  take a first, brief look at the Bayesian alternative.
- **Read:** Bolker Ch. 6 through section 6.5 (printed p. 227-267; stop
  at 6.6, "Comparing models"). Focus on 6.4 (p. 249-260) and 6.5
  (p. 260-267). Skim the Bayesian parts (6.2.2, p. 237-243, and 6.4.2,
  p. 258-260) for the main idea; Wednesday's lecture goes into Bayesian
  inference in more detail. Bring questions about the reading.
- **Review:** the Poisson distribution and its likelihood (last week's
  pika haypile example); log rules; derivatives, including second
  derivatives and the derivative of $\log x$ and of $1/x$.
- **Optional:** look up what a chi-squared ($\chi^2$) distribution is.
  You don't need its formula, only a rough idea of its shape.

# Overview (instructor only)

This file has two parts, both instructor-only. **Part 1** is read
beforehand. **Part 2** (after the page break, large print) is the
podium copy.

| Beat | Time |
|---|---|
| WOD: *Reading the Interval* (in class) | ~8 min |
| Lecture | ~35 min |
| Discussion | ~7 min |

Source material: Bolker 6.4-6.5 (the whole lecture follows his logic),
6.2.2 for the Bayesian aside; legacy `LECTURE4.Rmd` ("Estimating
parameter uncertainty" through "The usual way to generate CIs") for the
slice/profile/Hessian sequence; legacy `LECTURE6.Rmd` for the
prior $\times$ likelihood picture.

**Example choice.** The one-parameter thread reuses last Monday's
pika haypile data (Week 6, Lecture 1), so students see a familiar
likelihood curve and only the new idea (its *width*) is unfamiliar. The
numbers are friendly: the curvature at the MLE is 2, so the
standard error is $1/\sqrt{2}$. The two-parameter piece uses Bolker's
tadpole predation data (Figures 6.7-6.8, in the assigned reading). I
avoided the frog chorus data from Lab 5 Act 2 on purpose: if Act 2 went
home as a take-home, showing its fitted $\hat a, \hat b$ on Monday
would give away 2b before it's due.

# Where this sits in the week

- **Last week:** likelihood, the NLL, and the MLE (Monday); Lab 5 found
  an MLE by calculus (Act 1) and with `optim()` (Act 2). Lab 5 Act 1
  Part 3 already showed two NLL curves, 3 vs. 30 wetlands, with the
  same $\hat p = 0.4$ but different widths, and asked which survey
  says more about $p$. Today formalizes that answer.
- **Today:** the width and curvature of the NLL curve become a
  confidence interval, two ways (likelihood ratio, and the quadratic or
  Hessian approximation). Brief Bayesian contrast.
- **Tomorrow (Lab 6):** Bliss's beetle fumigation data. Act 1 (paper)
  draws a slice and a profile on a very tilted $(a, b)$ surface and
  uses a $2 \times 2$ Hessian to get an SE by hand. Act 2 fits the LD50
  directly and compares Hessian and profile intervals. Section 7 below
  is the preview, so keep it visual and don't work the beetle numbers.
- **Wednesday:** Bayesian inference in more detail (optimization moved
  to next week). Wednesday's take-home WOD, *Curvature by Hand*, has
  students derive the Poisson and binomial curvatures themselves, so
  today should show the curvature value (2) but **not** derive it on
  the board.

# WOD: Reading the Interval (~8 min)

File: `wod/WOD_Week7_Lecture1_ReadTheInterval.Rmd`. Deliberately light
(no algebra), since students have had several math-heavy WODs in a row.
It gives students the "2 units" rule of thumb *before* the lecture
explains it, so section 5 can pay it off.

Chytrid swabs: 2 positives in 20 frogs, binomial, printed NLL-minus-min
curve. About 4 min alone, 2 min with a neighbor, 2 min debrief.

- **1a.** $\hat p = 0.10$.
- **1b.** Within 2 units: about 0.02 to 0.28 (exact: 0.017 to 0.282).
- **1c.** Not symmetric: 0.08 below $\hat p$, 0.18 above. The curve
  shoots up near 0 because prevalence can't be negative and 2 positives
  already rule out values very close to 0; high values are ruled out
  more gradually.
- **2a.** 8 of 80: same $\hat p = 0.10$, curve 4 times steeper, range
  about 0.05 to 0.18 (narrower, and less lopsided).

Debrief lightly: confirm 1a and 1b, take one answer to 1c, then say
"the rest of today explains where the 2 comes from and why the curve's
width matters." Save the full asymmetry point for section 5, and
don't mention the Wald interval's negative lower bound for these data:
that's the payoff of 2d in Wednesday's take-home.

# Part 1: Lecture content

## 1. Big picture (~3 min)

Put the loop on the board:

$$\text{THEORY} \to \text{MODEL(S)} \to \text{DATA (fit)} \to \text{COMPARE} \to \text{repeat}$$

- So far in the "fit" step we have produced one number per parameter,
  the MLE. Bolker (p. 249): point estimates are useless without a
  measure of uncertainty.
- Today's question: *which parameter values are still plausible given
  the data?* Answer: all the information is in the shape of the
  likelihood curve near its peak. A narrow valley means the data rule out
  most values; a broad valley means they don't.
- The same likelihood feeds two schools of inference. Frequentists use
  it to build confidence intervals (today's focus). Bayesians combine
  it with a prior to get a probability distribution for the parameter
  (brief look today, more on Wednesday).
- Popperian tie-in, worth saying out loud: a confidence interval is
  the set of parameter values the data *fail to reject*. Values outside
  it are falsified (at the 5% level); values inside survive. None is
  proven, including the MLE.

## 2. Questions about the reading (~4 min)

Open the floor. If nobody bites, seed with one of these:

- "Bolker says slices are 'usually wrong.' Why would fixing the other
  parameters at their best values be a problem?" (Answer comes in
  section 7.)
- "Why does doubling the data make the *maximum* likelihood smaller
  but the estimate better?" (Bolker's coin footnote, p. 254: 1 head in
  2 flips has probability 0.5; 500 in 1000 has probability 0.025.
  That doesn't make the coin less fair.)
- "Where does the 1.92 come from?" (Section 5.)

Don't answer at length now if the question is covered later; say
"hold that, it's coming in ten minutes" and write it in the corner of
the board so you remember to close the loop.

## 3. Bayesian inference, in five minutes (~4 min)

Keep this conceptual. One equation, one picture.

$$\underbrace{P(\lambda \mid y)}_{\text{posterior}} =
  \frac{\mathcal{L}_6(\lambda) \, P(\lambda)}{\int \mathcal{L}_6(\lambda)\, P(\lambda)\, d\lambda}
  \;\propto\; \text{likelihood} \times \text{prior}$$

- The likelihood is *not* a probability distribution over $\lambda$
  (Week 5 WOD, question 2d). Multiplying by a prior and dividing by the
  total area turns it into one: the posterior.
- The denominator is just a constant that makes the area equal 1. In
  practice it is hard to compute, which is why MCMC exists (later in
  the semester). Wednesday picks this up.
- With a flat prior, the posterior has the same shape as the
  likelihood. An informative prior pulls it toward the prior.

```{r pika-setup}
haypiles <- c(2, 5, 3, 0, 4, 4)
n_plots <- length(haypiles)
lambda_hat <- mean(haypiles)

nll_pois <- function(lambda, y) {
  -sum(dpois(y, lambda = lambda, log = TRUE))
}

lambda_grid <- seq(0.01, 8, by = 0.01)
lik_vals <- sapply(lambda_grid, function(l) prod(dpois(haypiles, l)))
```

For the informative prior, suppose an earlier survey at a nearby talus
field suggests about 2 haypiles per plot, encoded as a Gamma prior with
mean 2 (shape 10, rate 5). Everything below is brute force on a grid,
which is what legacy `LECTURE6.Rmd` did and is easier to explain than
conjugacy.

```{r pika-bayes, fig.width=6.5, fig.height=2.8}
step <- diff(lambda_grid)[1]
post_from_prior <- function(prior_vals) {
  numer <- lik_vals * prior_vals
  numer / sum(numer * step)
}
prior_flat <- rep(1, length(lambda_grid))
prior_info <- dgamma(lambda_grid, shape = 10, rate = 5)
post_flat <- post_from_prior(prior_flat)
post_info <- post_from_prior(prior_info)
lik_scaled <- lik_vals / max(lik_vals) * max(post_flat)

par(mfrow = c(1, 2), mar = c(4, 4, 1.5, 0.5))
plot(lambda_grid, post_flat, type = "l", lwd = 2, ylim = c(0, 0.95),
     xlab = "lambda", ylab = "Density", main = "Flat prior", cex.main = 0.9)
lines(lambda_grid, lik_scaled, lwd = 2, lty = 2, col = "darkorange")
plot(lambda_grid, post_info, type = "l", lwd = 2, ylim = c(0, 0.95),
     xlab = "lambda", ylab = "Density", main = "Prior centered on 2",
     cex.main = 0.9)
lines(lambda_grid, prior_info, lwd = 2, col = "grey60")
lines(lambda_grid, lik_scaled, lwd = 2, lty = 2, col = "darkorange")
legend("topright", c("posterior", "prior", "likelihood (rescaled)"),
       lwd = 2, lty = c(1, 1, 2), col = c("black", "grey60", "darkorange"),
       bty = "n", cex = 0.7)
```

In the left panel the posterior and the rescaled likelihood sit on top
of each other. In the right panel the posterior mode moves from 3 to
`r round(lambda_grid[which.max(post_info)], 2)`, between the prior and the
data. (The likelihood is rescaled only so it fits on the same axes;
its own vertical scale means nothing, as last Monday's sticking point
said.)

The payoff for Bayesians: a 95% **credible interval** is read straight
off the posterior, and it *is* a probability statement about
$\lambda$. With the flat prior the posterior is Gamma(shape 19, rate 6),
so the equal-tailed interval is:

```{r pika-credible}
cred_int <- qgamma(c(0.025, 0.975), shape = sum(haypiles) + 1, rate = n_plots)
round(cred_int, 2)
```

Hold on to these numbers. They get compared with the frequentist
interval at the end of section 5.

**Sticking point:** students hear "prior" and think "bias." Say
plainly that with a flat prior and a reasonable amount of data the
Bayesian and likelihood answers nearly coincide; the frameworks differ
more in *interpretation* than in numbers for simple problems like this.

## 4. Curvature is information (~6 min)

Callback: Lab 5 Act 1, question 3a. Same $\hat p$, but the 30-wetland
curve was narrower. Now do the same thing with the pikas. Imagine the
biologist had surveyed 24 plots, with the same mean (each of the six
counts appearing four times).

```{r pika-two-curves, fig.width=5, fig.height=3.2}
haypiles_24 <- rep(haypiles, 4)
dnll_6 <- sapply(lambda_grid, nll_pois, y = haypiles) - nll_pois(3, haypiles)
dnll_24 <- sapply(lambda_grid, nll_pois, y = haypiles_24) - nll_pois(3, haypiles_24)
lr_cut <- qchisq(0.95, df = 1) / 2

par(mar = c(4, 4.5, 0.5, 0.5))
plot(lambda_grid, dnll_6, type = "l", lwd = 2, xlim = c(0.5, 7), ylim = c(0, 6),
     xlab = "lambda (haypiles per plot)", ylab = "NLL minus its minimum")
lines(lambda_grid, dnll_24, lwd = 2, lty = 2)
abline(h = lr_cut, col = "red")
legend(x = 4.0, y = 6.2, c("6 plots", "24 plots"), lwd = 2, lty = c(1, 2),
       bty = "n", cex = 0.8)
```

- Both bottom out at $\lambda = 3$.
- The NLL is a *sum* over observations, so 4 times the data means 4
  times the NLL, everywhere. The valley gets 4 times steeper (Bolker
  p. 254).
- Steeper valley = the fit gets worse faster as you move away from the
  MLE = the data rule out more values = a narrower interval.

**Sticking point:** the *height* of the minimum also changes with $n$
(more data means a smaller maximum likelihood, larger minimum NLL).
That's why the figure subtracts the minimum. Only *differences* in NLL
between parameter values on the same data mean anything.

## 5. The likelihood ratio interval (~7 min)

We need a rule for "how far up the valley wall is still plausible."

- **Informal version (Edwards, cited in Bolker p. 254):** this is the
  WOD's rule of thumb. Keep every
  value within 2 log-likelihood units of the best. That's any value
  whose likelihood is at least $1/e^2 \approx 1/7.4$ of the maximum.
- **Frequentist version (the likelihood ratio test):** the deviance,
  $2\,[\text{NLL}(\lambda) - \text{NLL}(\hat\lambda)]$, has
  approximately a $\chi^2$ distribution with 1 degree of freedom when
  $\lambda$ is the true value. Its 95th percentile is 3.84, so the
  cutoff on the NLL scale is $3.84 / 2 = 1.92$. (This is Wilks'
  theorem. Don't derive it; Bolker's footnote on p. 255 gives the
  intuition, which you can paraphrase: near the MLE the NLL is a
  parabola, and a squared normal deviate is $\chi^2_1$.)
- The 95% interval is every $\lambda$ whose NLL is within 1.92 of the
  minimum: where the curve crosses the red line in the figure above.

```{r pika-lr-interval}
lr_cut
f_cut <- function(l, y) nll_pois(l, y) - nll_pois(mean(y), y) - lr_cut
lr_int <- c(uniroot(f_cut, c(0.5, 3), y = haypiles)$root,
            uniroot(f_cut, c(3, 8), y = haypiles)$root)
lr_int_24 <- c(uniroot(f_cut, c(1, 3), y = haypiles_24)$root,
               uniroot(f_cut, c(3, 6), y = haypiles_24)$root)
round(rbind(plots_6 = lr_int, plots_24 = lr_int_24), 2)
```

- With 6 plots: $\lambda \in$ [`r round(lr_int[1], 2)`,
  `r round(lr_int[2], 2)`]. With 24: [`r round(lr_int_24[1], 2)`,
  `r round(lr_int_24[2], 2)`], about half as wide (4 times the data,
  $\sqrt{4} = 2$ times narrower).
- The 6-plot interval is **not symmetric** around 3 (it reaches
  `r round(lr_int[2] - 3, 2)` above but only `r round(3 - lr_int[1], 2)`
  below). That's real: counts can't go below zero, and the NLL curve is
  steeper on the low side. Keep this in mind for section 6.
- WOD callback: the chytrid curve was lopsided for the same reason,
  and more so, since $\hat p = 0.1$ sits closer to its boundary.
- Popperian reading: $\lambda = 1.5$ is falsified at the 5% level;
  $\lambda = 2$ survives; so does 4. The interval is the set of
  hypotheses that survive.

Now compare with the Bayesian credible interval from section 3:

| Method | 95% interval |
|---|---|
| Likelihood ratio | [`r round(lr_int[1], 2)`, `r round(lr_int[2], 2)`] |
| Bayesian, flat prior | [`r round(cred_int[1], 2)`, `r round(cred_int[2], 2)`] |

Close numbers, different meanings. The confidence interval is a
statement about the *procedure*: if we repeated the survey many times,
about 95% of intervals built this way would contain the true
$\lambda$. The credible interval is a statement about $\lambda$ itself,
given the data and the prior. This distinction feeds the discussion.

## 6. The quadratic approximation and the Hessian (~6 min)

The LR interval needs a search (here `uniroot()`; with many parameters,
a full profile, section 7). The shortcut every canned routine uses: near
its minimum, the NLL looks like a parabola.

A Taylor expansion around $\hat\lambda$ (Calc II), using the fact that
the first derivative is zero at the minimum:

$$\text{NLL}(\lambda) \approx \text{NLL}(\hat\lambda) +
  \tfrac{1}{2}\, \text{NLL}''(\hat\lambda)\,(\lambda - \hat\lambda)^2$$

Set the increase equal to the LR cutoff and solve:

$$\tfrac{1}{2}\, \text{NLL}''(\hat\lambda)\,(\lambda - \hat\lambda)^2 = 1.92
  \;\Rightarrow\; \lambda = \hat\lambda \pm \frac{1.96}{\sqrt{\text{NLL}''(\hat\lambda)}}$$

($\sqrt{3.84} = 1.96$, the familiar normal quantile.) So the
**standard error is $1/\sqrt{\text{curvature}}$**. Sharp valley, big
second derivative, small SE. This is Bolker's eq. 6.5.5.

For the pika data the curvature at $\hat\lambda = 3$ is 2 (from
`optim()` below). **Don't derive it on the board:** working out
$\text{NLL}''$ for the Poisson and binomial is Wednesday's take-home WOD.

```{r pika-hessian}
fit <- optim(par = 2, fn = nll_pois, y = haypiles, method = "Brent",
             lower = 0.01, upper = 20, hessian = TRUE)
fit$hessian                       # curvature at the MLE: 2
se_lambda <- 1 / sqrt(fit$hessian[1, 1])
wald_int <- fit$par + c(-1, 1) * qnorm(0.975) * se_lambda
round(c(se = se_lambda, lower = wald_int[1], upper = wald_int[2]), 2)
```

`hessian = TRUE` asks `optim()` for the second derivative, computed
numerically by finite differences. Nobody has to do the calculus.

```{r pika-quad-approx, fig.width=5, fig.height=3.2}
quad_approx <- 0.5 * fit$hessian[1, 1] * (lambda_grid - lambda_hat)^2
par(mar = c(4, 4.5, 0.5, 0.5))
plot(lambda_grid, dnll_6, type = "l", lwd = 2, xlim = c(0.5, 7), ylim = c(0, 5),
     xlab = "lambda (haypiles per plot)", ylab = "NLL minus its minimum")
lines(lambda_grid, quad_approx, lwd = 2, lty = 2, col = "blue")
abline(h = lr_cut, col = "red")
abline(v = lr_int, lty = 3)
abline(v = wald_int, lty = 3, col = "blue")
legend(x = 4.5, y = 1.3, c("true NLL", "quadratic"), lwd = 2, lty = c(1, 2), cex = 0.8,
       col = c("black", "blue"), bty = "n")
```

- Quadratic (Wald) interval: [`r round(wald_int[1], 2)`,
  `r round(wald_int[2], 2)`], symmetric by construction.
- LR interval: [`r round(lr_int[1], 2)`, `r round(lr_int[2], 2)`],
  shifted right, following the real curve.
- With 6 plots the parabola misses the curve's lopsidedness. With more
  data the NLL gets closer to a parabola and the two agree (both are
  large-sample approximations; the LR one is usually the better of
  the two).

**Sticking point:** why does a *second* derivative measure
uncertainty? The first derivative is zero at the MLE by definition, so
it carries no information about width. The second derivative says how
fast the slope picks up as you walk away, which is how fast the fit
degrades. Draw two parabolas with the same vertex, one narrow and one
wide, and ask which has the bigger second derivative.

## 7. More than one parameter: slices, profiles, the Hessian matrix (~4 min)

Keep this mostly visual. Bolker's tadpole data (Figures 6.7-6.8):
Holling type II predation, binomial number killed,
$p = a / (1 + a h N)$.

```{r tadpole-fit}
library(emdbook)
data(ReedfrogFuncresp)
initial_n <- ReedfrogFuncresp$Initial
killed <- ReedfrogFuncresp$Killed

nll_holling <- function(a, h) {
  p <- a / (1 + a * h * initial_n)
  -sum(dbinom(killed, size = initial_n, prob = p, log = TRUE))
}

tad_fit <- optim(c(0.5, 0.0125), function(par) nll_holling(par[1], par[2]),
                 method = "L-BFGS-B", lower = c(0.01, 0.0001),
                 upper = c(2, 0.1), hessian = TRUE)
a_hat <- tad_fit$par[1]
h_hat <- tad_fit$par[2]
round(tad_fit$par, 4)
```

```{r tadpole-surface, fig.width=6.5, fig.height=3.2}
a_seq <- seq(0.33, 0.8, length.out = 121)
h_seq <- seq(0.001, 0.035, length.out = 121)
nll_surf <- outer(a_seq, h_seq, Vectorize(nll_holling))
min_nll <- tad_fit$value

profile_h <- sapply(a_seq, function(a) {
  optimize(function(h) nll_holling(a, h), c(0.0001, 0.1))$minimum
})
profile_a <- sapply(seq_along(a_seq), function(i) nll_holling(a_seq[i], profile_h[i]))
slice_a <- sapply(a_seq, nll_holling, h = h_hat)

par(mfrow = c(1, 2), mar = c(4, 4.5, 0.5, 0.5))
contour(a_seq, h_seq, nll_surf - min_nll, levels = c(1, 2, 3, 5, 8, 12),
        xlab = "attack rate a", ylab = "handling time h", col = "grey50")
contour(a_seq, h_seq, nll_surf - min_nll, levels = qchisq(0.95, 2) / 2,
        add = TRUE, lwd = 2, drawlabels = FALSE)
abline(h = h_hat, lty = 2, col = "purple")
lines(a_seq, profile_h, lty = 1, lwd = 2, col = "darkgreen")
points(a_hat, h_hat, pch = 16)
plot(a_seq, slice_a - min_nll, type = "l", lwd = 2, lty = 2, col = "purple",
     ylim = c(0, 5), xlab = "attack rate a", ylab = "NLL minus its minimum")
lines(a_seq, profile_a - min_nll, lwd = 2, col = "darkgreen")
abline(h = lr_cut, col = "red")
legend(x = 0.63, y = 1.1, c("slice", "profile"), lwd = 2, lty = c(2, 1), cex = 0.8, seg.len = 1.2,
       col = c("purple", "darkgreen"), bty = "n")
```

- **Left:** the NLL surface. The valley is a tilted ellipse: estimates
  of $a$ and $h$ are correlated (a higher attack rate can be offset by
  a longer handling time). Thick contour: 95% joint region
  ($\chi^2_2$ cutoff, $5.99/2 \approx 3$).
- **Slice** (purple dashed): hold $h$ at $\hat h$, vary $a$. **Profile**
  (green): for each $a$, re-optimize $h$. The profile follows the
  floor of the tilted valley.
- **Right:** the slice is steeper, so it gives a falsely narrow
  interval. It ignores that $h$ can adjust to compensate. That's why
  Bolker calls slices "usually wrong." The profile is the right way to
  get an LR interval for one parameter.

Profiles cost one optimization per grid point. The quadratic shortcut
again: invert the Hessian matrix to get the variance-covariance matrix.

```{r tadpole-vcov}
vcov_tad <- solve(tad_fit$hessian)
round(sqrt(diag(vcov_tad)), 4)    # standard errors of a and h
round(cov2cor(vcov_tad), 2)       # correlation between a-hat and h-hat
```

The off-diagonal correlation (about `r round(cov2cor(vcov_tad)[1, 2], 2)`)
is the tilt of the ellipse, in one number. Don't go deeper into matrix
inversion today; the point is that the same "curvature = precision"
idea scales up.

## 8. Return to the big picture (~1 min)

- The MLE says where the valley bottom is; the curvature says how wide
  it is. Width = uncertainty.
- Three frequentist routes to an interval: LR on the curve (or
  profile), quadratic/Hessian (fast, symmetric, large-sample), and
  bootstrapping (not covered today). One Bayesian route: prior
  $\times$ likelihood, read the interval off the posterior.
- Next in the loop: COMPARE. The likelihood ratio test we used today
  to build an interval is the same test we'll use to compare models.

# Discussion (~7 min)

Pick one or two:

1. Our two 95% intervals for $\lambda$ are about [1.8, 4.6]
   (likelihood ratio) and [1.9, 4.7] (Bayesian, flat prior). A land
   manager asks, "So there's a 95% chance the true density is between
   1.8 and 4.6?" Which interval lets you say yes? What would you say
   instead for the other one?
2. Is $\lambda = 1.5$ haypiles per plot *disproven* by these data? Is
   $\lambda = 3$ *proven*? What would change your answer?
3. A reviewer says six plots is too few to trust any interval. Of the
   intervals we built today, which would you trust most with $n = 6$,
   and why? What would you do to check?

Question 1 is the one most likely to surface the confidence vs.
credible misconception. Question 2 reinforces the survive/falsify
vocabulary.

\newpage
\large

# PART 2: PODIUM CHEAT SHEET

\normalsize
\textit{Terse on purpose. Full rationale and code: Part 1.}
\large

**WOD ~8 min. Lecture ~35 min. Discussion ~7 min.**

## WOD: Reading the Interval (~8 min)

Chytrid: 2 of 20 positive. 4 min alone, 2 with neighbor.

- 1a: $\hat p = 0.10$
- 1b: about 0.02 to 0.28
- 1c: lopsided (0.08 below, 0.18 above); boundary at 0
- 2a: 8 of 80, same $\hat p$, narrower (about 0.05 to 0.18)
- Don't mention the Wald interval here (take-home 2d)

\newpage

## 1. Big picture (~3 min)

- Loop on board: THEORY $\to$ MODEL $\to$ DATA (fit) $\to$ COMPARE $\to$ repeat
- So far: MLE = one number. Useless without uncertainty (Bolker p. 249)
- Today: **shape of the likelihood = how sure we are**
- Same likelihood, two schools: CI (today), posterior (brief; more Wed.)
- CI = values that *survive*; outside = falsified; nothing proven

## 2. Reading questions (~4 min)

Seeds if silent:

- Why are slices "usually wrong"? (sec. 7)
- More data, *smaller* max likelihood, better estimate? (coin footnote)
- Where does 1.92 come from? (sec. 5)

Park answers on the board corner.

\newpage

## 3. Bayes in five minutes (~4 min)

$$P(\lambda \mid y) = \frac{\mathcal{L}_6(\lambda)\,P(\lambda)}{\int \mathcal{L}_6(\lambda)\,P(\lambda)\,d\lambda}$$

```{r podium-bayes, echo=FALSE, fig.width=6, fig.height=2.4}
par(mfrow = c(1, 2), mar = c(4, 4, 1.5, 0.5))
plot(lambda_grid, post_flat, type = "l", lwd = 2, ylim = c(0, 0.95),
     xlab = "lambda", ylab = "Density", main = "Flat prior", cex.main = 0.9)
lines(lambda_grid, lik_scaled, lwd = 2, lty = 2, col = "darkorange")
plot(lambda_grid, post_info, type = "l", lwd = 2, ylim = c(0, 0.95),
     xlab = "lambda", ylab = "Density", main = "Prior centered on 2",
     cex.main = 0.9)
lines(lambda_grid, prior_info, lwd = 2, col = "grey60")
lines(lambda_grid, lik_scaled, lwd = 2, lty = 2, col = "darkorange")
```

- Likelihood $\times$ prior, rescaled to area 1
- Flat prior: posterior shape = likelihood shape
- Denominator is hard: why MCMC exists (later)
- Credible interval: [`r round(cred_int[1], 2)`, `r round(cred_int[2], 2)`]. Hold for sec. 5

\newpage

## 4. Curvature is information (~6 min)

Pikas: **2, 5, 3, 0, 4, 4**. Same data $\times$ 4 = 24 plots.

```{r podium-two-curves, echo=FALSE, fig.width=4.5, fig.height=2.8}
par(mar = c(4, 4.5, 0.5, 0.5))
plot(lambda_grid, dnll_6, type = "l", lwd = 2, xlim = c(0.5, 7), ylim = c(0, 6),
     xlab = "lambda", ylab = "NLL - min")
lines(lambda_grid, dnll_24, lwd = 2, lty = 2)
abline(h = lr_cut, col = "red")
```

- Callback: Lab 5 Act 1, 3a (3 vs 30 wetlands)
- NLL is a sum: 4$\times$ data = 4$\times$ steeper
- **Sticking point:** height of minimum changes with $n$; only *differences* matter

\newpage

## 5. Likelihood ratio interval (~7 min)

- Edwards: within 2 log-lik units (the WOD rule)
- LRT: $2\,\Delta\text{NLL} \sim \chi^2_1$; 95% = 3.84; cutoff **1.92**
- 6 plots: [`r round(lr_int[1], 2)`, `r round(lr_int[2], 2)`] (asymmetric!)
- 24 plots: [`r round(lr_int_24[1], 2)`, `r round(lr_int_24[2], 2)`] ($\sqrt{4}$ = 2$\times$ narrower)
- $\lambda = 1.5$ falsified; 2 and 4 survive

| | 95% interval |
|---|---|
| LR | [`r round(lr_int[1], 2)`, `r round(lr_int[2], 2)`] |
| Bayes (flat) | [`r round(cred_int[1], 2)`, `r round(cred_int[2], 2)`] |

- Close numbers, different meaning: procedure vs. belief about $\lambda$

\newpage

## 6. Quadratic approx. / Hessian (~6 min)

$$\text{NLL}(\lambda) \approx \text{NLL}(\hat\lambda) + \tfrac12\,\text{NLL}''(\hat\lambda)(\lambda-\hat\lambda)^2$$
$$\Rightarrow\; \hat\lambda \pm \frac{1.96}{\sqrt{\text{NLL}''(\hat\lambda)}}, \qquad \text{SE} = \frac{1}{\sqrt{\text{curvature}}}$$

- Curvature = 2; SE = 0.71. **Don't derive** (Wed. take-home)
- `optim(..., hessian = TRUE)` gives 2 numerically
- Wald: [`r round(wald_int[1], 2)`, `r round(wald_int[2], 2)`] vs LR [`r round(lr_int[1], 2)`, `r round(lr_int[2], 2)`]

```{r podium-quad, echo=FALSE, fig.width=4.5, fig.height=2.6}
par(mar = c(4, 4.5, 0.5, 0.5))
plot(lambda_grid, dnll_6, type = "l", lwd = 2, xlim = c(0.5, 7), ylim = c(0, 5),
     xlab = "lambda", ylab = "NLL - min")
lines(lambda_grid, quad_approx, lwd = 2, lty = 2, col = "blue")
abline(h = lr_cut, col = "red")
```

- **Sticking point:** why the *second* derivative? Sketch narrow vs wide parabola

\newpage

## 7. More parameters (~4 min)

Bolker tadpoles, $p = a/(1 + ahN)$

```{r podium-tadpole, echo=FALSE, fig.width=6, fig.height=2.8}
par(mfrow = c(1, 2), mar = c(4, 4.5, 0.5, 0.5))
contour(a_seq, h_seq, nll_surf - min_nll, levels = c(1, 2, 3, 5, 8, 12),
        xlab = "a", ylab = "h", col = "grey50")
contour(a_seq, h_seq, nll_surf - min_nll, levels = qchisq(0.95, 2) / 2,
        add = TRUE, lwd = 2, drawlabels = FALSE)
abline(h = h_hat, lty = 2, col = "purple")
lines(a_seq, profile_h, lwd = 2, col = "darkgreen")
points(a_hat, h_hat, pch = 16)
plot(a_seq, slice_a - min_nll, type = "l", lwd = 2, lty = 2, col = "purple",
     ylim = c(0, 5), xlab = "a", ylab = "NLL - min")
lines(a_seq, profile_a - min_nll, lwd = 2, col = "darkgreen")
abline(h = lr_cut, col = "red")
```

- Tilted ellipse = correlated estimates (cor $\approx$ `r round(cov2cor(vcov_tad)[1, 2], 2)`)
- Slice (purple): other param fixed, too narrow
- Profile (green): re-optimize the others, correct
- Shortcut: `solve(hessian)` = var-cov matrix

## 8. Wrap (~1 min)

- Bottom of valley = estimate; width = uncertainty
- Same LRT next: COMPARE models

\newpage

## Discussion (~7 min)

1. "95% chance $\lambda$ is in [1.8, 4.6]?" Which interval lets you say yes?
2. Is $\lambda = 1.5$ disproven? Is 3 proven?
3. $n = 6$: which interval do you trust most? How would you check?
