---
title: "NRES 746 Code Demo"
author: "Emily Ferrell and Joe Coolidge"
date: "2025-10-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(repos = c(CRAN = "https://cloud.r-project.org"))
```


[click here](GLMM_Code_Demo_2025.R) for the R code for this lecture

# Multi-Level (Hierarchical) Regression

Today we will be discussing when, why, and how to use Multi-level (mixed effects) models in R with the packages 'lme4' and 'glmmTMB'.

These are frequentist tools for extensions of simple linear regression.

### Let's ground ourselves with this simple formula for linear regression 

where y is our response variable, x is our predictor variable, b0 is our intercept term, b1 is our slope term, and we have some error term epsilon.
$$y = \beta_0 + \beta_1*x_1 + \epsilon$$
We can also include multiple predictor variables.
$$y = \beta_0 + \beta_1*x_1 + \beta_2*x_2+ ...+\beta_n*x_n +\epsilon$$
And include interaction terms between multiple predictors.
$$y = \beta_0 + \beta_1*x_1 + \beta_2*x_2+ \beta_3*(x_1*x_2) +\epsilon$$
We will continue to build off this formula.

### Now we need to make sure all of our packages are loaded. Here are the ones you will need:
```{r}
#install.packages("lme4")
#install.packages("glmmTMB")
#install.packages("ggplot2")
#install.packages("broom.mixed")
#install.packages("performance")
#install.packages("DHARMa")
#install.packages("effects")
```

### Next we will get our data.
This is some data on juvenile owl begging calls that is included in the glmmTMB package. 
```{r}
library(glmmTMB)
data("Owls")
head(Owls)
```

#### Key variables we will use:
* SiblingNegotiation (count of begging calls)
* FoodTreatment (deprived vs satiated)
* SexParent (female/male)
* Nest

SiblingNegotiation will be our response variable, and FoodTreatment and SexParent are predictor variables. We could draft a simple model where
$$SiblingNegotiation = \beta_0 + \beta_1*(FoodTreatment) + \beta_2*(SexParent) + \epsilon$$
Or include an interaction term between our predictors.
$$SiblingNegotiation = \beta_0 + \beta_1*(FoodTreatment) + \beta_2*(SexParent) + \beta_3*(FoodTreatment*SexParent) + \epsilon$$
But what about our nest variable?

That is where **mixed effects models** come in!!

### How do we know a multi-level mixed effects model is the right choice?

Hierarchical models combine fixed effects with random effects. When you have data with some sort of grouping, repetition, or hierarchical structure (e.g., site, plot, nest, year, individual) with repeated measurements this is the right model for you. 

Observations within the same group are **not independent**. In other words, differences in our response variable can be partially explained by group identity. Mixed models explicitly account for that by allowing each group (nest, plot, etc.) to have its own random intercept or slope.

In this case, **nest is our grouping factor** with repeated measurements or visits to the same owl nest. So, we will treat nest as a random effect in our multi-level mixed effects models.

### Lets visualize our data!
```{r}
library(ggplot2)

ggplot(Owls, aes(SiblingNegotiation)) +
       geom_histogram(binwidth = 1) +
       labs(title = "Begging calls per visit", x = "Calls", y = "Frequency")
```

Already, it is clear that we are not dealing with a normally distributed response variable (this is common!). For the purpose of this demo, we will subset out the zeros (don't worry we will come back to them).

```{r}
owldata <- subset(Owls, SiblingNegotiation > 0) # remove zeros
summary(owldata$SiblingNegotiation) # double check distribution

ggplot(owldata, aes(SiblingNegotiation)) +
       geom_histogram(binwidth = 1) +
       labs(title = "Begging calls per visit", x = "Calls", y = "Frequency")
```

The data is still not normally distributed. It is right skewed, which is pretty typical when you are looking at count data from discrete events (like bird calls).

That this is count data narrows things down to a couple potential distributions: Poisson or negative binomial. 

A Poisson distribution has mean \lambda equal to variance \lambda.
$$E[Y] = \lambda, Var[Y] = \lambda$$
Whereas a negative binomial distribution has an additional dispersion parameter \theta such that variance can be greater than the mean.
$$E[Y] = \mu, Var[Y] = \mu + (\mu^2/\theta)$$
Let's check how the variance in our data compares to the mean.

```{r}
var(owldata$SiblingNegotiation) # Var(Y)
mean(owldata$SiblingNegotiation) # E(Y)
```

Clearly the observed variability is much greater than the expected. In this case, we can use a mixed-effects model with a **negative binomial distribution** to fit our data! The negative binomial introduces an extra dispersion parameter that allows variance to grow faster than the mean. 

## Multi-level (mixed effects) models

#### We want to analyze the effect of the experimental treatment and sex of the parent bringing food on the number of calls (SiblingNegotiation) with a random effect of nest:


$$SiblingNegotiation = \beta_0 + \beta_1*(FoodTreatment) + \beta_2*(SexParent) + \beta_3*(FoodTreatment*SexParent) + \epsilon_{Nest} + \epsilon$$
\[
\text{SiblingNegotiation} \sim \text{NegBinomial}(\mu, \theta)
\]

## Lets start with the package "lme4"
```{r}
library(lme4)
```

#### Within this package, there are many functions for modeling mixed-effects models. Here are some of the most common ones:
* glmer() fits GLMMs
* glmer.nb() fits negative binomial GLMMs
* lmer() fits linear mixed-effects models
* nlmer() fits nonlinear mixed-effects models

#### We will be using glmer.nb() to fit our model.
```{r}
owls_lme4_nb <- glmer.nb(
SiblingNegotiation ~ FoodTreatment * SexParent # This is shorthand for each predictor and the interaction between both

+ (1|Nest), # This is the syntax for including Nest as a random effect
data = owldata
)
summary(owls_lme4_nb)
```

**Note that glmer.nb() automatically uses a log link function**

#### Takeaways
* Random effects: there is relatively little variability between nests (Std. Dev. = 0.1846)
* Fixed effects: no meaningful effects of or interactions between FoodTreatment and SexParent (p >> 0.05)

This model is rather uninformative, but is it misspecified? Let's dive into some diagnostic tools to check.

#### The "DHARMa" package is a great way to get diagnostics plots when running mixed-effects models.
```{r}
library(DHARMa)
par(mfrow = c(2, 2))
res <- simulateResiduals(owls_lme4_nb, plot = FALSE)
plot(res)
```

#### Based on these plots, the model fits very well! 

#### QQ Plot
* shows generally good fit in the residual distribution
* negative binomial family captures variance structure
* no problematic outliers

#### Residuals by Predicted Bins
* shows no heteroscedasticity and no systematic pattern in residuals vs fitted values

So, the model is a good fit to the data (correctly specified), it's just not very informative (our predictors don't have "significant" effects). Maybe that's because we removed a large portion of the data...

### Evaluate performance of the model
```{r}
library(broom.mixed)
library(performance)
```
```{r}
tidy(owls_lme4_nb, effects = "fixed", conf.int = TRUE)
```

#### Based on the model, there is no strong evidence that either food treatment or parent sex meaningfully changes the rate of begging calls in this filtered (non-zero) dataset.

```{r}
r2_nakagawa(owls_lme4_nb)
```

#### Marginal $R^2$
* this value shows that the fixed predictors (FoodTreatment, SexParent, interaction) explain very little of the total variance in begging calls
* this is consistent with the non-significant outputs above

#### Conditional $R^2$
* the increase in this value indicates that most of the explained variation comes from nest-to-nest differences

## Now lets try the package "glmmTMB"

#### This package is much more model-specific than "lme4" and fits linear and generalized linear mixed models using Template Model Builder for Maximum Likelihood Estimates. The "glmmTMB" package is usually a better fit for 'weird' data, such as zero-inflated data.

Remember that lme4 has different functions for different distributions:

* lmer() --> Gaussian (normal)
* glmer() --> any standard GLM family
* glmer.nb() --> wrapper for glmer() that allows it to fit a negative binomial model

However, glmmTMB is much more flexible (at the sacrifice of longer run times). glmmTMB has more options to specify distribution family (including 2 types of negative binomial distributions), and it has options to specify a zero-inflation model or dispersion model! It can even handle hurdle models, but we won't steal Ryan and Catherine's thunder...

First, let's compare nbinom1 and nbinom2. Both handle overdispersion by including an additional dispersion parameter (compared to a Poisson distribution where mean = variance)

nbinom1 has the following variance formula:
$$Var = \mu * (1+\alpha)$$
And is useful for mild overdispersion.

nbinom2 has the following variance formula:
$$Var = \mu + (\mu^2/\theta)$$
And is useful for strong overdispersion.

#### Let's see how it works on our subsetted data...
```{r}
library(glmmTMB)
owls_tmb_nb1 <- glmmTMB(
SiblingNegotiation ~ FoodTreatment * SexParent + (1|Nest),
data = owldata,
family = nbinom1()
)
summary(owls_tmb_nb1)

owls_tmb_nb2 <- glmmTMB(
SiblingNegotiation ~ FoodTreatment * SexParent + (1|Nest),
data = owldata,
family = nbinom2()
)
summary(owls_tmb_nb2)
```

#### Now we can run our diagnostics plots
```{r}
res_tmb1  <- simulateResiduals(owls_tmb_nb1, plot = FALSE)
plot(res_tmb1)

res_tmb2  <- simulateResiduals(owls_tmb_nb2, plot = FALSE)
plot(res_tmb2)
```

#### Based on these plots, both "glmmTMB" models also fit! 

#### QQ Plot
* residuals are uniform
* negative binomial family captures variance structure
* no problematic outliers

#### Residuals by Predicted Bins
* shows no heteroscedasticity and no systematic pattern in residuals vs fitted values
* no bias at low or high predicted values

#### Let's evaluate performance again
```{r}
tidy(owls_tmb_nb1, effects = "fixed", conf.int = TRUE)
tidy(owls_tmb_nb2, effects = "fixed", conf.int = TRUE)
```

#### Based on this model, there is still no strong evidence that either food treatment or parent sex meaningfully changes the rate of begging calls. The results of the "glmmTMB" model are identical in direction and magnitude to the "lme4" model. We see the same results with the $R^2$ values below:

```{r}
r2_nakagawa(owls_tmb_nb1)
r2_nakagawa(owls_tmb_nb2)
```

### So if both our models fit well and have similar results, how can we test which one is actually the best?

#### Start by comparing the AIC
```{r}
AIC(owls_lme4_nb, owls_tmb_nb1, owls_tmb_nb2)
```

#### Compare our $R^2$ values
```{r}
r2_nakagawa(owls_lme4_nb)
r2_nakagawa(owls_tmb_nb1)
r2_nakagawa(owls_tmb_nb2)
```

#### We can also compare log-likelihood
```{r}
logLik(owls_lme4_nb)
logLik(owls_tmb_nb1)
logLik(owls_tmb_nb2)
```

#### We can go back and compare the diagnostics plots but I think we can safely call this a tie between models! (Maybe a slight victory for owls_tmb_nb1)

### However, there are some instances where this may not be the case. Here is a general overview of the difference between the two packages:

#### lme4 is better when...
1. You want simplicity and transparency.
2. You have standard response families (e.g., Gaussian, Binomial, Poisson...)
3. You don’t need zero-inflation or complex variance structures.
4. Your model is small/medium-sized and converges easily.

#### glmmTMB shines when...
1. Your data are zero-inflated or truncated.
2. You need flexible distributions.
3. You need to model the dispersion itself (heterogeneous variance).
4. You’re doing model selection or prediction at scale.
5. You need complex random structures or correlated random effects.

### Lets try the glmmTMB() function with our full dataset from earlier (zero-inflated)
```{r}
# Baseline NB (no ZI factor)
m_nb <- glmmTMB(SiblingNegotiation ~ FoodTreatment * SexParent + (1|Nest),
data = Owls, family = nbinom2())

# Zero-inflated NB: a ZI component (~ 1) adds extra zeros
m_zinb <- glmmTMB(SiblingNegotiation ~ FoodTreatment * SexParent + (1|Nest),
ziformula = ~ 1,
data = Owls, family = nbinom2())

# Hurdle NB: zero vs positive modeled separately
m_hurd <- glmmTMB(SiblingNegotiation ~ FoodTreatment * SexParent + (1|Nest),
ziformula = ~ 1,              # hurdle uses 'truncated_nbinom2' for count
data = Owls,
family = truncated_nbinom2())
```

#### Now we can compare the glmmTMB() models
```{r}
AIC(m_nb, m_zinb, m_hurd)
```

#### The glmmTMB() model with the zero inflation factor performed the best based on AIC. For your exercise you will use all evaluation criteria but we will skip that for now. 

#### Here is how you can interpret a zero-inflated model
* **Conditional model** (count part): effects on mean calls when calling is possible.
* **ZI model** (zero part): effects on the odds of structural zero.
```{r}
summary(m_zinb)$coefficients
```

#### Effect Display
```{r}
library(effects)
pred_zinb <- allEffects(m_zinb)
plot(pred_zinb, main = "ZI-NB model effects")

```

The plot shows how the predictors influence the **expected counts** when calling actually happens (non-zero).

## Now that we've gone through the differences between lme4() and glmmTMB() as well as how to include zero inflation... its your turn!

### Your goal: decide which model best explains begging calls in the full dataset and justify it with diagnostics.

1. Get the full Owls dataset.
2. Select your predictors (they should be different than the ones in the example).
3. Build three models with the Owls data:
  * NB GLMM with lme4()
  * NB GLMM with glmmTMB()
  * ZI-NB with glmmTMB()
4. Choose three types of model comparison (e.g. AIC, visual assessment of diagnostics, $R^2$, etc.) to determine which model is best.
5. Create a prediction figure from your chosen model.
6. Describe your model results in ecological context (a few sentences).

## Good Luck!




