---
title: "Lab 1, Act 3: Defend"
subtitle: "Does the evidence actually support our candidate?"
author: "NRES 746 -- name: ____________________    group: ______"
date: ""
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Data ---------------------------------
frogs <- read.csv("ReedfrogSizepred.csv")
mn <- aggregate(Kill ~ TBL, data = frogs, FUN = mean)

# Candidate Ricker curve (from Act 2) ---------------------------------
a <- 0.9; b <- 1/12
ricker <- function(x, a, b) a * x * exp(-b * x)
pred_ricker <- round(ricker(mn$TBL, a, b), 2)
grand_mean <- round(mean(frogs$Kill), 2)
```

<!-- ============================================================
INSTRUCTOR NOTES (delete this block before printing for students)

Pacing (~28 min total for Act 3):
 - 10 min individual: Part 1 (residuals & SSR for two rival models)
 - 10 min group discussion
 - 8 min revise + written reflection

Lab-wide time budget check (total lab period is 2h45m = 165 min, with
two 5-10 min breaks built in between Acts):
  Act 1 (45 min) + Act 2 (70 min) + Act 3 (28 min) = 143 min of content
  + 2 breaks (10-20 min total) = 153-163 min, under the 165 min cap.
Act 3 is intentionally short and tight (one individual task, not three)
to leave room for Acts 1-2's longer runtimes. If a section is running
long, the revise/reflect questions are the easiest to trim -- they're
individual written work, not group-dependent.
(NOTE: Act 2's own instructor-notes header currently says "45 min total"
but its actual breakdown sums to 70 min -- that mislabel should be fixed
there too, since it's what this 165-min budget above is built on.)

Dataset: same running dataset as Acts 1-2 (Vonesh & Bolker 2005 / Bolker
Fig. 3.1, via emdbook::ReedfrogSizepred).

THE TWIST (important -- this is the pedagogical point of the whole Act):
the Act 2 candidate (a=0.9, b=1/12) was chosen for a clean critical point
and a plausible-looking peak, NOT a good fit -- and it isn't one. It
loses badly (SSR ~27.2) to the trivial "ignore body size, always predict
the grand mean" null model (SSR ~11.4), because it overshoots badly for
TBL=21/25/37 (see Act 2's own instructor notes, which already flagged
this mismatch as a preview of Act 3). This is real, not manufactured --
don't be tempted to quietly pick a better a,b to make Ricker win. The
whole point is for students to discover that a plausible SHAPE with
badly-guessed PARAMETERS can lose to a dumb model, which is exactly the
motivation for Lab 2's formal fitting (nls/MLE) -- eyeballing isn't
enough. Students who complain "but the shape still looks right" are
having exactly the intended realization; push them toward Part 1's
second interpretive question and revise Q1.

Feel free to add a simple ANOVA-flavor aside for strong students during
group discussion (comparing SSR is literally the intuition behind R^2
and F-tests) but it's not required -- the lab is scoped to stay
calculator/algebra-level, no new formal machinery introduced here.
============================================================ -->

## Picking up where we left off

In Act 1, your group diagnosed the tadpole data as hump-shaped and picked the Ricker function as your top candidate. In Act 2, you put specific numbers on it ($a=0.9$, $b=1/12$) and derived its peak, $x^*=12$ mm. Today we put that candidate on trial: does it actually predict the real data well, or does it just *look* plausible?

```{r candidate-curve, fig.height=3.2, fig.width=5.5}
set.seed(1)
x_j <- frogs$TBL + runif(nrow(frogs), -0.6, 0.6)
plot(x_j, frogs$Kill, xlim = c(5, 40), ylim = c(-0.3, 5.3),
     xlab = "Tadpole size, TBL (mm)", ylab = "Number killed (out of 10)",
     pch = 21, bg = "grey70", cex = 1.4)
curve(ricker(x, a, b), from = 5, to = 40, lwd = 2, col = "steelblue", add = TRUE)
grid()
```

## Part 1: Does the shape earn its keep? (individual, ~10 min)

A model earns its keep by predicting better than a "dumb" baseline. One simple way to score any model: for each data point, compute the **residual** (observed $-$ predicted), square it (so overshoots and undershoots both count against the model, and big misses count extra -- this is the same squared-deviation idea from Act 1's variance calculation), and add up the squares. Smaller total = better fit. This total is often called the **sum of squared residuals (SSR)**.

**Model 1: our Ricker candidate.** Using $a=0.9$, $b=1/12$ from Act 2, fill in the missing residuals:

| TBL (mm) | Observed mean | Ricker predicted | Residual | Squared residual |
|:---:|:---:|:---:|:---:|:---:|
| 9  | 1.00 | 3.83 | ________ | 8.01 |
| 12 | 4.00 | 3.97 | ________ | 0.00 |
| 21 | 0.00 | 3.28 | ________ | 10.76 |
| 25 | 0.33 | 2.80 | ________ | 6.10 |
| 37 | 0.00 | 1.53 | ________ | 2.34 |
| | | | **SSR:** | **27.21** |

**Model 2: the "null" model.** The simplest possible model: ignore tadpole size entirely, and always predict the overall average. First, find that average (using either Act 1's data table or the mean of the five group means):

$\bar{y} =$ ________

Now fill in the rest of the table yourself:

| TBL (mm) | Observed mean | Null predicted | Residual | Squared residual |
|:---:|:---:|:---:|:---:|:---:|
| 9  | 1.00 | ________ | ________ | ________ |
| 12 | 4.00 | ________ | ________ | ________ |
| 21 | 0.00 | ________ | ________ | ________ |
| 25 | 0.33 | ________ | ________ | ________ |
| 37 | 0.00 | ________ | ________ | ________ |
| | | | **SSR:** | ________ |

**Compare.** Which model has the smaller (better) SSR -- the Ricker curve, or the null model that ignores body size entirely? Is this what you expected, given Acts 1 and 2?

____________________________________________________________
____________________________________________________________

What does today's result mean for the tadpole predation model (predation as a function of size)? What do you take away from the above result?

____________________________________________________________
____________________________________________________________

---

## Group discussion (~10 min)

Compare your work with your group:

- Compare your SSR(null) calculations. Did everyone get the same total? Resolve any arithmetic disagreements.
- Debate about whether you think the null model is the best you can do. Where does your group land? What would you try next?
- Look back at Act 1's variance table -- most groups had only 3 replicate tanks. Given that, how much would you trust any single group's mean? Would you be surprised if a different set of tanks gave noticeably different numbers, especially for TBL=25?
- As a group: are you ready to *defend* the Ricker function as the right model for this system, or do you want to reserve judgment? What (if anything) would change your mind?

## Revise and reflect (individual, ~8 min)

Revise any of your answers above based on the group discussion. Then answer:

1. In your own words: what's the difference between a model having the *wrong shape* versus the *right shape but poorly-chosen parameters*? Based on today's evidence, which one is the bigger problem for our Ricker candidate?

____________________________________________________________
____________________________________________________________

2. If you could plug in the *best possible* $a$ and $b$ for the Ricker shape -- not just eyeballed guesses -- do you think it could beat the null model? Why or why not?

____________________________________________________________
____________________________________________________________

3. Muddiest point: what's the one thing from this part of the lab you'd like to see explained again?

____________________________________________________________
____________________________________________________________
