---
title: "Lab 1"
author: "NRES 746"
date: "Fall 2023"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
#library(checkr)
#tutorial_options(exercise.timelimit = 60, exercise.checker = checkr::check_for_learnr)
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
gradethis::gradethis_setup()
```

```{r echo=FALSE}

############################################################
####                                                    ####  
####  NRES 746, Lab 1                                   ####
####                                                    ####
####  Kevin Shoemaker                                   #### 
####  University of Nevada, Reno                        ####
####                                                    #### 
############################################################


############################################################
####  Computational algorithms and standard statistics  ####
############################################################


# gradethis::grade_learnr

```

## Lab 1 overview

In this course we will rely heavily on the R programming language for statistical computing and graphics (e.g., this website is actually built in R using ['rmarkdown'](https://rmarkdown.rstudio.com/) and [rshiny](https://shiny.rstudio.com/)). The purpose of this first laboratory exercise is to develop the level of familiarity with R that is needed to succeed in this course – and ultimately, to establish a foundation for you to develop your data analysis skills using R throughout your scientific career. 

## Lab 1 details

Specifically, this lab will provide a basic introduction to the R programming Language and the use of R to perform statistics and programming tasks. Some of the materials from this lab (especially the regression example) were borrowed/modified from the previous instructor for this course (P. Weisberg). 

As with all lab reports, your answers will either take the form of R functions or short written responses. The R functions should be stored in an R script file ('.R' extension). To allow me to evaluate your work more efficiently, please name your R script using the following convention: "[your first name]_[your last name]_lab1.R". So my submission would be "kevin_shoemaker_lab1.R". _Please include only the requested functions in your submitted R script_. You will probably complete the lab exercises using a main script (that includes all your work), and then save a reduced copy (with only the requested functions) for submission. 

Please submit the R script and the Word document via WebCampus by midnight on the due date (one week after the final lab session allocated for this topic). You can work in groups but submit the materials individually. One student in each group should be an 'R guru' if at all possible!

Please provide any requested written answers as a Word document in WebCampus- keep your responses as brief as possible! Also, I will not be asgining yr scroes bsased on spellin, grammmer or the propreiety orf yur senstence cosnstrujctin.

This lab exercise will extend over two laboratory periods (Aug. 27 and Sep. 3), with your R script and Word document due Sept. 10.

## Let's get started!

1. Go to website [http://cran.r-project.org/](http://cran.r-project.org/). This is the source for downloading the free, public-domain R software and where you can access R packages, find help, access the user community, etc. If you haven't already installed R, do so now! 
2. Open the Rstudio software ([https://www.rstudio.com](https://www.rstudio.com). Change the working directory to a convenient directory (e.g.,  a subfolder called 'Lab 1' in your main course directory). NOTE- if you set up an Rstudio "project" (.Rproj extension) in this directory, the working directory is set automatically each time you load the project! I recommend you set up an R studio project for this course. 
3. Open a new, blank R script window in Rstudio. Using comments (anything preceded by a pound sign is not interpreted by R and can be used to provide human-readable notes and commentary so your code is more readable), add a header to the script to indicate that this is lab 1- include your name and the course number in the header.
4. Save the script to your working directory, using the naming convention: "[your first name]_[your last name]_lab1.R", all lower-case. So my script would be named: "kevin_shoemaker_lab1.R". You will be submitting this script when you have finished all the exercises. Again, please include only the requested functions in your submitted R script.  
5. While you're at it, also start a new Word document to record your written responses to the lab exercises (I will indicate where I expect a written response in Word vs. an R function). There is no specific naming convention for your Word documents.

## Take some time to get more familiar with R

From the R manual, ['Introduction to R'](http://cran.r-project.org/doc/manuals/R-intro.html) you will implement all the steps in Appendix A, located [here](http://cran.r-project.org/doc/manuals/R-intro.html#A-sample-session). This takes you far but without much explanation-- it is a way to jump into the deep end of the pool. Alternatively, or in addition, go to the [Links](http://naes.unr.edu/shoemaker/teaching/NRES-746/Links.html) page and pick a tutortial to run through (e.g., the Datacamp introductory R course). You can also check out the materials from [UNR's R "bootcamp"](http://naes.unr.edu/shoemaker/teaching/R-Bootcamp/index.html). Before you move on, make sure you have a basic understanding of how to work with R and RStudio. If you are working with someone who has used R before, don't be afraid to ask questions! Also, take your time with this- you might want to use the entire lab period on this if you have never used R before.

Another useful introductory R tutorial can be found [here](http://www.nceas.ucsb.edu/files/scicomp/Dloads/RProgramming/BestFirstRTutorial.pdf), courtesy of [NCEAS](https://www.nceas.ucsb.edu/). Please take the time to complete this tutorial, including going through the exercises.   

If you already have basic R expertise, this is your opportunity to help your peers to develop the level of comfort and familiarity with R that they will need to perform data analysis and programming tasks in this course.     

Depending on whether you are already familiar with R, you may also find the remainder of this document useful as you work your way through the course (and there are many other good introductory R resources available online... let me know if there is one you particularly like and I will add it to the course website ([Links page](http://naes.unr.edu/shoemaker/teaching/NRES-746/Links.html)). As you work your way through the tutorial(s) (on your own pace), please ask the instructor or your peers if you are uncertain about anything.  

## Interactive lab exercises!

Since there is no TA in this class, I have tried wherever possible to embed interactive text windows (which serve as a 'virtual TA' and were created using 'R shiny') in the course website where you can test your code and get instant feedback as to whether your answers are correct! Here's an example:

Write a function that takes a numeric input vector and computes the sum of all its elements.

Basically, all you need to do here is write 'sum(vector)' where it says '[add code here]'

NOTE: you don't need to submit this as part of your lab report! This is just practice...

```{r functest-solution}
myvec <- c(1:4)    # you don't need to modify this line- and if you do, revert before you submit since you'll get the wrong answer otherwise!!
myfunc <- function(vector){
    sum(vector)
}
myfunc(myvec)    

```

```{r functest, exercise=TRUE}
myvec <- c(1:4)    # you don't need to modify this line- and if you do, revert before you submit since you'll get the wrong answer otherwise!!
myfunc <- function(vector){
    # [add code here!]
}
myfunc(myvec)   

```

<div id="functest-hint">
**Hint:** You may want to use the `sum()` function.
</div>

```{r functest-check}
myvec <- c(1:4)
check_functest <- function(USER_CODE){
    code <- checkr::for_checkr(USER_CODE)
    func_call <- checkr::line_where(code,checkr::insist(F=="myfunc","Use function 'myfunc()'"))
    checkr::line_where(func_call, checkr::insist(V==10, "Your solution is {{V}}. This is not the correct result"),checkr::passif(TRUE, "Great job!"))
}
check_functest(USER_CODE)

# check_functest <- function(user_code){
#   test_result(user_code)
# }
# check_functest(grader_args)

```


## Demonstration: Central Limit Theorem (CLT)
To gain some familiarity with using R scripts and developing algorithms, complete the following exercise (you don't need to submit this demo as part of your lab write-up BUT you will be asked to use this code as part of exercise 1 (below):  

## Complete the following steps:
1. Review the meaning of the [Central Limit Theorem](https://en.wikipedia.org/wiki/Central_limit_theorem), which implies that the mean of a sufficiently large number of independent random variables (small samples from a much larger *population* of interest) will have a *sampling distribution* that is approximately normally distributed, regardless of whether the underlying data distribution is normally distributed.
2. The following code for illustrating the Central Limit Theorem was inspired by Teetor, Paul. 2011. R Cookbook. O-Reilly Media, Inc. (p. 45), with several modifications. Type (or paste) the following code into your R script window (or Rstudio script window):         

```{r results='hide'}

####################
# CENTRAL LIMIT THEOREM demonstration
####################

lots <- 100000       # number approximating infinity      

N_IND_SAMPLES <- 1000       # number of indepenent random samples to draw from the specified distribution
SAMPLESIZE <- 10          # sample size of each independent random sample


########
#  Define the random number distribution. Here, we will use a "uniform" random number generator with a min of 10 and max of 20.   
TRUEMIN <- 10            
TRUEMAX <- 20      
datafountain <- runif(lots,TRUEMIN,TRUEMAX)      # here we define the full set of possible random numbers to draw random samples from (the population of potential data)

#######
# Draw multiple samples from the pool (population) of possible data.  

samplemean <- numeric(N_IND_SAMPLES)     # set up storage vector
for(i in 1:N_IND_SAMPLES){      # for each replicate (independent random sample)
  sample <- sample(datafountain,SAMPLESIZE)   # draw an independent random sample from the population of interest
  samplemean[i] <- mean(sample)    # compute and record the sample mean
}

hist(datafountain,freq=F,ylim=c(0,1),main="",xlab="Value")      # plot out the distribution of sample means
hist(samplemean,freq=F,add=T,col="red")    # overlay the distribution of the underlying data from which we are drawing samples.
legend("topleft",fill=c("white","red"),legend=c("data population","sample means"),bty="n")     # add a legend to the plot

```

3. Experiment with executing this code in the following four ways:   
    * copy and paste from the script window directly into the R console;       
    * use <ctrl Enter> to execute line by line from within the script window in RStudio; (or on Macs, do something else!... sorry...)       
    * use <ctrl A> to select the whole code block, then <ctrl Enter> to execute all at once;  (or on Macs, do something else!)      
    * save the script to your working directory as a text file with .R extension, and then run the script using the "source()" function, e.g.:

```

source("H:\\Courses\\NRES746\\Lab1\\CentralLimitTheorem.R")     # load the R script!

```

**NOTE**: The pound (#) sign used above allows the R programmer to insert comments adjacent to snippets of code, which facilitates readability of code (and lets the programmer remember later on what he/she was thinking when coding things a certain way!). It is good practice, especially for the beginning programmer, to comment every line so as to describe its precise meaning, including all variables and functions. Make sure you fully understand the commented code for the central limit theorem demonstration above!


Here is an interactive window you can use to play around with the CLT code:

```{r cltdemo1, exercise=TRUE}

# [Add code here] 

```

<div id="cltdemo1-hint">
**Hint:** Try running with different values for the 'sample.size' parameter.
</div>


4. Now modify your R script to see how closely the *distribution of sample means* follows a normal distribution. Use a "quantile-quantile" (q-q) plot to visualize how closely the quantiles of the sampling distribution resemble the quantiles of a normal distribution. Use the "qqnorm()" function. To learn more about this function, type:

```{r eval=FALSE}

?qqnorm   # learn more about the "qqnorm()" function

```

Plot the q-q plot next to the histograms. The plot on the left should be the comparison of histograms (for population distribution and distribution of sample means) shown in the original script (above). The plot on the right should be the q-q plot. To produce side-by-side plots, you will need to add this line of code to the appropriate place in your script:

```{r eval=FALSE}

########
# Set graphical parameters for side by side plotting

par(mfrow=c(1,2)) # sets up two side by side plots as one row and two columns

## or alternatively...

layout(matrix(1:2,nrow=1))

```

In addition, run a *Shapiro-Wilk normality test*, which tests the null hypothesis that a set of numbers (in this case the vector of sample means) indeed comes from a normal distribution (so what does a low p-value mean??). Use the "shapiro.test()" function:

```{r eval=FALSE}

?shapiro.test

```


Again, here is an interactive window you can use to play around with the CLT code. 

```{r cltdemo2, exercise=TRUE}

# [Add code here] 

```

<div id="cltdemo2-hint">
**Hint:** Try running with different values for the 'sample.size' parameter.
</div>

So... what can you conclude from these tests? Can you conclude that the distribution of sample means is NOT normal/Gaussian??


## Exercise 1: home-made functions!

For the first part of this lab, you are asked to write several functions (and submit them as part of your lab script, of course!). The functions increase in order of complexity!

You should now know how to construct functions in R. If you don't, go back to the [NCEAS tutorial](https://www.nceas.ucsb.edu/files/scicomp/Dloads/RProgramming/BestFirstRTutorial.pdf) and review the section on writing functions.

## **Exercise 1a** (for R script) 

Write an R function called "CoefVar()" that takes a numeric vector (named 'vector' within the function but could have any name outside the function) as input, and computes (and returns) its coefficient of variation (CV; standard deviation as a proportion of the mean). To make sure it works, apply your function to the 'Height' vector in the 'trees' dataset that installs with R as sample data:

```{r answer1a, echo=FALSE}

######################
# LAB 1 (possible) answers
######################

#############
## exercise 1a

CoefVar <- function(vector){    
  cv <- sd(vector,na.rm=TRUE)/mean(vector,na.rm=TRUE)
  return(cv)     
}


```

```{r test1a, echo=FALSE, context='data'}

answer1a <- CoefVar(trees$Height)    # run your new function!
answer1a

```


```{r eval=FALSE}

############
# Explore the "trees" dataset

?trees

summary(trees)    # learn more about the data

trees$Height    # extract the "Height" column from the trees dataset.

CoefVar(trees$Height)    # run your new function!

```


You can use this "sandbox" (below) to develop and test your function!

```{r q1a, exercise=TRUE}
test <- trees$Height
CoefVar <- function(vector){
    # [add code here!]
}
CoefVar(test)   # note that the name of the vector can be different inside and outside the function. The name of the
                   # vector inside the function is just a placeholder for any vector the user wants to compute a cv for

```

<div id="q1a-hint">
**Hint:** You probably want to use the 'sd()' and 'mean()' functions.
</div>

```{r q1a-check}
test <- trees$Height
check_q1a <- function(USER_CODE){
    code <- checkr::for_checkr(USER_CODE)
    func_call <- checkr::line_where(code,checkr::insist(F== "CoefVar","Use CoefVar() function!"))
    t1 <- checkr::line_where(func_call, checkr::insist(all(V==answer1a), "Your solution is {{V}}. This is not the correct result"),checkr::passif(TRUE, "Great job!"))
}
check_q1a(USER_CODE)
```

```{r playwithcheckr, eval=FALSE, echo=FALSE}

code <- checkr::for_checkr(quote({test <- trees$Height;CoefVar <- function(vector){sd(vector)/mean(vector)};CoefVar(vector=test)}))
code
func_call <- line_calling(code, CoefVar, message = "Use CoefVar() to construct the model.")
checkr::line_where(code, checkr::insist(V==answer1a, "Your solution is {{V}}. This is not the correct result"),checkr::passif(TRUE, "Great job!"))
check(func_call, checkr::insist(dplyr::near(V,answer1a,tol=0.001), 
        message="Your solution is {{V}}. This is not the correct result"),
        checkr::passif(TRUE, "Great job!"))


```

```{r playwithtestthis1, eval=FALSE, echo=FALSE}

check_result(pass_if(answer1a),grader_args=list(user_quo = quo({test <- trees$Height;CoefVar <-            function(vector){sd(vector)/mean(vector)};CoefVar(vector=test)})))

test_result(grader_args=list(user_quo = quo({a=10;a}),solution_quo=quo(10)))    # use this syntax!



```

## * **Exercise 1b** (for R script) 

Write a function called "DrawLine()" for drawing a regression line through a scatter plot. This function should be specified as follows:

+ **input**:   
    + x = a numeric vector specifying the x-coordinates of the scatter plot   
    + y = a numeric vector specifying the y-coordinates of the scatter plot    

+ **suggested algorithm**:   
    + with the x and y coordinates, first produce a scatterplot (HINT: use the "plot()" function)
    + use the "lm()" function to regress the y variable on the x variable.
    + record the intercept and slope of the linear relationship between x and y (HINT: use the "coef()" function)
    + add a regression line to the scatter plot (HINT: use the "abline()" function)

+ **return**:  
    + coefs = a vector of length 2, storing the intercept and slope of the linear relationship


```{r answer1b, echo=FALSE}

#############
## exercise 1b

DrawLine <- function(x,y){  
  plot(y~x)
  mod <- lm(y~x)
  coefs <- coef(mod)
  abline(mod)
  return(coefs)
  
}


```

```{r test1b, echo=FALSE, fig.keep='none', context='data'}

answer1b <- DrawLine(trees$Height,trees$Volume)    # run your new function!

DrawLine(faithful$waiting,faithful$eruptions)    # test using the old faithful eruptions data



```

Apply this function to the 'Height' (x axis) and 'Volume' (y axis) vectors in the 'trees' dataset, and then to the 'waiting' (x axis) and 'eruptions' (y axis) vectors in the 'faithful' dataset.

```{r eval=FALSE}

?faithful

summary(faithful)

faithful$eruptions
faithful$waiting

```


You can use this "sandbox" (below) to develop and test your function!

```{r q1b, exercise=TRUE}

DrawLine <- function(x,y){
    # [add code here!]
}
DrawLine(trees$Height,trees$Volume)

```

<div id="q1b-hint">
**Hint:** You may want to use the `coef()` function.
</div>

```{r q1b-check}

check_q1b <- function(USER_CODE){
    code <- checkr::for_checkr(USER_CODE)
    func_call <- checkr::line_where(code,checkr::insist(F== "DrawLine","Use DrawLine() function!"))
    t1 <- checkr::line_where(func_call, checkr::insist(all(V==answer1b), "Your solution is {{V}}. This is not the correct result"),checkr::passif(TRUE, "Great job!"))
}
check_q1b(USER_CODE)

```



## * **Exercise 1c** (for R script) 

Write a function called "DrawLine2()" for (optionally) drawing a "smoothed" regression line through a scatter plot, making the smoothing span (degree of smoothness, or non-wiggliness of the line) a user-defined option. This function should be specified as follows:

+ **input**:  
    + x = a numeric vector specifying the x-coordinates of the scatter plot   
    + y = a numeric vector specifying the y-coordinates of the scatter plot
    + smooth = a logical (TRUE/FALSE) value defining whether or not to add a smoothed line or a simple regression line
    + span = a number indicating the degree of smoothness, or "non-wiggliness" of the smoothed line (only applies if smooth=TRUE)

+ **suggested algorithm**:   
    + with the x and y coordinates, first produce a scatterplot (HINT: use the "plot()" function)
    + if smooth is FALSE, then proceed to draw a straight line as before (and return the coefficients)
    + if smooth is TRUE, use the "scatter.smooth()" function to plot a smoothed, locally-weighted regression of the y variable on the x variable. Make sure you use the "span" argument!
    + if smooth is TRUE, use the "loess()" function to record the same smoothed, locally-weighted regression of the y variable on the x variable. Again, make sure you use the "span" argument!

+ **return**:    
    + out = the loess model (the output produced by running "loess()" (or the slope and intercept from the linear regression, if smooth=FALSE) 
        

```{r answer1c, echo=FALSE}

#############
## exercise 1c

DrawLine2 <- function(x,y,smooth=TRUE,span=1){  
  plot(y~x)
  if(smooth==FALSE){
    mod <- lm(y~x)
    coefs <- coef(mod)
    abline(coefs)
  }else{
    coefs <- loess(y~x,span=span)
    scatter.smooth(y~x,span=span)
  }
  return(coefs)     
}

```

```{r test1c, echo=FALSE, context='data', warning=FALSE, quietly=T,fig.keep='none',results=FALSE}
xvec <- c(1:10)
set.seed(100) 
yvec <- rnorm(length(xvec),c(2:6,7:3),2)
answer1c <- DrawLine2(xvec,yvec,smooth=T,span=0.5)    # run your new function!

DrawLine2(x=trees$Height,y=trees$Volume,smooth=F,span=NA)

DrawLine2(faithful$waiting,faithful$eruptions,span=.5)    # test using the old faithful eruptions data

DrawLine2(faithful$waiting,faithful$eruptions,span=.1)

```

You can use this "sandbox" (below) to develop and test your function!

```{r q1c, exercise=TRUE}
xvec <- c(1:10)
set.seed(100)    # don't change this line, or the answer checking function won't work!
yvec <- rnorm(length(xvec),c(2:6,7:3),2)
DrawLine2 <- function(x,y,smooth=F,span=1){
    # [add code here!]
}
DrawLine2(xvec,yvec,smooth=T,span=0.5)

```

<div id="q1c-hint">
**Hint:** You may want to use the `scatter.smooth()` function to draw the curve and the 'loess()' function to build the model.
</div>

```{r q1c-check}
check_q1c <- function(USER_CODE){
    code <- checkr::for_checkr(USER_CODE)
    func_call <- checkr::line_where(code,checkr::insist(F== "DrawLine2","Use DrawLine2() function!"))
    t1 <- checkr::line_where(func_call, 
        checkr::insist(all(V$fitted == answer1c$fitted), 
        "Sorry, that's not correct"),
        checkr::passif(TRUE, "Fantastic!"))
}
check_q1c(USER_CODE)
```



```{r playwithcheckr2, eval=FALSE, echo=FALSE}

code <- checkr::for_checkr(quote({xvec <- c(1:10)
                          set.seed(100)    # don't change this line, or the answer checking function won't work!
                          yvec <- rnorm(length(xvec),c(2:6,7:3),2)
                          DrawLine2 <- function(x,y,smooth=F,span=1){
                              plot(y~x)
                              if(smooth==FALSE){
                                mod <- lm(y~x)
                                coefs <- coef(mod)
                                abline(coefs)
                              }else{
                                coefs <- loess(y~x,span=span)
                                scatter.smooth(y~x,span=span)
                              }
                              return(coefs)    
                          }
                          DrawLine2(xvec,yvec,smooth=T,span=0.5)}))
code
#func_call <- line_calling(code, DrawLine2, message = "Use DrawLine2() to construct the model.")
func_call <- checkr::line_where(code,checkr::insist(F== "DrawLine2","Use DrawLine2() to construct the model."))
#func_call <- line_binding(code, DrawLine2, message = "Use DrawLine2() to construct the model.")
t1 <- checkr::line_where(func_call, 
        checkr::insist(all(V$fitted == answer1c$fitted) , "Sorry, that's not correct"),
        checkr::passif(TRUE, "Great job!"))


code <- checkr::for_checkr(quote(lm(mpg ~ hp, data= mtcars)))
code
lm_line <- line_calling(code, lm, message = "Use lm() to construct the model.")
lm_call <- arg_calling(lm_line, lm)
t1 <- data_arg(lm_call, 
                 checkr::insist(identical(V, mtcars), 
                        "Your data argument {{E}} was not `mtcars`."),
                 message = "You didn't supply a `data = ` argument to `lm()`.")

# t1 <- checkr::line_where(code,checkr::insist(V %same_as% answer1c, message = "Sorry, try again!"))
# checkr::passif(any(func_call$values[[length(func_call$values)]]$fitted==answer1c$fitted),"Right!")
# #t1 <- checkr::line_where(func_call, checkr::insist(V==8.49, "Sorry, this is not the correct result"),checkr::passif(TRUE, "Fantastic!"))



```

## **Exercise 1d** (for R script) 

Write a function called "CLTdemo()" based on the central limit theorem (CLT) demonstration code above. This function should be specified as follows:

+ **input**:  
    + n.samples = number of independent random samples to draw from the specified distribution (default = 1000)   
    + sample.size = sample size (length of each independent random sample) (default = 10)
    + min = lower bound of the uniform distribution to draw from (default=10)
    + max = upper bound of the uniform distribution (default=20)

+ **suggested algorithm**:   
    + see demonstration above!
    + generate side-by-side plots of the histogram of sample means (left) and a quantile-quantile plot to test for normality.

+ **return**:  
    + out = the Shapiro-Wilks normality test results (the output produced by running "shapiro.test()")  
    
        
```{r answer1d, echo=FALSE}

#############
## exercise 1d

####################
# CENTRAL LIMIT THEOREM function
####################


CLTdemo <- function(n.samples=1000,sample.size=10,min=10,max=20){
  lots <- 100000       # number approximating infinity      
  datafountain <- runif(lots,min,max)      # here we define the full set of possible random numbers to draw random samples from (the population of potential data)
  
  #######
  # Draw multiple samples from the pool (population) of possible data.  
  
  samplemean <- numeric(n.samples)     # set up storage vector
  i=1
  for(i in 1:n.samples){      # for each replicate (independent random sample)
    sample <- sample(datafountain,sample.size)   # draw an independent random sample from the population of interest
    samplemean[i] <- mean(sample)    # compute and record the sample mean
  }
  
  
  par(mfrow=c(1,2))
  hist(datafountain,freq=F,ylim=c(0,1),main="",xlab="Value")      # plot out the distribution of sample means
  hist(samplemean,freq=F,add=T,col="red")    # overlay the distribution of the underlying data from which we are drawing samples.
  legend("topleft",fill=c("white","red"),legend=c("data population","sample means"),bty="n")
  qqnorm(samplemean)
  
  
  out <- shapiro.test(samplemean)
  
  return(out)
}



```

```{r test1d, echo=FALSE, eval=FALSE, context='data', warning=FALSE, quietly=T,fig.keep='none',results=FALSE}

answer1d <- CLTdemo(n.samples=5000,sample.size=4,min=10,max=20)    # run your new function!


```

You can use this "sandbox" (below) to develop and test your function!

```{r q1d, exercise=TRUE}
CLTdemo <- function(n.samples=1000,sample.size=10,min=10,max=20){
    # [add code here!]
}
CLTdemo(n.samples=5000,sample.size=4,min=10,max=20)

```

<div id="q1d-hint">
**Hint:** Use the CLT Demo code from earlier in Lab 1! 
</div>

```{r q1d-check}
check_q1d <- function(USER_CODE){
    code <- checkr::for_checkr(USER_CODE)
    func_call <- checkr::line_where(code,checkr::insist(F== "CLTdemo","Use CLTdemo() function!"))
    t1 <- checkr::line_where(func_call, 
        checkr::insist(V$p.value <= 0.01, 
        "Sorry, that's not correct"),
        checkr::passif(TRUE, "Fantastic!"))
}
check_q1d(USER_CODE)
```


## **Exercise 1e**  (for Word doc)

Finally, test the function out for different parameter combinations to make sure it works! See if you can use this function to develop a reasonable rule for how big a sample size is necessary to ensure that the sample mean is normally distributed given that the underlying data population is distributed along a uniform distribution. *Please include your answer in your Word document*- and please justify your answer! [note: you don't need to write new code to answer this question, I am just looking for a thoughtful response, not a definitive answer! And just a few sentences is fine.]
    

```{r answer1e, echo=FALSE, eval=FALSE}

##############
# Exercise 1e

reps <- 50
sizes <- seq(3,10,1)
prop <- numeric(length(sizes))
i=sizes[1]
counter=1
for(i in sizes){
  okay <- logical(reps)
  j=1
  for(j in 1:reps){
    temp <- CLTdemo(n.samples=1000,sample.size=i,min=10,max=20,plot=FALSE)    # run your new function!
    okay[j] <- temp$p.value<=0.05
  }
  num <- length(which(okay))
  prop[counter] <- num/reps   # proportion rejected
  counter=counter+1
}

par(mfrow=c(1,1))
names(prop) <- sizes
barplot(prop,xlab="sample size",ylab="Normality rejected (proportion)")




```

## **Exercise 1f (optional)**

Optionally, Modify your CLTdemo() function (e.g., call it "CLTdemo2()") to try a different underlying data distribution (other than uniform). If you really want to test the limits of the CLT, try creating your own highly non-standard distribution and seeing if the CLT still holds (spoiler alert: it does!). For example:
    

```{r}
rlocodist <- function(n){
  vals <- c(1,7,10,35)         # possible data values
  probs <- c(1,2,5,0.5)     # relative probability of each data values
  probs <- probs/sum(probs)    
  vals[apply(rmultinom(n,1,probs),2,function(t) which(t==1))]     # sample from this made-up distribution
}

lots=10000
datafountain <- rlocodist(lots)

hist(datafountain, main="non-standard made-up distribution")
```

You can use this "sandbox" (below) to develop and test your function!

```{r q1f, exercise=TRUE}

CLTdemo2 <- function(n.samples=1000,sample.size=10,...){
    # [add code here!]
}
CLTdemo2(n.samples=5000,sample.size=4,...)

```

<div id="q1f-hint">
**Hint:** Use the CLT Demo code from earlier in Lab 1! 
</div>


## Aside: default values in functions

**NOTE:** to set default values, just use the equals sign when specifying your function. For example, say I wanted to write a function that adds numbers in a vector. It might look something like this:

```{r}

newsum <- function(x=c(1,2,4)){
  sm <- sum(x)
  return(sm)
}

newsum(x=c(5:10))   # specify x manually

newsum()    # use default value!

```

## Multiple Regression 1: Air Quality Data 

The following is a refresher on performing multiple regression analyses in R:

1. Type the following for a list of sample datasets that come with the core R package (some of these you have already encountered).  

```{r eval=FALSE}

library(help = "datasets")    # list of sample datasets that come with R

?airquality

```

2. Examine the 'airquality' dataset (use the 'head' and 'summary' functions). Note that there are *missing values* where ozone concentration data and solar radiation data were not collected.  

3. We could ignore the missing values and just go ahead with our regression analysis, since the default response of the "lm()" ('linear model') function is to omit cases with missing values in any of the specified parameters. However, to avoid problems later, we will omit them by constructing a new, 'cleaned' dataset as follows:

```{r}

air.cleaned <- na.omit(airquality)    # remove rows with missing data

```

4. Conduct a multiple linear regression of ozone concentration as a function of solar radiation, wind and temperature. Use the 'lm()' function to conduct an ordinary least squares (OLS) regression analysis.    

5. Explore the regression outputs using the 'summary' function, and explore regression diagnostics using, e.g. (depending on what you named the regression model object):

```{r include=FALSE}

model1 <- lm(Ozone~Solar.R+Wind+Temp,data=air.cleaned)

```

```{r}

par(mfrow=c(3,2))
plot(model1, which=c(1:4))   # diagnostic plots (NOTE: the plot function returns these plots by default when the input is a linear regression model)


hist(residuals(model1), breaks=10)   # histogram of residuals
plot(predict(model1) ~ air.cleaned$Ozone)     # plot predicted vs observed- should follow 1:1 line. Examine this for model biases.
abline(0,1)

```

NOTE: see [this website](https://data.library.virginia.edu/diagnostic-plots/) for more information on the diagnostic plots produced by lm().

If no one in your group knows why you are doing any of this or what it all means, ask the instructor! That’s why he’s hanging around the lab…  

6. Consider the possibility that there may be an important interaction effect between solar radiation and temperature on influencing ozone concentrations. Explore that with a simple scatter plot where symbol size is scaled to ozone concentration:

```{r}

symbols(air.cleaned$Temp, air.cleaned$Solar.R, circles=air.cleaned$Ozone/100, ylab="Solar Radiation", xlab="Temperature", main="Interaction Plot", inches=FALSE)

       # alternatively...
coplot(air.cleaned$Ozone~air.cleaned$Temp|air.cleaned$Solar.R,rows=1)


```

7. Now fit a second model that includes the interaction between solar radiation and temperature. Use the following formula to fit the interaction:

```{r}
formula2 <- "Ozone ~ Wind + Solar.R * Temp"   # you can name formulas...
```

```{r include=FALSE}

model2 <- lm(formula2,data=air.cleaned)

```

8. Explore regression outputs for the second model in the same way as you did for the first model without the interaction term.

9. Conduct an 'F Test' (or a Likelihood Ratio Test, LRT, if you prefer...) to formally test whether the more complex model (including the interaction term) fits the data significantly better than the reduced model (with fewer parameters) that lacks the interaction term. Recall that the $R^2$ value is inadequate for this purpose because $R^2$ will always increase with additional parameters! Use the following syntax,

```{r eval=FALSE}

anova(model1, model2, test="F")   # how would you run an LRT test instead?

```

You can use this "sandbox" (below) to play around with multiple regression in R!

```{r q2prelim, exercise=TRUE}

# your code here

```


## Exercise 2: regression in R (written responses)

Very briefly answer the following questions in your Word document:  

* **Exercise 2a**	On average, by how much does ozone concentration increase (or decrease) for each 10-unit increase in temperature? Use the 'no interaction' model from part 4 above to answer this question.    

*	**Exercise 2b** What is the (null) hypothesis that the *p*-values for the individual regression coefficients are designed to test?  



## Multiple Regression 2: Noble fir data
For this exercise, we will use tree data from a forest in the western Oregon Cascades.  

We will fit a multiple linear regression model that predicts forest tree biomass as a function of environmental variables (including a mix of continuous and categorical predictors) and stand age.

Obtain the TreeData.csv file from the "Data Sets" tab (or just download [here](http://naes.unr.edu/shoemaker/teaching/NRES-746/TreeData.csv))- save it to your working directory.

This describes a subset of forest inventory data from the Douglas-fir forests of western Oregon (n = 90, 0.1-ha sites).

Arranged in columns from left to right, variables are:

* Site: site identifier  
* Biomass: tree biomass (for all species) in Mg/ha, the response variable for Part 1 of the lab.  
* ABPR: Presence/absence of Abies procera (noble fir) on a given site (coded 1 for presence).  
* StandAge: Maximum tree age in the 0.1-ha plot. This variable will be used as a proxy for successional stage. We assume that stand-replacing fires are the dominant form of disturbance and that stand age is a reasonable proxy variable for time since the last fire.    
* X, Y: geographic coordinates – UTM easting and northing, respectively    
* Elev: elevation (m)      
* Northeastness: slope aspect that has been linearized using a cosine transformation so that the aspect of 45 degrees has value 1 and aspect of 225 degrees has value -1. In this study area, this variable is expected to reflect a moisture gradient from moister (NE) to drier (SW) aspects.      
* Slope: slope steepness (degrees)     
* SlopePos: slope position, a categorical variable (i.e. factor) with three values: Valley, Slope and Ridge.     

This is a comma-delimited (.csv) file, which is a common file format for importing data into R. Import the data into R as a data frame (R’s version of an excel spreadsheet), using the following command:

```{r}

NobleFir.df <- read.csv("TreeData.csv")

```

Inspect the resulting data object. Summarize it using the 'summary()' and 'plot()' functions.   

Obtain a correlation matrix for biomass and the four numeric predictor variables using the 'cor()' function and by subscripting column locations on the data frame (ask instructor for explanation of syntax if needed):

```{r eval=FALSE}

cor(NobleFir.df[,c(2,4,7:9)])

```

Are any of the predictor variables highly correlated?

Calculate *Box Plots* for the continuous predictor variables (excluding x and y coordinates) according to sites with or without noble fir. Use the 'boxplot()' function. What clear relationships, if any, emerge for how sites with and without noble fir differ with regard to their environmental setting? For example:

```{r echo=FALSE}

boxplot(NobleFir.df$Northeastness~NobleFir.df$ABPR, xlab="Presence of Noble Fir", ylab="Northeastness")   # for example

```


Use multiple linear regression to model *A. procera* biomass as a function of predictor variables (excluding spatial coordinates), using the same approach for regression fitting and diagnostics as we did previously.  

Re-run the regression to obtain standardized regression coefficients, allowing direct comparison of effect sizes for the continuous predictor variables (since all variables are then transformed to standard deviate units, i.e. mean centered on zero with standard deviation of one). The 'scale' function provides an easy way to implement this.

```{r}

Biomass_std.lm <- with(NobleFir.df,     # using the "with()" statement, we don't need to keep referencing the name of the data frame.
  lm(scale(Biomass) ~ scale(elev) + scale(Northeastness) + scale(Slope) + SlopePos + scale(StandAge))
)

```


Visually assess whether regression errors (residuals) are spatially autocorrelated using the 'symbols' function:

```{r}

with(NobleFir.df,
  symbols(x,y,circles=abs(residuals(Biomass_std.lm)), inches=0.3, ylab="Northing", xlab="Easting", main="Errors from Biomass Regression Model")
)

```


You can use this "sandbox" (below) to play around with this example in R!

```{r q3prelim, exercise=TRUE}

# your code here

```

## Exercise 3: noble fir regression

Answer the following questions (with brief justification) in your Word document:

* **Exercise 3a** Can forest biomass be reliably predicted by topographic variables and stand age? Explain your reasoning.     
* **Exercise 3b** Is there spatial variation in model goodness of fit? Explain your reasoning.  
* **Exercise 3c** Which of the environmental influences are most important in predicting forest biomass? Explain your reasoning.     


## Exercise 4: Algorithmic (brute force) t-test

Review the "brute-force t-test" code from the ["Why focus on algorithms" lecture](https://kevintshoemaker.github.io/NRES-746/LECTURE1.html). Then complete the following exercises: 

```{r answer4a0, echo=FALSE}

#############
# CHALLENGE 4: brute force t-tests!
#############

##### from lecture...

t.test.algorithm <- function(dat = reshape_df, group = "Treatment", value = "Mass" ){
  
  #############
  # Compute the sample statistic
  #############
  
  indexA <- which(dat[,group]=="A")     # rows representing treatment A
  indexB <- which(dat[,group]=="B")     # rows representing treatment B
  observed_dif <- mean(dat[indexA,value]) - mean(dat[indexB,value])
  
  sample.size <- length(indexA)
  
  #############
  # Simulate the STATISTICAL POPULATION under the null hypothesis
  #############
  
  lots <- 1000000  # large number approximating infinity 
  
  popMean_null <- mean(dat[,value])           # assume groups A and B come from a population with common mean 
  popSD_null <- sd(dat[,value])                      # and common standard deviation... 
  popData_null <- rnorm(n=lots,mean=popMean_null,sd=popSD_null)    # the statistical "population" of interest (under null model w no treatment effect)

  #################
  # Repeat sampling process (sampling from population) using a FOR loop
  #################
  
  reps <- 1000                 # set the number of replicates
  null_difs <- numeric(reps)       # initialize a storage structure to hold one anomaly (sampling error) per replicate
  
  for(i in 1:reps){            # for each replicate... 
    sampleA <- sample(popData_null,size=sample.size)      # draw a sample assuming no treatment effect       
    sampleB <- sample(popData_null,size=sample.size)      # draw a sample assuming no treatment effect (again!)
    null_difs[i] <- mean(sampleA)-mean(sampleB)           # compute and store the sampling error produced under the null hypothesis
  }
  
  ordered_difs <- sort(abs(null_difs))       # sort the vector of sampling errors 
  higher_anomaly <- length(which(ordered_difs>=abs(observed_dif)))       # how many of these sampling errors equal or exceed the sample statistic?
  p_value <- higher_anomaly/reps
  
  to_return <- list()   # initialize object to return
  
  to_return$null_difs <- null_difs
  to_return$p_value <- p_value
  to_return$observed_dif <- observed_dif
  
  return(to_return)

}


```


## Exercise 4a

What if we wanted to run a *one-tailed t-test*? That is, what if our alternative hypothesis were that lizards in treatment group "A" would tend to have lower body mass than treatment group "B"? Modify the function ("t.test.algorithm()") to do this! Name your new function "t.test.onetail()". 

To convince yourself that your new function works, try running your function for a case where the "group A" body mass data are generally higher than group B- the opposite of your (alternative) hypothesis!

NOTE: you may get tangled up with the null hypothesis/p-value concept, which is admittedly a difficult concept! A p-value always assumes the null hypothesis is true (given the null hypothesis is true, a p-value gives the probability of obtaining a test statistic that lends as much or more support to the alternative hypothesis than the observed test statistic). The alternative hypothesis for the 1-tailed test is that the mean for treatment A is less than the mean for treatment B. The null hypothesis, as always, is that there is no difference between treatments. Therefore, if you hypothesize that A (which is actually much larger) is less than B, then the p-value should reflect a near-complete inability to reject the null hypothesis (very high p-value) -- that is, nearly all possible datasets generated under the null hypothesis lend as much or more support to the idea that A is less than B than the observed data!   


```{r answer4a, echo=FALSE}

###############
# Question 4a

t.test.onetail <- function(dat = reshape_df, group = "Treatment", value = "Mass" ){
  
  #############
  # Compute the sample statistic
  #############
  
  indexA <- which(dat[,group]=="A")     # rows representing treatment A
  indexB <- which(dat[,group]=="B")     # rows representing treatment B
  observed_dif <- mean(dat[indexB,value]) - mean(dat[indexA,value])
  
  sample.size <- length(indexA)
  
  #############
  # Simulate the STATISTICAL POPULATION under the null hypothesis
  #############
  
  lots <- 1000000  # large number approximating infinity 
  
  popMean_null <- mean(dat[,value])           # assume groups A and B come from a population with common mean 
  popSD_null <- sd(dat[,value])                      # and common standard deviation... 
  popData_null <- rnorm(n=lots,mean=popMean_null,sd=popSD_null)    # the statistical "population" of interest (under null model w no treatment effect)

  #################
  # Repeat sampling process (sampling from population) using a FOR loop
  #################
  
  reps <- 1000                 # set the number of replicates
  null_difs <- numeric(reps)       # initialize a storage structure to hold one anomaly (sampling error) per replicate
  
  for(i in 1:reps){            # for each replicate... 
    sampleA <- sample(popData_null,size=sample.size)      # draw a sample assuming no treatment effect       
    sampleB <- sample(popData_null,size=sample.size)      # draw a sample assuming no treatment effect (again!)
    null_difs[i] <- mean(sampleB)-mean(sampleA)           # compute and store the sampling error produced under the null hypothesis
  }
  
  ordered_difs <- sort(null_difs)       # sort the vector of sampling errors 
  higher_anomaly <- length(which(ordered_difs>=observed_dif))       # how many of these sampling errors equal or exceed the sample statistic?
  p_value <- higher_anomaly/reps
  
  to_return <- list()   # initialize object to return
  
  to_return$null_difs <- null_difs
  to_return$p_value <- p_value
  to_return$observed_dif <- observed_dif
  
  return(to_return)

}


```


```{r test4a1}

df <- data.frame(
  TreatmentA = c(175, 168, 168, 190, 156, 181, 182, 175, 174, 179),
  TreatmentB = c(185, 169, 173, 173, 188, 186, 175, 174, 179, 180) 
)

```

Note that we need to reshape the data frame before entering it into our function. Since we will need to do this again, let's make a function to do this!

```{r test4a2, context="data", echo=TRUE}

reshapefunc <- function(df,varname){
  sample.size <- nrow(df)     # determine sample size 

  reshape_df <- data.frame(                # "reshape" the data frame so each observation gets its own row (standard format)
    Treatment = rep(c("A","B"),each=sample.size),
    Value = c(df[,1],df[,2])
  )
  
  names(reshape_df)[2] <- varname   # rename the value column to whatever the user specifies
  return(reshape_df)
}

```

```{r test4a3}

dfnew <- reshapefunc(df,"Mass")

ttest_onetail <- t.test.onetail(dat=dfnew,group = "Treatment", value = "Mass" )
ttest_onetail$p_value

ttest_twotail <- t.test.algorithm(dat=dfnew,group = "Treatment", value = "Mass" )    # try two-tailed test
ttest_twotail$p_value

t.test(df$TreatmentA,df$TreatmentB,alternative = "less")          # compare with true t-test
t.test(df$TreatmentA,df$TreatmentB,alternative = "two.sided")


```

```{r test4a4, echo=FALSE, context='data'}

df <- data.frame(
      TreatmentA = c(105, 118, 119, 112, 109),
      TreatmentB = c(138, 130, 150, 123, 144)
    )
df2<-reshapefunc(df,varname="Mass")

answer4a <- t.test.onetail(dat=df2,group = "Treatment", value = "Mass")


```

Include your function in your submitted r script!

You can use this "sandbox" (below) to develop and test your function!

```{r q4a, exercise=TRUE}
df <- data.frame(
      TreatmentA = c(105, 118, 119, 112, 109),
      TreatmentB = c(138, 130, 150, 123, 144)
    )
df2<-reshapefunc(df,varname="Mass")
t.test.onetail <- function(dat, group, value){
    # [add code here!]
}
t.test.onetail(dat=df2,group = "Treatment", value = "Mass")

```

<div id="q4a-hint">
**Hint:** No hints ... yet!
</div>


```{r q4a-check}
check_q4a <- function(USER_CODE){
    code <- checkr::for_checkr(USER_CODE)
    func_call <- checkr::line_where(code,checkr::insist(F== "t.test.onetail","Use t.test.onetail() function!"))
    t1 <- checkr::line_where(func_call, 
        checkr::insist(all(V$p_value <= 0.02), 
        "Sorry, that's not correct"),
        checkr::passif(TRUE, "Very nice!"))
}
check_q4a(USER_CODE)
```



## Exercise 4b

What if we wanted to *relax* the assumption of equal variances? That is, what if our null hypothesis were that each treatment group has the same mean, but they could differ in variance? Modify the function ("t.test.algorithm()") to do this! To enable automated grading, please name your new function "t.test.vardif()". To convince yourself that your new function works, try running the following code:

```{r answer4b, echo=FALSE}

###############
# A possible answer to question 4b

t.test.vardif <- function(dat = df.vardif2, group = "Treatment", value = "Mass" ){
  
  #############
  # Compute the sample statistic
  #############
  
  indexA <- which(dat[,group]=="A")     # rows representing treatment A
  indexB <- which(dat[,group]=="B")     # rows representing treatment B
  observed_dif <- mean(dat[indexA,value]) - mean(dat[indexB,value])
  
  sample.size <- length(indexA)
  
  varA <- sd(dat[indexA,value])    # estimate variance of each group separately
  varB <- sd(dat[indexB,value])
  
  #############
  # Simulate the STATISTICAL POPULATION under the null hypothesis
  #############
  
  lots <- 1000000  # large number approximating infinity 
  
  popMean_null <- mean(dat[,value])           # assume groups A and B come from a population with common mean 
  popData_nullA <- rnorm(n=lots,mean=popMean_null,sd=varA)    # the statistical "population" of interest (under null model w no treatment effect but allowing variances to differ)
  popData_nullB <- rnorm(n=lots,mean=popMean_null,sd=varB)    # the statistical "population" of interest (under null model w no treatment effect but allowing variances to differ)
  #################
  # Repeat sampling process (sampling from population) using a FOR loop
  #################
  
  reps <- 1000                 # set the number of replicates
  null_difs <- numeric(reps)       # initialize a storage structure to hold one anomaly (sampling error) per replicate
  
  for(i in 1:reps){            # for each replicate... 
    sampleA <- sample(popData_nullA,size=sample.size)      # draw a sample assuming no treatment effect       
    sampleB <- sample(popData_nullB,size=sample.size)      # draw a sample assuming no treatment effect (again!)
    null_difs[i] <- mean(sampleB)-mean(sampleA)           # compute and store the sampling error produced under the null hypothesis
  }
  
  ordered_difs <- sort(abs(null_difs))       # sort the vector of sampling errors 
  higher_anomaly <- length(which(ordered_difs>=abs(observed_dif)))       # how many of these sampling errors equal or exceed the sample statistic?
  p_value <- higher_anomaly/reps
  
  to_return <- list()   # initialize object to return
  
  to_return$null_difs <- null_difs
  to_return$p_value <- p_value
  to_return$observed_dif <- observed_dif
  
  return(to_return)

}


```


```

ttest_vardif <- t.test.vardif(dat=df.vardif)

```

where "df.vardif" is the following dataframe (where variance differs dramatically between the two groups):

```{r test4b1}

###########
# Test data for unequal variances (prior to reshaping)...

df.vardif <- data.frame(
  TreatmentA = c(135, 128, 139, 122, 126, 121, 128, 135, 134, 129),
  TreatmentB = c(215, 69, 143, 153, 218, 186, 125, 98, 271, 340)
)

summary(df.vardif)    # summarize!

```

Now we can test our new t-test function!

```{r test4b2, eval=FALSE, echo=FALSE}

df.vardif2 <- reshapefunc(df.vardif,"Mass")

ttest_vardif <- t.test.vardif(dat=df.vardif2, group = "Treatment", value = "Mass")    # test new function
ttest_vardif$p_value

ttest_varequal <- t.test.algorithm(dat=df.vardif2)    # versus original function
ttest_varequal$p_value

t.test(df.vardif$TreatmentA,df.vardif$TreatmentB, var.equal = FALSE, alternative="two.sided")   # versus built-in function



```

```{r test4b3, echo=FALSE, context='data'}

df <- data.frame(
      TreatmentA = c(1135, 1128, 1139, 1122, 1126),
      TreatmentB = c(1915, 69, 3143, 53, 1818)
    )
df2<-reshapefunc(df,varname="Mass")
    
answer4b <- t.test.vardif(dat=df2,group = "Treatment", value = "Mass")


```

Include your function in your submitted r script!

You can use this "sandbox" (below) to develop and test your function!

```{r q4b, exercise=TRUE}
df <- data.frame(
      TreatmentA = c(1135, 1128, 1139, 1122, 1126),
      TreatmentB = c(1915, 69, 3143, 53, 1818)
    )
df2<-reshapefunc(df,varname="Mass")
    
t.test.vardif <- function(dat, group, value){
    # [add code here!]
}
t.test.vardif(dat=df2,group = "Treatment", value = "Mass")

```

<div id="q4b-hint">
**Hint:** No hints ... yet!
</div>


```{r q4b-check}
check_q4b <- function(USER_CODE){
    code <- checkr::for_checkr(USER_CODE)
    func_call <- checkr::line_where(code,checkr::insist(F== "t.test.vardif","Use t.test.vardif() function!"))
    t1 <- checkr::line_where(func_call, 
        checkr::insist(all((V$p_value >=  (answer4b$p_value-0.2))&(V$p_value <=  (answer4b$p_value+0.2))), 
        "Sorry, that's not correct"),
        checkr::passif(TRUE, "Bravo!"))
}
check_q4b(USER_CODE)
```



## Exercise 4c 

Modify the above function ("t.test.vardif()") to allow for unequal sample sizes AND unequal variances. That is, our null hypothesis is the same (equal treatment means, potentially unequal variances), but now we enable each treatment group to have different sample sizes! To enable automated grading, please name your new function "t.test.ndif()". To convince yourself that your new function works, try running the following code:

```{r answer4c, echo=FALSE}

###############
# A possible answer to question 4c

t.test.ndif <- function(dat = df.ndif2, group = "Treatment", value = "Mass" ){
  
  #############
  # Compute the sample statistic
  #############
  
  levs <- levels(dat[,group])
  
  indexA <- which(dat[,group]==levs[1])     # rows representing treatment A
  indexB <- which(dat[,group]==levs[2])     # rows representing treatment B
  observed_dif <- mean(dat[indexA,value],na.rm=TRUE) - mean(dat[indexB,value],na.rm=T)
  
  sample.size <- tapply(dat[,value],dat[,group],function(t) length(which(!is.na(t))))

  
  varA <- sd(dat[indexA,value],na.rm = T)    # estimate variance of each group separately
  varB <- sd(dat[indexB,value],na.rm=T)
  
  #############
  # Simulate the STATISTICAL POPULATION under the null hypothesis
  #############
  
  lots <- 1000000  # large number approximating infinity 
  
  popMean_null <- mean(dat[,value],na.rm=TRUE)           # assume groups A and B come from a population with common mean 
  popData_nullA <- rnorm(n=lots,mean=popMean_null,sd=varA)    # the statistical "population" of interest (under null model w no treatment effect but allowing variances to differ)
  popData_nullB <- rnorm(n=lots,mean=popMean_null,sd=varB)    # the statistical "population" of interest (under null model w no treatment effect but allowing variances to differ)
  #################
  # Repeat sampling process (sampling from population) using a FOR loop
  #################
  
  reps <- 1000                 # set the number of replicates
  null_difs <- numeric(reps)       # initialize a storage structure to hold one anomaly (sampling error) per replicate
  
  for(i in 1:reps){            # for each replicate... 
    sampleA <- sample(popData_nullA,size=sample.size[1])      # draw a sample assuming no treatment effect       
    sampleB <- sample(popData_nullB,size=sample.size[2])      # draw a sample assuming no treatment effect (again!)
    null_difs[i] <- mean(sampleB)-mean(sampleA)           # compute and store the sampling error produced under the null hypothesis
  }
  
  ordered_difs <- sort(abs(null_difs))       # sort the vector of sampling errors 
  higher_anomaly <- length(which(ordered_difs>=abs(observed_dif)))       # how many of these sampling errors equal or exceed the sample statistic?
  p_value <- higher_anomaly/reps
  
  to_return <- list()   # initialize object to return
  
  to_return$null_difs <- null_difs
  to_return$p_value <- p_value
  to_return$observed_dif <- observed_dif
  
  return(to_return)

}


```

```

ttest_ndif <- t.test.ndif(dat=df.ndif)

```

where "df.ndif" is the following dataframe:

```{r test4c1}

###########
# Test data for unequal sample sizes (before reshaping)...

df.ndif <- data.frame(
  TreatmentA = c(135, 128, 139, 122, 126, 121, 128, 135, 134, 129, 134, 125, 130, 132, 125),
  TreatmentB = c(98, 271, 340, rep(NA,times=12))
)

summary(df.ndif)    # summarize!

```

```{r test4c2, eval=FALSE, echo=FALSE}

df.ndif2 <- reshapefunc(df.ndif,"Mass")

ttest_ndif <- t.test.ndif(dat=df.ndif2)    # test new function
ttest_ndif$p_value

```

```{r test4c3, echo=FALSE, context='data'}

df <- data.frame(
      TreatmentA = c(135, 128, 139, 122, 126,rep(NA,3)),
      TreatmentB = c(315, 69, 143, 53, 818,4,55,190)
    )
df2<-reshapefunc(df,varname="Mass")
answer4c <- t.test.ndif(dat=df2,group = "Treatment", value = "Mass")

```

Note: data frames require that all columns have the same number of elements. If one vector (column) is shorter than the others, R *recycles* the shorter vector, repeating the vector until it is the same length as the longer vector. To avoid this unwanted behavior, we can use the "NA" designation to tell R that no data exist. However, you will have to modify your function to ensure that it can handle missing values! 

Include your function in your submitted r script!

You can use this "sandbox" (below) to develop and test your function!

```{r q4c, exercise=TRUE}
df <- data.frame(
  TreatmentA = c(135, 128, 139, 122, 126,rep(NA,3)),
  TreatmentB = c(315, 69, 143, 53, 818,4,55,190)
)
df2<-reshapefunc(df,varname="Mass")
    
t.test.ndif <- function(dat, group, value){
    # [add code here!]
}
t.test.ndif(dat=df2,group = "Treatment", value = "Mass")

```

<div id="q4c-hint">
**Hint:** No hints ... yet!
</div>


```{r q4c-check}
check_q4c <- function(USER_CODE){
    code <- checkr::for_checkr(USER_CODE)
    func_call <- checkr::line_where(code,checkr::insist(F== "t.test.ndif","Use t.test.ndif() function!"))
    t1 <- checkr::line_where(func_call, 
        checkr::insist(all((V$p_value >=  (answer4c$p_value-0.15))&(V$p_value <=  (answer4c$p_value+0.15))), 
        "Sorry, that's not correct"),
        checkr::passif(TRUE, "Bravo!"))
}
check_q4c(USER_CODE)
```



## Exercise 5: Bootstrapping algorithm


```{r answer5a0, echo=FALSE}

#####################
# CHALLENGE 5: Bootstrapping regression coefficients!
#####################

# first, grab the code from the lecture (bootstrapping R-squared)

#########
# Function for returning a vector of R-squared statistics from models regressing a response variable on multiple possible predictor variables
   # here we assume that all columns in the input data frame that are NOT the response variable are potential predictor variables.

Rsquared <- function(df,responsevar="Volume"){    # univariate models only- interaction and multiple regression not implemented here
  response <- df[,responsevar]       # extract the response variable
  names <- names(df)                  
  rsq <- numeric(length(names))        # named storage vector
  names(rsq) <- names(df)               
  rsq <- rsq[names(rsq)!=responsevar]           # assume that all columns that are not the response variable are possible predictor variables
  for(i in names(rsq)){         # loop through predictors
      predictor <- df[,i]                  # extract this predictor
      model <- lm(response~predictor)       # regress response on predictor
      rsq[i] <- summary(model)$r.square       # extract R-squared statistic
  }
  return(rsq)     
}

boot_sample <- function(df,statfunc,n_samples,n_stats,responsevar="Volume"){
  indices <- c(1:nrow(df))
  output <- matrix(NA,nrow=n_samples,ncol=n_stats)        # storage object- to store a single bootstrapped sample from the original data
  
  for(i in 1:n_samples){              # for each bootstrap replicate:
    boot_rows <- sample(indices,size=nrow(df),replace=T)         # randomly sample observations with replacement
    newdf <- df[boot_rows,]                       # dataframe of bootstrapped observations
    output[i,] <- statfunc(newdf,responsevar)                 # generate statistics from the bootstrapped sample  (e.g., compute Rsquared after regressing y on all possible x variables)
  }
  return(output)
}

# boot <- boot_sample(df=trees,statfunc=Rsquared,n_samples=1000,n_stats=2)   # generate test statistics (Rsquared vals) for 1000 bootstrap samples
# confint <- apply(boot,2,function(t)  quantile(t,c(0.025,0.5,0.975)))       # summarize the quantiles to generate confidence intervals for each predictor variable
# colnames(confint) <- names(stat)
# t(confint)


```


Review the bootstrapping code from the ["Why focus on algorithms" lecture](https://kevintshoemaker.github.io/NRES-746/LECTURE1.html) (to generate confidence intervals for arbitrary test statistics). Then complete the following exercise:

## Exercise 5a

Generate a new R function, called "RegressionCoefs()" that takes a data frame as the first input and the name of the response variable as the second input, and returns the (univariate) regression coefficients ($\beta$) produced by regressing the response variable on each predictor variable (returning a vector of regression coefficients). You can use the "Rsquared()" function (from the lecture) as a reference!

```{r answer5a, echo=FALSE}

#############
## exercise 5a

RegressionCoefs <- function(df=trees,responsevar="Volume"){    # univariate models only- interaction and multiple regression not implemented here
  response <- df[,responsevar]       # extract the response variable
  names <- names(df)                  
  coefs <- numeric(length(names))        # named storage vector
  names(coefs) <- names(df)               
  coefs <- coefs[names(coefs)!=responsevar]           # assume that all columns that are not the response variable are possible predictor variables
  i=names(coefs)[1]
  for(i in names(coefs)){         # loop through predictors
      predictor <- df[,i]                  # extract this predictor
      model <- lm(response~predictor)       # regress response on predictor
      coefs[i] <- coefficients(model)["predictor"]       # extract slope term
  }
  return(coefs)     
}


```

```{r test5a, echo=FALSE, eval=FALSE}

RegressionCoefs(df=trees,responsevar="Volume")

```

```{r test5a2, echo=FALSE, context='data'}

df <- mtcars[,c(1,3,4,6)]
answer5a <- coefficients(lm(df$mpg~df$hp))[2]

```


Include your function in your submitted r script!

You can use this "sandbox" (below) to develop and test your function!

```{r q5a, exercise=TRUE}
df <- mtcars[,c(1,3,4,6)]
   
RegressionCoefs <- function(df,responsevar){
    # [add code here!]
}
RegressionCoefs(df,"mpg")

```

<div id="q5a-hint">
**Hint:** No hints ... yet!
</div>


```{r q5a-check}
check_q5a <- function(USER_CODE){
    code <- checkr::for_checkr(USER_CODE)
    func_call <- checkr::line_where(code,checkr::insist(F== "RegressionCoefs","Use RegressionCoefs() function!"))
    t1 <- checkr::line_where(func_call, 
        checkr::insist(any((V ==  answer5a)), 
        "Sorry, that's not correct"),
        checkr::passif(TRUE, "Bravo!"))
}
check_q5a(USER_CODE)
```



## **Exercise 5b:**      

Generate a new R function, called "BootCoefs()" that meets the following specifications:     

+ **inputs**:    
    + "df" = a data frame that includes the response variable and all possible predictor variables     
    + "statfunc" = a function for generating summary statistics (regression coefficients) from a data frame (which you already developed in part 1 of this challenge problem)  
    + "n_samples" = the number of bootstrapped samples to generate    
    + "n_stats" = the number of predictor variables     
    + "responseVar" = the name of the response variable    [NOTE: this is optional since it was added after the lab was first assigned!]    

+ **algorithm**:
    + with the data frame, first use the "boot_sample()" function provided in the lecture to generate summary statistics for multiple bootstrap samples.   
    + Then, generate confidence intervals for each variable as the 2.5%, 50% and 97.5% quantile of the summary statistic for each predictor variables.

+ **return**: a matrix (rows=predictor vars, cols=2.5%, 50%, and 97.5% quantiles). Try to give the rows and columns proper labels!


```{r answer5b, echo=FALSE}

#############
## exercise 5b

boot_sample <- function(df=trees,statfunc=Rsquared,n_samples=1000,n_stats=2,responsevar="Volume"){
  indices <- c(1:nrow(df))
  output <- matrix(NA,nrow=n_samples,ncol=n_stats)        # storage object- to store a single bootstrapped sample from the original data
  
  for(i in 1:n_samples){              # for each bootstrap replicate:
    boot_rows <- sample(indices,size=nrow(df),replace=T)         # randomly sample observations with replacement
    newdf <- df[boot_rows,]                       # dataframe of bootstrapped observations
    output[i,] <- statfunc(newdf,responsevar)                 # generate statistics from the bootstrapped sample  (e.g., compute Rsquared after regressing y on all possible x variables)
  }
  return(output)
}


BootCoefs <- function(df=trees,statfunc=RegressionCoefs,n_samples=1000,n_stats=2,responsevar="Volume"){    # univariate models only- interaction and multiple regression not implemented here
  boot <- boot_sample(df=df,statfunc=statfunc,n_samples=n_samples,n_stats=ncol(df)-1,responsevar=responsevar)   # generate test statistics (Rsquared vals) for 1000 bootstrap samples
  confint <- apply(boot,2,function(t)  quantile(t,c(0.025,0.5,0.975)))       # summarize the quantiles to generate confidence intervals for each predictor variable
  colnames(confint) <- paste("stat",c(1:n_stats),sep="")
  return(t(confint))     
}


```

You can test your function using the following code:

```{r test5b, eval=FALSE}

BootCoefs(df=trees,statfunc=RegressionCoefs,n_samples=1000,n_stats=2,responsevar="Volume")

```

```{r test5b2, echo=FALSE, context="data"}

df <- mtcars[,c(1,3,4,6)]
responsevar="mpg"
answer5b <- BootCoefs(
              df=df,
              statfunc=RegressionCoefs,
              n_samples=1000,
              n_stats=3,
              responsevar=responsevar
            )

#answer5b <- round(confint(lm(df$mpg~df$disp))[2,],2)[1]

```

Include your function in your submitted r script!

You can use this "sandbox" (below) to develop and test your function!

```{r q5b, exercise=TRUE}
df <- mtcars[,c(1,3,4,6)]
responsevar="mpg"

boot_sample <- function(df=trees,statfunc=Rsquared,n_samples=1000,n_stats=2,responsevar="Volume"){
  indices <- c(1:nrow(df))
  output <- matrix(NA,nrow=n_samples,ncol=n_stats)        # storage object- to store a single bootstrapped sample from the original data
  
  for(i in 1:n_samples){              # for each bootstrap replicate:
    boot_rows <- sample(indices,size=nrow(df),replace=T)         # randomly sample observations with replacement
    newdf <- df[boot_rows,]                       # dataframe of bootstrapped observations
    output[i,] <- statfunc(newdf,responsevar)                 # generate statistics from the bootstrapped sample  (e.g., compute Rsquared after regressing y on all possible x variables)
  }
  return(output)
}

BootCoefs <- function(df,statfunc,n_samples,n_stats,responsevar){
    # [add code here!]
}
BootCoefs(df=df,statfunc=RegressionCoefs,
              n_samples=1000,n_stats=3,responsevar=responsevar)

```

<div id="q5b-hint">
**Hint:** No hints ... yet!
</div>


```{r q5b-check}
check_q5b <- function(USER_CODE){
    code <- checkr::for_checkr(USER_CODE)
    func_call <- checkr::line_where(code,checkr::insist(F== "BootCoefs","Use BootCoefs() function!"))
    t1 <- checkr::line_where(func_call, 
        checkr::insist(all((V[1,3] >=  (answer5b[1,3]-0.2))&(V[1,3] <=  (answer5b[1,3]+0.2))), 
        "Sorry, that's not correct"),
        checkr::passif(TRUE, "Delightful!"))
}
check_q5b(USER_CODE)
```


## **Exercise 5c:**      

To test your new function(s), generate bootstrap confidence intervals around the regression parameters for your tree biomass regression. Compare these bootstrapped confidence intervals with the standard confidence intervals given by R in the "lm()" function. Are there any important differences? Explain your answer in your Word document. As always, please use your Word document submission to raise any points of confusion or ask me questions!  

You can use this "sandbox" (below) to develop and test your function!

```{r q5c, exercise=TRUE}

BootCoefs(df=trees,statfunc=RegressionCoefs,n_samples=1000,n_stats=2,responsevar="Volume")

confint.lm(lm(Volume~Girth,trees))    # compare with lm() 

```


```{r test5c, echo=FALSE, eval=FALSE}

############
# Exercise 5c

BootCoefs(df=trees,statfunc=RegressionCoefs,n_samples=1000,n_stats=2,responsevar="Volume")

confint.lm(lm(Volume~Girth,trees))    # compare with lm() 



###########
# end of lab 1
###########

```

## End of Lab 1!



```{r eval=FALSE, echo=FALSE}
#    output:
#   html_document:
#     theme: cerulean
#     toc: yes
#     toc_float: yes
#     
#     
# {r, echo=FALSE}
# 
# shinyApp(
# 
#   ui = fluidPage(
#     textInput("testq", "", value = "", width = NULL,
#            placeholder = "Your code here"),
#      verbatimTextOutput("value")
#   ),
# 
#   server = function(input, output) {
#     
#     output$phonePlot = renderPlot({
#       barplot(WorldPhones[,input$region]*1000,
#               ylab = "Number of Telephones", xlab = "Year")
#     })
#   },
# 
#   options = list(height = 500)
# )
# exercise.setup="setdata1"
```
