---
title: "Lab Instructions"
author: "NRES 746"
date: "Fall 2021"
output: 
  html_document: 
    theme: cerulean
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This semester we're trying something a little new (and dangerous!)- the labs are built as interactive documents using R-shiny. This is intended to give you instant feedback, because with no TA it can take me a while to give you feedback on your labs! Hopefully this helps your learning experience- please let me know what you think or how you think this can be improved!

## Interactive labs

The interactive labs are available via an *R package* called "NRES746labs". To access the interactive labs, use the following steps:

### Step 1a (Windows users): install 'RTools' (first time only)

RTools is not an R package but a standalone software that enables Windows users to compile and install R packages from source code. You can install RTools from this link: [https://cran.r-project.org/bin/windows/Rtools/](https://cran.r-project.org/bin/windows/Rtools/). Binary installers are available for 64 and 32 bit machines.

### Step 1b (Mac users): install Xcode from the Mac App Store

At least I think this is what you need to do. I am not a Mac user and may be less helpful in getting you started. But I'll try to help!

### Step 2: make sure that R recognizes RTools! (first time only)

For Windows users, you need to make sure that RTools is in your PATH variable. 

First check to see if R already recognizes where RTools is. Try running the following code:

```{r}
Sys.which("make")
## "C:\\rtools40\\usr\\bin\\make.exe"
```

If the 'Sys.which' command above returns something, you should be fine. If it returns nothing (an empty text string), you need to complete the following step:

Add RTools to your PATH variable. To do this, you can create a text file .Renviron in your Documents folder which appends your RTools directory onto your PATH variable. You can do this in R by running the following line:

```{r}
writeLines('PATH="${RTOOLS40_HOME}\\usr\\bin;${PATH}"', con = "~/.Renviron")
```

Now close R and RStudio and try the 'Sys.which' command again (see above). If this works, you can move on to the next step 

### Step 3: install package 'devtools' (first time only)

To do this, use the "Install" button on the "Packages" tab in Rstudio, or use the following command:

```{r eval=FALSE}

install.packages("devtools")

```

NOTE: you only have to do this once! Once you've installed a package on your machine it will always be there! Installing devtools for the first time can be time consuming, especially if (as they recommend) you install dependencies from source.

### Step 4: install the "gradethis" package from GitHub (first time only)

One of the packages that I use in the tutorials (to give you instant feedback) is still in development and is not yet available on CRAN. So, you need to install it from GitHub. To do this, use the following code:

```{r}
remotes::install_github("rstudio/gradethis")
```

NOTE: this can take a while the first time, and lots of messages will pop up. If you're asked to choose packages to update, I don't think it matters what option you choose.

Luckily, you only need to do this once!

### Step 5: install the "NRES746labs" package from GitHub (every time you are about to start a new lab)

Ultimately I may load this package to the CRAN repository, but for now it lives on GitHub (since I'll be updating it a lot!). You can use the following code to install the package from GitHub:

```{r eval=FALSE}

remotes::install_github("kevintshoemaker/NRES746labs")

```

NOTE: you should re-install the 'NRES746labs' package for every lab (using the 'install_github' function above), since I'll be updating frequently. For example, the package currently only includes lab 1 (and old versions of labs 2-4).

### Step 6: run the lab!

Now you have everything you need to run the lab. Use the following code to start the interactive tutorial:

```{r eval=FALSE}

learnr::run_tutorial("lab1","NRES746labs")

```

NOTE: the system doesn't seem to work for Microsoft Edge (at least it didn't back in 2019), so make sure your default browser is Chrome or Safari. 

NOTE: the shiny app will "use up" your R session, so if you want to work on your own script at the same time as you're running the app, you'll have to open up a second R session. I recommend you do this! 

Hope this works! Good luck!


### DEBUGGING

Here are some tips if things don't work right away:

1. Make sure you have the most recent version of R (R v4.1.0) and the most recent version of Rstudio (1.4.1717)  

2. If R throws errors like "can not remove package [...]" then close all R sessions and go into your local package directory and delete the package manually. Then go back to R and reinstall the package. To find your local package directory, enter the following in your R session:

    ```{r eval=F}
      Sys.getenv('R_LIBS_USER')
    ```

This seems to be easier said than done on a Mac!

    
3. If you try to run the tutorial through the base R GUI and not RStudio you may have to do the following:
    * Go into Rstudio and find your 'pandoc' directory:
    ```{r eval=FALSE}
       Sys.getenv('RSTUDIO_PANDOC')
    ```
    * Go back to your base R session and enter the following:
    ```{r eval=FALSE}
       Sys.setenv(RSTUDIO_PANDOC=[...your directory here...])
    ```

4. If you get an error stating that GitHub's API rate limit has been exceeded (GitHub has a rate limit of 60 requests per hour for unauthenticated requests), the easiest solution is to create a GitHub access token (PAT). You can get a PAT by entering 'usethis::browse_github_pat()' in your R console and following the instructions. Then, add your PAT to your .Renviron file by entering 'usethis::edit_r_environ()' in your R console. Finally, in the '.Renviron' file (the one that just opened up in Rstudio), type the following:

```
GITHUB_PAT=[your PAT here]
```

5. If nothing else seems to work, try using one of the lab computers!





