---
title: "Lab Instructions"
author: "NRES 746"
date: "Fall 2019"
output: 
  html_document: 
    theme: cerulean
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This semester I am trying something a little new- the labs are now built as interactive documents using R-shiny. This is intended to give you instant feedback, because with no TA it can take me a while to give you feedback on your labs! Hopefully this helps your learning experience- please let me know what you think!

## Interactive labs

The interactive labs are available via an *R package* called "NRES746labs". To access the interactive labs, use the following steps:

### Step 1: open R/Rstudio

### Step 2: install devtools

To do this, use the "Install" button on the "Packages" tab in Rstudio, or use the following command:

```{r eval=FALSE}

install.packages("devtools")

```

NOTE: you only have to do this once! That is, once you've installed a package on your machine it will always be there!  

### Step 3: install the "NRES746labs" package from GitHub

Ultimately I may load this package to CRAN, but for now it lives on GitHub (since I'll be updating it a lot!). You can use the following code:

```{r eval=FALSE}

devtools::install_github("kevintshoemaker/NRES746labs")

```

NOTE: this can take a while, and lots of messages will pop up. If you're asked to choose packages to update, I don't think it matters what option you choose.

### Step 4: run the lab!

Now you have everything you need to run the lab. Use the following code:

```{r eval=FALSE}

library(NRES746labs)
learnr::run_tutorial("lab1","NRES746labs")

```

NOTE: the system doesn't seem to work for Microsoft Edge, so make sure your default browser is Chrome or Safari. 


Hope this works! Good luck!



