---
title: "Model Selection"
author: "NRES 746"
date: "Fall 2025"
output: 
  html_document: 
    theme: cerulean
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(cmdstanr)
register_knitr_engine(override = TRUE)
options(mc.cores = 4) 
```

```{r echo=FALSE}

#  NRES 746, Lecture 8                     
#   University of Nevada, Reno                       
#   Model selection and multi-model inference    -------------------      


```


[Click here](LECTURE8.R) for the companion R script for this lecture.

**Model selection** or **model comparison** is a very common problem in science- that is, we often have multiple competing hypotheses about how our data were generated and we want to see which model is best supported by the available evidence. 

If we can describe our data generating process explicitly as a set of deterministic and stochastic components (i.e., likelihood functions), then we can use likelihood-based methods (e.g., LRT, AIC, BIC, Bayesian model selection) to infer which data generating model(s) could most plausibly have generated our observed data. 

## Principle of Parsimony ("Occam's razor")

We will discuss several alternative approaches to model selection in ecology. However, all approaches follow the same basic principle- that -- all things equal, we should prefer the simpler model over any more complex alternative. This is known as the principle of parsimony.

## Example data: Balsam fir data from NY

Bolker uses a study of balsam fir in New York to illustrate model selection. Perhaps it's time to move on from Myxomatosis anyway...

Let's load up the data first

```{r}

# Load the balsam fir dataset

library(emdbook)
data(FirDBHFec)
fir <- na.omit(FirDBHFec[,c("TOTCONES","DBH","WAVE_NON")])
fir$TOTCONES <- round(fir$TOTCONES)
head(fir)

```

We can examine the fecundity (total cones) as a function of tree size (DBH):

```{r}

plot(fir$TOTCONES ~ fir$DBH)   # fecundity as a function of tree size (diameter at breast height)

```

One additional point of complexity in this data set- some trees were sampled from areas in which these fir trees experience a phenomenon called wave regeneration, in which tree age follows a chronosequence as you move downwind (due to larger trees at the wave front dying due to frost damage, allowing for recruitment of new trees). Other trees were sampled from "normal" fir forests in which there is no wave regeneration. 

```{r}

# tree fecundity by size, categorized into two site-level categories: "wave" and "non-wave" 

ndx <- fir$WAVE_NON=="w"   # logical vector indicating which observations were from "wave" sites
plot(fir$TOTCONES[ndx] ~ fir$DBH[ndx],xlab="DBH",ylab="Tot Cones")
points(fir$DBH[!ndx],fir$TOTCONES[!ndx],pch=4,col="red")
legend("topleft",pch=c(1,4),col=c("black","red"),legend=c("Wave","Non-wave"),bty="n")

```

Let's assume (following Bolker) that fecundity increases as a power-law relationship with DBH:

$\mu = a\cdot DBH^{b}$

Let's also assume that the fecundity follows a negative binomial distribution:

$Y = NegBin(\mu,k)$

We can model each of these parameters (*a*, *b*, and *k*) separately for trees from wave and non-wave populations.

We can also run simpler models in which some or all of these parameters are modeled as the same for both populations.

Then we can ask the question: **which model is the "best model"?**

### FULL MODEL

Here is a likelihood function for the *full model* -- that is, the most complex model (6-dimensional likelihood surface):

```{r}

# build likelihood function for the full model: CONES ~ negBINOM( a(wave)*DBH^b(wave), dispersion(wave))

   
NLL_full <- function(params){
  wave.code <- as.numeric(fir$WAVE_NON)      # convert to ones and twos
  a <- c(params[1],params[2])[wave.code]     # a parameters (two for wave and one for non-wave)
  b <- c(params[3],params[4])[wave.code]      # b parameter (two for wave and one for non-wave)
  k <- c(params[5],params[6])[wave.code]       # over-dispersion parameters (two for wave and one for non-wave)
  expcones <- a*fir$DBH^b   # expected number of cones (deterministic component)
  -sum(dnbinom(fir$TOTCONES,mu=expcones,size=k,log=TRUE))     # add stochastic component: full data likelihood
}

params <- c(a.n=1,a.w=1,b.n=1,b.w=1,k.n=1,k.w=1)

NLL_full(params)

```

We can fit the full model using "optim" (using a derivative based optimization routine), just like we have done before:

```{r warning=FALSE}

## Find the MLE -----------------------

pars = c(a.n=1,a.w=1,b.n=1,b.w=1,k.n=1,k.w=1)
opt_full <- optim(pars,NLL_full,method="BFGS")

MLE_full <- opt_full$par
MinNLL_full <- opt_full$value

```


## REDUCED MODELS

Let's run a simpler model now. This time, let's model the b parameter as equal for wave and nonwave populations:

```{r}

# build likelihood function for a reduced model: CONES ~ negBINOM( a(wave)*DBH^b, dispersion(wave))

NLL_constb <- function(params){
  wave.code <- as.numeric(fir$WAVE_NON)      # convert to ones and twos
  a <- c(params[1],params[2])[wave.code]      # a parameters
  b <- params[3]                              # b parameter (not a function of wave/nonwave)
  k <- c(params[4],params[5])[wave.code]      # dispersion parameters
  expcones <- a*fir$DBH^b
  -sum(dnbinom(fir$TOTCONES,mu=expcones,size=k,log=TRUE))
}

params <- c(a.n=1,a.w=1,b=1,k.n=1,k.w=1)

NLL_constb(params)

```

And we can fit the full model using "optim":

```{r}

### Find the MLE

opt_constb <- optim(fn=NLL_constb,par=c(a.n=1,a.w=1,b=1,k.n=1,k.w=1),method="L-BFGS-B")

MLE_constb = opt_constb$par

MinNLL_constb = opt_constb$value

```

Let's compute *-2 times the log likelihood* ($-2*log(likelihood)$) of the two models at the MLE -- That is, we compute $-2*log(Lik@MLE)$ for each alternative model.

```{r}

# compute -2*loglik for each model at the MLE

ms_full <- 2*MinNLL_full     # this is 2 * min.nll = -2*logLik_at_MLE
ms_constb <- 2*MinNLL_constb

ms_full
ms_constb

```

Note here that the log-likelihood of the full model is lower ("better": that is, the data are more likely to be produced under this model) than the the reduced model. _This should always be the case_- if not, something is wrong. That is, the plausibility of producing the observed data under a model fitted to those data should *always* increase when more free parameters are added!

This is where the principle of parsimony comes into play! 

What if we wanted to test which model was better supported by the evidence? One way is to use our old friend, the Likelihood Ratio Test (LRT)!

### Likelihood-ratio test (frequentist solution)

We have encountered the LRT once before, in the context of generating confidence intervals from likelihood surfaces (in the neighborhood of the MLE). The same principle applies for model selection.  The LRT tests whether the extra goodness-of-fit of the full model is worth the extra complexity of the additional parameters.

As you recall, the likelihood ratio is defined (obviously) as the ratio of two likelihoods. The numerator represents the likelihood of a reduced model (fewer free parameters) that is nested within a full model -- which in turn serves as the denominator:

$\frac{Likelihood_{reduced}}{Likelihood_{full}}$

The raw likelihood ratio under the null hypothesis does not have a known distribution. BUT, if we first convert the likelihood ratio to **-2X the difference in log likelihoods**, we have a statistic that approximately follows a Chi-squared distribution.

$$-2ln(\frac{Likelihood_{reduced}}{Likelihood_{full}})$$

Which can also be written as:

$$-2ln(Likelihood_{reduced}) - -2ln(Likelihood_{full})$$

Which can also be written as:

$$2[ln(Likelihood_{full} - ln(Likelihood_{reduced})]$$
that is: 2 times the absolute difference in log likelihood between the full and reduced models. [note that the order of the likelihoods is often reversed because we tend to deal with negative log likelhoods instead of log likelihoods]

When expressed this way, this likelihood ratio statistic (asymptotically) should be approximately Chi-squared distributed with df equal to the number of fixed dimensions (difference in free parameters between the full and reduced model).
 
The LRT can be used for *two-way model comparison* as long as one model is nested within the other (full model vs. reduced model). If the models are not nested then the LRT doesn't really make sense.

```{r}

# Likelihood-Ratio test -----------------------

Deviance <- ms_constb - ms_full 
Deviance

Chisq.crit <- qchisq(0.95,1)
Chisq.crit

Deviance>=Chisq.crit   # perform the LRT

1-pchisq(Deviance,1)   # p-value

```

```{r}

# Visualize the likelihood ratio test- compare the observed deviance with the distribution of deviances expected under the null hypothesis

curve(dchisq(x,df=1),0,5)
abline(v=Deviance,col="red",lwd=4)

```

Clearly, the gain in model performance is not worth the extra complexity in this case (the observed deviance could easily be produced by random chance!). Therefore, we favor the reduced model. 

What about if we try a different reduced model? This time, we decide to fix the a, b, and k parameters, so the "wave" factor is not considered.

```{r}

# Try a different reduced model: CONES ~ negBINOM( a*DBH^b, dispersion)

NLL_nowave <- function(params){
  a <- params[1]      # a parameters
  b <- params[2]      # b parameter (not a function of wave/nonwave)
  k <- params[3]      # dispersion parameters
  expcones <- a*fir$DBH^b
  -sum(dnbinom(fir$TOTCONES,mu=expcones,size=k,log=TRUE))
}

params <- c(a=1,b=1,k=1)

NLL_nowave(params)

```


And we use "optim" to locate the maximum likelihood estimate:

```{r}

# Find the MLE

opt_nowave <- optim(fn=NLL_nowave,par=params,method="L-BFGS-B")

MLE_nowave = opt_nowave$par

MinNLL_nowave = opt_nowave$value

```


Now we can perform a LRT to see which model is better!

```{r}

# Perform LRT -- this time with three fewer free parameters in the reduced model

ms_full <- 2*MinNLL_full

ms_nowave <- 2*MinNLL_nowave

Deviance <- ms_nowave - ms_full 
Deviance

Chisq.crit <- qchisq(0.95,df=3)   # now three additional params in the more complex model!
Chisq.crit

Deviance>=Chisq.crit

1-pchisq(Deviance,df=3)   # p-value

```


```{r}

# Visualize the likelihood ratio test (test statistic and sampling distribution under the null)
curve(dchisq(x,df=3),0,15)
abline(v=Deviance,col="red",lwd=4)

```


Again, the difference in deviance does not justify the additional parameters. This difference in deviance between the full and restricted model could easily be produced easily by random sampling variation. 

Remember *this is a frequentist test*. The null hypothesis is that there is no difference between the restricted model and the more complex model. So we are imagining multiple alternative universes where we are collecting thousands of datasets under the null hypothesis (simpler model is correct-- that is, the extra parameters are junk) and determining a maximum likelihood estimate for the full model (with meaningless free params) and a reduced model where we *fix* the value of one or more of our junk free parameters at the **true** parameter value of zero. Even though the data generating process is the same each time, each dataset we collect will yield a slightly different MLE for the full model and the reduced model. The deviance (-2*log likelihood ratio) between the restricted model and the full model (under the null hypothesis) should be chi-squared distributed with df = number of dimensions that were "fixed"!

As you can imagine, there are a lot of pairwise comparisons that could be generated, even in this simple example. For instance, there are 15 pairwise comparisons that could be produced from even this simple example. What about more complex models? Clearly this can get a bit unwieldy!

In addition, not all models we wish to compare will necessarily be nested. For example, consider the model selection exercise we were performing in lab- comparing the M-M fit to the Ricker fit. Since these models aren't nested, there is not clearly a "reduced" and a "full" model and we can't perform a LRT.

### Information-theoretic metrics

Information-theoretic metrics for model comparison, like Akaike's Information Criterion (AIC), provide a way to get around the issues with LRT. These metrics allow us to make tables for comparing multiple models simultaneously. However, these metrics *have no strict frequentist interpretation*. 
 
Metrics like AIC represent (theoretically) the distance between some particular model and the "perfect" or "true" model (which fits the data perfectly). Information-theoretic metrics are composed of a negative-log-likelihood component (e.g., $-2\ell$, in which $\ell$ is the log likelihood of the model and lower values mean better fit) and a *parsimony penalty term* (for which lower values mean more parsimony). For AIC, the penalty term is twice the number of parameters ($2*k$). The models with the lowest values of the information criterion are considered the best model (marrying good fit and parsimony) 

#### Akaike's Information Criterion (AIC) 

NOTE: AIC is still widely used (and misused) so it is important to understand!

AIC is computed using the following equation:  

$$AIC = -2 \ell_{max} + 2k$$

AIC is the most commonly used information criterion. 

$\ell_{max}$ is the maximum of the log-likelihood surface

k is the number of free parameters in the model

As with all information-theoretic metrics, we look for the model associated with the minimum value (lowest value corresponds with best model fit). This is the "best model".

For small sample sizes, Burnham and Anderson (2002) recommend that a finite-size correction should be used:

  $AIC_c = AIC + \frac{2k(k+1)}{n-k-1}$
 
A common *rule of thumb* is that models within 2 AIC units (or AICc units) of the best model are "reasonable" (does this "rule of 2" sound familiar?)

#### Schwarz information criterion (BIC)

Another common I-T metric is the Schwarz, or *Bayesian* information criterion. The penalty term for BIC is (log n)*k.

$BIC = -2\ell_{max} + (log(n))\cdot k$

In general, BIC is more conservative than AIC- that is, more likely to select the simpler model (since the penalty term is generally greater).

Although the Schwarz criterion has a Bayesian justification (as does AIC), it is computed from a point estimate (using the log-likelihood at the MLE) and so it doesn't pass any real test for being Bayesian -- true Bayesian analyses don't treat parameters as points, but as full probability distributions. 

#### AIC in action

Let's return to the fir fecundity model, and use AIC to select among a set of models. Let's first fit a couple more candidate models...

This time, we decide to fix the *a*, and *k* parameters, so the "wave" factor is only considered for the *b* parameter.

```{r}

# Information-theoretic metrics for model-selection ------------------------------

# Akaike's Information Criterion (AIC)

## First, let's build another likelihood function: whereby only the "b" parameter differs by "wave" sites

NLL_constak <- function(params){
  wave.code <- as.numeric(fir$WAVE_NON)      # convert to ones and twos
  a <- params[1]                             # a parameters
  b <- c(params[2],params[3])[wave.code]                              # b parameter (not a function of wave/nonwave)
  k <- params[4]                               # dispersion parameters
  expcones <- a*fir$DBH^b
  -sum(dnbinom(fir$TOTCONES,mu=expcones,size=k,log=TRUE))
}

params <- c(a=1,b.n=1,b.w=1,k=1)  

NLL_constak(params)

```


And we can fit the full model using "optim":

```{r}

### Fit the new model

opt_constak <- optim(fn=NLL_constak,par=params)

MLE_constak= opt_constak$par

MinNLL_constak = opt_constak$value

ms_constak <- 2*MinNLL_constak

```


Finally, let's fit a model with no "wave" effect, but where we assume the error is Poisson distributed...

```{r}

### Now, let's build and fit one more final model- this time with no wave effect and a Poisson error distribution

PoisLik_nowave <- function(params){
  a <- params[1]      # a parameters
  b <- params[2]      # b parameter (not a function of wave/nonwave)
  expcones <- a*fir$DBH^b
  -sum(dpois(fir$TOTCONES,lambda=expcones,log=TRUE))
}

params <- c(a=1,b=1)

PoisLik_nowave(params)

opt_pois <- optim(fn=PoisLik_nowave,par=params)

MLE_pois= opt_pois$par

MinNLL_pois= opt_pois$value

ms_pois <- 2*MinNLL_pois

```

**Note** we could not compare the Poisson model to the Negative Binomial model using LRT- one is not nested within the other!

Now we can compare the five models we have run so far using AIC

```{r}

# Compare all five models using AIC!

AIC_constak <- ms_constak + 2*4
AIC_full <- ms_full + 2*6
AIC_constb <- ms_constb + 2*5
AIC_nowave <- ms_nowave + 2*3
AIC_pois <- ms_pois + 2*2

AICtable <- data.frame(
  Model = c("Full","Constant b","Constant a and k","All constant","Poisson"),
  AIC = c(AIC_full,AIC_constb,AIC_constak,AIC_nowave,AIC_pois),
  LogLik = c(ms_full/-2,ms_constb/-2,ms_constak/-2,ms_nowave/-2,ms_pois/-2),
  params = c(6,5,4,3,2),
  stringsAsFactors = F
)

AICtable$DeltaAIC <- AICtable$AIC-AICtable$AIC[which.min(AICtable$AIC)]

AICtable$Weights <- round(exp(-0.5*AICtable$DeltaAIC) / sum(exp(-0.5*AICtable$DeltaAIC)),3)

AICtable$AICc <- AICtable$AIC + ((2*AICtable$params)*(AICtable$params+1))/(nrow(fir)-AICtable$params-1)

AICtable[order(AICtable$AIC),c(1,7,2,5,6,4,3)]
```

This AIC table shows us that the simplest model is best! Despite the fact that the deviance is lowest for the full model (principle of parsimony at work) 


## Bayesian Model Selection

Can we do model selection in a Bayesian framework? The answer is yes! 
One metric that is used by Bayesians for model selection is the *Bayes Factor* (below). The Bayes factor is defined as the ratio of *marginal likelihoods* of two models. 

In addition, there are some I-T metrics for Bayesian analyses that make model selection pretty much as easy as building an AIC table!

### Bayes Factor

Recall that our I-T metrics, as well as likelihood ratio tests, used the value of the likelihood surface at the MLE. That is, we are only taking into account a single point on the likelihood surface to represent what our data have to say about our model.

Bayesians compute a posterior distribution that takes into account the entire likelihood surface (and the prior distribution)-- that is, we now are working with an entire posterior distribution rather than just a single point. 

The *marginal likelihood* represents the likelihood averaged across parameter space (where the prior serves to weight this average to emphasize the higher probability parts of the parameter space. 

$$\overline{\mathcal{L}} = \int \mathcal{L}(x)\cdot Prior(x) dx$$

The marginal likelihood represents the *average quality of fit of a given model to the data*. 

This should look familiar! It is the denominator of the Bayes rule formula -- also known as the probability of the data, or the **model evidence**.

The higher the marginal likelihood, the more likely that model structure is to produce the data. 

The ratio of marginal likelihoods is known as the **Bayes factor** and is an elegant method for comparing models in a Bayesian context. 

$$\overline{\mathcal{L}}_1 / \overline{\mathcal{L}}_2$$

This is interpreted as *the odds in favor of model 1 versus model 2*

This simple formula elegantly penalizes over-parameterization. Simpler models will generally have higher marginal likelihoods than more complex models after all of free parameter space is taken into account. 

Can you see why the Bayes Factor is different than the likelihood ratios we were working with earlier?

We have already seen why this might be. More complex models will always have a higher likelihood at and around the MLE, but generally will have much lower likelihoods across other parts of parameter space. 

In practice, computing marginal likelihoods can be tricky, involving multi-dimensional integration! Recall that we use MCMC partially to avoid computing the denominator in the Bayes Rule formula. Bayes factors are computed entirely from the denominators of Bayesian parameter estimation problems! 

Furthermore, the prior specification can have a huge impact on Bayes factors, which can be unsettling at best!

#### Bayes Factor Example

A simple binomial distribution example can illustrate Bayes factors quite nicely.

Imagine we conduct a tadpole experiment where we are interested in estimating the mortality rate *p* due to a pesticide, on the basis of the number of dead tadpoles observed out of 10 in a treated tank. We are interested in comparing a simple model where *p* is fixed at 0.5 (a 'point null' model) with a more complex model where *p* is a *free parameter* estimated from data. That is, we are testing the hypothesis that *p* is different from 50%. 

First let's look at the simple model. 

```{r}

# Bayes factor example  ---------------------

# take a basic binomial distribution with parameter p fixed at 0.5:

probs1 <- dbinom(0:10,10,0.5)          
names(probs1) = 0:10
barplot(probs1,ylab="probability")

## Q: What is the *marginal likelihood* under this simple model for an observation of 2 mortalities out of 10? 

## A:

dbinom(2,10,0.5)


```

**Q** What is the *marginal likelihood* under this simple model for an observation of 2 mortalities out of 10? How about 3 mortalities?

**A** It is exactly `r dbinom(2,10,0.5)`. That is, there is no marginalizing to do since there is no free parameter to marginalize over. 

Now let's consider a more 'complex' model, where *p* is a free parameter (one additional free parameter relative to the simple model). First, let's assume that the parameter *p* is assigned a uniform $beta(1,1)$ prior across parameter space:

```{r}

## Now we can consider a model whereby "p" is a free parameter
curve(dbeta(x,1,1))  # uniform prior on "p"

```

What is the marginal likelihood of observing 2 mortalities under this model? 

In other words, what is the average probability of observing 2 mortalities, averaged across all possible values of *p*??? 

Intuitively, because all values of *p* are equally likely, all possible observations (0 mortalities, 2 mortalities, N mortalities) should all be equally likely! That is, neither the likelihood function nor the prior distribution favors any particular observation (0 to 10) over any other.   

We can do this mathematically...

For two observed mortalities, the marginal likelihood is: 

```{r}

# Compute the marginal likelihood of observing 2 mortalities

# ?integrate
binom2 <- function(x) dbinom(x=2,size=10,prob=x)
marginal_likelihood <- integrate(f=binom2,0,1)$value    # use "integrate" function in R
marginal_likelihood  # equal to 0.0909 = 1/11

```

For three observed mortalities, the marginal likelihood (probability of observing a "3" across all possible values of p) is:

```{r}

# Compute the marginal likelihood of observing 3 mortalities

binom3 <- function(x) dbinom(x=3,size=10,prob=x)
marginal_likelihood <- integrate(f=binom3,0,1)$value    # use "integrate" function
marginal_likelihood   # equal to 0.0909 = 1/11

```

Basically, if p could be anything between 0 and 1, no particular observation is favored over any other prior to observing any real data:

```{r}

# simulate data from the model across all possible values of the parameter "p"

lots=1000000
a_priori_data <- rbinom(lots,10,prob=rbeta(lots,1,1))   # no particular observation is favored
for_hist <- table(a_priori_data)/lots
barplot(for_hist,xlab="Potential Observation",ylab="Marginal likelihood")

```

Therefore, since there are 11 possible observations (outcomes of the binomial distribution of size=10), the marginal likelihood of any particular data observation under the model with p as a free parameter should be 1/11 = `r 1/11`.

Here is a visualization of the marginal likelihood across all possible data observations:

```{r}

# Visualize the marginal likelihood of all possible observations

probs2 <- rep(1/11,times=11)          
names(probs2) = 0:10
barplot(probs2,ylab="probability",ylim=c(0,1))

```

Now let's overlay the marginal likelihoods for the simpler model:

```{r}

# Overlay the marginal likelihood of the simpler model, with p fixed at 0.5

probs2 <- rep(1/11,times=11)          
names(probs2) = 0:10
barplot(probs2,ylab="probability",ylim=c(0,1))

probs1 <- dbinom(0:10,10,0.5)          
names(probs1) = 0:10
barplot(probs1,ylab="probability",add=T,col="red",density=20)

```

Assuming we observed 2 mortalities, what is the Bayes Factor? Which model is better?

```{r}

# Finally, compute the bayes factor given that we observed 2 mortalities. Which model is better?

probs2 <- rep(1/11,times=11)          
names(probs2) = 0:10
barplot(probs2,ylab="probability",ylim=c(0,1))

probs1 <- dbinom(0:10,10,0.5)          
names(probs1) = 0:10
barplot(probs1,ylab="probability",add=T,col="red",density=20)

abline(v=3,col="green",lwd=4 )

```


```{r}

BayesFactor = (1/11)/dbinom(2,10,0.5)   
BayesFactor

```

Here, the Bayes factor of around 2 indicates that the data lend support to the more complex model!

What if the data instead were 3 mortalities? Which model is the best model?

```{r}

# Compute the bayes factor given that we observed 3 mortalities. Which model is better now?

probs2 <- rep(1/11,times=11)          
names(probs2) = 0:10
barplot(probs2,ylab="probability",ylim=c(0,1))

probs1 <- dbinom(0:10,10,0.5)          
names(probs1) = 0:10
barplot(probs1,ylab="probability",add=T,col="red",density=20)

abline(v=4.3,col="green",lwd=4 )

```


```{r}

BayesFactor = dbinom(3,10,0.5)/(1/11)
BayesFactor

```

This time, *The simpler model wins*!!! That would never happen if we compared maximum likelihood estimates- the likelihood of the (fitted) more highly-parameterized model will always be (equal or) lower (i.e., the likelihood of three successes at p=0.3 is higher than the likelihood of three successes at p=0.5)! That's why we have to penalize or *regularize* more highly-parameterized models.

We can visualize this! First of all, what is the maximum likelihood estimate for p under the model with 3 mortality observations?

```{r}

# Visualize the likelihood ratio

# probs2 <- rep(1/11,times=11)          
# names(probs2) = 0:10
# barplot(probs2,ylab="probability",ylim=c(0,1))

probs1 <- dbinom(0:10,10,0.5)          
names(probs1) = 0:10
barplot(probs1,ylab="probability",col="red",density=20,ylim=c(0,1))

probs3 <- dbinom(0:10,10,0.3)          
names(probs3) = 0:10
barplot(probs3,ylab="probability",add=T,col="green",density=10,angle = -25)

abline(v=4.3,col="green",lwd=4 )

```

So clearly the likelihood ratio favors the more complex model (fitted parameter "p") vs the simple, 0-parameter model!

What does the likelihood-ratio test say?

```{r}

# LRT: simple model (p fixed at 0.5) vs complex model (p is free parameter)

Likelihood_simple <- dbinom(3,10,0.5)
Likelihood_complex <- dbinom(3,10,0.3)
Likelihood_simple
Likelihood_complex
-2*log(Likelihood_simple)--2*log(Likelihood_complex)

qchisq(0.95,1)

pchisq(1.64,1)    # very high p value, simpler model is preferred

```

What about AIC?

```{r}

# AIC: simple model (p fixed at 0.5) vs complex model (p is free parameter)

AIC_simple <- -2*log(Likelihood_simple) + 2*0
AIC_complex <-  -2*log(Likelihood_complex) + 2*1

AIC_simple
AIC_complex    

```

What about AICc?

$AIC_c = AIC + \frac{2k(k+1)}{n-k-1}$


```{r}

### Alternatively, use AICc

AICc_simple <- -2*log(Likelihood_simple) + 0 + 0
AICc_complex <-  -2*log(Likelihood_complex) + 1 + ((2*2)/(3-1-1))

AICc_simple
AICc_complex    

```


What about BIC?

```{r}

# Alternatively, try BIC

BIC_simple <- -2*log(Likelihood_simple) + log(10)*0
BIC_complex <-  -2*log(Likelihood_complex) + log(10)*1

BIC_simple
BIC_complex    


```

All these methods give the same basic answer- that the simple model is better, even though the complex model fits better!


### Bayesian Information Theoretic Criteria

Deviance Information Criterion (DIC) is computed by default in JAGS and WinBUGS. It is analogous to other I-T metrics like AIC (and therefore easy to interpret and use)!

DIC has been largely superseded by other I-T criterion for MCMC-fitted Bayesian models: these are WAIC and LOOIC. We will compute these using stan.

The WAIC (also known as the Wattanabe-Akaike information criterion, or the 'widely applicable information criterion') metric is also interpretable just like AIC, BIC and DIC and allows us to compare models fitted in a Bayesian framework via MCMC. 

The WAIC metric, while not computed by default in JAGS, is applicable more widely than DIC.

Finally, the WAIC metric has largely been superseded by the 'loo' methods, including LOO-CV and LOOIC. LOOIC is often indistinguishable from WAIC. One advantage of these methods is that they provide additional model fit diagnostics based on a technique known as Pareto-smoothed importance sampling (PSIS). 

First, let's write a 'stan' model for the fir data

```{stan output.var="firmodel_full"}

data {
  int<lower = 1> N;
  array [N] int<lower=0> obs_cones;
  vector<lower=0> [N] DBH;
  array [N] int<lower=1,upper=2> wave_ndx; 
}

parameters {
  vector[2] loga, b, logbeta;
}

transformed parameters {
  vector[2] a = exp(loga);
  vector[2] beta = exp(logbeta);
}

model  {
  vector[N] mean_cones = exp(loga[wave_ndx] + b[wave_ndx] .* DBH);   // power function: a*DBH^b
  vector[N] alpha = mean_cones .* beta[wave_ndx];
  obs_cones ~ neg_binomial(alpha,beta[wave_ndx]);
}

generated quantities {   // need log_lik of each data point for model selection
  vector[N] log_lik; // N is the number of data points
  {
    real m2, a2, b2;
    for (n in 1:N) {
       m2 = exp(loga[wave_ndx[n]] + b[wave_ndx[n]] * DBH[n]);   // power function: a*DBH^b
       a2 = m2 * beta[wave_ndx[n]];
       log_lik[n] = neg_binomial_lpmf(obs_cones[n] | a2, beta[wave_ndx[n]]);
    }
  }
}



```

Then we need to package the data for stan

```{r}

# Package the data for stan

stan_data <- list(
  N = nrow(fir),
  obs_cones = fir$TOTCONES,
  wave_ndx = as.numeric(fir$WAVE_NON),
  DBH = fir$DBH
)
#data.package

```


Then we can run the model!

```{r warning=FALSE}

# Run the model in stan

library(cmdstanr)    # load packages
library(bayesplot)
library(posterior)
options(mc.cores = 4) 

# compile the model using:
firmodel_full <- cmdstan_model("firmodel_full.stan") # Compile stan model

fit_full <- suppressMessages( firmodel_full$sample(
  data = stan_data,
  chains = 4,
  iter_warmup = 200,
  iter_sampling = 500,
  refresh = 0
)  )

```


```{r}
# fit_full$summary()
samples <- fit_full$draws(format="draws_df")
bayesplot::mcmc_trace(samples,"a[1]")
bayesplot::mcmc_trace(samples,"b[2]")
bayesplot::mcmc_trace(samples,"beta[1]")
```


First of all, is there any evidence that the mean reproduction from wave sites is different than that of the non-wave sites?


```{r}

meanrep_wave = exp(samples$`loga[2]` + samples$`b[2]`*mean(stan_data$DBH))
meanrep_nonwave = exp(samples$`loga[1]` + samples$`b[1]`*mean(stan_data$DBH))

hist(meanrep_nonwave,ylab="Prob Density",xlab="number of cones",freq = F,xlim=c(25,60),ylim=c(0,0.15),main="")
hist(meanrep_wave,density=20,col="darkgreen",add=T,freq=F)
legend("topleft",col=c("darkgreen","white"),density=c(20,0),legend=c("wave","nonwave"),bty="n")

```

We can use the 'loo' package to compute WAIC for this model

```{r warning=FALSE}

# Extract the WAIC for the full model!

library(loo)

loglik_names <- sapply (1:stan_data$N,
                          function(t) sprintf("log_lik[%s]",t) )
lls = as.matrix(samples)[,loglik_names]

waic_full = loo::waic(lls)
waic_full$estimates["waic",]

```
Notice that unlike AIC, we not only get a point estimate for WAIC but we also get a standard error, indicating how sure we are!

An even easier (and probably better) method is to use the LOOIC metric with pareto smoothed importance sampling (PSIS). This is because this method provides additional useful diagnostics. Generally WAIC and LOO-CV/LOOIC will give similar results. 

```{r}

loo_full <- fit_full$loo()
loo_full

```

Notice that the LOOIC and WAIC values are nearly identical. 


Now let's build the reduced model and compare LOOIC values!


```{stan output.var="firmodel_reduced"}

data {
  int<lower = 1> N;
  array [N] int<lower=0> obs_cones;
  vector<lower=0> [N] DBH;
}

parameters {
  real loga, b, logbeta;
}

transformed parameters {
  real a = exp(loga);
  real beta = exp(logbeta);
}

model  {
  vector[N] mean_cones = exp(loga + b .* DBH);   // power function: a*DBH^b
  vector[N] alpha = mean_cones .* beta;
  obs_cones ~ neg_binomial(alpha,beta);
}

generated quantities {   // need log_lik of each data point for model selection
  vector[N] log_lik; // N is the number of data points
  {
    real m2, a2, b2;
    for (n in 1:N) {
       m2 = exp(loga + b * DBH[n]);   // power function: a*DBH^b
       a2 = m2 * beta;
       log_lik[n] = neg_binomial_lpmf(obs_cones[n] | a2, beta);
    }
  }
}

```


Then we can run the model!

```{r warning=FALSE}
# compile the model using:
firmodel_reduced <- cmdstan_model("firmodel_reduced.stan") # Compile stan model

fit_reduced <- suppressMessages( firmodel_reduced$sample(
  data = stan_data,
  chains = 4,
  iter_warmup = 200,
  iter_sampling = 500,
  refresh = 0
)  )

```


```{r}
# fit_full$summary()
samples_red <- fit_reduced$draws(format="draws_df")
bayesplot::mcmc_trace(samples_red,"a")
# bayesplot::mcmc_trace(samples_red,"b")
# bayesplot::mcmc_trace(samples_red,"beta")
```

What is the LOOIC for this model?

```{r}

# Compute LOOIC

loo_reduced <- fit_reduced$loo()

loo_reduced

```

Is there a good reason to prefer the full model now? 

### Explicit Bayesian model selection

Another cool, sometimes useful, method of Bayesian model selection is to write the model selection directly into the MCMC algorithm!  This technique, known as reversible jump MCMC (RJMCMC), can be a particularly elegant way to do model selection. However, this method is not compatible with HMC, so it is currently not possible to do RJMCMC in stan. 

Here is an example of how to do this in the BUGS language (e.g., that could be run in JAGS or NIMBLE). 

```

model  {

  ### Likelihood for model 1: full

  for(i in 1:n.obs){
    expected.cones[i,1] <- a1[wave[i]]*pow(DBH[i],b1[wave[i]])       # a*DBH^b
    spread.cones[i,1] <- r1[wave[i]]
    p[i,1] <- spread.cones[i,1] / (spread.cones[i,1] + expected.cones[i,1])
    observed.cones[i,1] ~ dnegbin(p[i,1],spread.cones[i,1])
    predicted.cones[i,1] ~ dnegbin(p[i,1],spread.cones[i,1])
    SE_obs[i,1] <- pow(observed.cones[i,1]-expected.cones[i,1],2)
    SE_pred[i,1] <- pow(predicted.cones[i,1]-expected.cones[i,1],2)
  }
  
  
  ### Priors, model 1
  for(j in 1:2){   # estimate separately for wave and non-wave
    a1[j] ~ dunif(0.001,2)
    b1[j] ~ dunif(0.5,4)
    r1[j] ~ dunif(0.5,5)
  }

  ### Likelihood for model 2: reduced

  for(i in 1:n.obs){
    expected.cones[i,2] <- a2*pow(DBH[i],b2)       # a*DBH^b
    spread.cones[i,2] <- r2
    p[i,2] <- spread.cones[i,2] / (spread.cones[i,2] + expected.cones[i,2])
    observed.cones[i,2] ~ dnegbin(p[i,2],spread.cones[i,2])
    predicted.cones[i,2] ~ dnegbin(p[i,2],spread.cones[i,2])
    SE_obs[i,2] <- pow(observed.cones[i,2]-expected.cones[i,2],2)
    SE_pred[i,2] <- pow(predicted.cones[i,2]-expected.cones[i,2],2)
  }
  
  
  ### Priors, model 2
  a2 ~ dunif(0.001,2)
  b2 ~ dunif(0.5,4)
  r2 ~ dunif(0.5,5)

  ### Likelihood for model 3: constant a and b

  for(i in 1:n.obs){
    expected.cones[i,3] <- a3*pow(DBH[i],b3)       # a*DBH^b
    spread.cones[i,3] <- r3[wave[i]]
    p[i,3] <- spread.cones[i,3] / (spread.cones[i,3] + expected.cones[i,3])
    observed.cones[i,3] ~ dnegbin(p[i,3],spread.cones[i,3])
    predicted.cones[i,3] ~ dnegbin(p[i,3],spread.cones[i,3])
    SE_obs[i,3] <- pow(observed.cones[i,3]-expected.cones[i,3],2)
    SE_pred[i,3] <- pow(predicted.cones[i,3]-expected.cones[i,3],2)
  }
  
  SSE_obs[1] <- sum(SE_obs[,1]) 
  SSE_pred[1] <- sum(SE_pred[,1])
  SSE_obs[2] <- sum(SE_obs[,2]) 
  SSE_pred[2] <- sum(SE_pred[,2])
  SSE_obs[3] <- sum(SE_obs[,3]) 
  SSE_pred[3] <- sum(SE_pred[,3])

  ### Priors, model 3
  for(j in 1:2){   # estimate separately for wave and non-wave
    r3[j] ~ dunif(0.5,5)
  }
  a3 ~ dunif(0.001,2)
  b3 ~ dunif(0.5,4)

  #####################
  ### SELECT THE BEST MODEL!!! 
  #####################

  for(i in 1:n.obs){
    observed.cones2[i] ~ dnegbin(p[i,selected],spread.cones[i,selected])
    predicted.cones2[i] ~ dnegbin(p[i,selected],spread.cones[i,selected])     # for posterior predictive check!
    SE2_obs[i] <- pow(observed.cones2[i]-expected.cones[i,selected],2)
    SE2_pred[i] <- pow(predicted.cones2[i]-expected.cones[i,selected],2)
  }
  
  SSE2_obs <- sum(SE2_obs[])
  SSE2_pred <- sum(SE2_pred[])


  ### Priors
  
    # model selection...
  prior[1] <- 1/3
  prior[2] <- 1/3     # you can put substantially more weight because fewer parameters (there are more rigorous ways to do this!!)
  prior[3] <- 1/3
  selected ~ dcat(prior[])   
  
  
}
    
```


#### Evaluate model fit

Now we can look at model fit! We will use the same method we used in lab- the posterior predictive check... 

We can write a for loop to extract the prediction results for each model (and for the model-averaged model):

First let's just look at the model predictions vs the observed data. First, model 1

```{r}

# Goodness of fit

N <- stan_data$N

plot(fir$TOTCONES~fir$DBH,ylim=c(0,900),cex=2)

# which.max(stan_data$DBH)
# which.min(stan_data$DBH)

a_param <- samples_red$a
b_param <- samples_red$b
beta_param <- samples_red$beta
mu_rep <- sapply(1:N,function(t) a_param * stan_data$DBH[t]^b_param)

sim_dat <- function(s){
  thismean = mu_rep[s,]
  thisbeta <- samples_red$beta[s]
  thisalpha <- thismean * thisbeta 
  thismu = thisalpha / thisbeta
  rnbinom(N,size=thisalpha,mu=thismu)
}

nMCMC = length(samples_red$b)

for(d in 1:N){
  dat= sim_dat(sample(1:nMCMC,1))
  points(stan_data$DBH,dat,pch=20,col="gray",cex=0.4)
}
points(fir$DBH,fir$TOTCONES,cex=2)

```

Let's run a Bayesian p-value check for the reduced model:

```{r}

# Posterior Predictive Checks!

nreps = 500 
SSE_obs= numeric(nreps)
SSE_sim = numeric(nreps)
r=1
for(r in 1:nreps){
  this= sample(1:nMCMC,1,replace = T)
  SSE_obs[r] = sum((stan_data$obs_cones - mu_rep[this,])^2)
  simdat = sim_dat(this)
  SSE_sim[r] = sum((simdat - mu_rep[this,])^2)
}

plot(SSE_sim~SSE_obs,xlab="SSE, real data",ylab="SSE, simulated data",main="Posterior Predictive Check")
abline(0,1,col="red")
p.value=mean(SSE_sim>SSE_sim)
p.value 

```


Interesting- the model seems to not fit very well! You can't even see the 1-to-1 line here. It's possible I made a coding error somewhere, so if I were doing this analysis for real, I would triple check everything. If this result holds I might try to find a better model specification!

Okay, one more topic before we end:

### Model averaging

The fact that model selection is such a big deal in ecology and environmental science indicates that we are rarely certain about which model is the best model. Even after constructing an AIC table we may be very unsure about which model is the "true" model. 

The AIC weights tell us in essence how much we "believe" in each model. This is a very Bayesian interpretation, but model averaging really is best thought of in a Bayesian context. 

One way to do model averaging relies on AIC weights. Basically we take the set of predictions from each model independently and weight them by the Akaike weight. There is a literature on this and R packages for helping (see package ['AICcmodavg'](https://cran.r-project.org/web/packages/AICcmodavg/AICcmodavg.pdf))

#### When should you use model-averaged parameter estimates

Really never in my opinion- model averaging can be useful for prediction but not useful (even misleading) if you are trying to estimate effect sizes!



[--go to next lecture--](LECTURE9.html) 



































