---
title: "Why focus on algorithms?"
author: "NRES 746"
date: "August 23, 2016"
output: 
  html_document: 
    theme: cerulean
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r include=FALSE}
library(dplyr)
library(tidyr)
```

## Algorithmic vs standard statistics: a brief demonstration

### Standard t-test

First, let's go through a very basic review.   

Here is a made-up data set. Let's imagine we are interested in seeing if the average mass of study organisms in Group A are different from those in Group B.  

```{r}

df <- data.frame(
  GroupA = c(175, 168, 168, 190, 156, 181, 182, 175, 174, 179),
  GroupB = c(185, 169, 173, 173, 188, 186, 175, 174, 179, 180) 
)

summary(df)

sample.size <- length(df$GroupA)
reshape_df <- data.frame(
  Group = rep(c("A","B"),each=sample.size),
  Mass = c(df$GroupA,df$GroupB)
)
plot(Mass~Group, data=reshape_df)

# boxplot(df$GroupA,df$GroupB,names=c("GroupA","GroupB"))  # (alternative method!)

observed_dif <- mean(df$GroupA) - mean(df$GroupB)
observed_dif

```

You probably recognize the standard t-test below. For now, we are assuming that the samples come from normally distributed populations with equal variances. 

```{r}
t.test(df$GroupA,df$GroupB, var.equal=TRUE, paired=FALSE)

```


### Brute-force t-test
But let's imagine now that we didn't know about the t-test- but we still want to accomplish the same goal.  

First, let's review the question and state the problem clearly:

We want to know if these two populations are different- specifically if the expected value (mean) differs. We have small random (but representative) samples from these two putative populations.  

Can we build an **algorithm** to generate a p-value? 

Let's start by simulating a data population under the null hypothesis. 

```{r}
lots <- 1000000  # large number approximating infinity in this example!

popMean_null <- mean(reshape_df$Mass)        # assume groups A and B come from a population with common mean 
popSD_null <- sd(reshape_df$Mass)
popData_null <- rnorm(n=lots,mean=popMean_null,sd=popSD_null)    # the statistical "population" of interest (under null model)

```

Then we can sample from that data population:

```{r}
sampleA <- sample(popData_null,size=sample.size)   
sampleB <- sample(popData_null,size=sample.size)

round(sampleA)
difference <- mean(sampleA)-mean(sampleB)   # difference between sample means
difference

```

Then we can generate a distribution of "anomalies" (differences between the group means) expected under the null hypothesis (there is fundamentally no difference between the two groups):

```{r}
reps <- 1000
null_difs <- numeric(reps)

for(i in 1:reps){
  sampleA <- sample(popData_null,size=sample.size)   
  sampleB <- sample(popData_null,size=sample.size)
  null_difs[i] <- mean(sampleA)-mean(sampleB)
}

hist(null_difs)
abline(v=observed_dif,col="green",lwd=3)

```


Now, all we need to do is compare this more quantitatively with the observed among-group difference...

```{r}
  ordered_difs <- sort(abs(null_difs))   
  higher_anomaly <- length(which(ordered_difs>=abs(observed_dif)))
  p_value <- higher_anomaly/reps  
  p_value

```

#### Exercise (intermediate)
Repeat the same exercise, but using a different underlying distribution- say, the binomial or the Poisson. Compare with a standard t-test. Which test do you trust more?

#### Exercise (intermediate)
Modify the above algorithm to allow for unequal sample sizes. Compare with a standard t-test. Can you identify any cases where the t-test results differ substantially from your algorithmic results? 

#### Exercise (advanced)
Now, what about if we wanted to relax the assumption of equal variances? How would you modify the above algorithm to do this?  

#### Take-home message
The value of the algorithmic, brute-force approach to statistics is the flexibility! We have to be aware of assumptions in all of our analyses, but when we have some computational tools at hand, we can readily relax these assumptions. 


### A non-parametric alternative: permutation test algorithm

What if we don't want to make any assumptions about the process that generated the data. The normal distribution can arise in many different ways, but many data-generating processes **don't** result in a normal distribution!

We might be able to intuit which of the many alternative distributions makes the most sense for our data. But many times we can't do this with any level of certainty. What can we do in this case? A permutation test provides one answer.

Let's build this algorithm together. 

Here is some pseudocode:

1. Define the number of permutations to run
2. Start loop through permutations
    1. randomize the data indices with respect to group
    2. compute the difference between the group means after randomizing the data indices
    3. store this value for later analysis
3. Plot a histogram of differences between group A and group B under the null hypothesis
4. Add a vertical line to indicate the observed difference


```{r echo=FALSE}
reps <- 5000
null_difs <- numeric(reps)
for (i in 1:reps){					
	newindices <- sample(c(1:nrow(reshape_df)))   			
  newGroup <- reshape_df$Group[newindices]			
	dif <- mean(reshape_df$Mass[newGroup=="A"])	- mean(reshape_df$Mass[newGroup=="B"])	
	null_difs[i] <- dif	
}
hist(null_difs)
abline(v=observed_dif,col="green",lwd=3)


```

Now we can compute a p-value, just as we did before:

```{r}
higher_anomaly <- length(which(abs(null_difs)>=abs(observed_dif)))
p_value <- higher_anomaly/reps  
p_value

```


## Bootstrapping a confidence interval

(NOTE: some materials borrowed from [link](http://people.tamu.edu/~alawing/materials/ESSM689/Btutorial.pdf))

Let's imagine we want to compare different predictor variables in terms of how strong the relationship is with a response variable. In this case, we will use the coefficient of determination ($R^2$) as a measure of how good a predictor is. However, we want to be able to say that one predictor is definatively *better* than another one -- for that, we would like a confidence interval around the $R^2$ value.

Let's use the "trees" dataset provided in base R:

```{r}
head(trees)   # use help(trees) for more information

```

Let's say that tree volume is our response variable. We want to see if girth or height are better predictors of volume. 

Let's first do some basic data explorations:

```{r}
plot(trees$Volume~trees$Height, main = 'Black Cherry Tree Height/Volume Relationship', xlab = 'Height', ylab = 'Volume', pch = 16, col ='blue')
plot(trees$Volume~trees$Girth, main = 'Black Cherry Tree Girth/Volume Relationship', xlab = 'Girth', ylab = 'Volume', pch = 16, col ='red')
```

Let's write a simple function that generates coefficients of determination given a response and some predictor variables:

```{r}

Rsquared <- function(df,responsevar="Volume"){    # interactions not yet implemented
  response <- df[,responsevar]
  names <- names(df)
  rsq <- numeric(length(names))
  names(rsq) <- names(df)
  rsq <- rsq[names(rsq)!=responsevar]
  for(i in names(rsq)){         # loop through predictors
      predictor <- df[,i]
      model <- lm(response~predictor)  
      rsq[i] <- summary(model)$r.square 
  }
  return(rsq)
}

```


Let's first compute the Rsquared values for all predictor variables:

```{r}
stat <- Rsquared(trees,"Volume")
stat

```


Now we can use "bootstrapping" to generate a confidence interval around these values...

Let's first write a function to generate bootstrap samples from a dataset:

```{r}

boot_sample <- function(df,statfunc,n_samples,n_stats){
  indices <- c(1:nrow(df))
  output <- matrix(NA,nrow=n_samples,ncol=n_stats)
  for(i in 1:n_samples){
    boot_rows <- sample(indices,size=nrow(df),replace=T)
    newdf <- df[boot_rows,]
    output[i,] <- statfunc(newdf)
  }
  return(output)
}

```

Now we can generate a bunch of "bootstrapped" statistics to compare with the ones calculated from the full dataset:

```{r}
boot <- boot_sample(trees,Rsquared,10,length(stat))
colnames(boot) <- names(stat)

boot
stat

```

Finally, we can use the quantiles of the bootstrap samples to generate bootstrap confidence intervals.

```{r}
boot <- boot_sample(trees,Rsquared,1000,length(stat))   # 1000 bootstrap samples
confint <- apply(boot,2,function(t)  quantile(t,c(0.025,0.5,0.975)))
colnames(confint) <- names(stat)
t(confint)
quantile(boot,c(0.025,0.975))

```


#### Exercise (intermediate):
Generate bootstrap confidence intervals around the regression parameters. Compare with the standard confidence intervals given by R in the "lm" package.


#### Exercise (advanced):
Modify the algorithm above to select the top model from among all possible models (including interaction terms)
















