---
title: "First Lecture"
author: "NRES 746"
date: "August 23, 2016"
output: 
  html_document: 
    theme: cerulean
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r include=FALSE}
library(dplyr)
library(tidyr)
```


## Algorithmic vs standard statistics: a brief demonstration

### standard t test

First, let's go through a very basic review.   

Here is a made-up data set. Let's imagine we are interested in seeing if the average height of study organisms in Group A are different from those in Group B.  

```{r}

data <- data.frame(
  GroupA = c(175, 168, 168, 190, 156, 181, 182, 175, 174, 179),
  GroupB = c(185, 169, 173, 173, 188, 186, 175, 174, 179, 180) 
)

summary(data)

sample.size <- length(data$GroupA)

reshape_data <- data.frame(
  Group = rep(c("A","B"),each=sample.size),
  Height = c(data$GroupA,data$GroupB)
)

plot(Height~Group, data=reshape_data)

observed_dif <- mean(data$GroupA) - mean(data$GroupB)
observed_dif

```

You probably recognize the standard t-test below. For now, we are assuming that the samples come from normally distributed populations with equal variances. 

```{r}
t.test(data$GroupA,data$GroupB, var.equal=TRUE, paired=FALSE)

```

But let's imagine now that we didn't know about the t-test- but we still want to accomplish the same goal.  

First, let's review the question and state the problem clearly:

We want to know if these two populations are different- specifically if the expected value (mean) differs. We have small random (but representative) samples from these two putative populations.  

Can we build an **algorithm** to generate a p-value? 

Let's start by simulating a data population under the null hypothesis. 

```{r}
lots <- 1000000  # large number approximating infinity in this example!

popMean_null <- mean(reshape_data$Height)        # assume groups A and B come from a population with common mean 
popSD_null <- sd(reshape_data$Height)
popData_null <- rnorm(n=lots,mean=popMean_null,sd=popSD_null)    # the statistical "population" of interest (under null model)

```

Then we can sample from that data population:

```{r}
sampleA <- sample(popData_null,size=sample.size)   
sampleB <- sample(popData_null,size=sample.size)

sampleA
difference <- mean(sampleA)-mean(sampleB)   # difference between sample means

```

Then we can generate a distribution of "anomalies" (differences between the group means) expected under the null hypothesis (there is fundamentally no difference between the two groups):

```{r}
reps <- 1000
null_difs <- numeric(reps)

for(i in 1:reps){
  sampleA <- sample(popData_null,size=sample.size)   
  sampleB <- sample(popData_null,size=sample.size)
  null_difs[i] <- mean(sampleA)-mean(sampleB)
}

hist(null_difs)
abline(v=observed_dif,col="green",lwd=3)

```


Now, all we need to do is compare this more quantitatively with the observed among-group difference...

```{r}
  ordered_difs <- sort(abs(null_difs))   # first order the distribution of null results in terms of absolute magnitude of difference
  higher_anomaly <- length(which(ordered_difs>=abs(observed_dif)))
  p_value <- higher_anomaly/reps   # probability of getting a value with anomaly as or more extreme than observed, given null hypothesis is true...
  p_value

```


Now, what about if we wanted to relax the assumptions -- allow for unequal sample sizes, or unequal variances, for example... 

or non-normal distributions?

The value of the algorithmic, brute-force approach to statistics is the flexibility! We have to be aware of assumptions in all of our analyses, but when we have some computational tools at hand, we can readily relax these assumptions. 


## Bootstrapping a confidence interval

(NOTE: some materials borrowed from [link](http://people.tamu.edu/~alawing/materials/ESSM689/Btutorial.pdf))

Let's imagine we want to compare different predictor variables in terms of how strong the relationship is with a response variable. In this case, we will use the coefficient of determination ($R^2$) as a measure of how good a predictor is. However, we want to be able to say that one predictor is definatively *better* than another one -- for that, we would like a confidence interval around the $R^2$ value.

Let's use the "trees" dataset provided in base R:

```{r}

head(trees)   # use help(trees) for more information

```

Let's say that tree volume is our response variable. We want to see if girth or height are better predictors of volume. 

Let's first do some basic data explorations:

```{r}
plot(trees$Volume~trees$Height, main = 'Black Cherry Tree Volume Relationship', xlab = 'Height', ylab = 'Volume', pch = 16, col ='blue')
plot(trees$Volume~trees$Girth, main = 'Black Cherry Tree Volume Relationship', xlab = 'Girth', ylab = 'Volume', pch = 16, col ='red')
```


Let's write a simple function that generates a coefficient of determination given a response and predictor variable:

```{r}

Rsquared <- function(response, predictor){
  model <- lm(response~predictor)
  rsq <- summary(model)$r.square
  return(rsq)
}

```


Let's first compute the Rsquared values for the two predictor variables:

```{r}
Rsquared(trees$Volume,trees$Girth)
Rsquared(trees$Volume,trees$Height)

```


Now we can use "bootstrapping" to generate a confidence interval around these values...

Let's first develop a function to develop a single bootstrap sample from a dataset:

```{r}

boot_sample <- function()

```


















