---
title: "Bayesian Analysis #2: MCMC"
author: "NRES 746"
date: "Fall 2025"
output: 
  html_document: 
    theme: cerulean
    toc: no
    toc_float: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r echo=FALSE}


#  NRES 746, Lecture 7   
#   Bayesian analysis #2: MCMC  ---------------------------------                   

```

For those wishing to follow along with the R-based demo in class, [click here](LECTURE7.R) for the companion R script for this lecture.

### Markov Chain Monte Carlo (MCMC)

MCMC is designed to sample from difficult-to-describe (e.g., multivariate, hierarchical) probability distributions. 

In many/most cases, the posterior distribution for ecological problems is a difficult-to-describe probability distribution. Therefore, MCMC is the most common method we use for doing Bayesian inference.

MCMC is kind of magical in that it allows you to sample from probability distributions that are impossible to fully define mathematically! The MCMC approach uses random jumps in parameter space that eventually (after a warm-up period) end up sampling from the joint posterior distribution! 

The key to MCMC is the following:

> The probability of a successful jump in parameter space from point A to point B is proportional to the ratio of the target probability densities at these two points. 

The probability of a successful jump in parameter space can be characterized as:

$P(jump) \cdot P(accept)$

That is: in order to jump, you need to *propose* a specific jump (according to some probability distribution) and then accept that jump in parameter space!

The ratio of jump probabilities can be characterized as:

$\frac{P(jump_{a\rightarrow b})\cdot P(accept b|a)}{P(jump_{b\rightarrow a})\cdot P(accept a|b)}$

For this procedure to (eventually) extract random samples from the posterior distribution, _this ratio MUST be equal to the ratio of the posterior probabilities at points A and B_:

$\frac{Posterior(B)}{Posterior(A)}$

If this rule is met, then in the long run the chain will explore each segment of parameter space in accordance with the posterior probability of that segment- thereby eventually capturing the entire posterior distribution (if run long enough).

Note: if the proposal distribution is symmetrical, then the probability of proposing a jump from A to B is the same as the probability of proposing a jump from B to A. Therefore, the proposal probabilities drop out of the above equation...

Surprisingly, MCMC is not that difficult to describe or even to implement (although implementing an efficient MCMC algorithm takes some advanced skills!). Let's look at a simple MCMC algorithm.

## Bivariate normal example

Remember that MCMC samplers are just a type of random number generator. We can use MCMC to develop our own random number generator for a simple known distribution. In this example, we use MCMC to generate random numbers from a standard bivariate normal probability distribution. I chose this example because its properties are fully worked out mathematically, so we can compare MCMC results with the "true" distribution. 

We certainly don't need an MCMC sampler for this example (we already have 'perfect' random number generators, like the one below). One way to do this would be to use the following code, which draws and visualizes an arbitrary number of independent samples from the bivariate standard normal distribution with a correlation parameter, $\rho$.

### Metropolis-Hastings algorithm

This algorithm is essentially the same as the *simulated annealing algorithm* we discussed in the "optimization" lecture! The main difference: the "temperature" doesn't decrease over time and the temperature parameter *k* is always set to 1.

The M-H algorithm can be expressed as:

$P(accept B|A) = min(1,\frac{Posterior(B)}{Posterior(A)}\cdot \frac{P(b\rightarrow a)}{P(a\rightarrow b)})$

**Q**: Does it make sense that this algorithm meets the basic rule of MCMC, that is, that the ratio of jump probabilities to/from any two points in parameter/model space is equal to the ratio of posterior probabilities evaluated at those two points? 

```{r}

# Simple example of MCMC sampling -----------------------

# first, let's build a function that generates random numbers from a bivariate standard normal distribution

rbvn<-function (n, rho){   #function for drawing an arbitrary number of independent samples from the bivariate standard normal distribution. 
        x <- rnorm(n, 0, 1)
        y <- rnorm(n, rho * x, sqrt(1 - rho^2))
        cbind(x, y)
}

# Now, plot the random draws from this distribution, make sure this makes sense!

bvn_true<-rbvn(10000,0.9)
par(mfrow=c(2,2))
plot(ts(bvn_true[,1]))
plot(ts(bvn_true[,2]))
hist(bvn_true[,1],40)
hist(bvn_true[,2],40)
par(mfrow=c(1,1))


```
These look like two normal distributions- but when we view them together we can visualize the correlation/dependence:

```{r}

plot(bvn_true[,1],bvn_true[,2])

```
Here's a function for visualizing the results from our various samplers:

```{r}
vis_mcmc2 <- function(mat){   # assume mat has 1 row per sample, and 2 columns (# params)
  par(mfrow=c(3,ncol(mat)))
  plot(mat,col=1:nrow(mat)) ; plot(mat,type="l")
  plot(ts(mat[,1])) ; plot(ts(mat[,2]))
  hist(mat[,1],40) ; hist(mat[,2],40)
  par(mfrow=c(1,1))
}

```



```{r}

## Metropolis-Hastings implementation of bivariate normal sampler...   ----------

proposal <- function(x) rnorm(1,x,0.4)
cond_prob <- function(this,xval) dnorm(this,rho*xval,1-rho^2) 
choose_mh <- function(prev,new,other){
  r = min(cond_prob(new,other)/cond_prob(prev,other),1)
  ifelse(r>runif(1),new,prev)
}

metropolisHastings <- function (n, rho){    # an MCMC bivariate random number generator
    mat <- matrix(NA,n,2)   # matrix for storing the random samples
    prev = c(0,0); mat[1,] <- prev       # initial values for all parameters
    for(i in 1:n){
      for(j in 1:2) prev[j] <- choose_mh(prev[j],proposal(prev[j]),prev[setdiff(1:2,j)] )
      mat[i,] <- prev
    }
    mat
}

```


Then we can use the M-H sampler to get random samples from this known distribution...

```{r}

# Test our M-H sampler

bvn_mh<-metropolisHastings(5000,0.9)
vis_mcmc2(bvn_mh)

```

**Q**: Why does the MCMC routine produce 'lower quality' random numbers than the first (bivariate normal) sampler we built?

The M-H algorithm is simple and widely applicable. The downside is computational. Many proposal steps are rejected, and this leads to low efficiency. Strong autocorrelation in the MCMC samples can also mean very low effective sample size (ess) despite a large number of MCMC samples. 

### Gibbs sampler

The Gibbs sampler is straightforward and efficient. Basically, the algorithm successively samples from the *full conditional* probability distribution -- that is, the posterior distribution for arbitrary parameter *i* conditional on known values for all other parameters in the model. Unlike the M-H algorithm, there is no rejection of proposed jumps when we use the Gibbs sampler!  

In many cases, we can't work out the full posterior distribution for our model directly, but we **CAN** work out the conditional posterior distribution analytically if all parameters except for the parameter in question were known with certainty. This is true if we use *conjugate priors* for our model specification (what BUGS/JAGS tries to do!). Even if not, the full conditional is often tractable. Nonetheless, even if it's not analytically tractable, we can use an M-H step as a last resort!

**Q**: Why is a Gibbs sampler usually more efficient than a pure M-H sampler?  

In this example (same as before), we use a Gibbs Sampler to generate random numbers from a standard bivariate normal probability distribution. Notice that the Gibbs sampler is in many ways more simple and straightforward than the M-H algorithm. 

```{r}

# Simple example of a Gibbs sampler ----------------

# first, recall our 'true' bivariate normal sampler

vis_mcmc2(bvn_true)

```



```{r}

## Now construct a Gibbs sampler  ---------------

conditional2 <- function(xval) rnorm(1, rho * xval, sqrt(1 - rho^2))   # sample random value from full conditional

gibbs<-function (n, rho){    # a Gibbs sampler for bivariate normal
    mat <- matrix(ncol = 2, nrow = n)   # matrix for storing the random samples
    prev <- c(0,0); mat[1, ] <- prev     # initialize the markov chain
    for (i in 2:n) {
      prev[1] <- conditional2(prev[2])   # sample from full conditional
      prev[2] <- conditional2(prev[1])
      mat[i,] <- prev
    }
    mat
}

```


Then we can use the Gibbs sampler to get random samples from this known distribution...

```{r}

# Test the Gibbs sampler ------------------

bvn_gbs <-gibbs(10000,0.9)
vis_mcmc2(bvn_gbs)

```

This looks pretty good! However, it is important to note that Gibbs samplers frequently result in chains with fairly high autocorrelation, where the effective sample size (ess) is much lower than the number of MCMC samples. 

### Hamiltonian Monte Carlo (HMC)

Hamiltonian Monte Carlo (HMC) is more mathematically complex than M-H or HMC. However, it has some very nice properties. Most importantly, the samples from HMC are much less correlated, so we need fewer samples in total. HMC is also very good at sampling even in high dimensional problems that are too slow or poorly mixed in a Gibbs sampler. The term "Hamiltonian" comes from physics, where the Hamiltonian equations describe an objects movement in space as a function of the relationship between kinetic energy and potential energy. Imagine the negative log-posterior as a frictionless surface we want to explore. We can think of the MCMC sample as a ball rolling in that space. We give the ball a little push in some direction (this is governed by the momemtum function), giving it kinetic energy. The ball might roll downhill, converting some of the potential energy into kinetic energy. At some point is starts rolling up again until all the kinetic energy is converted back into potential energy. 

Basically, this algorithm forces the MCMC sampler to stay within higher-probability regions of parameter space (called the "typical set"), making for a sampler that is efficient even when sampling from high dimensions (dealing with the curse of dimensionality). 

One restriction in HMC is that the parameter space must be continuous and differentiable. But there are tricks for using HMC even in cases where we have discrete elements of parameter space- basically we just need to integrate these discrete elements out of the likelihood. Which is not too difficult in many cases. 

The tuning parameters in HMC include the step size and the number of sub-steps between HMC draws, known as 'leap-frog' steps. There is a widely used algorithm called the 'no U-turn sampler' (NUTS) that is often used to define these tuning parameters in practice. 

The following code is borrowed from https://jonnylaw.rocks/posts/2019-07-31-hmc/. 
Law, Jonny. 2019. “Hamiltonian Monte Carlo in R.” July 31, 2019.

#### START HERE


Okay, let's try it for a non-trivial problem, like the Myxomatosis example from the Bolker book!

## Myxomatosis revisited (again!)

```{r}

# MCMC implementation of the Myxomatosis example from the Bolker book --------------

library(emdbook)

MyxDat <- MyxoTiter_sum
Myx <- subset(MyxDat,grade==1)
head(Myx)

```

Recall that we are modeling the distribution of measured titers (virus loads) for Australian rabbits. Bolker chose to use a Gamma distribution. Here is the empirical distribution:

```{r}

# Visualize the Myxomatosis data for the 100th time!

hist(Myx$titer,freq=FALSE)

```

We need to estimate the gamma shape and rate parameters that best fit this empirical distribution. Here is one example of a Gamma fit to this distribution:

Recall that the 2-D (log) likelihood surface looks something like this:

```{r}

# define 2-D parameter space!

shapevec <- seq(3,100,by=0.1)   
scalevec <- seq(0.01,0.5,by=0.001)

# define the likelihood surface  -------------

GammaLogLikelihoodFunction <- function(params){
  sum(dgamma(Myx$titer,shape=params['shape'],scale=params['scale'],log=T))
}
surface2D <- matrix(nrow=length(shapevec),ncol=length(scalevec))   # initialize storage variable

newparams <- c(shape=50,scale=0.2)
for(i in 1:length(shapevec)){
  newparams['shape'] <- shapevec[i]
  for(j in 1:length(scalevec)){
    newparams['scale'] <- scalevec[j]
    surface2D[i,j] <- GammaLogLikelihoodFunction(newparams) 
  }
}

# Visualize the likelihood surface

image(x=shapevec,y=scalevec,z=surface2D,zlim=c(-1000,-30),col=topo.colors(12))
contour(x=shapevec,y=scalevec,z=surface2D,levels=c(-30,-40,-80,-500),add=T)

```


Here is an implementation of the M-H algorithm to find the joint posterior distribution!

First, we need a likelihood function (our old friend!)- this time, we will return real probabilities-- NOT log-transformed probabilities 

```{r}

# Write a non-log-transformed likelihood function ------------

GammaLikelihoodFunction <- function(params){
  prod(dgamma(Myx$titer,shape=params['shape'],scale=params['scale'],log=F))   
}

  # and here's the log likelihood function
GammaLogLikelihoodFunction <- function(params){
  sum(dgamma(Myx$titer,shape=params['shape'],scale=params['scale'],log=T))   
}

params <- c(shape=40,scale=0.15) 
params
GammaLikelihoodFunction(params)
GammaLogLikelihoodFunction(params)

```

Then, we need a prior distribution for our parameters! Let's assign relatively flat priors for both of our parameters. In this case, let's assign a $gamma(shape=0.001,scale=1000)$ for both parameters (mean of 1 and very large variance):

```{r}

# Function for returning the prior probability density for any point in parameter space 

GammaPriorFunction <- function(params){
  prior <- c(shape=NA,scale=NA)
  prior['shape'] <- dgamma(params['shape'],shape=0.001,scale=1000)
  prior['scale'] <- dgamma(params['scale'],shape=0.001,scale=1000)
  # prior['shape'] <- dunif(params['shape'],3,100)        # alternative: could use uniform prior!
  # prior['scale'] <- dunif(params['scale'],0.01,0.5)
  return(prod(prior))
}

GammaLogPriorFunction <- function(params){
  prior <- c(shape=NA,scale=NA)
  prior['shape'] <- dgamma(params['shape'],shape=0.001,scale=1000,log=T)
  prior['scale'] <- dgamma(params['scale'],shape=0.001,scale=1000,log=T)
  # prior['shape'] <- dunif(params['shape'],3,100)        # alternative: could use uniform prior!
  # prior['scale'] <- dunif(params['scale'],0.01,0.5)
  return(sum(prior))
}

curve(dgamma(x,shape=0.001,scale=1000),3,100)

params <- c(shape=40,scale=0.15) 
params
GammaPriorFunction(params)

prior2D <- matrix(nrow=length(shapevec),ncol=length(scalevec))   # initialize storage variable

newparams <- c(shape=50,scale=0.2)
for(i in 1:length(shapevec)){
  newparams['shape'] <- shapevec[i]
  for(j in 1:length(scalevec)){
    newparams['scale'] <- scalevec[j]
    prior2D[i,j] <- GammaPriorFunction(newparams) 
  }
}

# Visualize the prior likelihood surface

image(x=shapevec,y=scalevec,z=prior2D,zlim=c(0.000000001,0.001),col=topo.colors(12))
#contour(x=shapevec,y=scalevec,z=prior2D,levels=c(-30,-40,-80,-500),add=T)



```

Note that we are also assuming that the shape and scale are *independent* in the prior (multiplicative probabilities for the joint prior). However, this does not impose the same assumption on the posterior- parameter correlations in the likelihood can and will be reflected in the posterior distribution. 

Then, we need a function that can compute the ratio of posterior probabilities for any given jump in parameter space. Because we are dealing with a *ratio* of posterior probabilities, _we do NOT need to compute the normalization constant_. 

Without the need for a normalization constant, we just need to compute the ratio of weighted likelihoods (that is, the likelihood weighted by the prior)

```{r}

# Function for computing the ratio of posterior densities -----------------

PosteriorRatio <- function(oldguess,newguess){
  oldLik <- max(1e-90,GammaLikelihoodFunction(oldguess))   # compute likelihood and prior density at old guess
  oldPrior <- max(1e-90,GammaPriorFunction(oldguess))
  newLik <- GammaLikelihoodFunction(newguess)             # compute likelihood and prior density at new guess
  newPrior <- GammaPriorFunction(newguess)
  return((newLik*newPrior)/(oldLik*oldPrior))          # compute ratio of weighted likelihoods
}

PosteriorRatio2 <- function(oldguess,newguess){
  oldLogLik <- GammaLogLikelihoodFunction(oldguess)   # compute likelihood and prior density at old guess
  oldLogPrior <- GammaLogPriorFunction(oldguess)
  newLogLik <- GammaLogLikelihoodFunction(newguess)             # compute likelihood and prior density at new guess
  newLogPrior <- GammaLogPriorFunction(newguess)
  return(exp((newLogLik+newLogPrior)-(oldLogLik+oldLogPrior)))          # compute ratio of weighted likelihoods
}

oldguess <- params
newguess <- c(shape=39,scale=0.15)

PosteriorRatio(oldguess,newguess)
PosteriorRatio2(oldguess,newguess)

```  

Then we need a function for making new guesses, or jumps in parameter space:

```{r}

# Define proposal distribution --------------------------
    #for jumps in parameter space (use normal distribution)!

     # function for making new guesses
newGuess <- function(oldguess){
  sdshapejump <- 3 #4
  sdscalejump <- 0.05
  corjump <- -0.6
  vcv <- diag(2)*c(sdshapejump^2,sdscalejump^2)
  vcv[2,1] <- vcv[1,2] <- corjump*sqrt(sdshapejump^2*sdscalejump^2)
  newguess <- mvtnorm::rmvnorm(1,oldguess,vcv)[1,] # c(shape=rnorm(1,mean=0,sd=sdshapejump),scale=rnorm(1,0,sdscalejump))
  # newguess <- abs(oldguess + jump)    # note: by taking the abs val to avoid negative numbers, our proposal jump probs are not strictly symmetrical, but this should not present a big issue in practice
  newguess <- ifelse(newguess<0.001,0.001,newguess)
  return(newguess)
}
  # set a new "guess" near to the original guess

newGuess(oldguess=params)   
newGuess(oldguess=params)
newGuess(oldguess=params)

```

Now we are ready to implement the Metropolis-Hastings MCMC algorithm:


We need a starting point:

```{r}

# Set a starting point in parameter space -------------------

startingvals <- c(shape=75,scale=0.28)    # starting point for the algorithm

```

Let's play with the different functions we have so far...

```{r}

# Try our new functions  ------------------

startingvals
newguess <- newGuess(startingvals)    # take a jump in parameter space
newguess

PosteriorRatio2(startingvals,newguess)   # difference in posterior ratio

```


Now let's look at the Metropolis-Hastings routine:

```{r}

# Visualize the Metropolis-Hastings routine: ---------------

chain.length <- 11
oldguess <- startingvals
guesses <- matrix(0,nrow=chain.length,ncol=2)
colnames(guesses) <- names(startingvals)
guesses[1,] <- startingvals
counter <- 2
while(counter <= chain.length){
  newguess <- newGuess(oldguess)
  post.rat <- PosteriorRatio2(oldguess,newguess)
  prob.accept <- min(1,post.rat)
  rand <- runif(1)
  if(rand<=prob.accept){
    oldguess <- newguess
    guesses[counter,] <- newguess 
  }else{
    guesses[counter,] <- oldguess 
  }
  counter=counter+1
}

# visualize!

image(x=shapevec,y=scalevec,z=surface2D,zlim=c(-1000,-30),col=topo.colors(12))
contour(x=shapevec,y=scalevec,z=surface2D,levels=c(-30,-40,-80,-500),add=T)
lines(guesses,col="red")

```


Let's run it for longer...


```{r}

# Get more MCMC samples --------------

chain.length <- 1000
oldguess <- startingvals
guesses <- matrix(0,nrow=chain.length,ncol=2)
colnames(guesses) <- names(startingvals)
guesses[1,] <- startingvals

counter <- 2
while(counter <= chain.length){
  newguess <- newGuess(oldguess)
  post.rat <- PosteriorRatio2(oldguess,newguess)
  prob.accept <- min(1,post.rat)
  rand <- runif(1)
  if(rand<=prob.accept){
    oldguess <- newguess
    guesses[counter,] <- newguess 
  }else{
    guesses[counter,] <- oldguess 
  }
  counter=counter+1
}

# visualize!

image(x=shapevec,y=scalevec,z=surface2D,zlim=c(-1000,-30),col=topo.colors(12))
contour(x=shapevec,y=scalevec,z=surface2D,levels=c(-30,-40,-80,-500),add=T)
lines(guesses,col="red")

```

How about for even longer??


```{r}

# And more... -------------------

chain.length <- 10000
oldguess <- startingvals
guesses <- matrix(0,nrow=chain.length,ncol=2)
colnames(guesses) <- names(startingvals)
guesses[1,] <- startingvals

counter <- 2
while(counter <= chain.length){
  newguess <- newGuess(oldguess)
  post.rat <- PosteriorRatio2(oldguess,newguess)
  prob.accept <- min(1,post.rat)
  rand <- runif(1)
  if(rand<=prob.accept){
    oldguess <- newguess
    guesses[counter,] <- newguess 
  }else{
    guesses[counter,] <- oldguess 
  }
  counter=counter+1
}

# visualize!

image(x=shapevec,y=scalevec,z=surface2D,zlim=c(-1000,-30),col=topo.colors(12))
contour(x=shapevec,y=scalevec,z=surface2D,levels=c(-30,-40,-80,-500),add=T)
lines(guesses,col="red")

```

This looks better! The search algorithm is finding the high-likelihood parts of parameter space pretty well!

Now, let's look at the chain for the "shape" parameter

```{r}

# Evaluate "traceplot" for the MCMC samples... ---------------------

## Shape parameter

plot(1:chain.length,guesses[,'shape'],type="l",main="shape parameter",xlab="iteration",ylab="shape")

```

And for the scale parameter...

```{r}

## Scale parameter

plot(1:chain.length,guesses[,'scale'],type="l",main="scale parameter",xlab="iteration",ylab="scale")

```

Can we say that these chains have converged on the posterior distribution for the shape parameter??


First of all, the beginning of the chain "remembers" the starting value, and is therefore not a stationary distribution. We need to remove the first part of the chain, called the **'burn-in'**.

```{r}

# Remove "burn-in" (allow MCMC routine some time to get to the posterior) --------------

burn.in <- 1000
MCMCsamples <- guesses[-c(1:burn.in),]

chain.length=chain.length-burn.in
plot(1:chain.length,MCMCsamples[,'shape'],type="l",main="shape parameter",xlab="iteration",ylab="shape")
plot(1:chain.length,MCMCsamples[,'scale'],type="l",main="scale parameter",xlab="iteration",ylab="scale")

```


But it still doesn't look all that great. Let's run it for even longer, and see if we get something that looks more like a proper random number generator (white noise)...


```{r}

# Try again- run for much longer ---------------------

chain.length <- 100000
oldguess <- startingvals
guesses <- matrix(0,nrow=chain.length,ncol=2)
colnames(guesses) <- names(startingvals)
guesses[1,] <- startingvals

counter <- 2
while(counter <= chain.length){
  newguess <- newGuess(oldguess)
  post.rat <- PosteriorRatio2(oldguess,newguess)
  prob.accept <- min(1,post.rat)
  rand <- runif(1)
  if(rand<=prob.accept){
    oldguess <- newguess
    guesses[counter,] <- newguess 
  }else{
    guesses[counter,] <- oldguess 
  }
  counter=counter+1
}

# visualize!

image(x=shapevec,y=scalevec,z=surface2D,zlim=c(-1000,-30),col=topo.colors(12))
contour(x=shapevec,y=scalevec,z=surface2D,levels=c(-30,-40,-80,-500),add=T)
lines(guesses,col="red")

```


Let's first remove the first 5000 samples as a burn-in

```{r}

# Use longer "burn-in" ------------------

burn.in <- 25000
MCMCsamples <- guesses[-c(1:burn.in),]
chain.length=chain.length-burn.in

```


Now, let's look at the chains again

```{r}

plot(1:chain.length,MCMCsamples[,'shape'],type="l",main="shape parameter",xlab="iteration",ylab="shape")
plot(1:chain.length,MCMCsamples[,'scale'],type="l",main="scale parameter",xlab="iteration",ylab="scale")

```


When evaluating these trace plots, we are hoping to see a "stationary distribution" that looks like white noise. This trace plot looks like it might have a little autocorrelation. One way to "fix" this is to thin the MCMC samples:

```{r}

# "thin" the MCMC samples  -----------------------

thinnedMCMC <- MCMCsamples[seq(1,chain.length,by=10),]
plot(1:nrow(thinnedMCMC),thinnedMCMC[,'shape'],type="l",main="shape parameter",xlab="iteration",ylab="shape")
plot(1:nrow(thinnedMCMC),thinnedMCMC[,'scale'],type="l",main="scale parameter",xlab="iteration",ylab="scale")

```



Now we can examine our posterior distribution!

```{r}

# Visualize the posterior!

plot(density(thinnedMCMC[,'scale']),main="scale parameter",xlab="scale")
plot(density(thinnedMCMC[,'shape']),main="shape parameter",xlab="shape")

```

And we can visualize as before.

 
```{r}

# More visual posterior checks... -----------------

par(mfrow=c(3,2))
plot(thinnedMCMC,col=1:10000)
plot(thinnedMCMC,type="l")
plot(ts(thinnedMCMC[,1]))
plot(ts(thinnedMCMC[,2]))
hist(thinnedMCMC[,1],40)
hist(thinnedMCMC[,2],40)
par(mfrow=c(1,1))

```

Hopefully it is clear that the Metropolis-Hastings MCMC method could be modified to fit arbitrary numbers of free parameters for arbitrary models. However, the M-H algorithm by itself is not necessarily the most efficient- and it can be very difficult to develop proposal distributions for multi-dimensional parameter space. The **Gibbs sampler** tends to be a more effective sampler for multi-dimensional joint posterior distributions. In lab we will play around with Gibbs samplers, mostly using a modeling language called **BUGS** (**B**ayesian **I**nference **U**sing **G**ibbs **S**ampling). 

NOTE: BUGS implementations (e.g., JAGS) actually tend to use a combination of M-H and Gibbs sampling! 

NOTE: M-H and Gibbs samplers aren't the only MCMC routines out there. For example, the popular program 'stan' uses a modification of M-H sampling called 'Hamiltonian Monte Carlo", which tends to explore parameter space more efficiently 




### Back to Myxomatosis!

#### Aside: the BUGS language

Finally, let's build a Gibbs sampler for our favorite Myxomatosis example! To do this, we will use the BUGS language (as implemented in JAGS), to help us!

The BUGS language looks similar to R, but there are several key differences:  

* First of all, BUGS is a compiled language, so the order of operations in your code doesn't really matter
* BUGS is not vectorized- you need to use FOR loops for just about everything!
* Several probability distributions are parameterized very differently in BUGS. Notably, the normal distribution is parameterized with a mean and a precision ($1/Variance$).


Here is the myxomatosis example, as implemented in the BUGS language:

NOTE: this code looks a little like R, but don't be confused- this is not R! 

```

model {
  
  #############
  # LIKELIHOOD
  ############
  for(obs in 1:n.observations){
    titer[obs] ~ dgamma(shape,rate)
  }
  
  #############
  # PRIORS
  ############
  shape ~ dgamma(0.001,0.001)
  scale ~ dgamma(0.001,0.001)
  rate <- 1/scale   # convert the scale parameter to a "rate" for BUGS
}

```

We can use the "cat" function in R to write out this model to a text file in your working directory:

```{r}

# Myxomatosis example in BUGS modeling language ---------------

# Write the BUGS model to file

cat("
  model {
    
    #############
    # LIKELIHOOD
    ############
    for(obs in 1:n.observations){
      titer[obs] ~ dgamma(shape,rate)
    }
    
    #############
    # PRIORS
    ############
    shape ~ dgamma(0.001,0.001)
    scale ~ dgamma(0.001,0.001)
    rate <- 1/scale
  }
", file="BUGSmodel.txt")


```

Now that we have the BUGS model packaged as a text file, we bundle the data into a single list object that contains all the relevant data referenced in the BUGS code:

```{r}

# Encapsulate the data into a single "list" object ------------------

myx.data.for.bugs <- list(
  titer = Myx$titer,
  n.observations = length(Myx$titer)
)

myx.data.for.bugs

```

Then we need to define the initial values for all parameters. It is convenient to define this as a function, so that each MCMC chain can be initialized with different starting values. This will become clear later!

```{r}

# Function for generating random initial values --------------

init.vals.for.bugs <- function(){
  list(
    shape=runif(1,20,100),
    scale=runif(1,0.05,0.3)
  )
}

init.vals.for.bugs()
init.vals.for.bugs()
init.vals.for.bugs()

```

Now we can call JAGS!

```{r}

# Run JAGS!!!!  ------------------

#library(R2jags)
library(jagsUI)

library(coda)

params.to.store <- c("shape","scale")

jags.fit <- jags(model="BUGSmodel.txt",data=myx.data.for.bugs,inits=init.vals.for.bugs,parameters.to.save=params.to.store,n.iter=5000,n.chains = 3,n.adapt = 100,n.burnin = 0)

jagsfit.mcmc <- jags.fit$samples   # extract "MCMC" object

summary(jagsfit.mcmc)

plot(jagsfit.mcmc)

```

### Assessing convergence

This is probably a good time to talk about convergence of MCMC chains on the stationary posterior distribution. The above plots don't look great. We want to see white noise, and we want to see chains that look similar to one another. 

The first check is just visual- we look for the following to assess convergence:

* The chains for each parameter, when viewed as a "trace plot" should look like white noise (fuzzy caterpillar!), or similar.
* Multiple chains with different starting conditions should look the same!! 

One way we might be able to do a better job here is to run the chains for longer and discard the initial samples as a *burn in*!

We can also try to reduce serial autocorrelation by thinning our chain- here we retain only 1 out of every 100 samples.

```{r}

# Run the chains for longer! -----------------

jags.fit <- jags(model="BUGSmodel.txt",data=myx.data.for.bugs,inits=init.vals.for.bugs,parameters.to.save=params.to.store,
                     n.iter = 100000,n.chains = 3,n.adapt = 1000,n.burnin = 10000,
                     n.thin=100,parallel=T )

jagsfit.mcmc <- jags.fit$samples   # convert to "MCMC" object (coda package)

summary(jagsfit.mcmc)

plot(jagsfit.mcmc)

```

Just visually, this looks better. Now we can use some more quantitative convergence metrics. 

#### The Gelman-Rubin diagnostic

One simple and intuitive convergence diagnostic is the *Gelman-Rubin diagnostic*, which assesses whether chains are more different from one another than they should be on the basis of simple Monte Carlo error:

```{r}

# Run convergence diagnostics  ------------------

gelman.diag(jagsfit.mcmc)

```

In general, values of 1.1 or higher are considered poorly converged. Compute the G-R diagnostic for all free parameters in your model. If your tests fail, you should try running longer chains!

So this model looks pretty good!!



## Useful links for MCMC

[Darren Winkinson's research blog](https://darrenjw.wordpress.com/2011/07/16/gibbs-sampler-in-various-languages-revisited/)   


[--go to next lecture--](LECTURE8.html) 



