---
title: "Figure Demo"
author: "NRES 746"
date: "Fall 2018"
output: 
  html_document: 
    theme: cerulean
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Publication-quality figures in R

NOTE: this demo was originally compiled by Dr. Elizabeth Hunter in fall 2016 and has been modified slightly from its original version.               


```{r, echo=FALSE}
#rm(list=ls())
def.par <- par(no.readonly = TRUE)
par(def.par)
```


#Read in the data
These data are nest fates for Seaside Sparrows that live in tidal marshes.  We're going to make a figure that shows both how nest height and tidal height affect Seaside Sparrow nest fates. See Hunter et al. 2016 Animal Behaviour for more information.
```{r}
tide = read.csv("tide_ALL_navd_hh.csv", header=TRUE)
tide$Date = as.Date(tide$Date, format="%m/%d/%Y")
tide$Year = substr(tide$Date, 3, 4)
tide.13 = tide[tide$Year==13,]
tide.14 = tide[tide$Year==14,]
tide.15 = tide[tide$Year==15,]

bn = read.csv("Nest_basic_ALL.csv", header=TRUE)
#Extract pertinent data
bn = bn[bn$Species=="SESP",]
bn = bn[bn$Fate_date!="",]
bn = bn[bn$Fate!="NA",]
bn = bn[bn$Fate!=4,]
bn = na.omit(bn)
#Put dates/times in correct format
bn$Date_found = as.Date(bn$Date_found, format="%m/%d/%Y")
bn$Fate_date = as.Date(bn$Fate_date, format="%m/%d/%Y")
bn$Start_date = as.Date(bn$Start_date, format="%m/%d/%Y")
bn$year = substr(bn$Start_date, 3, 4)
bn = bn[order(bn$year, -bn$Fate),]
bn.13 = bn[bn$year==13,]
bn.14 = bn[bn$year==14,]
bn.15 = bn[bn$year==15,]

#Take a look at data structure
head(bn)
```

# Creating figures
First, we'll start by making single-panel figures of tides and nest fates.
```{r}
#Plot tide height as a function of time
plot(tide.13$Water_Level ~ tide.13$Date, type="l")

#Fix up the plot a bit
plot(tide.13$Water_Level ~ tide.13$Date, type="l", ylab="Max Tidal Height (m)", lwd=1, main="2013", xlab="")
#Add segments to show when in the tidal cycle sparrow nests were flooded
segments(x0=as.Date("2013-05-26"), x1=as.Date("2013-05-26"), y0=0.6, y1=1.595, col=gray(0.5))

###
#Plot nest fate as a function of time and nest height
plot(x=bn.13$Fate_date, y=bn.13$Nest_height, xlim=c(min(tide.13$Date), max(tide.13$Date)), 
	ylim=c(min(bn$Nest_height), max(bn$Nest_height)), ylab="Nest Height (m)")
#"Jitter" to show overlapping points
plot(x=bn.13$Fate_date, y=jitter(bn.13$Nest_height), xlim=c(min(tide.13$Date), max(tide.13$Date)), 
	ylim=c(min(bn$Nest_height), max(bn$Nest_height)), ylab="Nest Height (m)")
#Color and shapes to distiguish fates
plot(x=bn.13$Fate_date, y=jitter(bn.13$Nest_height), xlim=c(min(tide.13$Date), max(tide.13$Date)), 
	ylim=c(min(bn$Nest_height), max(bn$Nest_height)), ylab="Nest Height (m)",
	col=ifelse(bn.13$Fate==1, "dodgerblue4", ifelse(bn.13$Fate==3, "palegreen3", "gold3")),
	pch=ifelse(bn.13$Fate==1, 0, ifelse(bn.13$Fate==3, 1, 17)))
#Cex to change size of pch
plot(x=bn.13$Fate_date, y=jitter(bn.13$Nest_height), xlim=c(min(tide.13$Date), max(tide.13$Date)), 
	ylim=c(min(bn$Nest_height), max(bn$Nest_height)), ylab="Nest Height (m)",
	col=ifelse(bn.13$Fate==1, "dodgerblue4", ifelse(bn.13$Fate==3, "palegreen3", "gold3")),
	pch=ifelse(bn.13$Fate==1, 0, ifelse(bn.13$Fate==3, 1, 17)), cex=1.5)

```

# Multipanel Figures
There are several ways to make multipanel figures in R. We'll start with the easiest (changing the par(mfrow) options), and then look at how to use layout() to make a publication-ready figure.
```{r}
#Multipanel plotting using par(mfrow) and par(mar) to change panel margins.
par(mfrow=c(2,3), mar=c(2, 4, 2, 0) + 0.1)
plot(tide.13$Water_Level ~ tide.13$Date, type="l", ylim=c(min(tide$Water_Level), max(tide$Water_Level)), ylab="Max Tidal Height (m)", lwd=1,
	xaxt="n", main="2013")
abline(1.5, 0, lty=2)
plot(tide.14$Water_Level ~ tide.14$Date, type="l", ylim=c(min(tide$Water_Level), max(tide$Water_Level)), ylab="", lwd=1,
	xaxt="n", yaxt="n", main="2014")
abline(1.5, 0, lty=2)
plot(tide.15$Water_Level ~ tide.15$Date, type="l", ylim=c(min(tide$Water_Level), max(tide$Water_Level)), ylab="", lwd=1,
	xaxt="n", yaxt="n", main="2015")
abline(1.5, 0, lty=2)
plot(x=bn.13$Fate_date, y=jitter(bn.13$Nest_height), col=ifelse(bn.13$Fate==1, "dodgerblue4", ifelse(bn.13$Fate==3, "palegreen3", "gold3")),
	pch=ifelse(bn.13$Fate==1, 0, ifelse(bn.13$Fate==3, 1, 17)), xlim=c(min(tide.13$Date), max(tide.13$Date)), 
	ylim=c(min(bn$Nest_height), max(bn$Nest_height)), ylab="Nest Height (m)", cex=1.5)
legend("topleft", legend=c("Flood", "Pred", "Fledge"), col=c("dodgerblue4", "palegreen3", "gold3"), pch=c(0,1,17), cex=1.1)
plot(x=bn.14$Fate_date, y=jitter(bn.14$Nest_height), col=ifelse(bn.14$Fate==1, "dodgerblue4", ifelse(bn.14$Fate==3, "palegreen3", "gold3")),
	pch=ifelse(bn.14$Fate==1, 0, ifelse(bn.14$Fate==3, 1, 17)), xlim=c(min(tide.14$Date), max(tide.14$Date)), 
	ylim=c(min(bn$Nest_height), max(bn$Nest_height)), ylab="", yaxt="n", cex=1.5)
plot(x=bn.15$Fate_date, y=jitter(bn.15$Nest_height), col=ifelse(bn.15$Fate==1, "dodgerblue4", ifelse(bn.15$Fate==3, "palegreen3", "gold3")),
	pch=ifelse(bn.15$Fate==1, 0, ifelse(bn.15$Fate==3, 1, 17)), xlim=c(min(tide.15$Date), max(tide.15$Date)), 
	ylim=c(min(bn$Nest_height), max(bn$Nest_height)), ylab="", yaxt="n", cex=1.5)

#This looks okay, but we want our panels to be closer together and for our y-axis labels to have more room.

#Multipanel plotting using layout()
temp.layout <- layout(matrix(c(1,2,3,4,5,6),2,3,byrow=TRUE), widths=c(lcm(6.1),lcm(5),lcm(5)), heights=c(lcm(4.7),lcm(4.7)))
#Take a look at the layout we've created.
layout.show(temp.layout)

#Here we will change the panel margins for each panel separately using par(mar)
par(mar=c(0, 4, 2, 0) + 0.1)
plot(tide.13$Water_Level ~ tide.13$Date, type="l", ylim=c(min(tide$Water_Level), max(tide$Water_Level)), ylab="Max Tidal Height (m)", lwd=1,
	xaxt="n", main="2013")
abline(1.5, 0, lty=2)
segments(x0=as.Date("2013-05-26"), x1=as.Date("2013-05-26"), y0=0.6, y1=1.595, col=gray(0.5))
segments(x0=as.Date("2013-06-06"), x1=as.Date("2013-06-06"), y0=0.6, y1=1.344, col=gray(0.5))
segments(x0=as.Date("2013-06-21"), x1=as.Date("2013-06-21"), y0=0.6, y1=1.638, col=gray(0.5))
segments(x0=as.Date("2013-07-25"), x1=as.Date("2013-07-25"), y0=0.6, y1=1.566, col=gray(0.5))
par(mar=c(0, 1, 2, 0) + 0.1)
plot(tide.14$Water_Level ~ tide.14$Date, type="l", ylim=c(min(tide$Water_Level), max(tide$Water_Level)), ylab="Max Tidal Height (m)", lwd=1,
	xaxt="n", yaxt="n", main="2014")
abline(1.5, 0, lty=2)
segments(x0=as.Date("2014-04-19"), x1=as.Date("2014-04-19"), y0=0.6, y1=1.579, col=gray(0.5))
segments(x0=as.Date("2014-04-29"), x1=as.Date("2014-04-29"), y0=0.6, y1=1.527, col=gray(0.5))
segments(x0=as.Date("2014-05-15"), x1=as.Date("2014-05-15"), y0=0.6, y1=1.526, col=gray(0.5))
segments(x0=as.Date("2014-06-16"), x1=as.Date("2014-06-16"), y0=0.6, y1=1.502, col=gray(0.5))
segments(x0=as.Date("2014-07-12"), x1=as.Date("2014-07-12"), y0=0.6, y1=1.406, col=gray(0.5))
par(mar=c(0, 1, 2, 0) + 0.1)
plot(tide.15$Water_Level ~ tide.15$Date, type="l", ylim=c(min(tide$Water_Level), max(tide$Water_Level)), ylab="Max Tidal Height (m)", lwd=1,
	xaxt="n", yaxt="n", main="2015")
abline(1.5, 0, lty=2)
segments(x0=as.Date("2015-04-19"), x1=as.Date("2015-04-19"), y0=0.6, y1=1.526, col=gray(0.5))
segments(x0=as.Date("2015-05-16"), x1=as.Date("2015-05-16"), y0=0.6, y1=1.491, col=gray(0.5))
par(mar=c(2, 4, 0, 0) + 0.1)
plot(x=bn.13$Fate_date, y=jitter(bn.13$Nest_height), col=ifelse(bn.13$Fate==1, "dodgerblue4", ifelse(bn.13$Fate==3, "palegreen3", "gold3")),
	pch=ifelse(bn.13$Fate==1, 0, ifelse(bn.13$Fate==3, 1, 17)), xlim=c(min(tide.13$Date), max(tide.13$Date)), 
	ylim=c(min(bn$Nest_height), max(bn$Nest_height)), ylab="Nest Height (m)", cex=1.5)
segments(x0=as.Date("2013-05-26"), x1=as.Date("2013-05-26"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2013-06-06"), x1=as.Date("2013-06-06"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2013-06-21"), x1=as.Date("2013-06-21"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2013-07-25"), x1=as.Date("2013-07-25"), y0=0.3, y1=1.15, col=gray(0.5))
legend("topleft", legend=c("Flood", "Pred", "Fledge"), col=c("dodgerblue4", "palegreen3", "gold3"), pch=c(0,1,17), cex=1.1)
par(mar=c(2, 1, 0, 0) + 0.1)
plot(x=bn.14$Fate_date, y=jitter(bn.14$Nest_height), col=ifelse(bn.14$Fate==1, "dodgerblue4", ifelse(bn.14$Fate==3, "palegreen3", "gold3")),
	pch=ifelse(bn.14$Fate==1, 0, ifelse(bn.14$Fate==3, 1, 17)), xlim=c(min(tide.14$Date), max(tide.14$Date)), 
	ylim=c(min(bn$Nest_height), max(bn$Nest_height)), ylab="Nest Height (m)", yaxt="n", cex=1.5)
segments(x0=as.Date("2014-04-19"), x1=as.Date("2014-04-19"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2014-04-29"), x1=as.Date("2014-04-29"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2014-05-15"), x1=as.Date("2014-05-15"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2014-06-16"), x1=as.Date("2014-06-16"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2014-07-12"), x1=as.Date("2014-07-12"), y0=0.3, y1=1.15, col=gray(0.5))
par(mar=c(2, 1, 0, 0) + 0.1)
plot(x=bn.15$Fate_date, y=jitter(bn.15$Nest_height), col=ifelse(bn.15$Fate==1, "dodgerblue4", ifelse(bn.15$Fate==3, "palegreen3", "gold3")),
	pch=ifelse(bn.15$Fate==1, 0, ifelse(bn.15$Fate==3, 1, 17)), xlim=c(min(tide.15$Date), max(tide.15$Date)), 
	ylim=c(min(bn$Nest_height), max(bn$Nest_height)), ylab="Nest Height (m)", yaxt="n", cex=1.5)
segments(x0=as.Date("2015-04-19"), x1=as.Date("2015-04-19"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2015-05-16"), x1=as.Date("2015-05-16"), y0=0.3, y1=1.15, col=gray(0.5))

#Asides: try adding these other arguments to the plot() function
#To change orientation of figure labels:  las=2
#To put the tick marks inside the plot:   tcl=0.5
#To get rid of the bounding box:          bty="l"

#To put text in the figure:
text(x=as.Date("2015-07-25"), y=1.1, "a")

#Multipanel plotting using split.screen()
#Check out: http://seananderson.ca/courses/11-multipanel/multipanel.pdf

```

Now we want to write our figure directly to a vector graphics file that has the correct dimensions for a double column figure in a publication (no further editing required!).

```{r}
###
#Write figure to a pdf (or other format) with the correct dimensions for publication
#Using pdf() will not plot within RStudio, but directly write a file to your working directory
pdf("Figure2.pdf", width=7.6, height=3.8)
layout(matrix(c(1,2,3,4,5,6),2,3,byrow=TRUE), widths=c(lcm(7),lcm(6),lcm(6)), heights=c(lcm(4.7),lcm(4.7)))
par(mar=c(0, 4, 2, 0) + 0.1)
plot(tide.13$Water_Level ~ tide.13$Date, type="l", ylim=c(min(tide$Water_Level), max(tide$Water_Level)), ylab="Max Tidal Height (m)", lwd=1,
	xaxt="n", main="2013")
abline(1.5, 0, lty=2)
segments(x0=as.Date("2013-05-26"), x1=as.Date("2013-05-26"), y0=0.6, y1=1.595, col=gray(0.5))
segments(x0=as.Date("2013-06-06"), x1=as.Date("2013-06-06"), y0=0.6, y1=1.344, col=gray(0.5))
segments(x0=as.Date("2013-06-21"), x1=as.Date("2013-06-21"), y0=0.6, y1=1.638, col=gray(0.5))
segments(x0=as.Date("2013-07-25"), x1=as.Date("2013-07-25"), y0=0.6, y1=1.566, col=gray(0.5))
par(mar=c(0, 1, 2, 0) + 0.1)
plot(tide.14$Water_Level ~ tide.14$Date, type="l", ylim=c(min(tide$Water_Level), max(tide$Water_Level)), ylab="Max Tidal Height (m)", lwd=1,
	xaxt="n", yaxt="n", main="2014")
abline(1.5, 0, lty=2)
segments(x0=as.Date("2014-04-19"), x1=as.Date("2014-04-19"), y0=0.6, y1=1.579, col=gray(0.5))
segments(x0=as.Date("2014-04-29"), x1=as.Date("2014-04-29"), y0=0.6, y1=1.527, col=gray(0.5))
segments(x0=as.Date("2014-05-15"), x1=as.Date("2014-05-15"), y0=0.6, y1=1.526, col=gray(0.5))
segments(x0=as.Date("2014-06-16"), x1=as.Date("2014-06-16"), y0=0.6, y1=1.502, col=gray(0.5))
segments(x0=as.Date("2014-07-12"), x1=as.Date("2014-07-12"), y0=0.6, y1=1.406, col=gray(0.5))
par(mar=c(0, 1, 2, 0) + 0.1)
plot(tide.15$Water_Level ~ tide.15$Date, type="l", ylim=c(min(tide$Water_Level), max(tide$Water_Level)), ylab="Max Tidal Height (m)", lwd=1,
	xaxt="n", yaxt="n", main="2015")
abline(1.5, 0, lty=2)
segments(x0=as.Date("2015-04-19"), x1=as.Date("2015-04-19"), y0=0.6, y1=1.526, col=gray(0.5))
segments(x0=as.Date("2015-05-16"), x1=as.Date("2015-05-16"), y0=0.6, y1=1.491, col=gray(0.5))
par(mar=c(2, 4, 0, 0) + 0.1)
plot(x=bn.13$Fate_date, y=jitter(bn.13$Nest_height), col=ifelse(bn.13$Fate==1, "dodgerblue4", ifelse(bn.13$Fate==3, "palegreen3", "gold3")),
	pch=ifelse(bn.13$Fate==1, 0, ifelse(bn.13$Fate==3, 1, 17)), xlim=c(min(tide.13$Date), max(tide.13$Date)), 
	ylim=c(min(bn$Nest_height), max(bn$Nest_height)), ylab="Nest Height (m)", cex=1.5)
segments(x0=as.Date("2013-05-26"), x1=as.Date("2013-05-26"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2013-06-06"), x1=as.Date("2013-06-06"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2013-06-21"), x1=as.Date("2013-06-21"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2013-07-25"), x1=as.Date("2013-07-25"), y0=0.3, y1=1.15, col=gray(0.5))
legend("topleft", legend=c("Flood", "Pred", "Fledge"), col=c("dodgerblue4", "palegreen3", "gold3"), pch=c(0,1,17), cex=1.1)
par(mar=c(2, 1, 0, 0) + 0.1)
plot(x=bn.14$Fate_date, y=jitter(bn.14$Nest_height), col=ifelse(bn.14$Fate==1, "dodgerblue4", ifelse(bn.14$Fate==3, "palegreen3", "gold3")),
	pch=ifelse(bn.14$Fate==1, 0, ifelse(bn.14$Fate==3, 1, 17)), xlim=c(min(tide.14$Date), max(tide.14$Date)), 
	ylim=c(min(bn$Nest_height), max(bn$Nest_height)), ylab="Nest Height (m)", yaxt="n", cex=1.5)
segments(x0=as.Date("2014-04-19"), x1=as.Date("2014-04-19"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2014-04-29"), x1=as.Date("2014-04-29"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2014-05-15"), x1=as.Date("2014-05-15"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2014-06-16"), x1=as.Date("2014-06-16"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2014-07-12"), x1=as.Date("2014-07-12"), y0=0.3, y1=1.15, col=gray(0.5))
par(mar=c(2, 1, 0, 0) + 0.1)
plot(x=bn.15$Fate_date, y=jitter(bn.15$Nest_height), col=ifelse(bn.15$Fate==1, "dodgerblue4", ifelse(bn.15$Fate==3, "palegreen3", "gold3")),
	pch=ifelse(bn.15$Fate==1, 0, ifelse(bn.15$Fate==3, 1, 17)), xlim=c(min(tide.15$Date), max(tide.15$Date)), 
	ylim=c(min(bn$Nest_height), max(bn$Nest_height)), ylab="Nest Height (m)", yaxt="n", cex=1.5)
segments(x0=as.Date("2015-04-19"), x1=as.Date("2015-04-19"), y0=0.3, y1=1.15, col=gray(0.5))
segments(x0=as.Date("2015-05-16"), x1=as.Date("2015-05-16"), y0=0.3, y1=1.15, col=gray(0.5))
dev.off()
```

