---
title: "Week 2, Lecture 1: Working with Functions"
subtitle: "Instructor notes -- Part 1: prep reading, Part 2: podium cheat sheet"
fontsize: 12pt
output:
  pdf_document: default
header-includes:
  - \usepackage{enumitem}
  - \setlist[itemize]{itemsep=4pt, parsep=0pt, topsep=4pt}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 5, fig.height = 3.5)
```

This single file has two parts, both instructor-only (not a student
handout). **Part 1** is read beforehand -- full worked examples and the
reasoning behind the flow. **Part 2** (after the page break, large
print) is the podium copy.

Source material: Bolker Ch. 3, sections 3.2-3.4 ("Deterministic
functions for ecological modeling") almost verbatim in places -- this
lecture leans harder on the assigned reading than Week 1's did, because
today's content is a hard prerequisite for a specific Lab 2 question
(see below). Also mined: this morning's WOD
(`wod/WOD_Week2_Lecture1_RSyntaxAndNotation.Rmd`), Lab 1 Act 2 (Ricker
derivative), and Lab 2 Acts 1-2 (grid search, flexible Ricker, Holling
IV).

# Where this sits in the week, and a hard dependency

- Monday (this lecture): properties of deterministic functions --
  limits, derivatives, critical points -- as tools, not yet the
  bestiary itself (that's Wednesday).
- Monday AM (WOD, already done): bug-hunts in R function syntax
  (argument order/matching) plus deriving and coding a seasonal sine
  function -- both are direct rehearsals for today.
- Tuesday (Lab 2): Act 1 is a grid-search optimizer for the Ricker
  function (algorithms, not functions per se -- a callback to Week 1).
  **Act 2 explicitly instructs students to use "the same 'throw out the
  lower-order terms' trick from Monday's lecture"** to find the
  limiting behavior of the Holling type IV function as $x\to\infty$.
  That phrase is not a coincidence -- it's this lecture. If Section 2
  below (limits) doesn't land clearly today, that Lab 2 question will
  stall.
- Wednesday: Bestiary of functions (Bolker 3.5) -- the specific named
  catalog (power, exponential, hyperbolic, sigmoid...). Today builds the
  toolkit; Wednesday populates it.

# Scientific workflow: the modeling loop (carried over from Week 1 Lecture 2)

This was meant to open Week 1 Lecture 2 but got skipped there for time
-- it belongs here instead, since it's foundational framing rather than
content specific to last lecture's algorithms-vs-closed-form material.

Worth being precise about where this comes from, since it's easy to
misremember as more textbook-sourced than it is. Bolker section 1.6,
"Outline of the modeling process" (Figure 1.5, printed p. 27 -- PDF page
35, remember the +8 front-matter offset), lays out an explicit flow:
*ask ecological questions $\to$ collect data $\to$ \{stochastic
model(s), deterministic model(s)\} $\to$ estimate parameters $\to$
estimate confidence regions $\to$ hypothesis testing/selection $\to$
answer questions*, looping back to the question step (with a second
dotted arrow: model selection can also loop back to the data-collection
step).

Two things about that figure shaped how it's adapted for the whiteboard:

- It genuinely starts with the *question*, not raw data. Bolker's own
  footnote on p. 26 says it outright: "In an ideal world, you would
  identify ecological questions before you designed your experiments
  and gathered data (!), but in this book I will assume you've already
  got data ... to work with." Data comes second in his figure only
  because the book assumes a dataset already in hand, not because data
  should conceptually precede the question.
- Given that, and given a preference for a Popperian/hypothetico-
  deductive framing over a data-first or strictly Bayesian one, the
  whiteboard loop is:
  $$\begin{aligned}
    \text{THEORY/QUESTION} \to \text{CANDIDATE MODEL(S)} \to
      \text{DATA (fit)} \\ \to \text{COMPARE} \to \text{repeat}
    \end{aligned}$$
  where "repeat" loops back to *revising or rejecting the theory*, not
  just re-tuning parameters. This is the same shape as Bolker's own loop
  closure (his step vi, "put results together / return to step \#1",
  returns to step i, "ask ecological questions") -- just read through a
  falsificationist lens instead of an inductive one, and arguably truer
  to his own stated ideal (question before data) than a literal reading
  of his figure would be.

**What "COMPARE" means, concretely:** this step does double duty --

1. *Goodness of fit:* does the model's prediction even resemble the
   data? (Lab 1's SSR calculation for the Ricker curve.)
2. *Comparison against rivals, including the null:* does this candidate
   earn its keep relative to a simpler competitor? (Lab 1 Act 3's
   Ricker-vs-null SSR comparison.)

In Popperian terms, this is the severe test: a model that survives
comparison against a null and against competing candidates is
*corroborated*, not proven; one that doesn't survive is *falsified* and
should be revised or discarded. This foreshadows the frequentist/
likelihood hypothesis-testing and model-selection machinery (LRT, AIC)
in Weeks 6-8, and previews Bolker's frequentist/likelihood/Bayesian tour
in section 1.4 (pp. 13-22) without going into it today.

On the whiteboard: sketch the loop, then move straight into today's
topic (reading a function itself) -- the analytical-vs-computational
dichotomy from last lecture doesn't need repeating here.

# Today's throughline (bridge from Week 1)

One-sentence bridge: last lecture was about *how* we get from a
candidate model to an answer (algorithms vs. closed form); today is
about *reading* a candidate function itself -- what it does at the
extremes, where it peaks, what its parameters mean -- before you ever
fit it to data.

Recall from Lab 1 Act 1: students already have an informal shape
vocabulary (monotonic increasing/decreasing, saturating, hump-shaped,
roughly flat) from eyeballing the tadpole data. Today's tools let you
*prove* a shape claim algebraically instead of just eyeballing a curve
-- the same distinction as Bolker 3.2/3.3 vs. 3.4: numerically
(plug in values, plot it, `curve()`) gives you a picture; analytically
(algebra and calculus) tells you *why* the picture looks that way, and
lets you reason about regions you didn't plot.

# Limits: throw out the lower-order terms (the required section)

This is Bolker's own phrasing (3.4.1), worth keeping close to verbatim
since Lab 2 Act 2 quotes it back: *"The basic principle is to throw out
lower-order terms. As $x$ grows, it will eventually grow much larger
than the largest constant term in the equation. Terms with larger
powers of $x$ will dwarf smaller powers, and exponentials will dwarf
any power. If $x$ is very small then you apply the same logic in
reverse; constants are bigger than (positive) powers of $x$, and
negative powers... are bigger than any constants."*

In short, a hierarchy of "who wins" as $x \to \infty$:
$$\text{exponential} \;\gg\; x^{\text{large power}} \;\gg\; x^{\text{small power}} \;\gg\; \text{constant}$$
and the *same* hierarchy reverses as $x \to 0$: constants beat positive
powers of $x$, and negative powers of $x$ (i.e. $1/x$, $1/x^2$, ...)
beat everything.

**Worked example (Bolker's own -- Michaelis-Menten), do this on the
board:** $f(x) = \dfrac{ax}{b+x}$

- As $x \to \infty$: $x \gg b$, so $b + x \approx x$, so
  $f(x) \approx ax/x = a$ -- the curve levels off at the constant $a$
  (a horizontal asymptote).
- As $x \to 0$: $b \gg x$, so $b + x \approx b$, so
  $f(x) \approx ax/b$ -- a straight line through the origin with slope
  $a/b$.

```{r mm-plot, echo=FALSE, fig.height=3}
a <- 10; b <- 5
curve(a * x / (b + x), from = 0, to = 40, lwd = 2, col = "steelblue",
      xlab = "x", ylab = "f(x)", ylim = c(0, 11.5))
abline(h = a, lty = 2, col = "grey50")
abline(v = b, lty = 3, col = "grey50")
points(b, a/2, pch = 19)
text(b, a/2, "  half-max at x = b", adj = c(0, 1.5))
text(20, a, "asymptote = a", adj = c(0, -0.5), col = "grey30")
```

**Sticking point to flag:** students will want to "plug in infinity."
Head that off explicitly -- this is qualitative dominance reasoning
(which term is negligible), not literal substitution. Also flag that
the *same* parameter can play opposite roles at the two ends: $b$ is
negligible as $x\to\infty$ but dominant as $x \to 0$.

**Fallback tool (mention, don't dwell):** for fraction limits where
both numerator and denominator vanish or blow up together (an
indeterminate $0/0$ or $\infty/\infty$ form, so the dominance trick
doesn't immediately apply), L'Hopital's rule says
$\lim \dfrac{a(x)}{b(x)} = \lim \dfrac{a'(x)}{b'(x)}$ -- take the
derivative of the top and bottom separately, not the derivative of the
whole fraction. One line, not a worked example today.

**Do not solve Holling IV's limit here.** $y = ax^2/(b+cx+x^2)$ is
literally Lab 2 Act 2's question 1c -- comparing the highest powers of
$x$ in numerator ($x^2$) and denominator ($x^2$) is the same dominance
trick, one level harder (both leading terms are degree 2, so the
constant $a$ from the ratio of leading coefficients survives). Let
students discover that tomorrow; today's example (Michaelis-Menten) is
deliberately a different, simpler rational function so the technique
transfers without the answer being pre-given.

# Derivatives and critical points (recap + formalize)

Students already computed one derivative and one critical point by hand
in Lab 1 Act 2 -- this section names what they did, generalizes it, and
contrasts it with a monotonic function that has no peak at all.

**Recap, don't re-derive from scratch:** the Ricker function
$y = a x e^{-bx}$ (Lab 1's running example: $a = 0.9$, $b = 1/12$).
Product rule:
$$\frac{dy}{dx} = a e^{-bx} + ax(-be^{-bx}) = a(1-bx)e^{-bx}$$
Setting this to zero: since $ae^{-bx} \neq 0$, the only way the product
is zero is $1 - bx = 0$, i.e. $x^* = 1/b$.

```{r ricker-peak}
a_r <- 0.9; b_r <- 1/12
x_star <- 1 / b_r
peak_height <- a_r * x_star * exp(-b_r * x_star)
x_star
peak_height
```

That's $x^* = `r x_star`$ mm and peak height $`r round(peak_height, 2)`$
-- matching Lab 1 Act 3's table exactly (TBL = 12, Ricker predicted =
3.97). This is the payoff of the "critical point" concept: it's not
new math, it's a name for exactly the calculation they already did.

**Formalize:** a critical point is any $x$ where $f'(x) = 0$ (or is
undefined) -- a candidate for a peak, a trough, or (at a saddle point) a
flat spot that isn't either. To classify which, plug the critical
$x$-value back into $f$ itself (peak/trough height) and reason about
the sign of $f'$ on either side (or use the second derivative).

**Contrast with a monotonic function:** try the same exercise on
Michaelis-Menten, $f(x) = ax/(b+x)$. Its derivative (quotient rule) is
$f'(x) = ab/(b+x)^2$ -- always positive (a sum of squares in the
denominator can't be negative), so the function is *monotonic
increasing* everywhere and *never* has a critical point. That's exactly
why it saturates toward an asymptote instead of peaking and coming back
down -- contrast this shape against Ricker's hump directly, on the
board, side by side.

**Sticking point to flag:** "critical point" (where $f'(x)=0$) is easy
to confuse with "root" (where $f(x)=0$). The Ricker function has both:
it's zero at $x=0$ (a root) and peaks at $x=1/b$ (a critical point) --
two different questions with two different answers from the same
function.

# R convention: `x` first, then parameters (with defaults)

Directly from Bolker's R supplement (3.3.1, p. 104): *"When you write
such an R function, you would typically make the input value ($x$) be
the first argument, followed by all of the other parameters. It often
saves time if you assign default values to the other parameters."*

This is exactly why `ricker <- function(x, a, b)` is the convention
used everywhere in the labs (and why WOD 1a's `function(a, b, x)` was
buggy this morning -- it broke that convention on purpose). Point this
out explicitly: the WOD wasn't testing an arbitrary rule, it was testing
*this* rule, which is about to matter again.

**Why it matters going forward, concretely:** `sapply(xvec, ricker)`
only works cleanly if `x` is the first positional argument -- `sapply`
always feeds the vector element into the first open slot. Also
preview: Lab 2 Act 2 asks students to write `flex_ricker` and
`holling4` "following the same pattern as the `ricker` function (`x`
first, then parameters)" -- today is where that instruction gets its
justification, not just its statement.

**Scale parameters, briefly (Bolker 3.4.2):** when a parameter appears
as $bx$ or $x/c$, its job is to stretch or shrink the curve along the
$x$-axis without changing its shape -- Bolker calls this a
*characteristic scale*. $b$ has units of inverse-$x$; $c = 1/b$ has the
same units as $x$ and is often easier to interpret (e.g. "the peak is
at $x=12$" instead of "$b = 1/12$"). This is exactly what Lab 2 Act 2's
flexible Ricker does: it reparameterizes $b$ as $p$ (the peak location
directly) precisely so the parameter is easier to interpret and to set.
One sentence today is enough -- Act 2 does the work of actually
reparameterizing.

# Take-home message

- Two ways to know a function: numerically (plot it, plug in numbers)
  and analytically (limits, derivatives) -- today was almost entirely
  the analytical toolkit.
- Limits at the extremes: throw out lower-order terms. Same hierarchy,
  reversed at the two ends (exponential $\gg$ big power $\gg$ small
  power $\gg$ constant $\gg$ negative power, as $x\to\infty$; flip it
  for $x\to0$).
- Critical points ($f'(x)=0$) find peaks/troughs; monotonic functions
  (no sign change in $f'$) never have any.
- R convention (`x` first, params after, with defaults) isn't
  arbitrary style -- it's what makes `sapply()` and vectorized code work
  cleanly, and it's the same convention Lab 2 asks you to follow when
  writing new bestiary members from scratch.

# Roadmap tie-in

- Wednesday: the bestiary itself (Bolker 3.5) -- linear, power,
  exponential, hyperbolic/Michaelis-Menten, sigmoid families, each
  characterized using exactly today's tools (limits, critical points,
  scale parameters).
- Tuesday's Lab 2 Act 2 needs today's limit technique directly (Holling
  IV, question 1c) -- flagged above, worth a one-line reminder at the
  end of lecture.
- Later in the semester: "throw out lower-order terms" reasoning
  resurfaces informally whenever we ask what a fitted model predicts far
  outside the range of the data (extrapolation risk) -- worth a passing
  mention now, not a digression.

# Discussion seed (if time remains)

Ask: *pick any function from your own field that saturates (levels off)
as some predictor gets large. What's the mechanism that caps it -- what
in the real world plays the role of "$a$" (the asymptote) in
Michaelis-Menten?* Good prompt for surfacing intuition about *why*
saturation happens biologically, not just how to compute it. (If it
comes up: "$b$" is a different question -- not what caps the curve, but
how fast it gets there.)

\newpage
\large

# PART 2 -- PODIUM CHEAT SHEET

\normalsize
*Podium copy -- terse on purpose. Full derivations, rationale, and
code: see Part 1 above.*
\large

## 0. Open (~2 min)

- Last time: *how* we get an answer (algorithm vs. closed form)
- Today: *reading* a function itself -- extremes, peaks, parameters
- Recall Lab 1's shape vocabulary (monotonic / saturating / hump) --
  today we prove those claims, not just eyeball them

## 1. Scientific workflow: the modeling loop (~5 min)

- **Loop (Popperian, not data-first):**
  THEORY/QUESTION $\to$ CANDIDATE MODEL(S) $\to$ DATA + FIT $\to$
  COMPARE $\to$ repeat
- repeat = revise/reject the **theory**, not just re-tune parameters
- COMPARE = (1) fit ok? *and* (2) beats rivals/null?
- $\to$ survive = corroborated; fail = falsified
- cf. Bolker Fig. 1.5 (p.27) -- his loop too, question-first

\newpage

## 2. Two ways to know a function (~2 min)

- **Numerically:** plug in values, `curve()`, plot it -- a picture
- **Analytically:** limits + derivatives -- *why* the picture looks
  that way
- Today = almost entirely analytical

\newpage

## 3. Limits: throw out the lower-order terms (~10-12 min)

**The rule** (Bolker's own words -- write it up):

- as $x\to\infty$: exponential $\gg$ big power $\gg$ small power
  $\gg$ constant
- as $x\to0$: reverse it -- constant $\gg$ positive power $\gg$
  nothing beats a negative power ($1/x$, $1/x^2$, ...)

**Board example -- Michaelis-Menten:** $f(x) = \dfrac{ax}{b+x}$

```{r mm-podium-fig, echo=FALSE, fig.width=3.3, fig.height=2.3}
par(mar = c(3.5, 3, 1, 0.5))
curve((a * x) / (b + x), from = 0, to = 40, lwd = 2, col = "steelblue",
      xlab = "x", ylab = "f(x)")
abline(h = a, lty = 2, col = "grey50")
abline(v = b, lty = 3, col = "grey50")
```

- $x\to\infty$: $b+x\approx x \Rightarrow f(x)\approx ax/x=a$
  (asymptote)
- $x\to0$: $b+x\approx b \Rightarrow f(x)\approx ax/b$ (line, slope
  $a/b$)
- **Not** "plug in infinity" -- dominance reasoning, term by term
- $b$ is negligible at one end, dominant at the other -- same
  parameter, opposite roles
- Fallback only, don't dwell: L'Hopital's rule for $0/0$ or
  $\infty/\infty$ forms -- derivative of top over derivative of bottom
- **Do not solve Holling IV today** -- that's Lab 2 Act 2, question 1c,
  tomorrow. Michaelis-Menten teaches the same trick without giving away
  the answer.

\newpage

## 4. Derivatives and critical points (~7-8 min)

**Recap (already done in Lab 1 Act 2), don't re-derive:** Ricker
$y=axe^{-bx}$, product rule $\to$
$dy/dx = a(1-bx)e^{-bx}$

- critical point: $1-bx=0 \Rightarrow x^*=1/b$
- with $a=0.9,\ b=1/12$: $x^*=`r x_star`$, height
  $=`r round(peak_height,2)`$ -- matches Lab 1 Act 3's table exactly
- **critical point:** name for exactly what they already computed

**Contrast -- Michaelis-Menten's derivative** (quotient rule):
$f'(x)=ab/(b+x)^2$ -- always positive $\to$ monotonic, **no** critical
point, saturates instead of peaking

- **Sticking point:** critical point ($f'=0$) $\neq$ root ($f=0$) --
  Ricker has both, at different $x$ (root at $x=0$, peak at $x=1/b$)

## 5. R convention + scale parameters (~5 min)

- Bolker's own rule: `x` first argument, params after, with defaults
- *this is why* `function(a, b, x)` in this morning's WOD 1a was buggy
- matters for `sapply()`: vector element fills the *first* open slot
- **preview:** Lab 2 Act 2 asks you to write new bestiary functions
  "following the same pattern (`x` first, then parameters)"
- **scale parameter:** $bx$ or $x/c$ stretches/shrinks the $x$-axis;
  Lab 2's flexible Ricker reparameterizes $b\to p$ (peak location
  directly) for exactly this reason

## 6. Roadmap (~2 min)

- Wednesday: the bestiary itself (linear, power, exponential,
  hyperbolic, sigmoid) -- same tools, now cataloging specific families
- **Reminder before they leave:** Lab 2 Act 2 needs today's limit trick
  for Holling IV

## If time: discussion seed

- *Pick a function from your own field that saturates. What plays the
  role of "$a$" (the asymptote) -- what caps it, mechanistically?*
