---
title: "Lab 3, Act 2: Answer Key"
subtitle: "Counting detections: binomial or Poisson?"
author: "NRES 746, Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 6, fig.height = 4)
f4 <- function(v) formatC(v, format = "f", digits = 4)
```

Each item repeats the handout's question, then gives the answer, a
short rationale, and mistakes to watch for. Questions with one correct
answer are answered directly. Open-ended items (the "plausible or not"
part of 1a, 2e, and the group discussion) get sample answers or
discussion points instead. All numbers are computed in R when this file
is knitted.

Setting: the carnivore is confirmed present; a camera detects it on a
given night with probability $p = 0.3$, independently from night to
night, over $N = 8$ nights.

## Part 1: Binomial by hand

**1a.**

> Why is "number of nights with a detection, out of 8" a reasonable
> binomial random variable here? Name the assumptions the binomial
> distribution requires, and say briefly why each one seems plausible
> (or not) for this camera.

**Answer.** Four assumptions:

- **Fixed number of trials** ($N = 8$ nights): true by design.
- **Two outcomes per trial** (detection or not on each night): true by
  how the variable is defined.
- **Independent trials:** plausible but questionable. A storm lasting
  several nights, a camera malfunction, or an animal returning to a
  marked spot links one night to the next.
- **Constant $p$ across trials:** also questionable. Detection can
  change with moon phase, weather, season, or the animal getting used
  to (or wary of) the camera.

**Why.** The first two hold by construction, so the substance of a
good answer is in the last two.

**Watch for.** Listing the four assumptions with no judgment about
which are shaky. Confusing "constant $p$" with "$p$ is known."

**1b.**

> Using the binomial pmf, $P(X=k) = \binom{N}{k}p^k(1-p)^{N-k}$,
> compute by hand the probability of exactly 3 detections in the 8
> nights ($N=8$, $k=3$, $p=0.3$). Show your work.

**Answer.**
$P(X = 3) = \binom{8}{3}(0.3)^3(0.7)^5 = 56 \times 0.027 \times 0.16807 = `r f4(dbinom(3, 8, 0.3))`$.

```{r binom-hand}
choose(8, 3) * 0.3^3 * 0.7^5
dbinom(3, 8, 0.3)   # check
```

**Why.** $\binom{8}{3} = \frac{8!}{3!\,5!} = 56$ ways to choose which 3
nights had detections; each such sequence has probability
$0.3^3 \times 0.7^5$.

**Watch for.** Leaving out $\binom{8}{3}$ (gives 0.0045, the
probability of one particular sequence). Using $(0.7)^3(0.3)^5$.
Arithmetic slips in $8!/(3!\,5!)$.

**1c.**

> The mean of this distribution is $Np = 8 \times 0.3 = 2.4$, but the
> tallest bar (the mode) sits at a different whole number. Why can't
> the mode land exactly on the mean here? And what does the direction
> of the gap between them (mode below the mean, or above?) tell you
> about the shape of this distribution?

**Answer.** The mode is 2. It can't equal 2.4 because the number of
detections must be a whole number, while the mean is an average and
can fall between whole numbers. The mode sits below the mean, so the
distribution is right-skewed: it's bounded at 0 on the left and has a
longer tail toward higher counts, which pulls the mean up.

```{r mode-mean}
(0:8)[which.max(dbinom(0:8, 8, 0.3))]   # mode
8 * 0.3                                 # mean
```

**Why.** In a right-skewed distribution, the few large values in the
long tail pull the mean above the most common value.

**Watch for.** "Because the distribution is skewed" as the only answer
to the first question (the whole-number reason comes first). Reading
the skew backward (mode below mean means a right tail, not a left one).

## Part 2: Comparing to Poisson

**2a.**

> Look closely at both pairs of bar charts. In Scenario A, where
> specifically do the binomial and Poisson bars disagree most: the
> center, the spread, or the tails? Does Scenario B show the same kind
> of disagreement, a different kind, or none at all?

**Answer.** Scenario A: the spread. The binomial is taller in the
middle (2 to 4 detections); the Poisson puts more probability in both
tails (0 and 1 detections, and 5 or more). Scenario B: almost no
disagreement anywhere.

```{r compare-ab}
round(rbind(binom_A = dbinom(0:8, 8, 0.3),
            binom_B = dbinom(0:8, 100, 0.024),
            poisson = dpois(0:8, 2.4)), 3)
```

**Why.** The Poisson is wider than Binomial(8, 0.3) even though both
have mean 2.4 (2b shows why). Binomial(100, 0.024) has nearly the same
spread as the Poisson.

**Watch for.** "The Poisson has a fatter right tail" alone; it's also
higher at 0 and 1. "The center differs": both means are 2.4.

**2b.**

> Compute the variance of each distribution by hand, using
> $\text{Var} = Np(1-p)$ for the binomial and $\text{Var} = \lambda$
> for the Poisson.

**Answer.**

- Scenario A binomial: $8 \times 0.3 \times 0.7 = 1.68$.
- Scenario B binomial: $100 \times 0.024 \times 0.976 = 2.3424$.
- Poisson (both scenarios): $\lambda = 2.4$.

**Why.** Direct substitution. The binomial variance equals the Poisson
variance times $(1 - p)$, so the two match only when $1 - p$ is close
to 1.

**Watch for.** Using $p$ instead of $1 - p$ in the binomial formula.
Computing only Scenario A.

**2c.**

> Does the pattern of variances you just computed explain what you saw
> by eye in 2a?

**Answer.** Yes. In Scenario A the Poisson's variance (2.4) is about
40% larger than the binomial's (1.68), matching its wider spread. In
Scenario B the variances are nearly equal (2.40 vs. 2.34), matching
the near-identical bars.

**Why.** With the same mean, the variance is what controls the width.

**Watch for.** "Yes" with no reference to the numbers.

**2d.**

> Finish this rule of thumb in your own words: *"A Poisson distribution
> is a good stand-in for a binomial distribution when..."*

**Answer.** "...the number of trials is large and the probability per
trial is small," so that $1 - p \approx 1$ and the binomial variance
$Np(1-p)$ is close to the Poisson's $\lambda = Np$.

**Why.** Scenario B (large $N$, small $p$) matched; Scenario A (small
$N$, $p = 0.3$) didn't.

**Watch for.** "When the means are equal" (they were equal in both
scenarios). "Always" based on Scenario B alone. Mentioning only large
$N$ or only small $p$; both are needed.

**2e.**

> Back in the original 8-night scenario ($p = 0.3$): name one realistic
> reason the constant, independent per-night detection probability
> assumption from 1a might be violated in the field.

**Sample answer.** Any one realistic mechanism, for example:

- A multi-night storm suppresses detections on several nights in a row
  (breaks independence).
- Moon phase or season changes the animal's activity (breaks constant
  $p$).
- The animal learns to avoid the camera, or is drawn to the site by
  scent (changes $p$ over the deployment).

**Why.** Each of these makes a night's detection depend on something
other than a single fixed $p$.

**Watch for.** Reasons that change whether the animal is present (the
Act assumes it is). Restating the assumption without a mechanism.

## Group discussion and revision

> Compare your variances and rule of thumb from 2d with your group.
> Record any changes, new insights, and muddiest points.

**Discussion points.** No fixed answer. Useful reflections connect the
variance formulas in 2b to the picture in 2a, or note why matching
means alone doesn't make two distributions interchangeable.
