---
title: "Lab 3, Act 1 -- Answer Key"
subtitle: "Trusting a positive detection"
author: "NRES 746 -- Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 6, fig.height = 4.5)

p_S  <- 1e-4   # prevalence
sens <- 0.98   # sensitivity
fpr  <- 0.01   # false positive rate
```

This key follows the same section structure as the student handout. Questions with a single
correct numeric or derived answer are answered directly; open-ended items (interpretation,
discussion) are given sample answers/discussion points rather than a fixed "correct" response --
use these to gauge whether a student's reasoning is on track, not as an exact-match rubric.
Act 1 is deliberately discovery-based (no formula is given up front), so expect more variation
in *how* students get to a correct answer than in later Acts.

## Part 1: Count your way to an answer

**1a.** Completed table ($N=100{,}000$), rounded to the nearest whole number:

```{r freq-table}
N <- 1e5
species     <- N * p_S
not_species <- N * (1 - p_S)
flag_S      <- species * sens
flag_notS   <- not_species * fpr
data.frame(
  row         = c("Actually the species", "Not the species", "Total"),
  flagged     = round(c(flag_S, flag_notS, flag_S + flag_notS)),
  not_flagged = round(c(species - flag_S, not_species - flag_notS,
                         (species - flag_S) + (not_species - flag_notS))),
  total       = c(species, not_species, N)
)
```

**1b.** PPV $=10/1010\approx0.0099\approx1\%$ -- pure counting, no formula needed.

**1c.** No fixed answer, but a correct response should register that $\approx1\%$ is
*surprisingly low* given how good 98% sensitivity / 1% false-positive rate sounded at the
start. A good response to the field team: a single flagged image alone isn't enough to act on
confidently -- look for corroborating evidence (a second independent detection, a track, scat)
before committing resources to that flag.

## Part 2: Turn your counting into a general recipe

**2a.** Flagged-and-truly-the-species count $=N\times\text{prevalence}\times\text{sensitivity}$.
Total flagged count $=N\times\text{prevalence}\times\text{sensitivity} +
N\times(1-\text{prevalence})\times\text{false positive rate}$.

**2b.** $N$ cancels from top and bottom, leaving:

```{r ppv-formula}
ppv <- (p_S * sens) / (p_S * sens + (1 - p_S) * fpr)
ppv
```

$$\text{PPV} = \frac{\text{prevalence}\times\text{sensitivity}}
{\text{prevalence}\times\text{sensitivity} + (1-\text{prevalence})\times\text{false positive rate}}
\approx 0.0097$$

**2c.** Any phrasing that captures the formula above is correct -- word-for-word notation
doesn't matter. This is Bayes' rule/PPV, arrived at without a name; don't volunteer the name
"Bayes' rule" when grading this -- that reveal is Wednesday's lecture.

**Common snags:** students who divide by $P(S)$ (the marginal probability of being the
species) instead of $P(F)$ (the marginal probability of being flagged) in 2b -- this produces a
PPV $>0.5$ at Site A, which is the tell. Students who try to cancel something other than $N$ in
2b, or who stall because they're trying to recall "the formula" instead of generalizing their
own Part 1 arithmetic.

## Part 3: Try it somewhere else

**3a.** Site B (prevalence $=0.01$, same sensitivity/false-positive rate):

```{r site-b}
p_S_B <- 0.01
ppv_B <- (p_S_B * sens) / (p_S_B * sens + (1 - p_S_B) * fpr)
ppv_B
```

PPV $\approx0.497\approx50\%$.

**3b.** No fixed answer, but a strong response identifies that the classifier's own properties
(sensitivity, false-positive rate) were *identical* at both sites -- only prevalence changed,
and yet PPV swung from $\approx1\%$ to $\approx50\%$. The intended takeaway: prevalence (the
"base rate") matters at least as much as the classifier's own accuracy, and a classifier that
sounds great in isolation can still be dominated by false positives when the target is rare.
Students who bring up the "base rate fallacy" by name (even loosely) should get full credit for
connecting today's exercise to that idea -- the formal term arrives Wednesday.

## Group discussion and revision

No fixed answer -- the point is for students to compare their Part 2 recipes and reconcile any
differences in notation or approach, and to flag their own muddiest points.
