---
title: "Lab 3, Act 1: Answer Key"
subtitle: "Trusting a positive detection"
author: "NRES 746, Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 6, fig.height = 4)

p_s <- 1e-4    # prevalence
sens <- 0.98   # sensitivity
fpr <- 0.01    # false positive rate
n_events <- 1e5

flag_s <- n_events * p_s * sens
flag_not <- n_events * (1 - p_s) * fpr
ppv_count <- round(flag_s) / (round(flag_s) + round(flag_not))
ppv_formula <- (p_s * sens) / (p_s * sens + (1 - p_s) * fpr)
p_s_b <- 0.01
ppv_b <- (p_s_b * sens) / (p_s_b * sens + (1 - p_s_b) * fpr)

f3 <- function(v, d = 3) formatC(v, format = "f", digits = d)
```

Each item repeats the handout's question, then gives the answer, a
short rationale, and mistakes to watch for. Questions with one correct
answer are answered directly. Open-ended items (1c, 2c, 3b, and the
group discussion) get sample answers or discussion points instead. All
numbers are computed in R when this file is knitted.

This Act is discovery-based (no formula is given up front), so expect
more variation in *how* students reach a correct answer than in later
Acts. Rates used throughout: prevalence = 0.0001, sensitivity = 0.98,
false positive rate = 0.01.

## Part 1: Count your way to an answer

**1a.**

> Suppose the cameras log $N = 100{,}000$ trigger events over a season.
> Fill in the table below using only the rates given above. Every blank
> cell is just one of those rates applied to a total you already know.
> Round to the nearest whole number.

**Answer.**

```{r freq-table, echo=FALSE}
species <- n_events * p_s
not_species <- n_events * (1 - p_s)
tab <- data.frame(
  row = c("Actually the species", "Not the species", "Total"),
  flagged = round(c(flag_s, flag_not, flag_s + flag_not)),
  not_flagged = round(c(species - flag_s, not_species - flag_not,
                        n_events - flag_s - flag_not)),
  total = c(species, not_species, n_events)
)
knitr::kable(tab, col.names = c("", "Flagged", "Not flagged", "Total"),
             format.args = list(big.mark = ","), align = "lccc")
```

**Why.** Flagged species events = 10 $\times$ 0.98 = 9.8 $\approx$ 10.
Flagged non-species events = 99,990 $\times$ 0.01 = 999.9 $\approx$
1,000. The "not flagged" cells are what's left in each row.

**Watch for.** Applying the false positive rate to the 10 species
events, or the sensitivity to the 99,990 non-species events. Rows that
don't add up to their totals.

**1b.**

> Of the images the classifier *flagged*, what fraction are actually the
> target species? No formula needed; just use the counts in your
> completed table.

**Answer.** $10 / 1{,}010 \approx `r f3(ppv_count, 4)`$, about 1%.

**Why.** The denominator is everything in the "Flagged" column (true
and false alarms together); the numerator is the true detections in
that column.

**Watch for.** Dividing by 10 (the species total) instead of by 1,010
(the flagged total), which gives 0.98, the sensitivity. That's the
central confusion of this Act: $P(\text{flagged} \mid \text{species})$
is not $P(\text{species} \mid \text{flagged})$.

**1c.**

> Is this number close to what you expected, given how good the
> classifier's sensitivity and false-positive rate sounded at the
> start? In a sentence or two, what would you tell the field team about
> how to handle a single flagged image?

**Sample answer.** No, about 1% is far lower than the 98% and 1% rates
suggest. A single flagged image is almost always a false alarm, so the
team should look for confirmation (a second independent detection,
tracks, scat, or a person checking the image) before acting on it.

**Why.** False alarms come from the huge pool of non-species events;
even a 1% false positive rate applied to 99,990 events swamps the 10
real ones.

**Watch for.** Answers that call the classifier "bad." Its error rates
are fine; the problem is how rare the species is. Saying so here sets
up 3b.

## Part 2: Turn your counting into a general recipe

**2a.**

> Rewrite the two counts you used in 1b (the number of flagged images
> that are truly the species, and the total number of flagged images)
> as expressions using $N$ and the rates given at the top (prevalence,
> sensitivity, false positive rate), instead of plugged-in numbers.

**Answer.**

- Flagged and truly the species: $N \times \text{prevalence} \times \text{sensitivity}$.
- All flagged: $N \times \text{prevalence} \times \text{sensitivity} + N \times (1 - \text{prevalence}) \times \text{false positive rate}$.

**Why.** Each count is "how many events are in that row" times "the
chance an event in that row gets flagged," as in the table.

**Watch for.** Leaving out the $(1 - \text{prevalence})$ factor, i.e.
applying the false positive rate to all $N$ events. With prevalence this
small the number barely changes, so ask the student which events can
produce a false alarm.

**2b.**

> Divide your two expressions from 2a to reconstruct the fraction from
> 1b. $N$ should appear in both the top and bottom; cancel it out. What
> are you left with?

**Answer.**
$$\text{PPV} = \frac{\text{prevalence} \times \text{sensitivity}}
{\text{prevalence} \times \text{sensitivity} + (1 - \text{prevalence}) \times \text{false positive rate}}
= `r f3(ppv_formula, 4)`$$

**Why.** $N$ multiplies every term, so it cancels. The value differs
slightly from 1b (`r f3(ppv_count, 4)`) only because 1a rounded 9.8 and
999.9 to whole numbers.

**Watch for.** Students who think the 0.0097 vs. 0.0099 difference
means an error. Cancelling a rate instead of $N$. Stalling while trying
to recall "the formula" instead of generalizing their own Part 1
arithmetic.

**2c.**

> State your result from 2b as a general rule, in your own words or
> symbols: given a classifier's sensitivity, false-positive rate, and
> the prevalence of the thing it's looking for, how do you compute the
> chance that a flagged event is real?

**Sample answer.** "The chance a flag is real equals the true flags
divided by all flags, where true flags = prevalence $\times$
sensitivity and false flags = (1 $-$ prevalence) $\times$ false
positive rate." Any wording or notation that captures the 2b formula is
correct.

**Why.** This is Bayes' rule (the positive predictive value), which
students built here before seeing its name in lecture.

**Watch for.** A rule that restates the 1b numbers instead of
generalizing them.

## Part 3: Try it somewhere else

**3a.**

> A second survey site uses the same classifier, but prevalence is
> 0.01 (1 in 100 trigger events). Using your recipe from Part 2 (or a
> fresh counting table), find this second site's PPV.

**Answer.** PPV $= \dfrac{0.01 \times 0.98}{0.01 \times 0.98 + 0.99 \times 0.01} = `r f3(ppv_b, 3)`$, about 50%.

**Why.** Prevalence is 100 times higher, so true flags (0.0098) are now
about as common as false flags (0.0099).

**Watch for.** Changing the sensitivity or false positive rate too (the
handout says they're the same). Students who use a counting table get
the same answer (980 true flags vs. 990 false flags out of 100,000).

**3b.**

> Same classifier, same sensitivity, same false-positive rate, but a
> different answer. What does comparing these two sites tell you about
> how much those two classifier properties matter versus how common
> (prevalent) the thing you're looking for actually is? Feel free to
> use the term "base rate" and speculate on how your findings relate to
> the "base rate fallacy."

**Sample answer.** The classifier's properties didn't change, yet the
chance a flag is real went from about 1% to about 50%. Prevalence (the
base rate) matters as much as the classifier's accuracy. The base rate
fallacy is judging a positive result by the test's accuracy alone and
ignoring how rare the target is.

**Why.** When the target is rare, almost every event is a non-target
event, so even a small false positive rate produces far more false
flags than true ones.

**Watch for.** Concluding that the classifier "works better" at the
second site. Give full credit for any reasonable connection to the base
rate fallacy, even if loosely stated.

## Group discussion and revision

> Compare your Part 2 recipe and your Part 3 answer with your group.
> Does everyone's recipe agree once you strip away notation
> differences? If not, find where they diverge. Record any changes,
> new insights, and muddiest points.

**Discussion points.** No fixed answer. Look for students who found the
sensitivity-vs-PPV confusion in their own or a groupmate's work, or
who reconciled different notations for the same recipe.
