---
title: "Lab 3, Act 3: Answer Key"
subtitle: "From counts to a continuous measurement"
author: "NRES 746, Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 5.5, fig.height = 3)

mu <- 4.5
sigma <- 0.6
f4 <- function(v) formatC(v, format = "f", digits = 4)
f2 <- function(v) formatC(v, format = "f", digits = 2)
```

Each item repeats the handout's question, then gives the answer, a
short rationale, and mistakes to watch for. Questions with one correct
answer are answered directly. Open-ended items (1d and the group
discussion) get sample answers or discussion points instead. All
numbers are computed in R when this file is knitted.

Setting: adult body mass is normal with mean $\mu = 4.5$ kg and
standard deviation $\sigma = 0.6$ kg.

## Part 1: Reading a density curve

**1a.**

> **Shade it by hand.** On the curve above, shade the region that
> represents $P(4.5 < \text{mass} < 5.5)$.

**Answer.** The area under the curve between 4.5 and 5.5 kg, from the
x-axis up to the curve:

```{r shade-answer, echo=FALSE}
par(mar = c(4, 4, 0.5, 0.5))
curve(dnorm(x, mu, sigma), from = 2.4, to = 6.6, lwd = 2,
      xlab = "Body mass (kg)", ylab = "Density", xaxs = "i")
xs <- seq(4.5, 5.5, length.out = 100)
polygon(c(4.5, xs, 5.5), c(0, dnorm(xs, mu, sigma), 0), col = "grey70", border = NA)
curve(dnorm(x, mu, sigma), add = TRUE, lwd = 2)
```

**Why.** For a continuous variable, probability is area under the
density over an interval. (This area is
`r f4(pnorm(5.5, mu, sigma) - 0.5)`.)

**Watch for.** Shading a thin strip at a single value, shading the
space above the curve, or shading everything to the right of 4.5.

**1b.**

> **Integral notation.** Given that the probability density function
> can be written $f(x)$, please write out the same region described
> above as a definite integral.

**Answer.**
$$P(4.5 < \text{mass} < 5.5) = \int_{4.5}^{5.5} f(x)\,dx$$

**Why.** A definite integral is the area under $f(x)$ between its
limits, which is what 1a shaded.

**Watch for.** Missing $dx$, limits reversed, or writing $f(5.5) -
f(4.5)$ (a difference of heights, not an area).

**1c.**

> What is $P(\text{mass} = 4.5)$ exactly? Why? What's different about a
> single point on a continuous curve versus an interval?

**Answer.** 0. A single point has no width, so the area under the
curve above it is zero, however tall the curve is there. Probability
for a continuous variable only builds up over an interval.

**Why.** $\int_{4.5}^{4.5} f(x)\,dx = 0$.

**Watch for.** Reading the curve's height at 4.5 (about
`r f2(dnorm(4.5, mu, sigma))`) as the probability. "Very small but not
zero."

**1d.**

> The height of the curve at $\text{mass} = 4.5$ is not zero (in fact,
> it's the highest point on the whole curve). If that height isn't a
> probability, what does it actually represent? Try to write out in
> words what a "probability density" means.

**Sample answer.** It's probability per unit of mass: how concentrated
the probability is near 4.5 kg. Multiplying the height by a small width
gives the approximate probability of landing in that small interval
(e.g. $0.665 \times 0.1 \approx 0.07$ for 4.45 to 4.55 kg).

**Why.** Density has units of 1/kg here, which is why it only becomes a
probability once multiplied by a width (or integrated over an
interval).

**Watch for.** "The most likely value" with no explanation of what the
height measures. Claims that a density can't exceed 1 (it can when
$\sigma$ is small).

## Part 2: Reading a CDF

**2a.**

> Using the CDF plot, estimate $P(\text{mass} > 5.5)$. You don't need
> to be precise, just in the right neighborhood.

**Answer.** About 0.05. The CDF at 5.5 reads about 0.95, and
$P(\text{mass} > 5.5) = 1 - F(5.5)$. Any estimate from about 0.03 to
0.08 is fine.

**Why.** The CDF gives $P(\text{mass} \le x)$; the probability above
$x$ is what's left over.

**Watch for.** Reporting $F(5.5) \approx 0.95$ itself as the answer
(forgetting to subtract from 1).

## Part 3: Standardize and look it up

**3a.**

> Compute the z-score for $x = 5.5$ kg.

**Answer.** $z = (5.5 - 4.5)/0.6 = `r f2((5.5 - mu) / sigma)`$
(1.667 unrounded).

**Why.** Subtract the mean, divide by the standard deviation.

**Watch for.** Dividing by the variance (0.36) instead of the standard
deviation. Computing $(4.5 - 5.5)/0.6$, which gives the right size
with the wrong sign.

**3b.**

> Look up $\Phi(z)$ for the z-score you just computed (round $z$ to the
> nearest 0.01 if needed), then use it to compute $P(\text{mass} > 5.5)$.
> How close was this to your graph-reading estimate from 2a?

**Answer.** $\Phi(1.67) = `r f4(pnorm(1.67))`$, so
$P(\text{mass} > 5.5) = 1 - `r f4(pnorm(1.67))` = `r f4(1 - pnorm(1.67))`$,
about 4.7%. It should be close to the 2a estimate.

```{r lookup}
1 - pnorm(1.67)            # table value
1 - pnorm(5.5, mu, sigma)  # unrounded z, for comparison
```

**Why.** The table gives the area below $z$; the probability above is
1 minus that. The small difference from the unrounded answer
(`r f4(1 - pnorm(5.5, mu, sigma))`) comes from rounding $z$ to 1.67.

**Watch for.** Reporting $\Phi(1.67)$ itself. Looking up the wrong cell
(row 1.6, column .07 is the right one).

**3c.**

> Search the table for the value closest to 0.90 and read off the
> z-score (90% quantile). Convert that z-score back into a body mass.

**Answer.** The closest table value is $\Phi(1.28) = `r f4(pnorm(1.28))`$,
so $z \approx 1.28$, and mass $= 4.5 + 1.28 \times 0.6 \approx
`r f2(mu + 1.28 * sigma)`$ kg. About 10% of adults are expected to
weigh more than this.

```{r reverse-lookup}
pnorm(c(1.28, 1.29))     # 0.8997 is closer to 0.90 than 0.9015
mu + 1.28 * sigma
qnorm(0.9, mu, sigma)    # exact, for comparison
```

**Why.** Reversing the z-score formula: $x = \mu + z\sigma$.

**Watch for.** Reporting $z = 1.28$ as the answer without converting
back to kg. Picking 1.29 (0.9015 is a little farther from 0.90).
Using $x = (z - \mu)/\sigma$ or $\mu - z\sigma$.

## Group discussion and revision

> Compare your Part 3 answers with your group. One thing worth
> discussing: would this z-table give the right answer if you
> standardized Act 2's binomial detection counts the same way? Record
> any changes, new insights, and muddiest points.

**Discussion points.** No, not reliably. Standardizing,
$z = (x - \mu)/\sigma$, works for any distribution with a mean and
standard deviation, but it only produces the standard normal curve
when the variable is normal to begin with. Act 2's counts are discrete
and right-skewed (for $N = 8$, $p = 0.3$), and standardizing changes
their location and scale, not their shape, so the table would give
the wrong probabilities.

**Watch for.**

- "It works because the distribution is symmetric." Symmetry isn't
  enough: a uniform distribution is symmetric, but its standardized
  shape is flat, not bell-shaped.
- "It's fine because we're far from 0." That's about whether a normal
  model is reasonable for these data in the first place, not about
  whether standardizing a non-normal variable gives normal
  probabilities.
