---
title: "Lab 3, Act 2 -- Answer Key"
subtitle: "Counting detections: binomial or Poisson?"
author: "NRES 746 -- Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 6, fig.height = 4)
```

This key follows the same section structure as the student handout. Questions with a single
correct numeric answer are answered directly; open-ended items (interpretation, discussion) are
given sample answers/discussion points rather than a fixed "correct" response.

## Part 1: Binomial by hand

**1a.** The binomial's full assumption list is exactly four items -- fixed $N$, binary
(Bernoulli) outcome per trial, independence across trials, constant $p$ across trials -- and a
complete answer names all four. What separates a thin answer from a strong one is engaging with
"plausible or not":

- **Fixed $N=8$** and **binary outcome per night** are true *by construction* (the deployment
  length is set in advance; "at least one detection" vs. none is how the variable is defined) --
  not much to interrogate in either one.
- **Independence** and **constant $p$** are the two actually worth scrutinizing. Independence
  could break down if a malfunction, a multi-night weather event, or the animal revisiting a
  scent-marked area links one night's outcome to the next's. Constant $p$ could drift with moon
  phase, temperature, food availability, or the animal habituating to (or growing wary of) the
  camera.

A student who only recites the four assumptions without engaging with which ones are actually
shaky hasn't fully answered 1a.

**1b.**

```{r binom-hand}
N <- 8; p <- 0.3; k <- 3
choose(N, k) * p^k * (1 - p)^(N - k)
dbinom(k, N, p)   # check
```

$P(X=3) = 0.2541$.

**1c.**

```{r mode-mean}
which.max(dbinom(0:8, 8, 0.3)) - 1   # mode, 0-indexed
8 * 0.3                              # mean
```

Mode $=2$; mean $=Np=2.4$. The mode can't land exactly on the mean because the mode must be a
whole number (a specific count, $k$), while the mean is a continuous expected value with no such
restriction -- they coincide only in special cases (e.g. a symmetric distribution where $Np$
happens to be an integer). Here the mean sits *above* the mode, which signals the distribution
is right-skewed: bounded at 0 on the left with more room to stretch toward higher counts on the
right, pulling the mean up above the single most common outcome.

## Part 2: Comparing to Poisson

**2a.** Scenario A (binomial vs. Poisson, both mean 2.4): the two bar charts disagree most in
the tails -- the Poisson has a visibly fatter right tail (it has no upper bound, while the
binomial is capped at 8) -- and somewhat in peak height. Scenario B: the two charts are nearly
indistinguishable across center, spread, and tails.

**2b.**

```{r variances}
N <- 8; p <- 0.3
N * p * (1 - p)          # Scenario A binomial variance
2.4                       # Scenario A/B Poisson variance (same lambda both scenarios)
100 * 0.024 * (1 - 0.024) # Scenario B binomial variance
```

Scenario A: binomial variance $=1.68$ vs. Poisson variance $=2.4$ -- mismatched. Scenario B:
binomial variance $=2.3424$ vs. Poisson variance $=2.4$ -- close match.

**2c.** Yes -- the variance mismatch in Scenario A (1.68 vs. 2.4) explains the visible
disagreement in 2a; the near-equal variances in Scenario B (2.3424 vs. 2.4) explain why those
charts looked nearly identical.

**2d.** *Poisson is a good stand-in for binomial when $N$ is large and $p$ is small*, so that
$(1-p)\approx1$ and $Np(1-p)\approx Np=\lambda$. Watch for students who conclude "Poisson is
just always fine" from Scenario B alone without registering that Scenario A disagreed -- push
them back to the variance mismatch as the reason it isn't universal.

**2e.** No fixed answer. Reasonable examples: a multi-night weather event (a storm suppressing
detections across several consecutive nights, breaking independence), seasonal or lunar changes
in animal activity (breaking constant $p$), or the animal habituating to (or growing wary of)
the camera over the deployment (also breaking constant $p$).

## Group discussion and revision

No fixed answer -- students compare variances and their Part 2d rule of thumb, and flag their
own muddiest points.
