---
title: "Lab 4, Act 1 -- Answer Key"
subtitle: "If we knew the truth, would our trend analysis find it?"
author: "NRES 746 -- Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 6, fig.height = 4)

I0 <- 60
sigma <- 0.2
lambda <- 0.75^(1/25)
b <- log(lambda)
tt_annual <- 1:25
tt_every2 <- seq(1, 25, by = 2)
Stt <- function(t) sum((t - mean(t))^2)
```

This key follows the same section structure as the student handout. Questions with a single
correct numeric or derived answer are answered directly; open-ended items (interpretation,
speculation, the plain-English procedure, reflections) are given sample answers and discussion
points rather than a fixed rubric. Because Part 4 is a Monte Carlo experiment, every group's
numbers will land somewhere in a range, not on an exact decimal -- the point of grading is
whether their numbers are in the right ballpark and their reasoning about direction (more/less
variable, does it agree) is sound, not exact-match precision.

## Part 1: The model

**1a.**

```{r part1a}
lambda <- 0.75^(1/25)
b <- log(lambda)
c(lambda = lambda, b = b)
```

$\lambda \approx 0.9886$, $b \approx -0.01151$. In words: the index falls by about 1.15% every
single year, and compounding that for 25 years produces the overall 25% loss.

**1b.** Only the two **log-scale** curves are straight lines. On the log scale,
$\log(\text{median}_t) = \log(I_0) + bt$, which is linear in $t$ with slope $b$. On the raw
scale, $\text{median}_t = I_0\lambda^t$ is an exponential function of $t$, so both raw-scale
curves are curved (concave up for the 25% decline, more sharply curved and flatter-looking for
the steeper 90% decline).

```{r part1b-figure, fig.width=6.5, fig.height=2.6}
par(mfrow = c(1, 2), mar = c(4, 4.5, 2, 1))
plot(NA, xlim = c(0, 25), ylim = c(0, 80), xaxs = "i", yaxs = "i",
     xlab = "Year", ylab = "Median index", main = "Raw scale")
curve(60 * 0.75^(x / 25), add = TRUE, lwd = 2)
curve(60 * 0.10^(x / 25), add = TRUE, lwd = 2, lty = 2)
points(0, 60, pch = 16, cex = 1.3)
plot(NA, xlim = c(0, 25), ylim = c(1, 4.5), xaxs = "i", yaxs = "i",
     xlab = "Year", ylab = "log(median index)", main = "Log scale")
curve(log(60) + (log(0.75) / 25) * x, add = TRUE, lwd = 2)
curve(log(60) + (log(0.10) / 25) * x, add = TRUE, lwd = 2, lty = 2)
points(0, log(60), pch = 16, cex = 1.3)
```

**1b, continued.** A band of $\pm\sigma$ on the log scale is a constant-width pair of parallel
lines (since $\sigma$ doesn't depend on $t$). Converted back to the raw scale, the band is the
median times $e^{\pm 0.2}$ (about $\times 1.2$ and $\times 0.8$), which **shrinks in absolute
width as the median shrinks** -- the band is a fixed *percentage* of the median, not a fixed
number of animals.

```{r part1b-band}
c(year0_lo = 60 * exp(-0.2), year0_hi = 60 * exp(0.2), year0_width = 60 * (exp(0.2) - exp(-0.2)))
c(year25_lo = 45 * exp(-0.2), year25_hi = 45 * exp(0.2), year25_width = 45 * (exp(0.2) - exp(-0.2)))
```

The raw-scale band is about 24.2 wide in year 0 (median 60) and only about 18.1 wide in year 25
(median 45, mild decline) -- both are exactly 75% of the median, confirming the band scales
proportionally rather than staying fixed. A straight-line fit to the raw counts assumes the
scatter around the line is the same size everywhere; the assumption being violated is
**constant variance (homoscedasticity)** of the residuals. This is exactly the setup behind the
optional Act 3's comparison of a raw-scale analyst and a log-scale analyst.

**1c.** Full raw-scale model:
$$I_t = \underbrace{I_0\,\lambda^t}_{\text{deterministic decline}} \cdot\ \underbrace{e^{\varepsilon_t}}_{\text{noise}} = I_0\,e^{bt}\,e^{\varepsilon_t} = I_0\,e^{bt+\varepsilon_t}, \qquad \varepsilon_t\sim\text{Normal}(0,\sigma)$$
Taking logs of both sides:
$$\log(I_t) = \log(I_0) + bt + \varepsilon_t$$

The **log-scale** version meets the assumptions of ordinary linear regression: it's linear in
$t$, and the noise $\varepsilon_t$ enters additively, independently, normally, and with
constant variance $\sigma$ every year. The **raw-scale** version does not: the noise enters
*multiplicatively* rather than additively, the mean function $I_0e^{bt}$ is nonlinear in $t$,
and (from 1b) its variance is not constant. This is the entire reason the analyst regresses
$\log(I_t)$ rather than $I_t$ -- it's the version that actually satisfies OLS's assumptions.

**Common mistakes to watch for:** writing the raw-scale model as $I_0\lambda^t + \varepsilon_t$
(additive noise on the raw scale) rather than $I_0\lambda^t\cdot e^{\varepsilon_t}$
(multiplicative) -- that's a different, unrelated model, not what 1c describes; forgetting that
$\lambda^t = e^{bt}$ and treating $\lambda^t$ and $e^{bt}$ as if they need separate terms in the
log-transformed model (they're two ways of writing the same term, so $\log(\lambda^t) = bt$
exactly).

## Part 2: What should the simulation show?

**2a.** Both designs should be centered at the **same** value, the true $b$ -- nothing about
*which* years get surveyed changes whether the estimate is centered correctly, only how
precisely it pins down $b$. (Full derivation isn't expected here, but for your own reference:
$\hat b$ is a weighted average of the log counts with weights that sum to zero and to one when
multiplied by $t$, regardless of which $t$'s are surveyed, so it's unbiased under any survey
schedule.)

The **every-other-year** design should give a **more variable** (wider-spread) set of $\hat b$
values, because it uses only 13 data points instead of 25 -- fewer independent observations to
average over, the same idea as the Week 4 WOD's $\text{Var}(\bar X)=\sigma^2/n$ shrinking as $n$
grows. A good answer just needs this direction and the "more data, more precision" reasoning;
Part 4C gives the exact formula and factor.

**Common misconception to watch for:** "the same 25-year span is covered either way, so
precision should be about the same" -- this misses that a regression slope's precision depends
on the *number* of data points used to estimate it, not just the time span they cover. Give
credit for correctly identifying annual as more precise even if the reasoning is incomplete;
full credit for citing sample size (or, as a bonus insight, that spreading points out in time
*also* helps precision, but not enough here to make up for having half as many).

**2b.** No single correct sketch, but a good one shows both curves centered at the *same* point
(near $b\approx-0.0115$), with the annual-survey curve taller and narrower and the
every-other-year curve shorter and wider, and both roughly bell-shaped with the same total area
underneath (each covers the axis at "1" of probability). For reference, here are the curves the
formula in 4C would actually produce:

```{r part2b-figure, fig.width=5.5, fig.height=2.6}
par(mar = c(4, 1, 1, 1))
sd_a <- sigma / sqrt(Stt(tt_annual))
sd_e <- sigma / sqrt(Stt(tt_every2))
xs <- seq(-0.03, 0.01, length.out = 400)
plot(xs, dnorm(xs, b, sd_a), type = "l", lwd = 2, yaxt = "n",
     xlab = expression(hat(b)), ylab = "", ylim = c(0, max(dnorm(xs, b, sd_a))))
lines(xs, dnorm(xs, b, sd_e), lwd = 2, lty = 2)
abline(v = b, col = "grey50", lty = 3)
legend("topleft", c("annual", "every other year"), lty = c(1, 2), lwd = 2, bty = "n", cex = 0.85)
```

**2c.**
$$\text{SE}(\widehat{\text{coverage}}) = \sqrt{\frac{p(1-p)}{R}} = \sqrt{\frac{0.9\times0.1}{R}} = \sqrt{\frac{0.09}{R}}$$
Setting this $\le 0.01$ and solving for $R$:

```{r part2c}
R_needed <- 0.9 * 0.1 / 0.01^2
R_needed
```

$R \ge 900$. The handout's default of 2,000 replicates clears this comfortably (expected MC SE
at $R=2000$ is $\sqrt{0.09/2000}\approx 0.0067$).

## Part 3: Plan the experiment

**3.** No single correct wording, but a complete procedure should include, in some form:

1. Repeat many times (e.g., $R=2{,}000$):
    a. Simulate one set of counts from the **known** data-generating model (using the *true*
       $b$ and $\sigma$) for the survey years being tested.
    b. Fit `lm(log(idx) ~ t)` to the simulated counts; extract the slope estimate $\hat b$ and its
       90% confidence interval.
    c. Record $\hat b$, and whether the interval contains the true $b$ (yes/no).
2. After all $R$ replicates: compute the mean and SD of the $R$ recorded $\hat b$ values (bias
   and variability), and the fraction of replicates whose interval contained the true $b$ (the
   estimated coverage) -- compare that fraction to 90%.

**Common omissions to watch for:** not being explicit that the *known* (true) $b$ and $\sigma$
are what generate the fake data -- that's the entire point of a "if we knew the truth" check;
forgetting to record whether each replicate's interval contains $b$ (needed for the coverage
question); describing only a single simulated dataset rather than many repeated ones.

## Group check-in

No fixed answer -- the point is to reconcile any disagreement in the 2a/2b speculation and
agree on one plain-English procedure and target $R$ before splitting into paths.

## Part 4A / Part 4B: Simulation results

Both paths run the identical experiment, so expected results are the same regardless of path.
Because this is Monte Carlo, every group's numbers will differ slightly; at the handout's
$R=2{,}000$, expect something close to (one example run shown, with a fixed seed):

```{r part4-example}
sim_one <- function(tt, sigma, b, I0 = 60, level = 0.9) {
  idx <- I0 * exp(b * tt + rnorm(length(tt), 0, sigma))
  m <- lm(log(idx) ~ tt)
  ci <- confint(m, "tt", level = level)
  c(bhat = unname(coef(m)["tt"]), lo = ci[1], hi = ci[2])
}
run_design <- function(tt, sigma, b, R, seed) {
  set.seed(seed)
  res <- t(replicate(R, sim_one(tt, sigma, b)))
  list(bhat = res[, "bhat"], mean_b = mean(res[, "bhat"]), sd_b = sd(res[, "bhat"]),
       coverage = mean(res[, "lo"] <= b & b <= res[, "hi"]))
}
res_a <- run_design(tt_annual, sigma, b, R = 2000, seed = 4746)
res_e <- run_design(tt_every2, sigma, b, R = 2000, seed = 4747)
data.frame(
  design = c("Annual", "Every other year"),
  mean_bhat = c(res_a$mean_b, res_e$mean_b),
  sd_bhat = c(res_a$sd_b, res_e$sd_b),
  coverage = c(res_a$coverage, res_e$coverage)
)
```

```{r part4-figure, fig.width=6.5, fig.height=2.8}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
hist(res_a$bhat, freq = FALSE, breaks = 30, main = "Annual", xlab = expression(hat(b)),
     col = "grey90", border = "white")
curve(dnorm(x, mean(res_a$bhat), sd(res_a$bhat)), add = TRUE, lwd = 2)
abline(v = b, col = "red", lwd = 2, lty = 2)
hist(res_e$bhat, freq = FALSE, breaks = 30, main = "Every other year", xlab = expression(hat(b)),
     col = "grey90", border = "white")
curve(dnorm(x, mean(res_e$bhat), sd(res_e$bhat)), add = TRUE, lwd = 2)
abline(v = b, col = "red", lwd = 2, lty = 2)
```

In both panels, the true $b$ (red dashed line) sits right at the center of the simulated
distribution, and a normal curve fit to the simulated mean/SD tracks the histogram well -- both
things Part 4C asks students to confirm quantitatively.

**What to look for, both paths:** mean $\hat b$ close to the true $b\approx-0.0115$ for both
designs (typically within about $\pm0.001$ at $R=2{,}000$); SD of $\hat b$ noticeably larger for
every-other-year than annual (roughly 30-40% larger); coverage somewhere in the neighborhood of
85-95% for both designs (the exact figure wobbles at $R=2{,}000$; see the MC SE from 2c). A
coverage estimate far outside that range (below ~80% or above ~98%) usually means a bug, not
bad luck -- see below.

**Common bugs to watch for (both paths, especially AI-generated code):**

- Passing survey-design arguments through `...` inside `replicate()`: `replicate()` wraps its
  expression in its own closure, so `function(...) Sim(...)` used loosely inside `replicate()`
  can silently drop or misroute arguments rather than erroring. Worth a spot check if results
  look implausible (e.g., the index always near zero).
- Calling `confint(m, level = 0.9)` without naming the coefficient (`"tt"` or `2`), which can
  grab the intercept's interval instead of the slope's.
- Rebuilding `tt` incorrectly for the every-other-year design (should be
  `seq(1, 25, by = 2)`, 13 points) -- e.g., accidentally reusing the annual `tt` for both runs,
  which would make the two designs' results identical.
- Simulating on the raw scale and fitting `lm(idx ~ tt)` instead of `lm(log(idx) ~ tt)` -- this
  silently answers a different question (the raw-scale analyst from the optional Act 3), not
  this Act's.

**Reflection prompts (4A/4B):** no fixed answer. A good AI-assisted reflection names a specific
place their own understanding of Parts 1-3 caught something the AI got wrong or wrote
ambiguously (e.g., which scale to simulate/fit on, or which coefficient's CI to use). A good
self-coded reflection names a specific sticking point (`confint()` syntax, indexing the
coefficient table, vectorizing `replicate()`) and how it got resolved.

## Part 4C: Compare with the formula

```{r part4c}
Stt_annual <- Stt(tt_annual); Stt_e2 <- Stt(tt_every2)
sd_annual <- sigma / sqrt(Stt_annual)
sd_e2 <- sigma / sqrt(Stt_e2)
c(annual = sd_annual, every_other_year = sd_e2, ratio = sd_e2 / sd_annual)
```

Formula SD: annual $\approx 0.00555$, every-other-year $\approx 0.00741$ -- the every-other-year
design is about **1.34 times** as variable as the annual design (matching the direction of the
2a speculation, and giving it an exact number).

**Do they agree?** A group's simulated SD (from `sd(bhat)` over $R=2{,}000$ replicates) should
typically land within a few percent of the formula value -- the relative standard error of a
sample SD from $n$ draws is about $1/\sqrt{2(n-1)}$, which is only about 1.6% at $n=2{,}000$, so
two independent SDs (`sd_a` from 4B and `0.00555` from the formula) agreeing to within roughly
$\pm5\%$ is a good match; agreement within $\pm15\%$ is still plausibly just Monte Carlo noise.
A discrepancy much larger than that (e.g., off by a factor of 1.5 or more, or in the wrong
direction) is much more likely to signal a bug than bad luck -- a good tell is rerunning the
simulation with a different seed: pure Monte Carlo noise moves around between reruns, while a
coding bug reproduces the same wrong answer every time.

## If you finish early ($\sigma=0.4$)

Since $\text{SD}(\hat b)=\sigma/\sqrt{S_{tt}}$ is *linear* in $\sigma$, doubling $\sigma$ exactly
doubles the formula SD:

```{r finish-early}
0.4 / sqrt(Stt_annual)
```

Formula SD $\approx 0.01109$ (versus $\approx 0.00555$ at $\sigma=0.2$). A simulation check
(R=5,000) gives SD $\approx 0.0111$ and mean $\hat b\approx-0.0116$ -- still centered on the
true $b$. **Coverage should stay close to 90%** even though the noise doubled: `confint()`
re-estimates the residual SD from each simulated dataset rather than assuming the true
$\sigma$, so the 90% interval remains valid regardless of how large $\sigma$ actually is (a
simulation check gives coverage $\approx 90.0\%$). This is worth flagging if a group predicted
coverage would drop -- doubling the noise makes each individual estimate less precise (wider
CI, more spread-out $\hat b$), but it doesn't break the *calibration* of the confidence
procedure itself.

## Group discussion and revision

No fixed answer -- the point is for students to reconcile any real disagreement between their
Monte Carlo results and the formula, and to record their own muddiest points.
