---
title: "Lab 4, Act 2 -- Answer Key"
subtitle: "How much monitoring is enough?"
author: "NRES 746 -- Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 6, fig.height = 4)

N0 <- 1000
pobs <- 0.02
lambda <- 0.75^(1/25)
```

This Act is a group activity throughout Part 2, so expect one shared set of answers and one
shared R script per group rather than individual variation -- grade the group's packet(s) and
script together. Part 1 stays individual-first, so some variation there is expected and fine.
Questions with a single correct numeric answer are answered directly; open-ended items
(1a's real-world question, 1c's reasoning, 1d's predictions, the Reflection) get discussion
points rather than a fixed rubric. Every number below comes from the four functions exactly as
specified in the handout, so a group's own numbers should land close to these (Monte Carlo
noise aside).

## Part 1: Plan the pieces

**1a.** Not detected on a single person-day: $1 - 0.02 = 0.98$. Not detected across $k$
independent person-days: $0.98^k$. So:
$$q(k) = 1 - 0.98^k$$

```{r part1a}
detect_prob <- function(k, pobs = 0.02) 1 - (1 - pobs)^k
detect_prob(3)
```

$q(3) \approx 0.0588$, matching the check value.

*Real-world differences between "5 observers, 1 day" and "1 observer, 5 days" (open-ended).*
Good answers name at least one of: a single observer working 5 days might get better at reading
sign/habitat as the days go on (learning), while 5 observers on 1 day bring more independent
eyes but no time to adapt; weather or animal activity can vary day to day, so spreading effort
over multiple days hedges against one bad day, while concentrating it in one day risks a single
unlucky day wasting the whole bout; animals might become wary of repeated visits (habituation or
avoidance) if the same day is revisited, which a single-day multi-observer bout avoids entirely.
The formula treats all of this as irrelevant because it only assumes each person-day is an
independent, identical 2% trial -- any of the above would violate that independence or identical-
probability assumption.

**1b.**
$$N_t = N_0 \lambda^t = 1000 \times \left(0.75^{1/25}\right)^t$$

```{r part1b}
true_abundance <- function(t, N0 = 1000, lambda = 0.75^(1/25)) N0 * lambda^t
true_abundance(c(0, 25))
```

$N_{25} = 750$, matching the check value (a clean number since $\lambda^{25} = 0.75$ exactly).

**1c.** **Binomial**, with size $N_t$ and probability $q(k)$: $I_t \sim \text{Binomial}(N_t, q(k))$.
Reasoning: at year $t$ there are $N_t$ animals present (a fixed number of "trials," conditional
on $N_t$), each independently detected or not with the same probability $q(k)$, and the survey
count is the number of "successes" out of those $N_t$ trials -- exactly the setup the binomial
was built for (same reasoning as the Week 5 WOD's nest-fate example, just with $N_t$ trials
instead of one).

**Common mistake to watch for:** proposing a Poisson here. Poisson is a reasonable answer if a
group treats $N_t$ as unknown/large and $q(k)$ as small (the usual binomial-to-Poisson limit),
and should get credit if they explain that reasoning -- but the more direct, exact answer, given
that $N_t$ is already known in this simulation, is binomial.

**1d.** *Two predictions (no fixed answer -- these are hunches to test in 2f).*

- **$k$ vs. $s$:** either direction is a reasonable guess, and a good answer just needs a
  reason ("more effort per bout should give a cleaner count," or "more surveys give the
  regression more points to work with"). The actual finding from 2f: in most of the design space
  the group will explore, $k$ and $s$ matter by *comparable* amounts, not one dominating the
  other. For instance, starting from $k=2, s=1$ (power $\approx 0.66$), doubling $k$ to 4 raises
  power to about $0.92$ (+0.26), while doubling $s$ to 2 lowers it to about $0.38$ ($-0.28$) --
  almost symmetric. (For instructor reference, not something groups need to derive: with $q(k)$
  small, the delta-method variance of a log count scales roughly as $1/k$, while
  $S_{tt}$ -- and hence $\text{Var}(\hat b)$ -- scales roughly as $1/s$, so the two enter the
  precision of $\hat b$ in a similar way.) Give full credit for a reasoned guess in either
  direction; the point of 1d is committing to a prediction before simulating, not getting it right.
- **False-decline rate at $\lambda=1$:** should **not** be exactly 5%. The rule requires the
  slope to be negative *and* significant. Under a truly flat population, a two-sided test is
  significant about 5% of the time, split roughly evenly between "significantly negative" and
  "significantly positive" slopes -- so requiring *both* negative and significant should roughly
  halve the false-decline rate to around 2.5%. A simulation check confirms this:

```{r part1d-null-check}
simulate_counts <- function(t_survey, k, N0 = 1000, lambda = 0.75^(1/25), pobs = 0.02) {
  N_t <- true_abundance(t_survey, N0, lambda)
  rbinom(length(t_survey), size = round(N_t), prob = detect_prob(k, pobs))
}
detect_decline <- function(t_survey, counts, alpha = 0.05) {
  m <- lm(log(counts) ~ t_survey)
  s <- summary(m)$coefficients[2, ]
  unname(s[1] < 0 && s[4] <= alpha)
}
set.seed(14)
t1 <- seq(1, 25, by = 1)
mean(replicate(3000, detect_decline(t1, simulate_counts(t1, k = 3, lambda = 1))))
```

About 2%, in line with the halved-5% reasoning above. **Common mistake to watch for:** a group
that predicts exactly 5% is forgetting the rule has two conditions, not one -- worth connecting
back to why 2d's decision rule needed both.

## Group check-in

No fixed answer -- the point is to reconcile 1a-1c and agree on both 1d predictions (even if the
predictions turn out wrong later) before opening laptops.

## Part 2: Build, test, and search

This section is graded as one shared group product: one filled-in result per item (from any
group member's packet) plus the group's single emailed R script. All four functions below are
given in full, both as a reference for checking a group's script and as a reference
implementation for reproducing every number in this key.

```{r part2-functions}
detect_prob <- function(k, pobs = 0.02) {
  1 - (1 - pobs)^k
}

true_abundance <- function(t, N0 = 1000, lambda = 0.75^(1/25)) {
  N0 * lambda^t
}

simulate_counts <- function(t_survey, k, N0 = 1000, lambda = 0.75^(1/25), pobs = 0.02) {
  N_t <- true_abundance(t_survey, N0, lambda)
  q <- detect_prob(k, pobs)
  rbinom(length(t_survey), size = round(N_t), prob = q)
}

detect_decline <- function(t_survey, counts, alpha = 0.05) {
  m <- lm(log(counts) ~ t_survey)
  s <- summary(m)$coefficients[2, ]
  unname(s[1] < 0 && s[4] <= alpha)
}
```

**2a.** `detect_prob(3)` $= 0.0588$ (from 1a).

**2b.** `true_abundance(0)` $= 1000$ (from 1b). As a **vector call**,
`true_abundance(c(0, 25))` should return $1000$ and $750$ *together*, confirming the function
handles a whole vector of years, not just one at a time -- essential, since `simulate_counts()`
hands it the full `t_survey` vector in 2c.

**Common mistake to watch for:** a `true_abundance()` written with an `if (length(t) == 1)`
branch, or one that otherwise only tested against single years, can look correct on the two
scalar checks above and still break (or silently return a recycled single value) once it's
handed a 25-element `t_survey` vector inside `simulate_counts()`. The vector check is what
catches this before it causes a confusing downstream bug.

**2c.** A representative run (`t_survey = seq(1, 25, by = 1)`, `k = 3`):

```{r part2c}
set.seed(11)
simulate_counts(seq(1, 25, by = 1), k = 3)
```

Counts should trend downward (from around 60-70 early on to around 40-50 by year 25) and stay
far below the true abundance ($N_0=1000$ falling to $N_{25}=750$) -- detection probability at
$k=3$ is under 6%, so only a small fraction of the population is ever counted. The output should
also be a **vector of 25 counts**: `length(simulate_counts(seq(1, 25, by = 1), k = 3))` $= 25$,
matching `length(t_survey)`.
**Common mistake to watch for:** a group whose counts *don't* trend downward at all, whose
counts are implausibly close to the true abundance, or whose output is a single recycled number
rather than 25 distinct values, likely has `detect_prob()` or `true_abundance()` wired in
backwards, has a `true_abundance()` that isn't actually vectorized (see 2b), or forgot to round
`N_t` before passing it to `rbinom()`'s `size` argument (which requires an integer and silently
truncates otherwise).

**2d.**

```{r part2d}
set.seed(12)
t1 <- seq(1, 25, by = 1)
mean(replicate(300, detect_decline(t1, simulate_counts(t1, k = 3))))
mean(replicate(300, detect_decline(t1, simulate_counts(t1, k = 3, lambda = 1))))
```

Under the real decline, `detect_decline()` should return `TRUE` a clear majority of the time
(roughly 80% at $k=3$, $s=1$ -- see 2e/2f); under $\lambda=1$ it should return `TRUE` only rarely
(around 2-3%, matching 1d). **Common mistake to watch for:** a `detect_decline()` that checks
only the $p$-value (forgetting the sign of the slope) will falsely detect a decline roughly twice
as often under the null as it should (closer to 5% than 2.5%), since it also counts significant
*increases* as "detected declines."

**2e.**

```{r part2e}
estimate_power <- function(k, s, R = 1000, lambda = 0.75^(1/25), alpha = 0.05,
                            N0 = 1000, pobs = 0.02) {
  t_survey <- seq(1, 25, by = s)
  mean(replicate(R, {
    counts <- simulate_counts(t_survey, k, N0, lambda, pobs)
    detect_decline(t_survey, counts, alpha)
  }))
}
set.seed(14)
fd_rate <- estimate_power(3, 1, R = 3000, lambda = 1)
fd_rate
sqrt(fd_rate * (1 - fd_rate) / 3000)   # Monte Carlo SE
```

False-decline rate $\approx 0.02$-$0.03$, consistent with 1d's halved-5% prediction.

**2f.**

```{r part2f-grid, fig.width=6, fig.height=4}
set.seed(746)
grid <- expand.grid(k = 1:15, s = 1:5)
grid$power <- mapply(estimate_power, grid$k, grid$s, MoreArgs = list(R = 1000))

plot(NA, xlim = c(1, 15), ylim = c(0, 1), xlab = "k (person-days per bout)",
     ylab = "Estimated power")
abline(h = 0.75, lty = 2, col = "grey40")
for (ss in 1:5) {
  d <- grid[grid$s == ss, ]
  lines(d$k, d$power, col = ss, lwd = 2)
  points(d$k, d$power, col = ss, pch = 16, cex = 0.6)
}
legend("bottomright", legend = paste("s =", 1:5), col = 1:5, lwd = 2, bty = "n", cex = 0.8)
```

**Pattern to expect:** power rises with $k$ for every survey interval $s$, and every curve shifts
down and to the right as $s$ increases -- annual surveys ($s=1$) cross the 0.75 line around
$k=3$, while every-5-year surveys ($s=5$) haven't reached it even at $k=15$. As discussed in
1d, $k$ and $s$ trade off at roughly comparable rates through most of this grid (neither
dominates outright), though $s=1$ is clearly the most power-efficient interval on its own.

**2g.**

```{r part2g-candidates}
cost_of <- function(k, s, nyears = 25) length(seq(1, nyears, by = s)) * (2000 + 200 * k)
grid$cost <- mapply(cost_of, grid$k, grid$s)
ok <- grid[grid$power >= 0.75, ]
ok <- ok[order(ok$cost), ]
head(ok, 6)
```

```{r part2g-recheck}
top3 <- head(ok, 3)
set.seed(999)
top3$power_recheck <- mapply(estimate_power, top3$k, top3$s, MoreArgs = list(R = 5000))
top3$mc_se <- sqrt(top3$power_recheck * (1 - top3$power_recheck) / 5000)
top3
```

| k | s | Cost | Power (re-checked) | MC SE |
|---|---|------|---------------------|-------|
| 10 | 4 | \$28,000 | $\approx 0.76$ | $\approx 0.004$-$0.006$ |
| 11 | 4 | \$29,400 | $\approx 0.80$-$0.81$ | $\approx 0.004$-$0.006$ |
| 7  | 3 | \$30,600 | $\approx 0.74$-$0.75$ | $\approx 0.004$-$0.006$ |

**This is the payoff of the "re-check with more replicates" instruction.** The cheapest design
from the 1,000-replicate grid ($k=10, s=4$, \$28,000) sits *right on* the 0.75 threshold; a high-
precision check (R = 10,000) puts it at about 0.764 -- it does clear the bar, but with almost no
margin. The third-cheapest candidate ($k=7, s=3$, \$30,600) looked like it cleared 0.75 in the
1,000-replicate grid (0.753) but drops to about **0.743 -- below threshold** -- under a
high-precision re-check. That's not Monte Carlo bad luck resolving itself; it's a genuine case
where the coarse grid estimate was on the wrong side of the line.

**Recommended design: $k=11$, $s=4$ (\$29,400).** It costs \$1,400 more than the cheapest
candidate but clears 0.75 with real margin (power $\approx 0.80$-$0.81$, several Monte Carlo
standard errors above the threshold), rather than sitting on the boundary where the "true" power
could plausibly be under 0.75. **Grading note:** a group that recommends $k=10, s=4$ isn't
necessarily wrong -- it's a defensible "minimum cost that clears the bar" choice -- but full
credit requires them to have actually re-checked it with more replicates and to say something
about why they're comfortable (or not) with how close to the threshold it sits. A group that
skips the re-check step entirely and just reports the 1,000-replicate numbers should be asked to
go back and do it -- that's the whole point of 2g's instruction.

**If you finish early (environmental noise).**

```{r part2h-env}
true_abundance_env <- function(t, N0 = 1000, lambda = 0.75^(1/25), sd_env = 0.05) {
  maxt <- max(t)
  eps <- rnorm(maxt, 0, sd_env)
  logN <- log(N0) + cumsum(rep(log(lambda), maxt) + eps)
  c(N0, exp(logN))[t + 1]
}
simulate_counts_env <- function(t_survey, k, N0 = 1000, lambda = 0.75^(1/25),
                                 pobs = 0.02, sd_env = 0.05) {
  N_t <- true_abundance_env(t_survey, N0, lambda, sd_env)
  rbinom(length(t_survey), size = round(N_t), prob = detect_prob(k, pobs))
}
estimate_power_env <- function(k, s, R = 2000, lambda = 0.75^(1/25), alpha = 0.05, sd_env = 0.05) {
  t_survey <- seq(1, 25, by = s)
  mean(replicate(R, {
    counts <- simulate_counts_env(t_survey, k, lambda = lambda, sd_env = sd_env)
    detect_decline(t_survey, counts, alpha)
  }))
}
set.seed(748)
estimate_power_env(11, 4)       # power with environmental noise, recommended design
set.seed(750)
estimate_power_env(11, 4, lambda = 1)   # false-decline rate with environmental noise
```

For the recommended design, power with environmental noise drops from about 0.80 to roughly
**0.50**, and the false-decline rate rises from about 2-3% to roughly **15%**. The assumption
from Act 1 that's now violated is **independence of the year-to-year noise terms** -- Act 1's
model had $\varepsilon_t$ drawn independently every year, but persistent shocks mean this year's
deviation carries into next year's abundance, so consecutive counts are correlated rather than
independent. This is exactly why the false-decline rate roughly quintuples: a single bad
(or good) run of years now looks like a real trend to a method that assumes independent noise.
The takeaway worth drawing out: real monitoring data plausibly needs a model that accounts for
this kind of temporal autocorrelation (a preview of time-series/state-space content later in the
course), not just more replicate surveys of the same naive model.

## Reflection

No fixed answers for R1-R4 -- these are individual (per the house note at the top of the Act)
and should be graded for engagement, not a specific "correct" response. A few things worth
recognizing as strong answers:

- **R1:** names the actual recommended design and cost, and identifies a real assumption (e.g.,
  independent detection across person-days, no environmental persistence, known and constant
  $N_0$) rather than a generic "the model could be wrong."
- **R2:** written for a non-technical reader -- no jargon like "power," "$p$-value," or
  "Monte Carlo" without a plain-language gloss.
- **R3:** no single correct answer, but a strong response recognizes *why* grid search is
  reasonable here specifically, rather than either defending or dismissing it in the abstract:
  the design space is small (75 combinations of $k$ and $s$) and both variables are naturally
  discrete -- you can't run "3.7 person-days" or survey "every 2.3 years" -- so exhaustive search
  is cheap and directly respects the real decision space. Grid search also produces the whole
  power surface (2f's plot), which is useful for seeing trade-offs and explaining the
  recommendation to a non-technical stakeholder, not just a single "optimal" point.

  For "better ways," reasonable answers include: an adaptive/staged search that runs every grid
  point cheaply first and only re-checks the near-threshold candidates with many more
  replicates (this is literally what 2g already does, just not generalized into a search
  strategy); exploiting that power is roughly monotonic in both $k$ and $s$ to search more
  cleverly (e.g., something like a binary search along each row instead of evaluating every
  cell); or a numerical optimizer such as `optim()` if $k$ and $s$ were treated as continuous
  and then rounded, though for a search space this small that adds complexity without an
  obvious payoff. Worth surfacing explicitly: grid search stops being practical once the design
  space grows past a couple of dimensions (adding, say, survey season or number of sites would
  make the grid explode combinatorially), which is when a smarter method actually earns its
  keep. **Common misconception to watch for:** claiming grid search is "obviously worse" than
  some other named method without weighing that claim against the size and discreteness of
  *this* particular design space -- there is no universally better search method, only one
  better suited to a given problem's dimensionality and structure.

- **R4:** names something concrete the group had to fix or push back on (a wrong argument order,
  a forgotten `round()`, an AI-suggested $s$ range that didn't match the handout's), not just
  "it worked great."
