---
title: "Lab 5, Act 2: Answer Key"
subtitle: "When Calculus Runs Out"
author: "NRES 746, Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 6, fig.height = 3.5)

chorus_data <- read.csv("../../data/FrogChorus.csv")
log_fact <- sum(lfactorial(chorus_data$males))
```

Each item repeats the handout's question, then gives the answer, a
short rationale, and mistakes to watch for. Questions with one correct
answer are answered directly. Open-ended items (1c, 1e, the judgment
parts of 2b and 2d, 2e, the early-finisher question, and the
reflection) get sample answers or discussion points instead. All
numbers are computed in R when this file is knitted.

Part 2 labels each coding question "Recommended: no AI" (2a, 2b, 2d) or
"AI optional/encouraged" (2c, the early-finisher question), and students circle how
they used AI after each one. Students turn in their filled-in answers
and the 2c figure, not code. AI use itself doesn't cost points; only
undisclosed use does. A student who circles "wrote the code" on a
"no AI" item is being honest, so treat it as something to talk about,
not a penalty.

Because students may keep or drop the $\log(y_i!)$ constant, NLL
values are given both ways where it matters. The constant is
$\sum \log(y_i!) = `r round(log_fact, 2)`$ for these data. Differences
between NLLs are the same either way.

Model: $y_i \sim \text{Poisson}(\lambda_i)$, with
$\lambda_i = a\left(1 - e^{-b d_i}\right)$.

## Part 1: Plan it on paper

**1a.**

> What do $a$ and $b$ mean biologically, and what are their units?

**Answer.** $a$ is the expected number of calling males heard far from
the road, where the chorus has fully recovered (units: males per
5-minute count). $b$ is how quickly the chorus recovers with distance
(units: 1/km). Larger $b$ means faster recovery.

**Why.** As $d \to \infty$, $e^{-bd} \to 0$ and $\lambda \to a$. The
exponent $bd$ must be unitless, so $b$ has units of 1/distance. A
useful translation: the chorus reaches half its full size at
$d = \ln 2 / b$.

**Watch for.**

- Calling $a$ "the count at the road." At $d = 0$ the model predicts
  zero, not $a$.
- Giving $b$ the same units as $d$ (km).

**1b.**

> Act 1 used a binomial distribution. Why is a Poisson distribution a
> more reasonable choice for these counts?

**Answer.** The counts have no fixed number of trials or upper limit.
The binomial needs a known $N$ (Act 1 had 10 visits). Here we don't
know how many males are present, only how many we heard.

**Why.** The Poisson describes counts of independent events with no
upper bound, with mean (and variance) $\lambda$.

**Watch for.** "Because the counts are bigger" or "because the data
are counts." Both distributions model counts. The key difference is
whether there is a known number of trials.

**1c.**

> Use the figure below and your function to estimate starting values
> for $a$ and $b$. Show how you got your estimated starting value for
> $b$. *Hint for $b$:* from the figure, estimate the distance $d_{1/2}$
> at which the chorus is halfway to its maximum, so that
> $\lambda(d_{1/2}) = a/2$. Then solve that equation for $b$. [...]

**Answer (any reasonable values).** $a \approx 12$ (where the points
level off, roughly 11 to 13). For $b$: the curve reaches half of $a$
(about 6) at roughly $d \approx 0.7$ to 1 km. Setting
$a(1 - e^{-bd}) = a/2$ gives $e^{-bd} = 1/2$, so
$b = \ln 2 / d \approx 0.7$ to 1.

**Why.** Starting values only need to be in the right neighborhood.
Every start tried in 2b ($a$ from 10 to 15, $b$ from 0.2 to 2) reaches
the same answer.

**Watch for.**

- Stopping at $d_{1/2}$ and reporting it as $b$, e.g. "$b \approx 1$
  km." It's a rate, so it needs the $\ln 2/d_{1/2}$ step.
- Algebra slips solving $1 - e^{-bd} = 1/2$: taking the log of both
  sides before isolating $e^{-bd}$, or losing the minus sign
  ($e^{-bd} = 1/2 \Rightarrow -bd = \ln(1/2) = -\ln 2$).
- Setting $a$ to the single highest point (14) rather than the level
  where the points cluster. Acceptable as a start, but worth a comment.

**1d.**

> Write the negative log-likelihood, $\text{NLL}(a, b)$, for the 15
> ponds using the Poisson pmf. Simplify it as far as you can. As in
> Act 1 (1c), you can drop any term that doesn't depend on $a$ or $b$,
> since it won't change where the minimum is.

**Answer.**
$$\text{NLL}(a, b) = -\sum_{i=1}^{15}\left[y_i \log \lambda_i - \lambda_i - \log(y_i!)\right]
\;\longrightarrow\; \sum_{i=1}^{15}\left[\lambda_i - y_i \log \lambda_i\right]$$
with $\lambda_i = a\left(1 - e^{-b d_i}\right)$. Fully expanded:
$$\text{NLL}(a, b) = \sum_{i=1}^{15}\left[a\left(1 - e^{-b d_i}\right) - y_i \log a - y_i \log\left(1 - e^{-b d_i}\right)\right]$$

**Why.** The Poisson pmf is $\lambda^y e^{-\lambda}/y!$. Taking the log
gives $y \log\lambda - \lambda - \log y!$. Negate and sum. The
$\log(y_i!)$ term doesn't involve $a$ or $b$, so it can be dropped.

**Watch for.**

- Sign errors when negating (the $\lambda_i$ term should end up
  positive).
- Writing $\log \lambda_i$ as $\log a \cdot \log(1 - e^{-bd_i})$
  instead of a sum.
- Using a single $\lambda$ for all ponds, forgetting the subscript $i$.

**1e.**

> In plain English (not R code), describe step by step what a function
> must do to compute the NLL for one candidate pair $(a, b)$. Then
> describe how you would use that function to find the MLE.

**Sample answer.** Given a candidate $a$ and $b$: (1) for each pond,
compute the expected count $\lambda_i$ from its distance using the
monomolecular formula; (2) for each pond, compute the log of the
Poisson probability of the observed count given $\lambda_i$; (3) add
these up across ponds; (4) return the negative of the sum. To find the
MLE, give this function and a starting guess (from 1c) to an optimizer,
which tries new $(a, b)$ pairs, moving toward lower NLL, until it can't
improve. A grid search over $(a, b)$ also works.

**Watch for.**

- Plans that compute one $\lambda$ for all ponds.
- Plans that maximize the NLL, or that forget the minus sign.
- R syntax instead of plain English (the handout asks for English).

## Part 2: Fit the model

**2a.**

> *(Recommended: no AI.)* Write a function `nll_mono(par, y, d)` that implements your NLL from
> 1d, where `par` holds $a$ and $b$. Check it: `nll_mono(c(10, 1),
> males, dist_km)` should return about 33.84 if you used `dpois()`, or
> about $-121.31$ if you coded the formula yourself and dropped the
> constant term. (Why can the second number be negative?)

**Answer.**

```{r part2a}
dist_km <- chorus_data$dist_km
males <- chorus_data$males

nll_mono <- function(par, y, d) {
  lambda <- par[1] * (1 - exp(-par[2] * d))
  -sum(dpois(y, lambda = lambda, log = TRUE))
}

nll_mono_nc <- function(par, y, d) {
  lambda <- par[1] * (1 - exp(-par[2] * d))
  sum(lambda - y * log(lambda))
}

nll_mono(c(10, 1), males, dist_km)
nll_mono_nc(c(10, 1), males, dist_km)
```

**Why negative?** Dropping the constant subtracts
$\sum\log(y_i!) = `r round(log_fact, 2)`$ from every NLL value. What's
left is no longer the negative log of a probability, so nothing keeps
it above zero. Only differences between NLL values matter.

**Watch for.**

- `dpois(lambda, y)` with the arguments swapped.
- `par$a` on an unnamed vector (returns an error or `NULL`).
- Forgetting the minus sign in front of `sum()`.

**2b.**

> *(Recommended: no AI.)* Evaluate the NLL at your starting values from 1c. Then use `optim()`,
> starting from those values, to find the MLE. [...] Did `optim()`
> print any warnings? If so, what do you think caused them, and do they
> mean the answer is wrong?

**Answer.**

```{r part2b}
start <- c(12, 0.8)
nll_mono(start, males, dist_km)

fit <- optim(start, nll_mono, y = males, d = dist_km)
fit$par
fit$value
fit$convergence
```

$\hat{a} \approx `r round(fit$par[1], 2)`$,
$\hat{b} \approx `r round(fit$par[2], 2)`$, NLL at the MLE
$\approx `r round(fit$value, 2)`$ (or
$`r round(fit$value - log_fact, 2)`$ without the constant),
convergence code 0.

Other reasonable starts land on the same answer:

```{r part2b-starts}
starts <- list(c(10, 1), c(14, 0.5), c(15, 2), c(12, 0.2))
t(sapply(starts, function(s) round(optim(s, nll_mono, y = males, d = dist_km)$par, 3)))
```

**Warnings (discussion).** Most starting values produce "NaNs
produced" warnings. While searching, `optim()` tries some negative
values of $a$ or $b$, which make $\lambda_i$ negative (or zero), and
`dpois()` returns `NaN` there. The warnings do **not** mean the final
answer is wrong: convergence code 0 and estimates that match the plot
say the search recovered. Two standard fixes, worth mentioning to
students who ask: bounds (`method = "L-BFGS-B"` with `lower`), or
fitting $\log a$ and $\log b$ so the parameters can't go negative.

**Watch for.**

- Ignoring the convergence code.
- Concluding the fit failed because of the warnings.
- Starting values far off (e.g. $b = 50$) that send `optim()` to a
  flat region. If a student gets a strange answer, check their start.

**2c.**

> *(AI optional/encouraged.)* Plot the data with your fitted curve and your hand-guessed curve from
> 1c. How far from the road does the expected chorus reach half its
> full size?

**Answer.** $\ln 2 / \hat{b} \approx `r round(log(2) / fit$par[2], 2)`$ km.

```{r part2c, fig.width=5.5, fig.height=3.2}
par(mar = c(4, 4.5, 0.5, 0.5))
plot(dist_km, males, pch = 16, xlab = "Distance from highway (km)",
     ylab = "Calling males heard", ylim = c(0, 16))
curve(fit$par[1] * (1 - exp(-fit$par[2] * x)), add = TRUE, lwd = 2)
curve(start[1] * (1 - exp(-start[2] * x)), add = TRUE, lwd = 2, lty = 2)
abline(v = log(2) / fit$par[2], lty = 3)
legend("bottomright", c("MLE fit", "hand guess (12, 0.8)"),
       lwd = 2, lty = c(1, 2), bty = "n")
```

**Why.** Half of $a$ means $1 - e^{-bd} = 1/2$, so $d = \ln 2 / b$
(the same step as 1c, now with $\hat{b}$).

**Watch for.**

- Reporting $1/\hat{b}$ (about `r round(1 / fit$par[2], 2)` km). That's
  where the chorus reaches $1 - e^{-1} \approx 63\%$ of full size, not
  half.
- Reading the half-size distance off the raw data (e.g. the first pond
  with 6 or more males) instead of the fitted curve.

**2d.**

> *(Recommended: no AI.)* A skeptic says the road has no effect: every pond has the same
> expected count $\lambda$, whatever its distance. What is the MLE of
> $\lambda$ under that model, and what is its NLL? (Keep or drop the
> constant term the same way you did in 2a, so the two NLLs are
> comparable.) Which model do the data support more? Does the size of
> the difference convince you?

**Answer.**

```{r part2d}
lambda_hat <- mean(males)
nll_null <- -sum(dpois(males, lambda_hat, log = TRUE))
c(lambda_hat = lambda_hat, nll_null = nll_null,
  nll_null_nc = nll_null - log_fact, difference = nll_null - fit$value)
```

$\hat{\lambda} = \bar{y} = 106/15 \approx `r round(lambda_hat, 2)`$.
NLL $\approx `r round(nll_null, 2)`$ (or
$`r round(nll_null - log_fact, 2)`$ without the constant). The road
model's NLL is lower by about `r round(nll_null - fit$value, 1)`
units, so the data support it much more.

**Why (discussion).** The Poisson MLE for a constant mean is the
sample mean (the pika example from Monday's lecture). A difference of
`r round(nll_null - fit$value, 1)` NLL units is a likelihood ratio of
about $e^{`r round(nll_null - fit$value, 1)`} \approx
`r signif(exp(nll_null - fit$value), 1)`$. Most students should find
this convincing. Good answers also note a caution: the road model has
one more parameter, so it would fit at least somewhat better even if
the road did nothing. How to account for that formally comes later in
the semester.

**Watch for.**

- Mixing conventions: a `dpois()` NLL for one model and a
  constant-dropped NLL for the other (difference of about 155,
  meaningless).
- Saying the smaller NLL model is "proven" or "correct." The road
  model is better supported (corroborated), not proven.
- Using the median or another summary for $\hat{\lambda}$.

**2e.**

> Where did you use AI, if anywhere, and what did you have to check or
> fix? If you didn't use it, what was hardest to get working?

**Sample points (no fixed answer).**

- **Used AI.** Most likely on 2c (the plot). Look for students who
  checked AI output against something they knew, such as the 2a check
  value or the curve passing through the data. Common AI divergences
  to listen for: `nls()` or least squares instead of a likelihood, or
  `glm()`, which can't fit this curve directly.
- **No AI.** Common answers: getting `dpois()` arguments in the right
  order, passing extra arguments (`y`, `d`) through `optim()`, and
  understanding the warnings.

**Watch for.** Answers that don't match the circled AI-use lines.
Worth a conversation, not a penalty.

**If you finish early.**

> *(AI optional/encouraged.)* A colleague prefers a different saturating curve,
> $\lambda(d) = \dfrac{a d}{b + d}$. Fit it the same way and compare its
> NLL with the monomolecular model's. What do you conclude?

**Answer.**

```{r early}
nll_mm <- function(par, y, d) {
  -sum(dpois(y, lambda = par[1] * d / (par[2] + d), log = TRUE))
}
fit_mm <- optim(c(15, 1), nll_mm, y = males, d = dist_km)
round(c(a = fit_mm$par[1], b = fit_mm$par[2], nll = fit_mm$value,
        diff = fit_mm$value - fit$value), 3)
```

The two NLLs differ by only about `r round(fit_mm$value - fit$value, 2)`
units. These data can't tell the two curves apart.

**Why (discussion).** Both curves rise and level off, and both have two
parameters. They differ mostly beyond 8 km (the Michaelis-Menten curve
approaches its asymptote more slowly, here
$\hat{a} \approx `r round(fit_mm$par[1], 1)`$), where there are no
data. Choosing between them would take more ponds far from the road,
or a biological reason to prefer one shape.

**Watch for.** Declaring the monomolecular model the winner because its
NLL is slightly lower.

## Reflection

> In Act 1, a single 5-minute visit detected this frog only about 40%
> of the time, even at sites where it was known to be present. With
> that in mind, what do the counts in this Act measure? Could the road
> lower the counts near it for reasons other than fewer frogs living
> there? Could these data tell those explanations apart?
>
> Also record any changes you made, new insights, and any "muddiest
> points" or topics you want to review.

**Answer.** The counts measure males *heard*, not males present. The
road could lower counts by (1) fewer frogs living near it, (2) males
calling less, or (3) noise masking calls so the observer misses them.
One listen per pond can't tell these apart.

**Why (discussion).** The expected count is roughly (males present)
$\times$ (probability a male calls) $\times$ (probability a call is
heard), and all three factors could fall near the road. The fitted
$a$ is an index of the chorus as heard, not abundance. Separating the
explanations would take repeat visits with a detection model like
Act 1's, or a separate measure such as recorders that pick up calls
through the noise.

**Watch for.**

- "The counts are an index of abundance," with no link to the road.
- Naming only one mechanism and treating it as settled.
- Claiming the 40% from Act 1 applies directly here. It was estimated
  at other sites, with no road noise.

Muddiest points worth noting for the next lecture: why the constant
can be dropped, what the warnings mean, and how big an NLL difference
has to be to matter.
