---
title: "Understanding Random Intercepts and Slopes in Mixed-Effects Models using lme4 and glmmTMB"
author: "Grecia, Joshua and Lawrence"
output: html_document
date: "11-04-2025"
---

[click here for R script](Mixed_Effects_Demo.R)

### **Background**

Multi-level regression, also known as a hierarchical linear model, is a statistical technique used to analyse data with a hierarchical, nested, or clustered structure, such as students within classrooms or employees within companies. It extends standard regression by accounting for the fact that observations from the same group are not independent, allowing for differences in intercepts and slopes across groups. This approach acknowledges the data structure, prevents biased standard errors, and allows examination of effects at different levels. Real-world data often have hierarchical or repeated structures, and ignoring grouping leads to biased estimates and underestimated uncertainty. Mixed models allow for group-specific variation, partial pooling, and better generalization across groups.


### **Why multi-levels matter**
1.	Data from the same class will perform more similarly than data from different contexts.
    - Children in the same class will perform more similarly than children from different classes, potentially because they have the same teacher and use the same materials.
    - People treated in the same clinics should be more similar in their responses than those treated at different clinics, even if they use the same protocols, etc.
2.	Lack of Independence
    - Violates the assumption of errors being independent
    - Bias, SEs, CIs, and p-values

### **Fixed vs Random Effects**
Fixed effects represent population-level parameters that estimate specific coefficients applicable to the entire dataset, such as treatment effects or time trends. In contrast, random effects capture group-level variation by modeling variance components that reflect how individual groups, like schools, subjects, or sites, deviate from the overall average. While fixed effects provide estimates of average trends, random effects account for the variability among groups, allowing for partial pooling and more accurate inference.

### **Nested vs Crossed Random Effects**
  -	Nested: One factor is contained within another Syntax: (1 | school/classroom)
  -	Crossed: Factors interact independently Syntax: (1 | student) + (1 | teacher)


### **Benefits of multi-level models**
1. Modelling variability in effects across contexts (quantify variation among groups)
    - Model the variability in intercepts
    - Model the variability in slopes
2. Gives more realistic predictions (we don’t force every group to behave identically)
3. Reduces bias in estimated effects and standard errors.



Mixed-effects models are used when data are **grouped or clustered**. For example, students nested within schools, patients measured by different doctors, or repeated measures on individuals over time.

These models allow us to:

1. **Model population-level effects (fixed effects)**: e.g., the *overall* effect of treatment.
2. **Model group-level variation (random effects)**: e.g., how the treatment effect or baseline differs across different groups.

These characteristics makes them particularly valuable when we expect differences among groups but want to share information across them.

#### **When to use which?**



| Scenario | Recommended Package |
|----------|---------------------|
| Simple Gaussian or Poisson GLMM | lme4 |
| Zero-inflated or overdispersed counts | glmmTMB |
| Need for hurdle or beta models | glmmTMB | 
| Fast prototyping with standard families | lme4 |
| Diagnostics with simulated residuals | glmmTMB + DHARMa | 


---

- #### **We will focus on** `lmer` **to explain random slopes and intercepts**

## **Random Intercepts and Random Slopes**

When we fit a mixed-effects model using `lmer`, we can include **random intercepts** and **random slopes**.

- **Random intercept**: Each group has its own baseline outcome.
- **Random slope**: Each group can differ in how strongly a predictor (like treatment) affects the outcome.

Mathematically:

$$
y_{ij} = (\beta_0 + u_{0j}) + (\beta_1 + u_{1j})x_{ij} + \epsilon_{ij}
$$

Where:

- $\beta_0$, $\beta_1$ = fixed (overall) intercept and slope  
- $u_{0j}$, $u_{1j}$ = random deviations for group *j*  
- $\epsilon_{ij}$ = residual error  

---

## **Syntax in `lme4`**

| Model Type | Formula | Meaning |
|-------------|----------|---------|
| Random intercept only | `(1 | group)` | Groups have different intercepts |
| Random intercept and slope | `(1 + treatment | group)` | Groups have different intercepts and treatment effects |
| Crossed random effects | `(1 + treatment | A) + (1 | B)` | Effects vary across two independent grouping factors |

---

## **Correlation Between Intercepts and Slopes**

When we specify `(1 + treatment | group)`, the model also estimates how correlated the random intercepts and slopes are.

$$
\rho = \frac{\sigma_{u_0, u_1}}{\sigma_{u_0} \sigma_{u_1}}
$$

The estimated correlation can ether be positive or negative.

- A **positive correlation** means schools with higher intercepts also tend to have stronger treatment effects.  
- A **negative correlation** means schools with higher intercepts tend to have smaller treatment effects.

---

### Let us work through examples for nested and crossed random effects
We will do the following

- Go through data simulation for running both model types
- Extract and interpret results
- Brace up for a real data. It's going to be fun

## **Workflow 1: Nested Random Effects (Schools and Students)**

- We’ll simulate a dataset where **students are nested within schools**, and each school has its own baseline and treatment effect. Look at this as students inside school.
- We are working with two hypothetical treatments namely **control** and **treated**.
- **control** and **treated** can be interpreted as two distinct sets of students (e.g., male vs female; undergraduate vs graduate, etc)


```{r setup, include=T, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(lme4)
library(glmmTMB)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(broom.mixed)
```


### **1.1 Data Generation for Nested Model**

```{r Data simulation}
set.seed(123)
n_schools <- 5
n_students <- 50

df_nested <- data.frame(
  school = factor(rep(1:n_schools, each = n_students)),
  treatment = factor(rep(rep(c("control","treated"), each = n_students/2), times = n_schools))
)

# Random intercepts and slopes per school
school_intercepts <- rnorm(n_schools, 0, 1)
school_slopes <- rnorm(n_schools, 2, 0.5)

df_nested$y <- 10 +
  school_intercepts[as.numeric(df_nested$school)] +
  school_slopes[as.numeric(df_nested$school)] * (df_nested$treatment == "treated") +
  rnorm(nrow(df_nested), 0, 1)

head(df_nested)
str(df_nested)
```

**y** = Response variable

**treatment** = Predictor variable

**school** = Grouping variable

---

### **1.2 Fitting the Nested Model**

 We will fit a mix effect model using *school* as the grouping variable and specifying a random intercept and slope for each school.

```{r Fit model}
m_nested <- lmer(y ~ treatment + (1 + treatment | school), data = df_nested)
summary(m_nested)
```

#### **Interpretation**

- **Fixed effects:** Average (overall) intercept and treatment effect.  
- **Random effects:** Variance of intercepts and slopes across schools, plus their correlation.  
- **Residual variance:** Variation not explained by school or treatment.

---

### **1.3 Extracting Random Effects**

```{r Extracting varying slope}
fix <- fixef(m_nested)
ranefs <- ranef(m_nested)$school

school_effects <- data.frame(
  school = rownames(ranefs),
  intercept = fix["(Intercept)"] + ranefs[,"(Intercept)"],
  treatment_effect = fix["treatmenttreated"] + ranefs[,"treatmenttreated"]
)

school_effects
```

Each row shows how the intercept and treatment effect differ per school. treatment effect is obtained by adding random effects to fixed effect.

---

### **1.4 Visualizing Random Slopes**

```{r Dot plot of treatment effect}
ggplot(school_effects, aes(x = school, y = treatment_effect)) +
  geom_point(size = 3) +
  geom_hline(yintercept = fix["treatmenttreated"], linetype = "dashed", color = "red") +
  labs(y = "Estimated treatment effect", title = "School-specific treatment effects") +
  theme_minimal()
```

> The dashed red line shows the *average* treatment effect, while the points show each school’s deviation.

---

### **1.5 Predicted Lines per School**

```{r Spaghetti plot of fitted line, warning=FALSE}
df_nested$pred <- predict(m_nested)

ggplot(df_nested, aes(x = treatment, y = y, group = school, color = school)) +
  stat_summary(fun = mean, geom = "point", size = 3, position = position_dodge(0.3)) +
  stat_summary(fun = mean, geom = "line", aes(group = school), size = 1) +
  labs(y = "Outcome", title = "Varying treatment effects by school") +
  theme_minimal()
```

> Each line shows how a school responds differently to treatment — some respond more strongly, some less.

---

### **1.6 Correlation Between Random Intercepts and Slopes**

```{r Correlation}
vc <- VarCorr(m_nested)
vc
attr(vc$school, "correlation")[1, 2]
```

> This value shows how much the intercepts and slopes co-vary. A positive number means that schools with higher baselines tend to have stronger treatment effects, and vice versa. Here, we have a strong negative correlation, meaning that schools with lower baselines will have stronger treatment effects.

---

## **Workflow 2: Crossed Random Effects (Students × Raters)**

Now we simulate a **crossed** structure, specifying that each student is rated by multiple raters. Just like we've done, we will:

- Simulate a dataset where **each student can be rated by multiple raters**.
- We will maintain the two hypothetical treatments namely **control** and **treated**.
- Remeber, **control** and **treated** can be interpreted as two distinct sets of students (e.g., male vs female; undergraduate vs graduate, etc)

### **2.1 Data Generation for Crossed Model**

```{r Data simulation Crossed}
set.seed(123)
n_students <- 40
n_raters <- 6

df_crossed <- expand.grid(
  student = factor(1:n_students),
  rater = factor(1:n_raters),
  treatment = factor(c("control","treated"))
)

rater_intercepts <- rnorm(n_raters, 0, 2)
rater_slopes <- rnorm(n_raters, 3, 0.7)

df_crossed$score <- 50 +
  rater_intercepts[as.numeric(df_crossed$rater)] +
  rater_slopes[as.numeric(df_crossed$rater)] * (df_crossed$treatment == "treated") +
  rnorm(nrow(df_crossed), 0, 2)

head(df_crossed)
str(df_crossed)
```

**score** = Response variable

**treatment** = Predictor variable as before

**student** = Grouping variable *A*

**rater** = Grouping variable *B*

---

### **2.2 Fitting the Crossed Model**

We will specify tw independent grouping factors, necessitate a crossed model structure

```{r Fit model Crossed}
m_crossed <- lmer(score ~ treatment + (1 + treatment | rater) + (1 | student), data = df_crossed)
summary(m_crossed)
```

#### **Interpretation**

- Random intercepts/slopes for raters mean each rater interprets the treatment differently.
- Random intercept for students adjusts for repeated measures per student.

---

### **2.3 Extracting Rater Effects**

```{r Extracting varying slope Crossed}
fix <- fixef(m_crossed)
ranefs <- ranef(m_crossed)$rater

rater_effects <- data.frame(
  rater = rownames(ranefs),
  intercept = fix["(Intercept)"] + ranefs[,"(Intercept)"],
  treatment_effect = fix["treatmenttreated"] + ranefs[,"treatmenttreated"]
)

rater_effects
```

---

### **2.4 Visualizing Rater Treatment Effects**

```{r Dot plot of treatment effect Crossed}
ggplot(rater_effects, aes(x = rater, y = treatment_effect)) +
  geom_point(size = 3) +
  geom_hline(yintercept = fix["treatmenttreated"], linetype = "dashed", color = "red") +
  labs(y = "Estimated treatment effect", title = "Rater-specific treatment effects") +
  theme_minimal()
```

---

### **2.5 Predicted Scores by Rater**

```{r Spaghetti plot of fitted line Crossed}
df_crossed$pred <- predict(m_crossed)

ggplot(df_crossed, aes(x = treatment, y = pred, group = rater, color = rater)) +
  stat_summary(fun = mean, geom = "point", size = 3, position = position_dodge(0.3)) +
  stat_summary(fun = mean, geom = "line", aes(group = rater), size = 1) +
  labs(y = "Predicted score", title = "Varying treatment effects by rater") +
  theme_minimal()
```

> Each rater interprets “treatment” differently, leading to variability in slopes.

---

### **2.6 Random Effect Correlation**

```{r Correlation Crossed}
VarCorr(m_crossed)
```

> The covariance section again shows how intercepts and slopes for raters co-vary.Here, there is a weak positive correlation, meaning that raters with higher baselines have stronger treatment effects.

---

## **Summary**

Let's do a quick recap:

- **Random intercepts** capture baseline variability across groups.  
- **Random slopes** capture variability in the *effect of a predictor* across groups.  
- Correlation between intercepts and slopes tells whether groups with higher baselines show stronger or weaker responses.  
- Mixed models balance flexibility (allowing variation) with efficiency (borrowing strength across groups).

---

## **Hands-On Exercise**

Let's do a quick class work to demonstrate what we have learnt so far. Here, we are going to use a real dataset present in the **lme4** package.

### **Objective:**
Fit and interpret a mixed-effects model with random slopes and intercepts using a new dataset.

### **Dataset:**
Use the built-in `sleepstudy` dataset in `lme4`. Run `?sleepstudy` to understand that dataset.

```{r}
data("sleepstudy")
head(sleepstudy)
```

### **Tasks:**

1. Fit a model predicting `Reaction` as a function of `Days` with random intercepts and slopes for `Subject`:
   ```r
   m_exercise <- lmer(Reaction ~ Days + (1 + Days | Subject), data = sleepstudy)
   summary(m_exercise)
   ```

2. Interpret:
   - What do the fixed effects mean?
   - How do random intercepts and slopes vary across subjects?
   - What does the correlation between them imply?

3. Plot subject-specific slopes using:
   ```r
   coef(m_exercise)$Subject
   ```

4. Discuss:  
   - Why do some subjects show stronger or weaker effects of `Days` on `Reaction`?  
   - What happens if we remove the random slope term?

---

