---
title: "Lab 1, Act 1: Diagnose"
subtitle: "How does predation risk depend on tadpole size?"
author: "NRES 746 -- name: ____________________    group: ______"
date: ""
output:
  word_document: default
  # pdf_document: default   # uncomment (and comment out word_document) to knit to PDF instead
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Data ---------------------------------
frogs <- read.csv("../data/ReedfrogSizepred.csv")
```

<!-- ============================================================
INSTRUCTOR NOTES (delete this block before printing for students)

Pacing (45 min total for Act 1):
 - 15 min individual: pp. 1-3 (~8 min Part 1 numeric diagnosis, ~4 min
   Part 2 visual diagnosis, ~3 min Part 3 candidate shapes)
 - 15 min group: compare answers, esp. shape classification + candidate functions
 - 15 min revise + written reflection

Dataset: Vonesh & Bolker (2005) size-predation experiment, via Bolker's
emdbook::ReedfrogSizepred. TBL = tadpole body length (mm), Kill = number
of 10 tadpoles killed by dragonfly larvae in a 3-day trial. n=15 (5 size
classes x 3 replicate tanks). This is the same dataset behind Bolker Fig.
3.1 -- students will see it again in Lab 2 when we fit the Ricker curve
to it formally.

Part 1 is deliberately light on rote computation: students are handed the
group means/variances directly rather than computing all 5 by hand, and
instead have to interpret and compare them (does body size matter? do
variances differ, and why might some be zero?). This mirrors the reduced-
scaffolding philosophy used in Act 2 -- some productive struggle and
disagreement here is fine and gets resolved at the 15-min group
discussion. The 1a "pick one row and verify by hand" spot-check is
deliberately light-touch and will likely differ by student -- that's a
feature, not a gap: groups can pool their spot-checks at discussion time
to cross-validate more of the table together than any one student did
alone.

Feel free to swap in blank grid paper for the histogram/scatter panels if
you'd rather students sketch on separate graph paper.
============================================================ -->

## The scenario

Vonesh and Bolker (2005) studied *Hyperolius spinigularis* tadpoles in Tanzania, which face a trade-off: hatching earlier means hatching smaller, and smaller tadpoles may be more vulnerable to predation by dragonfly larvae -- or they may be small enough to be ignored. To find out, they ran a field experiment: tadpoles of five different sizes were placed into tanks with a dragonfly predator, and after three days the researchers counted how many of the 10 tadpoles in each tank had been killed.

**The data below are real measurements from that experiment.** Today we'll use them to practice the math and data-exploration tools you'll need throughout the semester -- starting with the most basic question a data analyst always asks first: *what does the pattern look like?*

```{r data-table}
knitr::kable(frogs, col.names = c("Tadpole size, TBL (mm)", "Number killed (out of 10)"),
             caption = "Vonesh & Bolker (2005) size-predation data")
```


## Part 1: Diagnose numerically (individual, ~8 min)

It's tempting to jump straight to a plot, but real data analysis usually starts with the numbers. Below is the mean and variance of "number killed" for each of the five tank sizes (across the 3 replicate tanks per size).

```{r summary-table}
kv <- aggregate(Kill ~ TBL, data = frogs, FUN = function(x) paste(x, collapse = ", "))
mn <- aggregate(Kill ~ TBL, data = frogs, FUN = mean)
vr <- aggregate(Kill ~ TBL, data = frogs, FUN = var)
tab <- data.frame(TBL = kv$TBL, Kill = kv$Kill, Mean = round(mn$Kill, 2), Var = round(vr$Kill, 2))
knitr::kable(tab, col.names = c("TBL (mm)", "Kill values (3 tanks)", "Mean", "Variance"),
             caption = "Summary statistics by tadpole size")
```

**1a. Spot-check.** Pick *one* row above and verify its mean and variance by hand below, just to confirm where these numbers come from. (Reminder: $\bar{x} = \frac{1}{n}\sum x_i$, &nbsp; $s^2 = \frac{1}{n-1}\sum (x_i-\bar{x})^2$.)

Row I'm checking: TBL = ________

____________________________________________________________
____________________________________________________________

**1b.** Does body size (TBL) appear to matter for predation risk? Describe the pattern in the means across the five sizes.

____________________________________________________________
____________________________________________________________

**1c.** Now compare the variances. Are they roughly similar across groups, or do some vary a lot more than others? Do you notice any relationship between a group's mean and its variance? Any guesses why some groups have variance $=0$?

____________________________________________________________
____________________________________________________________
____________________________________________________________

**1d. Sketch a histogram.** Using the full set of 15 individual tank counts (not the group means), sketch a histogram of "number killed" below, then describe its shape in a sentence or two.

```{r blank-histogram, fig.height=3, fig.width=5}
plot(NA, xlim = c(-0.5, 5.5), ylim = c(0, 8), xlab = "Number killed (out of 10)",
     ylab = "Frequency", xaxt = "n", yaxt = "n")
axis(1, at = 0:5)
axis(2, at = 0:8)
abline(v = seq(-0.5, 5.5, by = 1), col = "grey85")
abline(h = 0:8, col = "grey85")
box()
```

____________________________________________________________
____________________________________________________________

## Part 2: Diagnose visually (individual, ~4 min)

Below is a scatterplot of all 15 tanks: tadpole size (TBL) on the x-axis, number killed on the y-axis. (Points are jittered slightly left-right so overlapping tanks are visible.)

```{r scatter, fig.height=3.2, fig.width=5.5}
set.seed(1)
x_j <- frogs$TBL + runif(nrow(frogs), -0.6, 0.6)
plot(x_j, frogs$Kill, xlim = c(5, 41), ylim = c(-0.3, 5.3),
     xlab = "Tadpole size, TBL (mm)", ylab = "Number killed (out of 10)",
     pch = 21, bg = "grey70", cex = 1.4)
grid()
```

**Sketch a smooth curve by hand directly on the plot above**, showing what you think the underlying trend looks like (ignore tank-to-tank noise -- draw the *pattern*, not a connect-the-dots line).

**Classify the pattern.** Circle whichever of these best describes what you drew:

> monotonic increasing &nbsp;&nbsp;|&nbsp;&nbsp; monotonic decreasing &nbsp;&nbsp;|&nbsp;&nbsp; saturating (rises then flattens) &nbsp;&nbsp;|&nbsp;&nbsp; hump-shaped (rises then falls) &nbsp;&nbsp;|&nbsp;&nbsp; roughly flat / no clear trend

In 1-2 sentences, what *ecological* story would explain the pattern you circled? (Hint: think about what small vs. large tadpoles might be able to do that changes their risk.)

____________________________________________________________
____________________________________________________________
____________________________________________________________
____________________________________________________________


## Part 3: Name some candidates

We haven't learned the equations for these yet (that's Lab 2!) -- but you've almost certainly seen shapes like these before. Here's a quick reference gallery of function shapes that come up constantly in ecology:

```{r function-gallery, fig.height=3.6, fig.width=7}
par(mfrow = c(2, 3), mar = c(2, 2, 2, 1))
x <- seq(0.01, 10, length.out = 200)

# Linear ---------------------------------
plot(x, 1 + 0.5 * x, type = "l", lwd = 2, main = "Linear",
     xlab = "", ylab = "", xaxt = "n", yaxt = "n")

# Exponential decay ---------------------------------
plot(x, 5 * exp(-0.4 * x), type = "l", lwd = 2, main = "Exponential decay",
     xlab = "", ylab = "", xaxt = "n", yaxt = "n")

# Saturating (Michaelis-Menten) ---------------------------------
plot(x, 5 * x / (2 + x), type = "l", lwd = 2, main = "Saturating",
     xlab = "", ylab = "", xaxt = "n", yaxt = "n")

# Hump-shaped (Ricker) ---------------------------------
plot(x, 3 * x * exp(-0.4 * x), type = "l", lwd = 2, main = "Hump-shaped (Ricker)",
     xlab = "", ylab = "", xaxt = "n", yaxt = "n")

# Sigmoid / logistic ---------------------------------
plot(x, 5 / (1 + exp(-(x - 5))), type = "l", lwd = 2, main = "Sigmoid",
     xlab = "", ylab = "", xaxt = "n", yaxt = "n")

# U-shaped ---------------------------------
plot(x, 0.15 * (x - 5)^2 + 0.5, type = "l", lwd = 2, main = "U-shaped",
     xlab = "", ylab = "", xaxt = "n", yaxt = "n")
```

Based on the shape you circled in Part 2, which **one or two** of these gallery shapes seem like plausible candidates for the tadpole data? List them, and briefly say why you ruled out the others:

____________________________________________________________
____________________________________________________________
____________________________________________________________
____________________________________________________________

---

## Group discussion (15 min)

Compare your answers with your group:

- Compare which row each of you spot-checked in Part 1 -- between your group, how many of the 5 rows did you collectively verify, and did your numbers agree?
- Did everyone read the same pattern in the means and variances? Resolve any disagreements about what the numbers are telling you.
- Did everyone classify the pattern the same way in Part 2? If not, what in the data is each person pointing to?
- As a group, agree on your top candidate shape(s) from Part 3. What would it take to convince a skeptic that this shape -- and not one of the others -- is the right one?

## Revise and reflect (individual, ~15 min)

Revise any of your answers above based on the group discussion (it's fine to cross out and rewrite -- I want to see your thinking evolve). Then answer:

1. What is your group's best candidate function shape for this pattern, and what one feature of the data is the strongest evidence for it?

____________________________________________________________
____________________________________________________________

2. What would you still need to know (mathematically) to go from "this looks hump-shaped" to an actual equation you could fit to the data? (No wrong answers -- "I don't know" is a fine answer here.)

____________________________________________________________
____________________________________________________________

3. Muddiest point: what's the one thing from this part of the lab you'd like to see explained again?

____________________________________________________________
____________________________________________________________
