---
title: "Lab 1, Act 2"
subtitle: "Turning a candidate shape into a testable equation"
author: "NRES 746 -- name: ____________________    group: ______"
date: ""
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 0.3)

# Data ---------------------------------
frogs <- read.csv("../data/ReedfrogSizepred.csv")

# Candidate Ricker curve ---------------------------------
a <- 0.9; b <- 1/12
ricker <- function(x, a, b) a * x * exp(-b * x)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}
```

## Picking up where we left off

In Act 1, your group likely diagnosed the tadpole predation data as **hump-shaped** and named the Ricker function, $y = a\,x\,e^{-bx}$, as a plausible candidate. Today we make that candidate concrete: we'll pick specific numbers for $a$ and $b$, evaluate the function, and use calculus to find exactly where it peaks -- then check whether that matches what you eyeballed last time.

Before we get there, a quick warm-up on the algebra tool we'll lean on all semester: logarithms and exponents.

## Part 1: Log & exponent warm-up (individual, ~10-15 min)

**Useful rules:**

$$\log(xy) = \log x + \log y \qquad\qquad \log\!\left(\frac{x}{y}\right) = \log x - \log y \qquad\qquad \log(x^k) = k\log x$$

$$e^{\ln x} = x \qquad\qquad \ln(e^x) = x \qquad\qquad \ln x \approx 2.303 \times \log_{10} x$$

**1a. Apply the rules.** Simplify each:

$\log_{10}(4) + \log_{10}(25) =$ ________

$\log_{10}(1000) - \log_{10}(10) =$ ________

$\log_{10}(2^{10}) =$ ________ &nbsp;&nbsp;(you'll need $\log_{10}(2) \approx 0.301$)

**1b. Solve for x:**

$4^x = 64 \;\Rightarrow\; x =$ ________ &nbsp;&nbsp;&nbsp;&nbsp; $3^x = 81 \;\Rightarrow\; x =$ ________ &nbsp;&nbsp;&nbsp;&nbsp; $10^x = 0.001 \;\Rightarrow\; x =$ ________

$e^x = 20 \;\Rightarrow\; x =$ ________

**1c. Linearizing an exponential decay.** This is the trick that will let us fit exponential-type models with plain linear regression later this semester. Take natural logs of both sides of $y = a e^{-bx}$ and simplify using the rules above. Show your work:

```{r fig.height=0.3*2}
blank_space()
```

$\ln y =$ ________________________

What form is this, as a function of $x$? What do the slope and intercept correspond to, in terms of $a$ and $b$?

```{r}
blank_lines()
```

Now check your answer against a small numeric example: $y = 10\,e^{-0.5x}$ (so $a=10$, $b=0.5$). Fill in the table below.

*Useful values:* $e^{-1} \approx 0.368$, &nbsp; $e^{-2} \approx 0.135$, &nbsp; $e^{-3} \approx 0.050$, &nbsp; $\ln(10) \approx 2.303$

| $x$ | $y = 10e^{-0.5x}$ | $\ln(y)$ |
|:---:|:---:|:---:|
| 0 |  |  |
| 2 |  |  |
| 4 |  |  |
| 6 |  |  |

Compute the change in $\ln(y)$ each time $x$ increases by 2: ________. Does it match what you predicted for the slope above? ________

## Part 2: Derive properties of the Ricker function (individual, ~20-25 min)

**Step 1: Today's candidate.** Let's put specific numbers on the Ricker shape from Act 1:

$$y = a\,x\,e^{-bx}, \qquad a = 0.9, \qquad b = \frac{1}{12} \text{ mm}^{-1}$$

(These are illustrative, chosen to [very!] roughly match the pattern you saw -- not a formal fit. We'll learn how to actually fit a curve to data soon!)

```{r candidate-curve, fig.height=3.4, fig.width=5.5}
set.seed(1)
x_j <- frogs$TBL + runif(nrow(frogs), -0.6, 0.6)
plot(x_j, frogs$Kill, xlim = c(5, 40), ylim = c(-0.3, 5.3),
     xlab = "Tadpole size, TBL (mm)", ylab = "Number killed (out of 10)",
     pch = 21, bg = "grey70", cex = 1.4)
curve(ricker(x, a, b), from = 5, to = 40, lwd = 2, col = "steelblue", add = TRUE)
grid()
```

**Step 2: Evaluate by hand.** Using $a=0.9$ and $b=1/12$, compute $y$ at three tadpole sizes. *Useful values:* $e^{-0.5}\approx 0.607$, $e^{-1}\approx 0.368$, $e^{-2}\approx 0.135$.

$y(6) =$ ________ &nbsp;&nbsp;&nbsp;&nbsp; $y(12) =$ ________ &nbsp;&nbsp;&nbsp;&nbsp; $y(24) =$ ________

Do these three values rise then fall, consistent with the hump shape? ________

**Step 3: Take the derivative.** Find $dy/dx$ for $y = a\,x\,e^{-bx}$. This is a *product* of two pieces of $x$, so you'll need the product rule together with the chain rule for the exponential term.

*Reminders:* $\dfrac{d}{dx}(uv) = u'v + uv'$ &nbsp;&nbsp;&nbsp;&nbsp; $\dfrac{d}{dx}\,e^{g(x)} = g'(x)\,e^{g(x)}$

Show your work:

```{r fig.height=0.3*3}
blank_space()
```

$\dfrac{dy}{dx} =$ ________________________

What is the derivative of f(x) at x=0? What does this tell you about the meaning of the parameter "a" in the Ricker equation?

```{r}
blank_lines()
```

**Step 4: Find the critical point.** Set $dy/dx = 0$ and solve for the critical point $x^*$, using our candidate's $b = 1/12$. (Hint: think about which factor(s) in your derivative could actually equal zero.)

Show your work:

```{r fig.height=0.3*2}
blank_space()
```

$x^* =$ ________ mm

**Step 5: Interpret.** In a sentence, what does $x^*$ mean ecologically?

```{r fig.height=0.3*2}
blank_lines(2)
```

How do you know $x^*$ is a peak and not a valley? (Hint: you could use the second derivative test but you don't need to!)

```{r fig.height=0.3*2}
blank_lines(2)
```

If you were to compute the second derivative at the critical point, would it be positive or negative?

```{r fig.height=0.3}
blank_lines()
```

## Part 3: A first taste of integration (individual, ~8-12 min)

Calculus has two main tools: derivatives (rates of change, which we just used to find a peak) and **integrals** (accumulation -- adding up a rate over some interval to get a total). We'll use this idea much more when we work with continuous probability distributions.

**Worked example.** Suppose -- just for this simple example, forget the Ricker curve -- that dragonfly larvae capture tadpoles at a *constant* rate of $r = 2$ tadpoles/day for the entire $T=3$-day trial. The total number killed is rate $\times$ time:

$$\text{Total killed} = r \times T = 2 \times 3 = 6 \text{ tadpoles}$$

```{r constant-rate, fig.height=2.6, fig.width=4.5}
plot(NA, xlim = c(0, 3), ylim = c(0, 3), xlab = "Time (days)",
     ylab = "Capture rate (tadpoles/day)", xaxt = "n", yaxt = "n")
axis(1, at = 0:3); axis(2, at = 0:3)
polygon(c(0, 3, 3, 0), c(0, 0, 2, 2), col = "grey85", border = NA)
segments(0, 2, 3, 2, lwd = 2)
box()
```

Geometrically, that's just the **area of the shaded rectangle**: base (time) $\times$ height (rate). "Area under the rate curve = cumulative total" is exactly the idea behind integration -- we'll just need it for curves that aren't flat lines.

**Your turn.** Now suppose the capture rate isn't constant -- it ramps up *linearly*, from 0 tadpoles/day at $t=0$ to 2 tadpoles/day at $t=3$ (dragonflies take some time to find the tadpoles). Sketch this rate on the blank axes below, shade the region representing the total killed, and compute the total using the area of a triangle.

```{r linear-rate-blank, fig.height=2.6, fig.width=4.5}
plot(NA, xlim = c(0, 3), ylim = c(0, 3), xlab = "Time (days)",
     ylab = "Capture rate (tadpoles/day)", xaxt = "n", yaxt = "n")
axis(1, at = 0:3); axis(2, at = 0:3)
abline(v = 0:3, col = "grey85"); abline(h = 0:3, col = "grey85")
box()
```

$\text{Total killed} =$ ________

---

## Group discussion and revision (10-15 min)

Compare your work with your group -- especially your derivative and critical point $x^*$. Feel free to revise any of your answers above based on the group discussion (it's fine to cross out and rewrite -- I like to see your thinking evolve).

Please take a minute to record any changes you made and any new insights gained during the group discussion. You can also use this space to reflect on any areas of confusion ("muddiest points") or topics you want to review or dust off going forward.

One additional discussion point: do you think you could find better-fitting parameters for the Ricker function than the ones we guessed above? 

```{r fig.height=0.3*8}
blank_lines(8)
```
