---
title: "Lab 3, Act 1"
subtitle: "Trusting a positive detection"
author: "NRES 746 -- name: _______________   group: _____________________"
date: ""
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}

# draws a table with given headers/body for handwritten answers ("" = blank
# cell); headers/values may be plotmath expressions for math; pair with
# fig.height = header_height + nrow(body) * row_height
blank_table <- function(headers, body, col_widths = NULL,
                         header_height = 0.35, row_height = 0.5,
                         cex = 0.95, bold_rows = integer(0)) {
  nc <- length(headers); nr <- length(body)
  if (is.null(col_widths)) col_widths <- rep(1, nc)
  col_widths <- col_widths / sum(col_widths)
  cx <- c(0, cumsum(col_widths))
  h_total <- header_height + nr * row_height
  par(mar = c(0, 0, 0, 0), family = "serif")
  plot(NA, xlim = c(0, 1), ylim = c(0, h_total), axes = FALSE, xlab = "", ylab = "",
       xaxs = "i", yaxs = "i")
  y <- h_total
  segments(0, y, 1, y, lwd = 1.5)
  y <- y - header_height
  segments(0, y, 1, y, lwd = 1.5)
  for (i in seq_len(nr)) {
    y <- y - row_height
    segments(0, y, 1, y, lwd = if (i == nr) 1.5 else 0.5,
             col = if (i == nr) "black" else "grey70")
  }
  for (v in cx) segments(v, 0, v, h_total, lwd = 1)
  yh <- h_total - header_height / 2
  for (j in seq_len(nc)) text((cx[j] + cx[j + 1]) / 2, yh, labels = headers[[j]], font = 2, cex = cex)
  y <- h_total - header_height
  for (i in seq_len(nr)) {
    yc <- y - row_height / 2
    fnt <- if (i %in% bold_rows) 2 else 1
    for (j in seq_len(nc)) {
      val <- body[[i]][[j]]
      if (!identical(val, "")) text((cx[j] + cx[j + 1]) / 2, yc, labels = val, font = fnt, cex = cex)
    }
    y <- y - row_height
  }
}
```

## The scenario

A regional wildlife agency is monitoring a rare forest carnivore using
motion-triggered camera traps. Most trigger events aren't the target
species at all (deer, rabbits, blowing branches, and hikers set off the
cameras far more often than the target animal). To keep field
biologists from having to review every single image by eye, the agency
uses an automated image classifier that flags each trigger event as
either "target species detected" or not.

The classifier isn't perfect. For this exercise, suppose:

```{r params-table}
params <- data.frame(
  Quantity = c("Prevalence: fraction of all trigger events that are actually the target species",
               "Sensitivity: P(flagged | actually the species)",
               "False positive rate: P(flagged | not the species)"),
  Value = c("1 in 10,000 (0.0001)", "0.98", "0.01")
)
knitr::kable(params, col.names = c("Quantity", "Value"))
```

That sounds like a pretty good classifier, since it catches 98% of true
detections and only mistakenly flags a "background" image not
containing the species 1% of the time. **If the classifier flags an
image, how worried should the field team be that it's a false alarm?**

You haven't been given a formula for this yet, but let's see if we can figure it out together!

## Part 1: Count your way to an answer (individual, ~10 min)

Formulas are easier to trust once you've built one yourself. Start
instead with plain counting. Suppose the cameras log $N=100{,}000$
trigger events over a season.

**1a.** Fill in the table below using only the rates given above. Every
blank cell is just one of those rates applied to a total you already
know (e.g., how many of the 10 true-species events do you expect the
classifier to flag, given its sensitivity?). Round to the nearest whole
number.

```{r fill-freq-table, fig.height=0.35+3*0.5, fig.width=5.5}
blank_table(
  headers = list("", "Flagged", "Not flagged", "Total"),
  body = list(
    list("Actually the species", "", "", "10"),
    list("Not the species", "", "", "99,990"),
    list("Total", "", "", "100,000")
  ),
  col_widths = c(1.6, 1, 1.1, 0.9),
  cex = 0.85
)
```

**1b.** Of the images the classifier *flagged*, what fraction are
actually the target species? No formula needed -- just use the counts
in your completed table.

```{r fig.height=0.3*2}
blank_lines(2)
```

**1c.** Is this number close to what you expected, given how good the
classifier's sensitivity and false-positive rate sounded at the start?
In a sentence or two, what would you tell the field team about how to
handle a single flagged image?

```{r fig.height=0.3*3}
blank_lines(3)
```

## Part 2: Turn your counting into a general recipe (individual, ~10 min)

Your table only works for exactly 100,000 events. Let's turn the same
reasoning into something that works for *any* number of events.

**2a.** Rewrite the two counts you used in 1b -- the number of flagged
images that are truly the species, and the total number of flagged
images -- as expressions using $N$ and the rates given at the top
(prevalence, sensitivity, false positive rate), instead of plugged-in
numbers.

```{r fig.height=0.3*6}
blank_space()
```

**2b.** Divide your two expressions from 2a to reconstruct the fraction
from 1b. $N$ should appear in both the top and bottom -- cancel it out.
What are you left with?

```{r fig.height=0.3*6}
blank_space()
```

**2c.** State your result from 2b as a general rule, in your own words
or symbols: given a classifier's sensitivity, false-positive rate, and
the prevalence of the thing it's looking for, how do you compute the
chance that a flagged event is real?

```{r fig.height=0.3*3}
blank_lines(3)
```

## Part 3: Try it somewhere else (individual, ~5 min)

A second survey site monitors the same species with the *same*
classifier (same sensitivity and false-positive rate), but the species
is well-established there rather than rare: prevalence $=0.01$ (1 in
100 trigger events), instead of the original site's 1 in 10,000.

**3a.** Using your recipe from Part 2 (or a fresh counting table, if
you'd rather), find this second site's PPV -- the chance that a flagged
image is real.

```{r fig.height=0.3*6}
blank_space()
```

**3b.** Same classifier, same sensitivity, same false-positive rate, but a different answer. What does comparing these two sites tell you about how much those two classifier properties matter versus how common (prevalent) the thing you're looking for actually is? We haven't talked about it yet, but feel free to use the term "base rate" to describe the prevalence (e.g., of the target predator) and speculate on how your findings today relate to a common logical fallacy known as the "base rate fallacy". 

```{r fig.height=0.3*3}
blank_lines(3)
```

## Group discussion and revision (~5 min)

Compare your Part 2 recipe and your Part 3 answer with your group.
Does everyone's recipe agree once you strip away notation differences?
If not, find where they diverge.

Feel free to revise any of your answers above based on the discussion.

Please take a minute to record any changes you made and any new
insights gained during the group discussion. You can also use this
space to reflect on any areas of confusion ("muddiest points") or
topics you want to review or dust off.

```{r fig.height=0.3*6}
blank_lines(6)
```
