---
title: "Lab 3, Act 1"
subtitle: "Trusting a positive detection"
author: "NRES 746 -- name: ____________________    group: ______"
date: ""
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}

# draws a table with given headers/body for handwritten answers ("" = blank
# cell); headers/values may be plotmath expressions for math; pair with
# fig.height = header_height + nrow(body) * row_height
blank_table <- function(headers, body, col_widths = NULL,
                         header_height = 0.35, row_height = 0.5,
                         cex = 0.95, bold_rows = integer(0)) {
  nc <- length(headers); nr <- length(body)
  if (is.null(col_widths)) col_widths <- rep(1, nc)
  col_widths <- col_widths / sum(col_widths)
  cx <- c(0, cumsum(col_widths))
  h_total <- header_height + nr * row_height
  par(mar = c(0, 0, 0, 0), family = "serif")
  plot(NA, xlim = c(0, 1), ylim = c(0, h_total), axes = FALSE, xlab = "", ylab = "",
       xaxs = "i", yaxs = "i")
  y <- h_total
  segments(0, y, 1, y, lwd = 1.5)
  y <- y - header_height
  segments(0, y, 1, y, lwd = 1.5)
  for (i in seq_len(nr)) {
    y <- y - row_height
    segments(0, y, 1, y, lwd = if (i == nr) 1.5 else 0.5,
             col = if (i == nr) "black" else "grey70")
  }
  for (v in cx) segments(v, 0, v, h_total, lwd = 1)
  yh <- h_total - header_height / 2
  for (j in seq_len(nc)) text((cx[j] + cx[j + 1]) / 2, yh, labels = headers[[j]], font = 2, cex = cex)
  y <- h_total - header_height
  for (i in seq_len(nr)) {
    yc <- y - row_height / 2
    fnt <- if (i %in% bold_rows) 2 else 1
    for (j in seq_len(nc)) {
      val <- body[[i]][[j]]
      if (!identical(val, "")) text((cx[j] + cx[j + 1]) / 2, yc, labels = val, font = fnt, cex = cex)
    }
    y <- y - row_height
  }
}
```

## The scenario

A regional wildlife agency is monitoring a rare forest carnivore using
motion-triggered camera traps. Most trigger events aren't the target
species at all (deer, rabbits, blowing branches, and hikers set off
the cameras more often than the target animal.
To keep field biologists from having to review every single image by
eye, the agency uses an automated image classifier that flags each
trigger event as either "target species detected" or not.

The classifier isn't perfect. For this exercise, suppose:

```{r params-table}
params <- data.frame(
  Quantity = c("Prevalence: fraction of all trigger events that are actually the target species",
               "Sensitivity: P(flagged | actually the species)",
               "False positive rate: P(flagged | not the species)"),
  Value = c("1 in 10,000 (0.0001)", "0.98", "0.01")
)
knitr::kable(params, col.names = c("Quantity", "Value"))
```

That sounds like a pretty good classifier, since it catches 98% of true
detections and only mistakenly flags a 'background' image not containing the species 
1% of the time. **If the classifier flags an image, how worried should the field team be
that it's a false alarm?**

## Part 1: Compute it with Bayes' rule (individual, ~15 min)

Let $S$ = "trigger event is actually the target species" and $F$ =
"classifier flags the image." You want $P(S \mid F)$.

**1a.** Compute the two joint probabilities you'll need: $P(F \cap S)$
and $P(F \cap \lnot S)$.

```{r fig.height=0.3*5}
blank_space()
```

**1b.** Use the law of total probability to combine your two joint
probabilities from 1a into the unconditional probability of a flag,
$P(F)$.

```{r fig.height=0.3*3}
blank_space()
```

**1c.** Now apply Bayes' rule to compute $P(S \mid F)$ -- that is, the
probability that a flagged image is a true detection.

```{r fig.height=0.3*4}
blank_space()
```

**1d.** Is this number close to what you expected given the
sensitivity and false-positive rate reported above? In a sentence or
two, what would you tell the field team about how to handle a single
flagged image?

```{r fig.height=0.3*3}
blank_lines(3)
```

## Part 2: The same answer, a different way (individual, ~5-8 min)

Bayes' rule can feel abstract. Natural frequencies -- working with
whole counts of events instead of probabilities -- often make the same
calculation feel more concrete. Suppose the cameras log 100,000 trigger
events over a season. Fill in the table below using the prevalence,
sensitivity, and false-positive rate given above (round to the nearest
whole number where needed).

```{r fill-freq-table, fig.height=0.35+3*0.5, fig.width=5.5}
blank_table(
  headers = list("", "Flagged", "Not flagged", "Total"),
  body = list(
    list("Actually the species", "", "", "10"),
    list("Not the species", "", "", "99,990"),
    list("Total", "", "", "100,000")
  ),
  col_widths = c(1.6, 1, 1.1, 0.9),
  cex = 0.85
)
```

**2a.** From your completed table, read off $P(S \mid F)$ directly (true
detections among flagged images, divided by all flagged images). Does
it match your answer to 1c?

```{r fig.height=0.3*2}
blank_lines(2)
```

## Group discussion and revision (~5 min)

Compare your answers and your natural-frequency table with your group.
One thing worth discussing: prevalence is doing almost all the work in
this result, more than sensitivity or false-positive rate -- does that
match your intuition, and can you point to exactly where in the
calculation that shows up?

Feel free to revise any of your answers above based on the discussion
(it's fine to cross out and rewrite -- I like to see your thinking
evolve).

Please take a minute to record any changes you made and any new
insights gained during the group discussion. You can also use this
space to reflect on any areas of confusion ("muddiest points") or
topics you want to review or dust off going forward.

```{r fig.height=0.3*6}
blank_lines(6)
```
