---
title: "Lab 3, Act 3"
subtitle: "From counts to a continuous measurement"
author: "NRES 746 -- name: _______________   group: _____________________"
date: ""
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 0.3)

mu <- 4.5
sigma <- 0.6

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}
```

## Picking up where we left off

So far this lab has treated detections as discrete counts (0, 1, 2, ...
detections). But some of the quantities the field team records are
continuous. Suppose adult body mass for this carnivore, estimated from
reference objects visible in camera-trap images, is well described by a
normal distribution with mean $\mu = 4.5$ kg and standard deviation
$\sigma = 0.6$ kg.

## Part 1: Reading a density curve (individual, ~10 min)

Below is the probability density function (pdf) for this normal
distribution.

```{r pdf-blank, fig.height=3.2, fig.width=5.5}
curve(dnorm(x, mu, sigma), from = 2.4, to = 6.6, lwd = 2,
      xlab = "Body mass (kg)", ylab = "Density", xaxs = "i")
abline(v = mu, lty = 3, col = "grey60")
```

**1a. Shade it by hand.** On the curve above, shade the region that
represents $P(4.5 < \text{mass} < 5.5)$.

**1b. Integral notation.** Given that the probability density function
can be written $f(x)$ where x is body mass and f(x) is the normal (Gaussian)
density function (you can ignore the specific mathematical equation for the 
Gaussian density function for now), please write out the same region described above as a 
definite integral. 

```{r fig.height=0.3*3}
blank_space()
```

**1c.** What is $P(\text{mass} = 4.5)$ exactly? Why -- what's different
about a single point on a continuous curve versus an interval?

```{r fig.height=0.3*3}
blank_lines(3)
```

**1d.** The height of the curve at $\text{mass}=4.5$ is not zero (in
fact, it's the highest point on the whole curve). If that height isn't
a probability, what does it actually represent? HINT: the technical answer is a 
'probability density'. Try to write out in words what a 'probability density' means.  

```{r fig.height=0.3*3}
blank_lines(3)
```

## Part 2: Reading a CDF (individual, ~5-8 min)

Below is the cumulative distribution function (CDF), $F(x) = P(\text{mass} \le x)$,
for the same distribution.

```{r cdf-blank, fig.height=3.2, fig.width=5.5}
curve(pnorm(x, mu, sigma), from = 2.4, to = 6.6, lwd = 2,
      xlab = "Body mass (kg)", ylab = "F(x)", xaxs = "i")
abline(v = 5.5, lty = 3, col = "grey60")
abline(h = c(0, 1), lty = 3, col = "grey85")
```

**2a.** Using the CDF plot, estimate $P(\text{mass} > 5.5)$. Record your estimate -- you
don't need to be precise, just in the right neighborhood.

```{r fig.height=0.3*3}
blank_lines(3)
```

## Part 3: Standardize and look it up (individual, ~10-12 min)

Graph-reading only gets you so far. The **standard normal table**
(Table 1, at the end of this Act) gives exact values of
$\Phi(z) = P(Z \le z)$ for the *standard* normal distribution ($\mu=0$,
$\sigma=1$). To use it for a normal distribution with a different mean
and standard deviation, first convert your value of interest to a
**z-score**:
$$z = \frac{x - \mu}{\sigma}$$
This rescales $x$ into "number of standard deviations away from the
mean," (standardized difference from the mean) which is exactly what the table expects.

**3a.** Compute the z-score for $x=5.5$ kg.

```{r fig.height=0.3*3}
blank_space()
```

**3b.** Look up $\Phi(z)$ for the z-score you just computed (round $z$
to the nearest 0.01 if needed), then use it to compute
$P(\text{mass}>5.5)$. How close was this to your graph-reading estimate
from 2a?

```{r fig.height=0.3*4}
blank_space()
```

**3c.** Now go the other direction. Search the table for
the value closest to $0.90$, and read off the z-score (90% quantile). 
This is (approximately) the z-score marking the heaviest 10%
of the population (only 10% of individuals expected to exceed this standardized value). 
In the space below, convert that z-score back into a body mass.

```{r fig.height=0.3*5}
blank_space()
```

## Group discussion and revision (~5 min)

Compare your Part 3 answers with your group. One thing worth
discussing: Would this z-table give the right answer if you standardized 
Act 2's binomial detection counts the same way?

Feel free to revise any of your answers above based on the discussion
(it's fine to cross out and rewrite).

Please take a minute to record any changes you made and any new
insights gained during the group discussion. You can also use this
space to reflect on any areas of confusion ("muddiest points") or
topics you want to review or dust off.

```{r fig.height=0.3*6}
blank_lines(6)
```

\newpage

## Table 1: Standard normal cumulative probabilities, $\Phi(z) = P(Z \le z)$

Row = $z$ to one decimal place; column = second decimal place. Example:
for $z=1.67$, read the row "1.6" and the column ".07".

```{r ztable}
z_int <- seq(0, 3.4, by = 0.1)
z_dec <- seq(0, 0.09, by = 0.01)
ztab <- outer(z_int, z_dec, function(a, b) pnorm(a + b))
ztab <- round(ztab, 4)
rownames(ztab) <- sprintf("%.1f", z_int)
colnames(ztab) <- sprintf(".%02d", 0:9)
knitr::kable(ztab, align = "c")
```
