---
title: "Lab 3, Act 2"
subtitle: "Counting detections: binomial or Poisson?"
author: "NRES 746 -- name: ____________________    group: ______"
date: ""
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}
```

## Picking up where we left off

In Act 1 you dealt with *uncertainty about a single flagged image*.
Now set that aside -- assume genetic testing has confirmed the target
carnivore is genuinely present at a particular site, so there's no more
classifier doubt to worry about. The new question is about *detection*:
even when the animal is really there, a camera doesn't catch it every
single night. Suppose, based on many past deployments at similar sites,
biologists estimate that on any given night there's a $p=0.3$
probability the camera captures at least one image of the animal,
independent from night to night.

## Part 1: Binomial by hand (individual, ~15 min)

A camera is deployed for $N=8$ nights.

**1a.** Why is "number of nights with a detection, out of 8" a
reasonable binomial random variable here? Name the assumptions the
binomial distribution requires, and say briefly why each one seems
plausible (or not) for this camera.

```{r fig.height=0.3*4}
blank_lines(4)
```

**1b.** Using the binomial pmf, $P(X=k) = \binom{N}{k}p^k(1-p)^{N-k}$,
compute by hand the probability of exactly 3 detections in the 8
nights ($N=8$, $k=3$, $p=0.3$). Show your work.

```{r fig.height=0.3*8}
blank_space()
```

Below is the full distribution of detections across all 9 possible
outcomes ($k=0$ to $8$), for you to check your answer against and read
off the most likely outcome.

```{r binom-full, fig.height=3, fig.width=5}
barplot(dbinom(0:8, 8, 0.3), names.arg = 0:8, ylab = "Probability",
        xlab = "Number of detections (out of 8 nights)",
        main = "Binomial(N = 8, p = 0.3)")
```

**1c.** Does the bar at $k=3$ match your hand calculation from 1b?
Which number of detections is most likely (the tallest bar -- the
*mode*)?

```{r fig.height=0.3*2}
blank_lines(2)
```

## Part 2: Comparing to Poisson (individual, ~15-20 min)

A colleague suggests it would be simpler to just model detections with
a Poisson distribution instead, using the same mean,
$\lambda = 8\times0.3=2.4$. Below are two scenarios, each comparing a
binomial distribution against a Poisson with the same mean.

```{r pois-compare, fig.height=3.4, fig.width=6.5}
par(mfrow = c(1, 2))
barplot(dbinom(0:8, 8, 0.3), names.arg = 0:8, ylab = "Probability",
        main = "Scenario A:\nBinomial(N=8, p=0.3)", cex.main = 0.85)
barplot(dpois(0:8, 2.4), names.arg = 0:8, ylab = "Probability",
        main = "Scenario A:\nPoisson(lambda=2.4)", cex.main = 0.85)
```

```{r pois-compare2, fig.height=3.4, fig.width=6.5}
par(mfrow = c(1, 2))
barplot(dbinom(0:8, 100, 0.024), names.arg = 0:8, ylab = "Probability",
        main = "Scenario B:\nBinomial(N=100, p=0.024)", cex.main = 0.85)
barplot(dpois(0:8, 2.4), names.arg = 0:8, ylab = "Probability",
        main = "Scenario B:\nPoisson(lambda=2.4)", cex.main = 0.85)
```

**2a.** Just by eye: which scenario's pair of bar charts looks more
alike -- A or B?

```{r fig.height=0.3*2}
blank_lines(2)
```

**2b.** Compute the variance of each distribution by hand, using
$\text{Var} = Np(1-p)$ for the binomial and $\text{Var}=\lambda$ for
the Poisson.

```{r fig.height=0.3*8}
blank_space()
```

**2c.** Does the pattern of variances you just computed explain what
you saw by eye in 2a?

```{r fig.height=0.3*3}
blank_lines(3)
```

**2d.** Finish this rule of thumb in your own words: *"A Poisson
distribution is a good stand-in for a binomial distribution when..."*

```{r fig.height=0.3*3}
blank_lines(3)
```

**2e.** Back in the original 8-night scenario ($p=0.3$): name one
realistic reason the constant, independent per-night detection
probability assumption from 1a might be violated in the field (think
about weather, the animal's behavior, or the camera itself).

```{r fig.height=0.3*3}
blank_lines(3)
```

## Group discussion and revision (~5 min)

Compare your variances and rule of thumb from 2d with your group.

Feel free to revise any of your answers above based on the discussion
(it's fine to cross out and rewrite -- I like to see your thinking
evolve).

Please take a minute to record any changes you made and any new
insights gained during the group discussion. You can also use this
space to reflect on any areas of confusion ("muddiest points") or
topics you want to review or dust off going forward.

```{r fig.height=0.3*6}
blank_lines(6)
```
