---
title: "Lab 3, Act 3"
subtitle: "From counts to a continuous measurement"
author: "NRES 746 -- name: ____________________    group: ______"
date: ""
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 0.3)

mu <- 4.5
sigma <- 0.6

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}
```

## Picking up where we left off

So far this lab has treated detections as discrete counts (0, 1, 2, ...
detections). But some of the quantities the field team records are
continuous. Suppose adult body mass for this carnivore, estimated from
reference objects visible in camera-trap images, is well described by a
normal distribution with mean $\mu = 4.5$ kg and standard deviation
$\sigma = 0.6$ kg.

## Part 1: Reading a density curve (individual, ~10 min)

Below is the probability density function (pdf) for this normal
distribution.

```{r pdf-blank, fig.height=3.2, fig.width=5.5}
curve(dnorm(x, mu, sigma), from = 2.4, to = 6.6, lwd = 2,
      xlab = "Body mass (kg)", ylab = "Density", xaxs = "i")
abline(v = mu, lty = 3, col = "grey60")
```

**1a. Shade it by hand.** On the curve above, shade the region that
represents $P(4.5 < \text{mass} < 5.5)$.

**1b.** What is $P(\text{mass} = 4.5)$ exactly? Why -- what's different
about a single point on a continuous curve versus an interval?

```{r fig.height=0.3*3}
blank_lines(3)
```

**1c.** The height of the curve at $\text{mass}=4.5$ is not zero (in
fact, it's the highest point on the whole curve). If that height isn't
a probability, what does it actually represent?

```{r fig.height=0.3*3}
blank_lines(3)
```

## Part 2: Reading a CDF (individual, ~5-8 min)

Below is the cumulative distribution function (CDF), $F(x) = P(\text{mass} \le x)$,
for the same distribution.

```{r cdf-blank, fig.height=3.2, fig.width=5.5}
curve(pnorm(x, mu, sigma), from = 2.4, to = 6.6, lwd = 2,
      xlab = "Body mass (kg)", ylab = expression(F(x) == P(mass <= x)), xaxs = "i")
abline(v = 5.5, lty = 3, col = "grey60")
abline(h = c(0, 1), lty = 3, col = "grey85")
```

**2a.** Using the CDF plot, estimate $P(\text{mass} > 5.5)$ (read off
$F(5.5)$ as best you can, then subtract). Record your estimate -- you
don't need to be precise, just in the right neighborhood.

```{r fig.height=0.3*3}
blank_lines(3)
```

## Group check-in (group, ~5 min)

Compare your shaded region (1a) and your estimate (2a) with your group
and agree on a shared answer for 2a before opening laptops.

```{r fig.height=0.3*2}
blank_lines(2)
```

---

**Laptops open now.**

---

## Part 3: Check your estimate (laptops, ~10 min)

**3a.** Use `pnorm()` to compute $P(\text{mass} > 5.5)$ exactly. How
close was your group's graph-reading estimate from 2a?

```{r fig.height=0.3*2}
blank_lines(2)
```

**3b.** Use `qnorm()` to find the body mass value that marks the
heaviest 10% of the population (i.e., only 10% of individuals are
expected to be heavier than this value).

```{r fig.height=0.3*2}
blank_lines(2)
```

## Group discussion and revision (~5 min)

Compare your Part 3 answers with your group. One thing worth
discussing: `dnorm()`, `pnorm()`, and `qnorm()` all describe the same
distribution but answer different questions -- can your group state, in
one sentence each, what question each one answers?

Feel free to revise any of your answers above based on the discussion
(it's fine to cross out and rewrite -- I like to see your thinking
evolve).

Please take a minute to record any changes you made and any new
insights gained during the group discussion. You can also use this
space to reflect on any areas of confusion ("muddiest points") or
topics you want to review or dust off going forward.

```{r fig.height=0.3*6}
blank_lines(6)
```
