---
title: "Lab 4, Act 1"
subtitle: "If we knew the truth, would our trend analysis find it?"
author: "NRES 746 -- name: _______________   group: _____________________"
date: ""
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
header-includes:
  - \usepackage{needspace}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}

# keeps a prompt and its answer space on one PDF page: place `r nq(...)` on its
# own line just above the prompt. Emits nothing in non-LaTeX output. lines =
# lines of prompt text, rows = rows of blank space below it (0.3 in each),
# extra = any other height (in inches), e.g. a figure or table
nq <- function(lines, rows = 0, extra = 0) {
  if (!knitr::is_latex_output()) return("")
  knitr::asis_output(sprintf("\n\\needspace{%.2fin}\n", 0.23 * lines + 0.15 + 0.3 * rows + extra))
}
```

## The scenario

A state wildlife agency is planning a long-term monitoring program for a
rare, secretive toad in a remote Great Basin spring complex. The
management goal is to detect a population decline of 25% or more over 25
years. 

Here is the 'truth' for this made-up scenario, which the analyst of real-world data never sees:
each year a survey produces a count index (# individuals captured or observed in each survey)
$I_t$ for years $t = 1, 2, \ldots, 25$. The population declines by the
same proportion every year, and each year's count differs from the
trend by a random multiplicative factor (proportional observation error). This scatter reflects noise in
how well the survey captures the truth (weather, effort, or observer
conditions can shift a whole year's count up or down together -- this is called "observation error"), not the
population itself fluctuating (we could model this too- this is called "process error").  


```{r truth-table}
truth <- data.frame(
  Quantity = c("Expected index in year 0",
               "Total change from year 0 to year 25",
               "Year-to-year scatter (SD of the log of the multiplicative factor)",
               "Surveys"),
  Value = c("I_0 = 60",
            "-25% (falls to 75% of its year-0 value)",
            "sigma = 0.2",
            "one per year, t = 1, ..., 25")
)
knitr::kable(truth, col.names = c("Quantity", "Value"))
```

The analyst (treating this as real-world data and not knowing the 'truth) sees only the 25 observed counts, fits a straight line to the *log*
counts using `lm()`, and reports the slope with a 90% confidence
interval. **Across many replicate monitoring programs, does that slope
hit the truth on average, how variable is it, and does the interval
contain the truth 90% of the time?**

Please close your laptop. 

Parts 1-3 and the group check-in that follows are pen-and-paper only- but feel free to use a calculator or use your phone as a calculator.

`r nq(4, 3, extra = 0.6)`

## Part 1: The model (individual, ~9 min)

**1a.** A decline of 25% over 25 years means the index value in year 25 is
0.75 times the index value in year 0. If the index is multiplied by the same
factor $\lambda$ every year, what is $\lambda$? Let $b = \ln \lambda$.
What is $b$ (also known as intrinsic rate of exponential growth/decline)?

```{r fig.height=0.3*2}
blank_space()
```

$\lambda =$ ________________ &nbsp;&nbsp;&nbsp;&nbsp; $b =$ ________________


`r nq(5, 0, extra = 3.2)`

**1b.** Sketch the trajectory of the index for a 25%
decline and, as a dashed curve, for a 90% decline. Do this on the
raw-scale axes at left, then again on the log-scale axes at right. Both
start at $I_0 = 60$ in year 0 (the dot).

```{r sketch-axes, fig.width=6.5, fig.height=2.6}
par(mfrow = c(1, 2), mar = c(4, 4.5, 2, 1))
plot(NA, xlim = c(0, 25), ylim = c(0, 80), xaxs = "i", yaxs = "i",
     xlab = "Year", ylab = "Median index", main = "Raw scale")
abline(v = seq(0, 25, 5), h = seq(0, 80, 10), col = "grey85")
box(); points(0, 60, pch = 16, cex = 1.3)
plot(NA, xlim = c(0, 25), ylim = c(1, 4.5), xaxs = "i", yaxs = "i",
     xlab = "Year", ylab = "log(median index)", main = "Log scale")
abline(v = seq(0, 25, 5), h = 1:4, col = "grey85")
box(); points(0, log(60), pch = 16, cex = 1.3)
```

Which of your curves are straight lines, and why?

```{r fig.height=0.3*2}
blank_lines(2)
```

`r nq(6, 0, extra = 2)`

**1b, continued.** The scatter around the trend has the same spread on
the log scale, $\sigma = 0.2$, every single year. On the log-scale axes, add a band of $\pm\sigma$ around your
**25% decline** line: a pair of lines running parallel to it, each about
0.2 log-units above and below (constant vertical distance apart the
whole way across). Now sketch what that same band looks like (approximately) on the
raw-scale axes once it's converted back (roughly: multiply the center point by
$e^{0.2}\approx 1.2$ for the top and $e^{-0.2}\approx 0.8$ for the
bottom). [note: technically the center point is the median, since the expectation of an exponentiated random variable is actually larger than the exponential of the expectation, by Jensen's inequality - but don't worry if you don't see that just yet]

If we used the raw response (abundance index) in a linear regression, what assumption(s) might we be violating?

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(9, 0, extra = 1.5)`

**1c.** In 1a you found $\lambda$ (with $b = \ln\lambda$) so that the
population is multiplied by $\lambda$ every year, with no randomness --
that per-year multiplication, applied $t$ times, is the **deterministic
decline**: $\lambda^t$, equivalently $e^{bt}$. Combine it with a random
multiplicative factor $e^{\varepsilon_t}$, $\varepsilon_t \sim
\text{Normal}(0, \sigma)$, independent across years, to write the full
model for the index $I_t$ on the raw scale: $I_0$ times the deterministic
decline, times the random factor. Then take the log of both sides to get
a model for $\log(I_t)$. Write out this model - does this model meet the basic assumptions of linear regression?

```{r fig.height=0.3*5}
blank_space()
```

`r nq(10, 0, extra = 0.9)`

## Part 2: What should the simulation show? (individual, ~8 min)

The analyst regresses $z_t = \log(I_t)$ on $t$ and gets a slope estimate
$\hat{b}$. Because the counts are random, every simulated monitoring
program (replicate) gives a slightly different $\hat{b}$, even though the underlying
trend is exactly the same every time -- $\hat{b}$ is itself a random
variable, with its own distribution across replicate monitoring programs.

`r nq(9, 4)`

**2a.** Suppose we simulate many replicate monitoring programs under two survey
designs: (i) an annual survey ($t = 1, \ldots, 25$, 25 surveys) and (ii)
a survey only every other year ($t = 1, 3, \ldots, 25$, 13 surveys). For
each design, where would you expect the resulting values of $\hat{b}$,
across many replicate monitoring programs, to be centered? Would you
expect the two designs to have the same center? Which design do you
expect to give a more variable (spread-out) set of $\hat{b}$ values, and
how different do you expect the variation (standard error) to be? 

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(5, 0, extra = 2.3)`

**2b.** On the axes below, sketch what you expect the *distribution* of
$\hat{b}$ to look like under each design: a solid curve for annual
surveys, a dashed curve for every-other-year surveys. Mark the true
value $b$ (from 1a). You don't need exact numbers -- just show where
each curve is centered and how wide or narrow it is, relative to the
other. Remember that the area under each of these curves should be 1. 

```{r bhat-axes, fig.width=5.5, fig.height=1.9}
par(mar = c(4, 1, 1, 1))
plot(NA, xlim = c(-0.03, 0.01), ylim = c(0, 1), yaxt = "n", xaxs = "i", yaxs = "i",
     xlab = expression(hat(b)), ylab = "")
abline(v = seq(-0.03, 0.01, 0.01), col = "grey85")
box()
```

`r nq(5, 5)`

**2c.** In a moment you will estimate the coverage of the 90% interval
as the fraction of $R$ simulated programs whose interval contains $b$.
Treat each program/replicate as a Bernoulli trial with success probability
$p = 0.9$. Using the rules for the variance of a mean (and the fact that the variance of a Bernoulli r.v. is p*(1-p)), what
is the standard error of the estimated coverage, as a function of $R$?
How large must $R$ be for that standard error to be less than or equal to 0.01?

```{r fig.height=0.3*5}
blank_space()
```

`r nq(4, 0, extra = 0.5)`

## Part 3: Plan the experiment (individual, ~3 min)

**3.** Recall the questions posed at the start: across many replicate
monitoring programs, does the slope estimate $\hat{b}$ hit the truth
($b$) on average, how variable is it, and does the 90% confidence
interval contain the truth 90% of the time? Write, in plain English, the
step-by-step procedure for a simulation experiment that answers these questions. Do
not write R code- instead just write in plain English. State what gets repeated, what you keep track of on each
repeat, and what you compute at the end.

```{r fig.height=0.3*9}
blank_lines(9)
```

`r nq(14, 0, extra = 0.5)`

## Group check-in: finalize your plan, then choose your path (group, ~7 min)

Compare your answers to Parts 2 and 3 with your group. Agree on one
version of your speculation about $\hat{b}$'s distribution under each
design, your sample-size calculation, and the plain-English procedure.
Where you differ, argue it out.

Your group must now pick **one** of the two paths below for running the
experiment. Both are equally valid; decide as a group before opening any
laptops.

- **Path A (AI-assisted):** write a prompt describing your procedure and the model, and have an AI assistant generate the R code.
- **Path B (self-coded):** write the R code yourselves, using R's documentation and each other, but no AI tools.

Our group is using: &nbsp;&nbsp; Path A (AI-assisted) &nbsp;&nbsp;&nbsp;&nbsp; / &nbsp;&nbsp;&nbsp;&nbsp; Path B (self-coded) &nbsp;&nbsp; *(circle one)*

---

**Laptops open now.** Complete only the section below that matches your
group's chosen path.

---

`r nq(7, 0, extra = 0.6)`

## Part 4A: AI-assisted path (laptops, ~20 min)

Use these settings: $I_0 = 60$, $\sigma = 0.2$, $b$ from 1a, and 2,000 simulated programs per design. The analysis is `lm()` of $\log(I_t)$ on $t$, with a 90% confidence interval for the slope from `confint()`.

Write a prompt describing your group's procedure and asking the AI to write R code that (1) runs the experiment for annual surveys ($t = 1, \ldots, 25$), (2) reports the mean and SD of $\hat{b}$ and the estimated coverage of the 90% interval, (3) draws a histogram of the $\hat{b}$ values with the true $b$ marked and a normal curve (using the simulated mean and SD) overlaid, and (4) repeats the experiment for surveys every other year ($t = 1, 3, \ldots, 25$). **You don't need to submit your code for this path** -- just your prompt notes and results below.

`r nq(3, 4)`

**Notes on the prompt you used and any changes you had to make to get the AI to do the right thing** (record your prompt and what happened, not the AI's code):

```{r fig.height=0.3*4}
blank_space()
```

**Results, annual surveys.** Mean of $\hat{b}$ $=$ ______ &nbsp;&nbsp; SD of $\hat{b}$ $=$ ______

Coverage of the 90% interval $=$ ______ $\pm$ ______ (Monte Carlo standard error)

**Results, every other year.** Mean of $\hat{b}$ $=$ ______ &nbsp; SD of $\hat{b}$ $=$ ______

`r nq(2, 3)`

How do the simulated results compare with your speculation in 2a and your sketch in 2b?

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(3, 4)`

**Reflection.** Did the AI's first attempt do what your procedure said? How did your own understanding or lack of understanding of the model from Parts 1-2 affect your ability to get AI to do what you wanted? 

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(9, 0, extra = 0.6)`

## Part 4B: Self-coded path (laptops, ~20 min)

Use these settings: $I_0 = 60$, $\sigma = 0.2$, $b$ from 1a, and 2,000 simulated programs per design. The analysis is `lm()` of $\log(I_t)$ on $t$, with a 90% confidence interval for the slope from `confint()`.

Write the R code for your group's procedure (R's built-in help, documentation, and human-written websites and neighbors are fair game; AI tools are not). Run the experiment for annual surveys ($t = 1, \ldots, 25$), then for surveys every other year ($t = 1, 3, \ldots, 25$). Draw a histogram of the $\hat{b}$ values for annual surveys with the true $b$ marked. **You don't need to write your code out by hand** -- one member should email your group's R script to the instructor, with everyone in your group listed on the email.

**Results, annual surveys.** Mean of $\hat{b}$ $=$ ______ &nbsp;&nbsp; SD of $\hat{b}$ $=$ ______

Coverage of the 90% interval $=$ ______ $\pm$ ______ (Monte Carlo standard error)

**Results, every other year.** Mean of $\hat{b}$ $=$ ______ &nbsp; SD of $\hat{b}$ $=$ ______

`r nq(2, 3)`

How do the simulated results compare with your speculation in 2a and your sketch in 2b?

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(2, 4)`

**Reflection.** What part of turning your procedure into working R gave your group the most trouble? How did you resolve it?

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(9, 0, extra = 1.6)`

## Part 4C: Compare with the formula (both paths, ~5 min)

It turns out there is an exact formula for the
sampling distribution of $\hat{b}$:

$$\hat{b} \sim \text{Normal}\!\left(b,\ \frac{\sigma}{\sqrt{S_{tt}}}\right), \qquad S_{tt} = \sum_t (t - \bar{t})^2.$$

In words: $\hat{b}$ is exactly unbiased (centered on the true $b$), and
its standard deviation shrinks as $S_{tt}$ (how spread out the survey
years are) grows. For our two designs, $S_{tt} = 1300$ for annual
surveys and $S_{tt} = 728$ for every-other-year surveys.

**4C.** Using $\sigma = 0.2$, compute the formula's SD of $\hat{b}$ for
each design, and compare with your simulated (Monte Carlo) SD from your
Results above.

Annual: formula SD $=$ ______ &nbsp;&nbsp; simulated SD $=$ ______

Every other year: formula SD $=$ ______ &nbsp;&nbsp; simulated SD $=$ ______

```{r fig.height=0.3*2}
blank_lines(2)
```

`r nq(5, 4)`

Do they agree well? If a formula SD and a simulated SD don't
agree closely, is that more likely to be Monte Carlo error (not enough
replicates) or a mistake somewhere in the simulation code? How could you
tell the difference?

```{r fig.height=0.3*4}
blank_lines(4)
```

---

## Group discussion and revision (~5 min)

Compare your results with a different group (ideally one that took the
other path). Where the simulation disagreed with a prediction, find out
why before moving on.

Feel free to revise any of your answers above based on the discussion
(cross out and rewrite; I like to see your thinking evolve).

Please take a minute to record any changes you made and any new
insights gained during the group discussion. You can also use this
space to reflect on any areas of confusion ("muddiest points") or
topics you want to review or dust off.

```{r fig.height=0.3*6}
blank_lines(6)
```
