---
title: "Lab 4, Act 2"
subtitle: "The model is wrong. Does it matter?"
author: "NRES 746 -- name: _______________   group: _____________________"
date: ""
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
header-includes:
  - \usepackage{needspace}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}

# draws a table with given headers/body for handwritten answers ("" = blank
# cell); headers/values may be plotmath expressions for math; pair with
# fig.height = header_height + nrow(body) * row_height
blank_table <- function(headers, body, col_widths = NULL,
                         header_height = 0.35, row_height = 0.5,
                         cex = 0.95, bold_rows = integer(0)) {
  nc <- length(headers); nr <- length(body)
  if (is.null(col_widths)) col_widths <- rep(1, nc)
  col_widths <- col_widths / sum(col_widths)
  cx <- c(0, cumsum(col_widths))
  h_total <- header_height + nr * row_height
  par(mar = c(0, 0, 0, 0), family = "serif")
  plot(NA, xlim = c(0, 1), ylim = c(0, h_total), axes = FALSE, xlab = "", ylab = "",
       xaxs = "i", yaxs = "i")
  y <- h_total
  segments(0, y, 1, y, lwd = 1.5)
  y <- y - header_height
  segments(0, y, 1, y, lwd = 1.5)
  for (i in seq_len(nr)) {
    y <- y - row_height
    segments(0, y, 1, y, lwd = if (i == nr) 1.5 else 0.5,
             col = if (i == nr) "black" else "grey70")
  }
  for (v in cx) segments(v, 0, v, h_total, lwd = 1)
  yh <- h_total - header_height / 2
  for (j in seq_len(nc)) text((cx[j] + cx[j + 1]) / 2, yh, labels = headers[[j]], font = 2, cex = cex)
  y <- h_total - header_height
  for (i in seq_len(nr)) {
    yc <- y - row_height / 2
    fnt <- if (i %in% bold_rows) 2 else 1
    for (j in seq_len(nc)) {
      val <- body[[i]][[j]]
      if (!identical(val, "")) text((cx[j] + cx[j + 1]) / 2, yc, labels = val, font = fnt, cex = cex)
    }
    y <- y - row_height
  }
}

# keeps a prompt and its answer space on one PDF page: place `r nq(...)` on its
# own line just above the prompt. Emits nothing in non-LaTeX output. lines =
# lines of prompt text, rows = rows of blank space below it (0.3 in each),
# extra = any other height (in inches), e.g. a figure or table
nq <- function(lines, rows = 0, extra = 0) {
  if (!knitr::is_latex_output()) return("")
  knitr::asis_output(sprintf("\n\\needspace{%.2fin}\n", 0.23 * lines + 0.15 + 0.3 * rows + extra))
}
```

## Picking up where we left off

In Act 1 the analyst fit a straight line to the *log* counts, which
matches how the data were generated, and your simulations showed the
slope estimate was unbiased with the spread you predicted. Real
analysts do not get to pick the analysis that matches the truth.
Suppose a second analyst, following the agency's old spreadsheet
protocol, fits a straight line to the *raw* counts instead. That model
says the index falls by the same *number* of animals every year. The
truth says it falls by the same *proportion*.

We will compare the two analysts under two truths. In both, the
expected index in year 0 is $N_0 = 60$, the scatter is $\sigma = 0.2$ on
the log scale, and surveys are annual ($t = 1, \ldots, 25$):

```{r scenario-table}
sc <- data.frame(
  Scenario = c("Mild", "Severe"),
  Decline = c("25% over 25 years", "90% over 25 years")
)
knitr::kable(sc, col.names = c("Scenario", "Total decline, year 0 to year 25"))
```

Both analysts report the same management quantity: the estimated
**percent decline over the 25 years**. The log-scale analyst computes it
as $1 - e^{25\hat{b}}$. The raw-scale analyst computes it as
$1 - \hat{y}_{25} / \hat{y}_{0}$, using the fitted values of the
straight line at $t = 25$ and $t = 0$.

**Does the wrong model give the wrong answer? And can the data alone
tell us the model is wrong?**

**Before you do anything else: close your laptop and put your phone
away.** Parts 1-3 and the group check-in that follows are pen-and-paper
only.

`r nq(4, 0, extra = 3.1)`

## Part 1: What does each model assume about scatter? (individual, ~10 min)

**1a.** The *variance function* of a distribution says how the variance
changes with the mean $\mu$. Fill in the blanks. A gamma distribution
with shape $\alpha$ and scale $s$ has mean $\alpha s$ and variance $\alpha s^2$. Bolker
Ch. 4 is fair game.

```{r variance-table, fig.height=0.35+5*0.5, fig.width=5.8}
blank_table(
  headers = list("Distribution", "Var(Y) in terms of the mean", "Constant CV?"),
  body = list(
    list("Normal (SD sigma)", "", ""),
    list("Poisson", "", ""),
    list("Negative binomial (mean, size k)", "", ""),
    list("Gamma (mean, shape alpha)", "", ""),
    list("Lognormal (log-SD s)", "(exp(s^2) - 1) x mean^2", "")
  ),
  col_widths = c(1.9, 1.9, 1),
  cex = 0.85
)
```

`r nq(5, 0, extra = 2.5)`

**1b.** For a lognormal variable whose log has mean $m$ and SD $s$, the
mean of the variable is $e^{m + s^2/2}$ and its SD is
$\sqrt{e^{s^2} - 1}$ times that mean. In our model $s = \sigma = 0.2$ and
the median in year 0 is $e^{m} = 60$. Fill in the table (a calculator is
fine).

```{r sd-table, fig.height=0.35+3*0.5, fig.width=5.8}
blank_table(
  headers = list("Year and scenario", "Median", "Mean", "SD"),
  body = list(
    list("Year 0", "60", "", ""),
    list("Year 25, mild", "45", "", ""),
    list("Year 25, severe", "6", "", "")
  ),
  col_widths = c(1.6, 1, 1, 1),
  cex = 0.9
)
```

`r nq(5, 0, extra = 3.5)`

**1c.** For the severe scenario, use the axes below to sketch (i) the
median trajectory, (ii) a band showing roughly $\pm 2$ SD around it,
and (iii) the straight line an analyst would fit through the counts. Where
does your line cross zero, and why is that a problem for a count?

```{r raw-axes, fig.width=5.5, fig.height=2.6}
par(mar = c(4, 4.5, 1, 1))
plot(NA, xlim = c(0, 25), ylim = c(-10, 80), xaxs = "i", yaxs = "i",
     xlab = "Year", ylab = "Count index")
abline(v = seq(0, 25, 5), h = seq(0, 80, 20), col = "grey85")
abline(h = 0, lwd = 1.5)
box()
```

```{r fig.height=0.3*2}
blank_lines(2)
```

`r nq(3, 0, extra = 1.5)`

## Part 2: Commit to predictions (individual, ~4 min)

**2a.** For each analyst and scenario, will the estimated percent
decline be unbiased, too high, or too low? Write a word or two in each
cell. Then say briefly why.

```{r bias-pred-table, fig.height=0.35+2*0.5, fig.width=5.5}
blank_table(
  headers = list("", "Mild (25%)", "Severe (90%)"),
  body = list(
    list("Log-scale analyst", "", ""),
    list("Raw-scale analyst", "", "")
  ),
  col_widths = c(1.6, 1, 1),
  cex = 0.9
)
```

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(8, 3, extra = 0.7)`

## Part 3: A check that does not need the truth (individual, ~5 min)

In real life we never know the true model. What we can do is ask
whether the data look like they could have come from the *fitted*
model: fit the model, simulate many new datasets from the fitted model,
and compare a summary statistic of the real data with the same
statistic across the simulated datasets.

Our statistic is the number of **sign changes** in the residuals over
the 25 years (the number of times neighboring residuals switch between
positive and negative). If the fitted line captures the pattern and the
residuals are independent, about half of the 24 neighboring pairs
should switch. If the true pattern is curved and the line is straight,
residuals come in long runs of one sign, and very few switch.

**3a.** Write, in plain English, the procedure for a check that returns
a $p$-value: the fraction of simulated datasets with as few sign changes
as the real data. Do not write R code. Be specific about what you fit
to each simulated dataset and why.

```{r fig.height=0.3*8}
blank_lines(8)
```

`r nq(3, 0, extra = 1.5)`

**3b.** How often do you expect this check to flag each model in each
scenario (rarely, sometimes, or usually)?

```{r gof-pred-table, fig.height=0.35+2*0.5, fig.width=5.5}
blank_table(
  headers = list("", "Mild (25%)", "Severe (90%)"),
  body = list(
    list("Log-scale model", "", ""),
    list("Raw-scale model", "", "")
  ),
  col_widths = c(1.6, 1, 1),
  cex = 0.9
)
```

`r nq(9, 0, extra = 0.5)`

## Group check-in: finalize your plan, then choose your path (group, ~8 min)

Compare your answers to Parts 1-3 with your group. Agree on one
procedure for the check in 3a. Where your predictions in 2a and 3b
differ, argue it out.

Your group must now pick **one** of the two paths below for running the
experiments. Both are equally valid; decide as a group before opening any
laptops.

- **Path A (AI-assisted):** write a prompt describing your procedure and the model, and have an AI assistant generate the R code.
- **Path B (self-coded):** write the R code yourselves, using R's documentation and each other, but no AI tools.

Our group is using: &nbsp;&nbsp; Path A (AI-assisted) &nbsp;&nbsp;&nbsp;&nbsp; / &nbsp;&nbsp;&nbsp;&nbsp; Path B (self-coded) &nbsp;&nbsp; *(circle one)*

---

**Laptops open now.** Both paths run the same three experiments. Choose
the instructions for your path.

---

`r nq(8, 0, extra = 1.5)`

## Part 4: Run the experiments (laptops, ~30 min)

Use these settings: $N_0 = 60$, $\sigma = 0.2$, $t = 1, \ldots, 25$,
$b_{\text{mild}} = \ln(0.75)/25$ and $b_{\text{severe}} = \ln(0.10)/25$.
The two helper functions below are ready-made. Type them into your R
script.

```{r helpers, echo=TRUE, eval=FALSE}
# number of times neighboring residuals switch sign
sign_changes <- function(r) {
  s <- sign(r)
  sum(s[-1] != s[-length(s)])
}

# t: years, y: observed counts, sims: matrix of simulated counts
# (one row per year, one column per simulated data set)
plot_envelope <- function(t, y, sims) {
  lo <- apply(sims, 1, quantile, 0.05)
  hi <- apply(sims, 1, quantile, 0.95)
  plot(t, y, type = "n", ylim = range(y, lo, hi), xlab = "Year", ylab = "Count index")
  polygon(c(t, rev(t)), c(lo, rev(hi)), col = "grey85", border = NA)
  points(t, y, pch = 16)
  abline(h = 0, lty = 3)
}
```

The three experiments:

1. **Two analysts.** For each scenario, simulate 2,000 datasets from the
   model. Fit both analysts to every dataset and compute each one's
   estimated percent decline. Record the mean and SD of the estimates and
   the share of estimates above 100%.
2. **An envelope.** Simulate one dataset from the severe scenario and fit
   the raw-scale model. Simulate 500 new datasets from the fitted raw-scale
   model (Normal with mean equal to the fitted values and SD equal to the
   residual SD). Draw the envelope with `plot_envelope()`.
3. **The check.** Implement your procedure from 3a as a function. For the
   log-scale model, run the same check on $\log(y)$, with the simulated
   datasets also generated on the log scale. Use 200 simulated datasets
   inside the check. Apply it to 100 fresh datasets for each scenario and
   each model, and record how often it returns $p < 0.05$.

`r nq(6, 0, extra = 0.6)`

### Path A: AI-assisted

Write a prompt describing your group's procedure and the model, and asking
the AI to write R code for the three experiments. Record your prompt and
what you had to change to make the code do what you intended (do not paste
the AI's code).

```{r fig.height=0.3*4}
blank_space()
```

`r nq(3, 4)`

**Reflection.** Did the AI's first attempt do what your procedure said? Where
did your own understanding of the models from Parts 1-3 help you catch
or fix something?

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(5, 0, extra = 0.3)`

### Path B: Self-coded

Write the R code yourselves (R's built-in help, documentation, and
human-written websites and neighbors are fair game; AI tools are not).

**Reflection.** What part of turning your procedure into working R gave your group
the most trouble? How did you resolve it?

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(8, 0, extra = 2.5)`

### Results (both paths)

**Experiment 1.** Estimated percent decline over 25 years.

```{r results1-table, fig.height=0.35+4*0.5, fig.width=5.8}
blank_table(
  headers = list("Scenario", "Analyst", "Mean", "SD", "Share > 100%"),
  body = list(
    list("Mild (25%)", "Log-scale", "", "", ""),
    list("Mild (25%)", "Raw-scale", "", "", ""),
    list("Severe (90%)", "Log-scale", "", "", ""),
    list("Severe (90%)", "Raw-scale", "", "", "")
  ),
  col_widths = c(1.3, 1.1, 0.9, 0.9, 1.2),
  cex = 0.85
)
```

`r nq(3, 0, extra = 1.5)`

**Experiment 3.** Share of datasets for which the check returns $p < 0.05$.

```{r results3-table, fig.height=0.35+2*0.5, fig.width=5.5}
blank_table(
  headers = list("", "Mild (25%)", "Severe (90%)"),
  body = list(
    list("Log-scale model", "", ""),
    list("Raw-scale model", "", "")
  ),
  col_widths = c(1.6, 1, 1),
  cex = 0.9
)
```

`r nq(4, 4)`

In one or two sentences, describe what the envelope in Experiment 2 shows
that a single glance at the raw counts might not.

```{r fig.height=0.3*3}
blank_lines(3)
```

---

**If you finish early:** simulate 100,000 values of $y_0$ from the model.
Compare `mean(log(y0))` with `log(mean(y0))`. The gap should match
$\sigma^2/2$ from 1b. Why does the average of the logs differ from the log
of the average?

`r nq(8, 0, extra = 0.5)`

## Group discussion and revision (~5 min)

Compare your results with a different group (ideally one that took the
other path). How did your predictions in 2a and 3b compare with the
results? In which scenario did the wrong model matter, and could the
check have warned you?

Feel free to revise any of your answers above based on the discussion
(cross out and rewrite; I like to see your thinking evolve).

Please take a minute to record any changes you made and any new
insights gained during the group discussion. You can also use this
space to reflect on any areas of confusion ("muddiest points") or
topics you want to review or dust off.

```{r fig.height=0.3*6}
blank_lines(6)
```
