---
title: "Lab 4, Act 1"
subtitle: "If we knew the truth, would our trend analysis find it?"
author: "NRES 746 -- name: _______________   group: _____________________"
date: ""
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
header-includes:
  - \usepackage{needspace}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}

# keeps a prompt and its answer space on one PDF page: place `r nq(...)` on its
# own line just above the prompt. Emits nothing in non-LaTeX output. lines =
# lines of prompt text, rows = rows of blank space below it (0.3 in each),
# extra = any other height (in inches), e.g. a figure or table
nq <- function(lines, rows = 0, extra = 0) {
  if (!knitr::is_latex_output()) return("")
  knitr::asis_output(sprintf("\n\\needspace{%.2fin}\n", 0.23 * lines + 0.15 + 0.3 * rows + extra))
}
```

## The scenario

A state wildlife agency is planning a long-term monitoring program for a
rare, secretive toad in a remote Great Basin spring complex. The
management goal is to detect a population decline of 25% or more over 25
years. In Act 3 you will design and evaluate the full program. Before
that, we need to know whether the statistical machinery we plan to use
can be trusted at all. Real monitoring data cannot tell us, because we
never know the true rate of decline. A simulation can, because we get to
choose the truth.

Today's simplified world: each year a survey produces a count index
$y_t$ for years $t = 1, 2, \ldots, 25$. The population declines by the
same proportion every year, and each year's count differs from the
trend by a random multiplicative factor. Here is the truth, which the
analyst never sees:

```{r truth-table}
truth <- data.frame(
  Quantity = c("Expected index in year 0",
               "Total change from year 0 to year 25",
               "Year-to-year scatter (SD of the log of the multiplicative factor)",
               "Surveys"),
  Value = c("N_0 = 60",
            "-25% (falls to 75% of its year-0 value)",
            "sigma = 0.2",
            "one per year, t = 1, ..., 25")
)
knitr::kable(truth, col.names = c("Quantity", "Value"))
```

The analyst sees only the 25 counts, fits a straight line to the *log*
counts using `lm()`, and reports the slope with a 90% confidence
interval. **Across many repeated monitoring programs, does that slope
hit the truth on average, how variable is it, and does the interval
contain the truth 90% of the time?**

**Before you do anything else: close your laptop and put your phone
away.** Parts 1-3 and the group check-in that follows are pen-and-paper
only.

`r nq(4, 3, extra = 0.6)`

## Part 1: The model (individual, ~7 min)

**1a.** A decline of 25% over 25 years means the index in year 25 is
0.75 times the index in year 0. If the index is multiplied by the same
factor $\lambda$ every year, what is $\lambda$? Let $b = \ln \lambda$.
What is $b$?

$\lambda =$ ________________ &nbsp;&nbsp;&nbsp;&nbsp; $b =$ ________________

Show your work:

```{r fig.height=0.3*3}
blank_space()
```

`r nq(5, 0, extra = 3.2)`

**1b.** Sketch the expected (median) trajectory of the index for a 25%
decline and, as a dashed curve, for a 90% decline. Do this on the
raw-scale axes at left, then again on the log-scale axes at right. Both
start at $N_0 = 60$ in year 0 (the dot).

```{r sketch-axes, fig.width=6.5, fig.height=2.6}
par(mfrow = c(1, 2), mar = c(4, 4.5, 2, 1))
plot(NA, xlim = c(0, 25), ylim = c(0, 65), xaxs = "i", yaxs = "i",
     xlab = "Year", ylab = "Expected index", main = "Raw scale")
abline(v = seq(0, 25, 5), h = seq(0, 60, 10), col = "grey85")
box(); points(0, 60, pch = 16, cex = 1.3)
plot(NA, xlim = c(0, 25), ylim = c(1, 4.5), xaxs = "i", yaxs = "i",
     xlab = "Year", ylab = "log(expected index)", main = "Log scale")
abline(v = seq(0, 25, 5), h = 1:4, col = "grey85")
box(); points(0, log(60), pch = 16, cex = 1.3)
```

Which of your four curves are straight lines, and why?

```{r fig.height=0.3*2}
blank_lines(2)
```

`r nq(6, 0, extra = 1.5)`

**1c.** Write the model for the index $y_t$ on the raw scale as
$N_0$ times a deterministic decline, times a random factor $e^{\varepsilon_t}$
with $\varepsilon_t \sim \text{Normal}(0, \sigma)$, independent across years.
Then take logs to get a model for $\log(y_t)$. Which of the two versions
is a straight-line regression model, and in what?

```{r fig.height=0.3*5}
blank_space()
```

`r nq(10, 0, extra = 0.9)`

## Part 2: What should the simulation show? (individual, ~10 min)

The analyst regresses $z_t = \log(y_t)$ on $t$. The least-squares slope
can be written as a weighted sum of the observations,

$$\hat{b} = \sum_{t} w_t z_t \qquad w_t = \frac{t - \bar{t}}{S_{tt}} \qquad S_{tt} = \sum_t (t - \bar{t})^2,$$

where the sums run over the years surveyed. You may use two facts about
the weights: $\sum_t w_t = 0$ and $\sum_t w_t\, t = 1$. The rules from
the Week 4 WOD for expectations and variances of sums of independent
random variables apply.

`r nq(3, 6)`

**2a.** Show that $\mathbb{E}[\hat{b}] = b$. What does this say about
bias?

```{r fig.height=0.3*6}
blank_space()
```

`r nq(3, 8)`

**2b.** Show that $\text{Var}(\hat{b}) = \sigma^2 / S_{tt}$. (First show
that $\sum_t w_t^2 = 1/S_{tt}$.)

```{r fig.height=0.3*8}
blank_space()
```

`r nq(4, 3, extra = 0.4)`

**2c.** For annual surveys ($t = 1, \ldots, 25$), $S_{tt} = 1300$. Now
consider surveying only every other year ($t = 1, 3, 5, \ldots, 25$,
13 surveys). Compute $S_{tt}$ for this design, then predict the SD of
$\hat{b}$ for both designs.

Annual: $S_{tt} = 1300$, SD$(\hat{b}) =$ ________________

Every other year: $S_{tt} =$ ____________, SD$(\hat{b}) =$ ________________

```{r fig.height=0.3*3}
blank_space()
```

`r nq(5, 0, extra = 2.2)`

**2d.** Suppose we simulate 2,000 monitoring programs with annual
surveys and record $\hat{b}$ each time. On the axes below, mark the true
value $b$ and sketch the histogram you expect to see. Label the center
and the spread on the horizontal axis.

```{r bhat-axes, fig.width=5.5, fig.height=1.9}
par(mar = c(4, 1, 1, 1))
plot(NA, xlim = c(-0.03, 0.01), ylim = c(0, 1), yaxt = "n", xaxs = "i", yaxs = "i",
     xlab = expression(hat(b)), ylab = "")
abline(v = seq(-0.03, 0.01, 0.01), col = "grey85")
box()
```

`r nq(6, 5)`

**2e.** In a moment you will estimate the coverage of the 90% interval
as the fraction of $R$ simulated programs whose interval contains $b$.
Treat each program as a Bernoulli trial with success probability
$p = 0.9$. Using the rules for the variance of a mean (Week 4 WOD), what
is the standard error of the estimated coverage, as a function of $R$?
How large must $R$ be for that standard error to be at most 0.01?

```{r fig.height=0.3*5}
blank_space()
```

`r nq(3, 9)`

## Part 3: Plan the experiment (individual, ~3 min)

**3.** Write, in plain English, the step-by-step procedure for an
experiment that answers the question at the end of the scenario. Do not
write R code. Say what gets repeated, what you keep track of on each
repeat, and what you compute at the end.

```{r fig.height=0.3*9}
blank_lines(9)
```

`r nq(14, 0, extra = 0.5)`

## Group check-in: finalize your plan, then choose your path (group, ~7 min)

Compare your answers to Parts 2 and 3 with your group. Agree on one
version of the derivation, the predicted SDs, and the plain-English
procedure. Where you differ, argue it out.

Your group must now pick **one** of the two paths below for running the
experiment. Both are equally valid; decide as a group before opening any
laptops.

- **Path A (AI-assisted):** write a prompt describing your procedure and the model, and have an AI assistant generate the R code.
- **Path B (self-coded):** write the R code yourselves, using R's documentation and each other, but no AI tools.

Our group is using: &nbsp;&nbsp; Path A (AI-assisted) &nbsp;&nbsp;&nbsp;&nbsp; / &nbsp;&nbsp;&nbsp;&nbsp; Path B (self-coded) &nbsp;&nbsp; *(circle one)*

---

**Laptops open now.** Complete only the section below that matches your
group's chosen path.

---

`r nq(7, 0, extra = 0.6)`

## Part 4A: AI-assisted path (laptops, ~20 min)

Use these settings: $N_0 = 60$, $\sigma = 0.2$, $b$ from 1a, and 2,000 simulated programs per design. The analysis is `lm()` of $\log(y_t)$ on $t$, with a 90% confidence interval for the slope from `confint()`.

Write a prompt describing your group's procedure and asking the AI to write R code that (1) runs the experiment for annual surveys ($t = 1, \ldots, 25$), (2) reports the mean and SD of $\hat{b}$ and the estimated coverage of the 90% interval, (3) draws a histogram of the $\hat{b}$ values with the true $b$ marked and the normal curve implied by your answers to 2a-2c overlaid, and (4) repeats the experiment for surveys every other year ($t = 1, 3, \ldots, 25$).

`r nq(3, 4)`

**Notes on the prompt you used and any changes you had to make to get the AI to do the right thing** (record your prompt and what happened, not the AI's code):

```{r fig.height=0.3*4}
blank_space()
```

**Results, annual surveys.** Mean of $\hat{b}$ $=$ ______ &nbsp;&nbsp; SD of $\hat{b}$ $=$ ______

Coverage of the 90% interval $=$ ______ $\pm$ ______ (Monte Carlo standard error)

**Results, every other year.** Mean of $\hat{b}$ $=$ ______ &nbsp; SD of $\hat{b}$ $=$ ______

`r nq(2, 3)`

How do the simulated results compare with your predictions in 2a-2e and your sketch in 2d?

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(3, 4)`

**Reflection.** Did the AI's first attempt do what your procedure said? Where did your own understanding of the model from Parts 1-2 help you catch or fix something?

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(9, 0, extra = 0.6)`

## Part 4B: Self-coded path (laptops, ~20 min)

Use these settings: $N_0 = 60$, $\sigma = 0.2$, $b$ from 1a, and 2,000 simulated programs per design. The analysis is `lm()` of $\log(y_t)$ on $t$, with a 90% confidence interval for the slope from `confint()`.

Write the R code for your group's procedure (R's built-in help, documentation, and human-written websites and neighbors are fair game; AI tools are not). Run the experiment for annual surveys ($t = 1, \ldots, 25$), then for surveys every other year ($t = 1, 3, \ldots, 25$). Draw a histogram of the $\hat{b}$ values for annual surveys with the true $b$ marked and the normal curve implied by your answers to 2a-2c overlaid.

**Results, annual surveys.** Mean of $\hat{b}$ $=$ ______ &nbsp;&nbsp; SD of $\hat{b}$ $=$ ______

Coverage of the 90% interval $=$ ______ $\pm$ ______ (Monte Carlo standard error)

**Results, every other year.** Mean of $\hat{b}$ $=$ ______ &nbsp; SD of $\hat{b}$ $=$ ______

`r nq(2, 3)`

How do the simulated results compare with your predictions in 2a-2e and your sketch in 2d?

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(2, 4)`

**Reflection.** What part of turning your procedure into working R gave your group the most trouble? How did you resolve it?

```{r fig.height=0.3*4}
blank_lines(4)
```

---

**If you finish early:** double $\sigma$ to 0.4. Predict the SD of $\hat{b}$ and the coverage before you rerun, then check.

## Group discussion and revision (~5 min)

Compare your results with a different group (ideally one that took the
other path). Where the simulation disagreed with a prediction, find out
why before moving on.

Feel free to revise any of your answers above based on the discussion
(cross out and rewrite; I like to see your thinking evolve).

Please take a minute to record any changes you made and any new
insights gained during the group discussion. You can also use this
space to reflect on any areas of confusion ("muddiest points") or
topics you want to review or dust off.

```{r fig.height=0.3*6}
blank_lines(6)
```
