---
title: "Lab 4, Act 3"
subtitle: "How much monitoring is enough?"
author: "NRES 746 -- name: _______________   group: _____________________"
date: ""
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
header-includes:
  - \usepackage{needspace}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}

# draws a table with given headers/body for handwritten answers ("" = blank
# cell); headers/values may be plotmath expressions for math; pair with
# fig.height = header_height + nrow(body) * row_height
blank_table <- function(headers, body, col_widths = NULL,
                         header_height = 0.35, row_height = 0.5,
                         cex = 0.95, bold_rows = integer(0)) {
  nc <- length(headers); nr <- length(body)
  if (is.null(col_widths)) col_widths <- rep(1, nc)
  col_widths <- col_widths / sum(col_widths)
  cx <- c(0, cumsum(col_widths))
  h_total <- header_height + nr * row_height
  par(mar = c(0, 0, 0, 0), family = "serif")
  plot(NA, xlim = c(0, 1), ylim = c(0, h_total), axes = FALSE, xlab = "", ylab = "",
       xaxs = "i", yaxs = "i")
  y <- h_total
  segments(0, y, 1, y, lwd = 1.5)
  y <- y - header_height
  segments(0, y, 1, y, lwd = 1.5)
  for (i in seq_len(nr)) {
    y <- y - row_height
    segments(0, y, 1, y, lwd = if (i == nr) 1.5 else 0.5,
             col = if (i == nr) "black" else "grey70")
  }
  for (v in cx) segments(v, 0, v, h_total, lwd = 1)
  yh <- h_total - header_height / 2
  for (j in seq_len(nc)) text((cx[j] + cx[j + 1]) / 2, yh, labels = headers[[j]], font = 2, cex = cex)
  y <- h_total - header_height
  for (i in seq_len(nr)) {
    yc <- y - row_height / 2
    fnt <- if (i %in% bold_rows) 2 else 1
    for (j in seq_len(nc)) {
      val <- body[[i]][[j]]
      if (!identical(val, "")) text((cx[j] + cx[j + 1]) / 2, yc, labels = val, font = fnt, cex = cex)
    }
    y <- y - row_height
  }
}

# keeps a prompt and its answer space on one PDF page: place `r nq(...)` on its
# own line just above the prompt. Emits nothing in non-LaTeX output. lines =
# lines of prompt text, rows = rows of blank space below it (0.3 in each),
# extra = any other height (in inches), e.g. a figure or table
nq <- function(lines, rows = 0, extra = 0) {
  if (!knitr::is_latex_output()) return("")
  knitr::asis_output(sprintf("\n\\needspace{%.2fin}\n", 0.23 * lines + 0.15 + 0.3 * rows + extra))
}
```

## Picking up where we left off

In Act 1 you showed that a straight-line fit to log counts recovers the
rate of decline, with a spread you could predict from a formula. In Act
2 you saw that a wrong model can be harmless for a mild decline, badly
wrong for a severe one, and that you can sometimes catch it by checking
the data against datasets simulated from the fitted model. Now we use all
of that to answer the question the agency actually cares about: what
does it take to run a monitoring program that reliably detects the
decline of the toad?

## The design problem

The true abundance in year 0 is $N_0 = 1000$ (assume it is known). Each
animal present is detected independently with probability 0.02 per
person-day of searching. A monitoring **bout** is defined by the number
of observers $P$ and the number of days $D$, and bouts are repeated every
$s$ years ($s = 1, 2, 3, 4$, or $5$), starting in year 1, for as long as
they fit within the 25-year window ($t = 1, 1+s, 1+2s, \ldots \le 25$). Costs
are \$200 per person-day, plus \$2,000 to mount each field expedition,
regardless of its length.

Each bout produces one count, $y_t$. The analysis is the one from Act 1:
regress $\log(y_t)$ on $t$ with `lm()`. A decline is **detected** if the
slope estimate is negative and its two-sided $p$-value is at most 0.05.
The management goal is to have **at least a 75% chance of detecting a
decline of 25% or more over 25 years.**

**What is the cheapest design that meets the goal, and how sure are we?**

Try Part 1 on paper first, before you open a laptop or start typing
code. Resist the urge to jump straight to R or an AI assistant. Once
your plan is finalized, AI tools, `?help`, and web searches are all fair
game for Part 2.

`r nq(4, 4, extra = 0.5)`

## Part 1: Predict before you simulate (~30 min)

**1a.** For a single animal, what is the probability of being detected
during a bout with $P$ observers searching for $D$ days? (Assume every
observer-day is an independent 0.02 chance.) Evaluate it for
$P = 1$, $D = 3$, and give the expected count of a bout in year 0 and its
variance. (This is where the index $N_0 = 60$ in Acts 1 and 2 came from.)

```{r fig.height=0.3*5}
blank_space()
```

`r nq(4, 4)`

**1b.** Your formula depends on $P$ and $D$ only through one combined
quantity. What is it? Is that also true of the *cost* of a bout? What
does that imply for how many design dimensions we really have, and what
might make 5 observers for 1 day differ from 1 observer for 5 days in the
real world?

```{r fig.height=0.3*5}
blank_lines(5)
```

`r nq(8, 0, extra = 0.5)`

Let $k = P \times D$ be the number of person-days per bout.

**1c.** *Precision.* For a positive random variable $Y$ with mean $\mu$
and variance $v$, the delta method gives $\text{Var}(\log Y) \approx v/\mu^2$
(Bolker Appendix A.7). Apply this to a bout count with $N$ animals
present and per-animal detection probability $q$, and find the SD of
$\log(y_t)$ as a function of $N$ and $q$. Then evaluate it for $k = 3$
and $N = 860$ (about the abundance at the midpoint of the 25 years).

```{r fig.height=0.3*5}
blank_space()
```

`r nq(5, 0, extra = 3.2)`

**1d.** For $m$ surveys spaced $s$ years apart, $S_{tt} = s^2 m(m^2-1)/12$.
Fill in the table, and use your Act 1 result $\text{SD}(\hat{b}) = \sigma/\sqrt{S_{tt}}$,
where $\sigma$ is now your answer to 1c, to get the SD of the slope estimate for
$k = 3$. (A calculator is fine.)

```{r stt-table, fig.height=0.35+5*0.5, fig.width=5.5}
blank_table(
  headers = list("Interval s", "Surveys m", "S_tt", "SD of slope, k = 3"),
  body = list(
    list("1", "25", "1300", ""),
    list("2", "13", "728", ""),
    list("3", "", "", ""),
    list("4", "", "", ""),
    list("5", "", "", "")
  ),
  col_widths = c(1, 1, 1, 1.5),
  cex = 0.9
)
```

`r nq(6, 0, extra = 3.2)`

**1e.** *Power.* A decline is detected when the slope estimate is negative
and larger in magnitude than about 1.96 of its SDs. Ignoring the tiny
chance of a significant *increase*, that gives
$$\text{power} \approx \Phi\!\left(\frac{|b|}{\text{SD}(\hat{b})} - 1.96\right),$$
where $\Phi$ is the standard normal cdf and $b = \ln(0.75)/25 = -0.0115$
is the true rate of decline. Some values of $\Phi$: $\Phi(0) = 0.50$,
$\Phi(0.25) = 0.60$, $\Phi(0.5) = 0.69$, $\Phi(0.75) = 0.77$,
$\Phi(1) = 0.84$, $\Phi(1.25) = 0.89$, $\Phi(1.5) = 0.93$, $\Phi(2) = 0.98$.
Approximate the power for $k = 3$ with annual surveys ($s = 1$) and with
surveys every other year ($s = 2$).

```{r fig.height=0.3*5}
blank_space()
```

`r nq(4, 4)`

**1f.** Design A is $k = 3$, $s = 1$. Design B is $k = 6$, $s = 2$. Which do
you predict has higher power? (Hint: when $q$ is small, how does the SD
from 1c change when $k$ doubles? How does $S_{tt}$ change when $s$
doubles?) Compute the total cost of each design. What does this suggest
about how to spend a monitoring budget?

```{r fig.height=0.3*6}
blank_space()
```

`r nq(5, 3)`

**1g.** Suppose the population is actually *stable*. How often will this
analysis declare a decline? It is not 5%. Explain, and predict the number.

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(5, 0, extra = 2.9)`

**1h.** In plain English (no R code), write the procedure you would follow
to estimate the power of one design ($k$ and $s$) by simulation, and the
extra step you need to find the cheapest design that meets the goal.

```{r fig.height=0.3*8}
blank_lines(8)
```

`r nq(4, 4)`

**Checkpoint: finalize your plan.** Reread your procedure from 1h and your
predictions from 1e-1g. Would your plan work if you started typing it up
right now? Revise anything incomplete or unclear before you write code.

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(6, 0, extra = 1)`

## Part 2: Run the simulations (~60 min)

Use these settings: $N_0 = 1000$, per-animal detection probability
0.02 per person-day, $\lambda = 0.75^{1/25}$ so that the expected
abundance is $N_t = \text{round}(N_0 \lambda^t)$, and analysis by
`lm()` of $\log(y_t)$ on $t$. Detection is binomial: each bout count is a
Binomial draw with size $N_t$ and detection probability $1 - 0.98^{k}$.

**2a.** Write functions that (i) simulate the counts from one monitoring
program, given $k$ and $s$; (ii) analyze them and report whether a decline was detected; and
(iii) estimate the power of a design from $R$ simulated programs. Record
the prompt you used if you asked an AI for help, but not the AI's code.

**2b.** *Validate before you trust.* Set $\lambda = 1$ (no decline) and run
$R = 2{,}000$ programs of the design $k = 3$, $s = 1$. Then compare your
simulated power for $k = 3$ with $s = 1$ and $s = 2$ against your
approximations in 1e.

Null-simulation detection rate: ______ $\pm$ ______ (Monte Carlo SE)

My prediction in 1g: ______

Simulated power, $s = 1$: ________ &nbsp;&nbsp; Approximation from 1e: ________

Simulated power, $s = 2$: ________ &nbsp;&nbsp; Approximation from 1e: ________

`r nq(5, 4)`

Does the null rate agree with your prediction? The approximations in 1e
should be somewhat off. In which direction, and why? (Think about what
the approximation ignored.)

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(5, 3)`

**2c.** *The design space.* Estimate the power for $k = 1, \ldots, 15$ and
$s = 1, \ldots, 5$ (1,000 programs per design is enough). Plot power against $k$, with one
line for each $s$ and a horizontal line at 0.75. Describe the pattern. Which
of your predictions in 1f held up?

```{r fig.height=0.3*5}
blank_lines(5)
```

`r nq(5, 0, extra = 2.4)`

**2d.** *Cost.* Compute the cost of every design in your grid. Among designs with
estimated power at or above 0.75, find the cheapest few. Because each power
estimate has Monte Carlo error, re-estimate your top candidates with 5,000
programs each.

```{r candidates-table, fig.height=0.35+3*0.5, fig.width=5.5}
blank_table(
  headers = list("k", "Interval s", "Cost ($)", "Power (5,000 runs)", "MC SE"),
  body = list(
    list("", "", "", "", ""),
    list("", "", "", "", ""),
    list("", "", "", "", "")
  ),
  col_widths = c(0.8, 1, 1, 1.5, 0.9),
  cex = 0.85
)
```

Which design do you recommend, and why that one and not the cheapest one
that squeaked past 0.75?

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(5, 4)`

**2e.** *Does the analysis scale matter here?* Rerun your recommended
design, but analyze the raw counts, $y_t$, with `lm()` instead of the log counts.
Does the power change noticeably? Connect your answer to what you saw in
Act 2 (hint: which scenario is this closer to, mild or severe?).

Power with raw counts: ________ &nbsp;&nbsp; Power with log counts: ________

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(8, 0, extra = 0.5)`

**2f. Optional extension: environmental noise.** So far the only randomness
in the abundance trajectory is sampling (detection) error. Real populations
also experience year-to-year shocks that *persist*, such as a drought year
that reduces the population, which the next year starts from. Modify your simulation so that
$N_t = N_{t-1}\,\lambda\,e^{\varepsilon_t}$ with $\varepsilon_t \sim \text{Normal}(0, 0.05)$,
and then rerun (i) the power of your recommended design and (ii) the
detection rate when $\lambda = 1$. What happened, and which assumption
from Act 1 is no longer true? What does this suggest about the kind of
statistical model you would want for real monitoring data?

Power with environmental noise: ________ &nbsp;&nbsp; Detection rate when $\lambda = 1$: ________

```{r fig.height=0.3*5}
blank_lines(5)
```

`r nq(8, 5)`

## Reflection (~10 min)

**R1.** Summarize your recommended design, its cost, and its estimated power.
Which assumption of the simulation would you worry about most, and how would
you find out whether it matters?

```{r fig.height=0.3*5}
blank_lines(5)
```

`r nq(6, 5)`

**R2.** Look back at the scenario in Act 1. Which result or derivation from
Act 1 did you rely on most in this Act, and how?

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(6, 5)`

**R3.** In three or four sentences, explain your recommendation to the
agency's director, who has no statistical training. Then say what
result would make *you* doubt your own recommendation.

```{r fig.height=0.3*5}
blank_lines(5)
```

`r nq(4, 4)`

**R4.** If you used an AI tool, how did you use it, and what did you have to
check or correct? If you did not, what was hardest?

```{r fig.height=0.3*4}
blank_lines(4)
```
