---
title: "Lab 5, Act 2: finding the peak when calculus is not practical!"
subtitle: "Highway noise vs. frog chorus"
author: "NRES 746, name: _______________   group: _____________________"
date: ""
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
header-includes:
  - \usepackage{needspace}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}

# keeps a prompt and its answer space on one PDF page: place `r nq(...)` on its
# own line just above the prompt. Emits nothing in non-LaTeX output. lines =
# lines of prompt text, rows = rows of blank space below it (0.3 in each),
# extra = any other height (in inches), e.g. a figure or table
nq <- function(lines, rows = 0, extra = 0) {
  if (!knitr::is_latex_output()) return("")
  knitr::asis_output(sprintf("\n\\needspace{%.2fin}\n", 0.23 * lines + 0.15 + 0.3 * rows + extra))
}

chorus_data <- read.csv("../data/FrogChorus.csv")
dist_km <- chorus_data$dist_km
males <- chorus_data$males

nll_mono <- function(par, y, d) {
  -sum(dpois(y, lambda = par[1] * (1 - exp(-par[2] * d)), log = TRUE))
}
```

## The scenario

Males of the frog species from Act 1 call to attract mates, and traffic
noise can interfere with frog choruses. A highway runs through the
spring complex. To see how far the road's effect reaches, you visit 15 ponds at different
distances from the road and count the calling males you hear during
one 5-minute listen at each pond. Here are your observed measurements:

```{r chorus-data}
chorus <- rbind(c("Distance (km)", format(dist_km, drop0trailing = TRUE, trim = TRUE)),
                c("Males heard", as.character(males)))
knitr::kable(chorus, col.names = c("Pond", seq_along(dist_km)),
             align = c("l", rep("c", length(dist_km))))
```

A reasonable model is that the expected count rises with distance and
levels off far from the road:
$$\lambda(d) = a\left(1 - e^{-b d}\right)$$
This is the monomolecular function from Bolker's bestiary (Ch. 3).
Bolker's worked examples in Ch. 6 (p. 243-249) are a useful guide.

**How far from the road does the chorus recover, and how strong is the
evidence that the road matters at all?**

Try Part 1 on paper first, before you open a laptop or start typing
code.

`r nq(3, 3)`

## Part 1: Plan it on paper (individual, ~25 min)

**1a.** What do $a$ and $b$ mean biologically, and what are their
units?

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(3, 3)`

**1b.** Act 1 used a binomial distribution. Why is a Poisson
distribution a more reasonable choice for these counts?

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(7, 7, extra = 2.9)`

**1c.** Use the figure below and your function to estimate starting
values for $a$ and $b$. Show how you got your estimated starting value for $b$.

*Hint for $b$:* from the figure, estimate the distance $d_{1/2}$ at
which the chorus is halfway to its maximum, so that
$\lambda(d_{1/2}) = a/2$. Then solve that equation for $b$. Bolker's
bestiary of functions (Ch. 3) shows how parameters like these relate
to features of the curve.

```{r chorus-plot, fig.width=5, fig.height=2.9}
par(mar = c(4, 4.5, 0.5, 0.5))
plot(NA, xlim = c(0, 8), ylim = c(0, 16), xaxs = "i", yaxs = "i",
     xlab = "Distance from highway (km)", ylab = "Calling males heard")
abline(v = 0:8, h = seq(0, 16, 2), col = "grey85")
points(dist_km, males, pch = 16)
box()
```

```{r fig.height=0.3*7}
blank_space()
```

$a \approx$ ________ &nbsp;&nbsp;&nbsp;&nbsp; $b \approx$ ________

`r nq(3, 8)`

**1d.** Write the negative log-likelihood, $\text{NLL}(a, b)$, for the
15 ponds using the Poisson pmf. Simplify it as far as you can. As in
Act 1 (1c), you can drop any term that doesn't depend on $a$ or $b$,
since it won't change where the minimum is.

```{r fig.height=0.3*8}
blank_space()
```

`r nq(7, 6)`

In Act 1, calculus gave a formula for $\hat{p}$. That rarely happens
with real models: for this one, no algebra trick produces a formula
for $\hat{a}$ and $\hat{b}$. In practice we usually let a computer
search for the minimum numerically instead!

**1e.** In plain English (not R code), describe step by step what a
function must do to compute the NLL for one candidate parameter set $(a, b)$.
Then describe how you would use that function to find the MLE.

```{r fig.height=0.3*6}
blank_lines(6)
```

`r nq(8, 1)`

## Checkpoint: finalize your plan (~5 min)

Look back over 1c-1e and fix anything that no longer looks right
before you start coding.

`r nq(10, 0)`

## Part 2: Fit the model (laptop, ~50 min)

Each coding question is labeled with a recommendation:

- **Recommended: no AI:** try to build this yourself. R's help pages, web search, and
  classmates are fine.
- **AI optional/encouraged:** I'd recommend trying to solve this with an AI assistant. This implies the results are important but
I don't think you absolutely need to know how to code it!

After each one, circle how you actually used AI. Using AI never costs
points (but please disclose appropriately!).

**What to turn in:** your filled-in answers and your 2c figure as an image (upload in webcampus). You
don't need to submit your code.

The data are in `FrogChorus.csv` (columns `pond`, `dist_km`,
`males`), or you can paste them into R directly:

```{r data-code, echo=TRUE, eval=FALSE}
dist_km <- c(0.1, 0.2, 0.3, 0.5, 0.7, 1, 1.2, 1.5, 2, 2.5, 3, 4, 5, 6, 8)
males   <- c(1, 0, 1, 2, 6, 4, 8, 11, 13, 5, 12, 6, 11, 12, 14)
```

```{r data-code-check}
# the pasted vectors above must match the CSV
stopifnot(all.equal(dist_km, c(0.1, 0.2, 0.3, 0.5, 0.7, 1, 1.2, 1.5, 2, 2.5, 3, 4, 5, 6, 8)),
          all.equal(as.numeric(males), c(1, 0, 1, 2, 6, 4, 8, 11, 13, 5, 12, 6, 11, 12, 14)))
```

`r nq(4, 0)`

**2a.** *(Recommended: no AI.)* Write a function `nll_mono(par, y, d)` that implements your NLL
from 1d, where `par` holds $a$ and $b$. Check it: `nll_mono(c(10, 1),
males, dist_km)` should return about `r round(nll_mono(c(10, 1), males, dist_km), 2)`
if you used `dpois()`, or about
$`r round(nll_mono(c(10, 1), males, dist_km) - sum(lfactorial(males)), 2)`$
if you coded the formula yourself and dropped the constant term. (Why
can the second number be negative?)

Your value: ________

AI use (circle one): none / help or debugging / wrote the code

`r nq(7, 3)`

**2b.** *(recommended: no AI.)* Evaluate the NLL at your starting values from 1c. Then use
`optim()`, starting from those values, to find the MLE.

NLL at your starting values: ________

$\hat{a} =$ ________ &nbsp;&nbsp; $\hat{b} =$ ________ &nbsp;&nbsp;
NLL at the MLE: ________ &nbsp;&nbsp; convergence code: ________

Did `optim()` print any warnings? If so, what do you think caused
them, and do they mean the answer is wrong?

```{r fig.height=0.3*3}
blank_lines(3)
```

AI use (circle one): none / help or debugging / wrote the code

`r nq(5, 3)`

**2c.** *(AI optional/encouraged.)* Plot the data with your fitted curve and your hand-guessed
curve from 1c. How far from the road does the expected chorus reach
half its full size?

Half-size distance: ________ km

```{r fig.height=0.3*3}
blank_lines(3)
```

AI use (circle one): none / help or debugging / wrote the code

`r nq(7, 4)`

**2d.** *(recommended: no AI.)* A skeptic says the road has no effect: every pond has the same
expected count $\lambda$, whatever its distance. What is the MLE of
$\lambda$ under that model, and what is its NLL? (Keep or drop the
constant term the same way you did in 2a, so the two NLLs are
comparable.) Which model do the data support more? Does the size of the
difference convince you?

$\hat{\lambda} =$ ________ &nbsp;&nbsp; NLL: ________

```{r fig.height=0.3*4}
blank_lines(4)
```

AI use (circle one): none / help or debugging / wrote the code

`r nq(3, 5)`

**2e.** Where did you use AI, if anywhere, and what did you have to
check or fix? If you didn't use it, what was hardest to get working?

```{r fig.height=0.3*5}
blank_lines(5)
```

`r nq(7, 4)`

**If you finish early.** *(AI optional/encouraged.)* A colleague prefers a
different saturating curve, $\lambda(d) = \dfrac{a d}{b + d}$. Fit it
the same way and compare its NLL with the monomolecular model's. What
do you conclude?

```{r fig.height=0.3*4}
blank_lines(4)
```

AI use (circle one): none / help or debugging / wrote the code

`r nq(5, 6)`

## Reflection (~10 min)

In Act 1, a single 5-minute visit detected this frog only about 40% of
the time, even at sites where it was known to be present. With that in
mind, what do the counts in this Act measure? Could the road lower the
counts near it for reasons other than fewer frogs living there? Could
these data tell those explanations apart?

Also record any changes you made, new insights, and any "muddiest
points" or topics you want to review.

```{r fig.height=0.3*6}
blank_lines(6)
```
