---
title: "Lab 5, Act 1: Finding the Peak"
subtitle: "What detection probability do the survey data support best?"
author: "NRES 746, name: _______________   group: _____________________"
date: ""
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
header-includes:
  - \usepackage{needspace}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}

# keeps a prompt and its answer space on one PDF page: place `r nq(...)` on its
# own line just above the prompt. Emits nothing in non-LaTeX output. lines =
# lines of prompt text, rows = rows of blank space below it (0.3 in each),
# extra = any other height (in inches), e.g. a figure or table
nq <- function(lines, rows = 0, extra = 0) {
  if (!knitr::is_latex_output()) return("")
  knitr::asis_output(sprintf("\n\\needspace{%.2fin}\n", 0.23 * lines + 0.15 + 0.3 * rows + extra))
}
```

## The scenario

A rare frog in a Sierra Nevada spring complex calls only on some nights,
so a survey can miss it even when it's present at a site. To estimate how often a
single visit detects it, you visit three distinct headwaters where the species is
*known to be present*, ten times each. On each visit you listen for 5
minutes and record whether you heard at least one call. Here are your observed measurements:

```{r detection-data}
detect <- data.frame(
  Wetland = c("A", "B", "C"),
  Visits = c(10, 10, 10),
  Detections = c(3, 2, 7)
)
knitr::kable(detect, col.names = c("Wetland", "Visits", "Visits with a detection"),
             align = "ccc")
```

Assume every wetland has the same single-visit detection probability
$p$, and that visits are independent.

**What value of $p$ do these data support best, and can we find it
without running a brute-force grid search?** 

Please keep your laptop closed for this section (calculator/phone is okay!).

`r nq(4, 6)`

## Part 1: The likelihood (individual, ~15 min)

**1a.** Write the likelihood of all three wetlands' data,
$\mathcal{L}_3(p)$, using product notation.

```{r fig.height=0.3*6}
blank_space()
```

`r nq(3, 8)`

**1b.** Take the log and simplify $\ell_3(p) = \log \mathcal{L}_3(p)$
into a sum, as far as you can.

```{r fig.height=0.3*8}
blank_space()
```

`r nq(4, 3)`

**1c.** Your answer to 1b includes binomial coefficients such as
$\binom{10}{3}$. If you left them out, would that change which value of
$p$ maximizes $\ell_3(p)$? Why or why not?

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(5, 12)`

## Part 2: Find the peak (individual, ~20 min)

**2a.** Find the derivative with respect to $p$ of either
$\mathcal{L}_3(p)$ or $\ell_3(p)$, your choice (both peak at the same
$p$). In general, the log-likelihood is easier to differentiate.

```{r fig.height=0.3*12}
blank_space()
```

`r nq(3, 9)`

**2b.** Set your derivative equal to zero and solve for $\hat{p}$, the
maximum likelihood estimate. Try to do this on your own first, but you can get help by
looking at the Bolker book (beginning of Ch. 6). 

```{r fig.height=0.3*8}
blank_space()
```

$\hat{p} =$ ________

`r nq(4, 8)`

**2c.** Now suppose you surveyed $n$ wetlands, each visited $N$ times,
with $k_i$ detections at wetland $i$. Write a general formula for
$\hat{p}$. Does it make sense?

```{r fig.height=0.3*5}
blank_space()
```

```{r fig.height=0.3*2}
blank_lines(2)
```

`r nq(5, 4, extra = 3)`

## Part 3: How sure are we? (individual, ~10 min)

A colleague ran the same survey at 30 spring complexes where the species is
known to be present (10 visits each) and
heard the frog on 120 of the 300 visits. The figure shows the negative
log-likelihood (NLL) curve for each survey, each shifted down so its
minimum sits at zero.

```{r two-curves, fig.width=5, fig.height=2.9}
p_grid <- seq(0.1, 0.99, by = 0.001)
nll_3 <- -(12 * log(p_grid) + 18 * log(1 - p_grid))
nll_30 <- -(120 * log(p_grid) + 180 * log(1 - p_grid))
par(mar = c(4, 4.5, 0.5, 0.5))
plot(NA, xlim = c(0.1, 1), ylim = c(0, 6),
     xlab = "p", ylab = "NLL minus its minimum")
abline(v = seq(0.1, 1, 0.1), h = 0:6, col = "grey85")
lines(p_grid, nll_3 - min(nll_3), lwd = 2)
lines(p_grid, nll_30 - min(nll_30), lwd = 2, lty = 2)
legend(x = 0.74, y = 6.1, legend = c("3 wetlands", "30 wetlands"),
       lwd = 2, lty = c(1, 2), cex = 0.8, bg = "white", box.col = "white")
```

**3a.** Do the two surveys give the same $\hat{p}$? Which survey pins
$p$ down more tightly, and what feature of the curve shows it?

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(4, 6)`

## Group discussion and revision (~10 min)

Compare your answers with your group, especially your derivative in 2a
and your answer to 3a, and revise as needed. Record any changes you
made, new insights, and any "muddiest points" or topics you want to
review.

```{r fig.height=0.3*6}
blank_lines(6)
```
