---
title: "Lab 6, Act 1: Reading the Surface"
subtitle: "How much do fumigation trials tell us about a lethal dose?"
author: "NRES 746, name: _______________   group: _____________________"
date: ""
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
header-includes:
  - \usepackage{needspace}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}

# keeps a prompt and its answer space on one PDF page: place `r nq(...)` on its
# own line just above the prompt. Emits nothing in non-LaTeX output. lines =
# lines of prompt text, rows = rows of blank space below it (0.3 in each),
# extra = any other height (in inches), e.g. a figure or table
nq <- function(lines, rows = 0, extra = 0) {
  if (!knitr::is_latex_output()) return("")
  knitr::asis_output(sprintf("\n\\needspace{%.2fin}\n", 0.23 * lines + 0.15 + 0.3 * rows + extra))
}

beetles <- read.csv("../data/BeetleFumigation.csv")
x <- beetles$log10_dose
n_exposed <- beetles$beetles
killed <- beetles$killed

nll_ab <- function(a, b, xx = x) {
  -sum(dbinom(killed, size = n_exposed, prob = plogis(a + b * xx), log = TRUE))
}
fit <- optim(c(-60, 34), function(p) nll_ab(p[1], p[2]), method = "BFGS",
             control = list(maxit = 1000, reltol = 1e-12))
fit_c <- optim(c(1, 34), function(p) nll_ab(p[1], p[2], x - 1.8), method = "BFGS",
               hessian = TRUE, control = list(reltol = 1e-12))
lr_cut <- qchisq(0.95, df = 1) / 2
```

## The scenario

Flour beetles (*Tribolium confusum*) infest stored grain, and fumigants
are a standard control. In a classic toxicology experiment, batches of
about 60 beetles were exposed to one of eight concentrations of carbon
disulphide gas, and the number killed in each batch was recorded
(Bliss 1935). Dose-response experiments like this one are used across
agriculture, environmental chemistry, pharmacology, and ecotoxicology.
Here are the observed measurements:

```{r beetle-data}
beetle_tab <- data.frame(
  x = sprintf("%.4f", x),
  mgl = sprintf("%.1f", 10^x),
  n = n_exposed,
  k = killed
)
knitr::kable(beetle_tab, align = "cccc",
             col.names = c("log10 dose, x", "Dose (mg/L)", "Beetles exposed", "Beetles killed"))
```

A standard model is logistic in log dose. The probability that a beetle
exposed to log dose $x$ dies is
$$p(x) = \frac{1}{1 + e^{-(a + b x)}}, \qquad\text{equivalently}\qquad
\log\frac{p(x)}{1 - p(x)} = a + b x,$$
and the number killed at each dose is binomial with that probability.
The MLEs are $\hat a = `r round(fit$par[1], 1)`$ and
$\hat b = `r round(fit$par[2], 1)`$:

```{r beetle-fit, fig.width=5, fig.height=2.8}
par(mar = c(4, 4.5, 0.5, 0.5))
plot(x, killed / n_exposed, pch = 16, ylim = c(0, 1), xlim = c(1.68, 1.9),
     xlab = "log10 dose, x", ylab = "Proportion killed")
curve(plogis(fit$par[1] + fit$par[2] * x), add = TRUE, lwd = 2)
```

**What dose kills half the beetles, and how uncertain is that
estimate?** This Act works on the second half of that question,
using the shape of the likelihood.

Please keep laptops closed for this Act (a calculator or phone is fine).

`r nq(3, 0, extra = 4.5)`

## Part 1: The likelihood surface (individual, ~20 min)

The figure shows the negative log-likelihood (NLL) over $(a, b)$, minus
its minimum. The thick curve connects parameter pairs whose NLL is 1.92
above the minimum (the likelihood-ratio cutoff from Monday's lecture).
The dot is the MLE.

```{r surface-ab, fig.width=6.5, fig.height=4.2}
a_grid <- seq(-76, -46, length.out = 301)
b_grid <- seq(27, 42, length.out = 301)
z_ab <- outer(a_grid, b_grid, Vectorize(function(a, b) nll_ab(a, b))) - fit$value
par(mar = c(4, 4.5, 0.5, 0.5))
plot(NA, xlim = range(a_grid), ylim = range(b_grid), xaxs = "i", yaxs = "i",
     xlab = "Intercept, a", ylab = "Slope, b")
abline(v = seq(-76, -46, by = 2), h = seq(27, 42, by = 1), col = "grey85")
contour(a_grid, b_grid, z_ab, levels = lr_cut, lwd = 2.5, drawlabels = FALSE, add = TRUE)
points(fit$par[1], fit$par[2], pch = 16, cex = 1.1)
box()
```

`r nq(6, 5)`

**1a.** Estimate two intervals for $b$ by eye from the figure (rough
values are fine). The **slice** interval: hold $a$ at $\hat a$ and find
the range of $b$ that stays inside the thick curve. The **profile**
interval: the full range of $b$ values the thick curve reaches. Which
interval is honest, and why?

Slice: ________ to ________ \qquad Profile: ________ to ________

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(5, 6)`

**1b.** The intercept $a$ is the log-odds of death at what dose (as $x$,
and in mg/L)? Use your answer to explain why the valley is a long, thin
diagonal: why does every good fit with a steeper slope also need a more
negative intercept?

```{r fig.height=0.3*6}
blank_lines(6)
```

`r nq(10, 0, extra = 3.2)`

## Part 2: Curvature with two parameters (individual, ~20 min)

With one parameter, $\text{SE} = 1/\sqrt{\text{curvature}}$. With two
parameters, `optim(..., hessian = TRUE)` returns the
**Hessian**, a $2 \times 2$ table of second derivatives of the NLL at
the MLE:
$$H = \begin{pmatrix} H_{aa} & H_{ab} \\ H_{ab} & H_{bb} \end{pmatrix}.$$
The diagonal entries are the curvature along each parameter's axis. The
off-diagonal entry $H_{ab}$ measures how the slope in one direction
changes as you move in the other direction (the tilt of the valley).

Now **center** the dose: let $x' = x - 1.8$ and fit
$\log\frac{p}{1-p} = a' + b x'$. This describes the same set of
curves (with $a' = a + 1.8\,b$), so the fitted curve and $\hat b$ don't
change. Here is the new surface:

```{r surface-centered, fig.width=5, fig.height=3}
ac_grid <- seq(0.5, 1.45, length.out = 201)
bc_grid <- seq(26, 43, length.out = 201)
z_c <- outer(ac_grid, bc_grid, Vectorize(function(a, b) nll_ab(a, b, x - 1.8))) - fit_c$value
par(mar = c(4, 4.5, 0.5, 0.5))
plot(NA, xlim = range(ac_grid), ylim = range(bc_grid), xaxs = "i", yaxs = "i",
     xlab = "Centered intercept, a'", ylab = "Slope, b")
abline(v = seq(0.5, 1.45, by = 0.05), h = seq(26, 43, by = 1), col = "grey85")
contour(ac_grid, bc_grid, z_c, levels = lr_cut, lwd = 2.5, drawlabels = FALSE, add = TRUE)
points(fit_c$par[1], fit_c$par[2], pch = 16, cex = 1.1)
box()
```

For the centered fit, `optim()` reports (rows and columns in the order
$a'$, $b$):
$$H = \begin{pmatrix} 58.5 & -1.26 \\ -1.26 & 0.145 \end{pmatrix}.$$

`r nq(14, 7)`

Here are two candidate standard errors for $\hat b$:
$$\text{SE}_1 = \frac{1}{\sqrt{H_{bb}}}
\qquad\qquad
\text{SE}_2 = \frac{1}{\sqrt{H_{bb}}} \times \frac{1}{\sqrt{1 - r^2}},
\quad\text{where } r = \frac{-H_{ab}}{\sqrt{H_{aa} H_{bb}}}.$$
$\text{SE}_1$ uses only $b$'s own curvature, $H_{bb}$, and ignores the
off-diagonal entry. $\text{SE}_2$ uses the
off-diagonal entry $H_{ab}$ through $r$.

**2a.** Compute $\text{SE}_1$, $r$, and $\text{SE}_2$, and the
approximate 95% interval $\hat b \pm 1.96\,\text{SE}$ for each (use
$\hat b = `r round(fit$par[2], 2)`$). Which do you think better
characterizes our uncertainty about $b$, and why? Relate your answer to
Part 1.

```{r fig.height=0.3*7}
blank_space()
```

$\text{SE}_1 =$ ______ \quad $r =$ ______ \quad $\text{SE}_2 =$ ______

Interval using $\text{SE}_1$: ______ to ______ \qquad Using $\text{SE}_2$: ______ to ______

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(6, 7)`

**2b.** For the original, uncentered fit, $H_{bb} = 185$ and
$r = -0.99968$. Compute both standard errors and their intervals again.
Compare them with your slice and profile intervals from 1a, and with
your answers to 2a. Centering didn't change the fitted curve, so what
did it change?

```{r fig.height=0.3*4}
blank_space()
```

$\text{SE}_1 =$ ______ \quad $\text{SE}_2 =$ ______

Interval using $\text{SE}_1$: ______ to ______ \qquad Using $\text{SE}_2$: ______ to ______

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(4, 6)`

## Group discussion and revision (~10 min)

Compare your answers with your group, especially your explanation in 1b
and your answer to 2a, and revise as needed. Record any changes you
made, new insights, and any "muddiest points" or topics you want to
review.

```{r fig.height=0.3*6}
blank_lines(6)
```
