---
title: "Lab 6, Act 2: An Interval for the Lethal Dose"
subtitle: "How uncertain is our estimate of the LD50?"
author: "NRES 746, name: _______________   group: _____________________"
date: ""
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
header-includes:
  - \usepackage{needspace}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}

# keeps a prompt and its answer space on one PDF page: place `r nq(...)` on its
# own line just above the prompt. Emits nothing in non-LaTeX output. lines =
# lines of prompt text, rows = rows of blank space below it (0.3 in each),
# extra = any other height (in inches), e.g. a figure or table
nq <- function(lines, rows = 0, extra = 0) {
  if (!knitr::is_latex_output()) return("")
  knitr::asis_output(sprintf("\n\\needspace{%.2fin}\n", 0.23 * lines + 0.15 + 0.3 * rows + extra))
}

beetles <- read.csv("../data/BeetleFumigation.csv")
x <- beetles$log10_dose
n_exposed <- beetles$beetles
killed <- beetles$killed

nll_ld50 <- function(par) {
  -sum(dbinom(killed, size = n_exposed, prob = plogis(par[2] * (x - par[1])), log = TRUE))
}
check_val <- nll_ld50(c(1.75, 30))
check_val_nc <- check_val - sum(lchoose(n_exposed, killed))
```

## The scenario

A pest manager has seen the fumigation results from Act 1 and wants to
know the dose that kills half the beetles (the **LD50**) in mg/L, with
a 95% interval. The model from Act 1 has no LD50 parameter; the LD50 is
a function of $a$ and $b$. In this Act you rewrite the model so that
$L$, the LD50 on the log10 dose scale, appears as a parameter. You then
build two intervals for $L$ (one from the Hessian, one from the
likelihood profile) and convert them to mg/L at the end
($\text{LD50} = 10^L$).

**How uncertain is the LD50, and do the two methods agree?**

Try Part 1 on paper first, before you open a laptop or start typing
code.

`r nq(4, 7)`

## Part 1: Plan it on paper (individual, ~20 min)

**1a.** Act 1 used the model
$$\log\frac{p}{1-p} = a + b\,x,$$
where $p$ is the probability that a beetle dies at log10 dose $x$. Let
$L$ be the log10 dose at which $p = 0.5$. Find $L$ in terms of $a$ and
$b$, and show that this model can be rewritten as
$\log\frac{p}{1-p} = b\,(x - L)$. Then use Act 1's MLEs
($\hat a = -60.72$, $\hat b = 34.27$) to estimate $L$, and convert your
estimate to mg/L.

```{r fig.height=0.3*7}
blank_space()
```

$\hat L \approx$ ________ (log10 dose) \qquad LD50 $= 10^{\hat L} \approx$ ________ mg/L

`r nq(3, 8)`

**1b.** Write the negative log-likelihood, $\text{NLL}(L, b)$, for the
eight doses. You can drop terms that don't depend on $L$ or $b$.

```{r fig.height=0.3*8}
blank_space()
```

`r nq(5, 7)`

**1c.** In plain English (not R code), describe step by step how you
would compute the likelihood profile for $L$ and read a 95% interval
off it. (Recall how the profile differed from the slice in Act 1.)

```{r fig.height=0.3*7}
blank_lines(7)
```

`r nq(4, 0)`

## Checkpoint: finalize your plan (~5 min)

Look back over 1a-1c and fix anything that no longer looks right
before you start coding.

`r nq(12, 0, extra = 0.8)`

## Part 2: Fit and compare (laptop, ~50 min)

Each coding question is labeled with a recommendation:

- **Recommended: no AI:** try to build this yourself. R's help pages,
  web search, and classmates are fine.
- **AI optional/encouraged:** the result matters, but writing this code
  by hand isn't the point.

After each one, circle how you actually used AI. Using AI never costs
points; only undisclosed use does.

**What to turn in:** your filled-in answers and your 2c figure as an
image (upload in WebCampus). You don't need to submit your code.

The data are in `BeetleFumigation.csv` (columns `log10_dose`,
`beetles`, `killed`), or you can paste them into R directly:

```{r data-code, echo=TRUE, eval=FALSE}
log10_dose <- c(1.6907, 1.7242, 1.7552, 1.7842, 1.8113, 1.8369, 1.8610, 1.8839)
beetles    <- c(59, 60, 62, 56, 63, 59, 62, 60)
killed     <- c(6, 13, 18, 28, 52, 53, 61, 60)
```

```{r data-code-check}
# the pasted vectors above must match the CSV
stopifnot(all.equal(x, c(1.6907, 1.7242, 1.7552, 1.7842, 1.8113, 1.8369, 1.8610, 1.8839)),
          all.equal(n_exposed, c(59, 60, 62, 56, 63, 59, 62, 60)),
          all.equal(killed, c(6, 13, 18, 28, 52, 53, 61, 60)))
```

`r nq(12, 1)`

**2a.** *(Recommended: no AI.)* Write an NLL function whose first
argument is `par = c(L, b)`. Check it: at $L = 1.75$, $b = 30$ it should
return about `r round(check_val, 2)` if you used `dbinom()`, or about
`r round(check_val_nc, 2)` if you coded the formula yourself and dropped
the constant. Then fit it with `optim(..., hessian = TRUE)`, starting
from your values in 1a. Use `solve()` on the Hessian to get the
variance-covariance matrix; the square roots of its diagonal are the
standard errors. Report the 95% quadratic interval,
$\hat L \pm 1.96\,\text{SE}$.

$\hat L =$ ________ \quad $\hat b =$ ________ \quad NLL at the MLE: ________

$\text{SE}(\hat L) =$ ________ \quad interval for $L$: ________ to ________

AI use (circle one): no AI / AI helped (e.g., debugging) / AI wrote the code

`r nq(4, 3)`

**2b.** *(Recommended: no AI.)* Implement your profile from 1c and
report the 95% profile interval for $L$. How does it compare with 2a?

Profile interval for $L$: ________ to ________

```{r fig.height=0.3*3}
blank_lines(3)
```

AI use (circle one): no AI / AI helped (e.g., debugging) / AI wrote the code

`r nq(6, 3)`

**2c.** *(AI optional/encouraged.)* Plot your profile. Put $L$ on the
x-axis. On the y-axis, put the profile NLL at each value of $L$ minus
the smallest profile NLL, so the lowest point of the curve is at zero.
Add a horizontal line at 1.92 and mark both intervals (from 2a and 2b).
Then convert $\hat L$ and the profile interval to mg/L.

LD50 $= 10^{\hat L}$: ________ mg/L, from ________ to ________ mg/L

AI use (circle one): no AI / AI helped (e.g., debugging) / AI wrote the code

`r nq(3, 4)`

**2d.** Where did you use AI, if anywhere, and what did you have to
check or fix? If you didn't use it, what was hardest to get working?

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(7, 3)`

**If you finish early.** *(AI optional/encouraged.)* The dose that kills
99% of beetles, $L_{99}$, satisfies
$\log\frac{p}{1-p} = b\,(x - L_{99}) + \log\frac{0.99}{0.01}$. Profile
$L_{99}$ the same way and compare the width of its interval with the
LD50's.

```{r fig.height=0.3*3}
blank_lines(3)
```

AI use (circle one): no AI / AI helped (e.g., debugging) / AI wrote the code

`r nq(8, 6)`

## Reflection (~10 min)

Suppose the pest manager wants a dose that kills 99% of beetles instead
of 50%. Without computing anything, would you expect its interval to be
wider or narrower than the LD50's? (Look at where the data sit in Act
1's figure.) The Hessian and profile intervals nearly agreed here. When
would you expect them to disagree? Think back to the chytrid curve from
Monday's WOD.

Also record any changes you made, new insights, and any "muddiest
points" or topics you want to review.

```{r fig.height=0.3*6}
blank_lines(6)
```
