---
title: "Week 3, Lecture 2 WOD -- Answer Key"
subtitle: "A Probability Puzzle"
author: "NRES 746 -- Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 6, fig.height = 4)
```

This WOD is ungraded, so this key isn't a grading rubric -- it's a quick reference for
circulating during the 5-8 minute window, and a fully worked version of the Part 3 teaser for
whenever you reveal it in class.

## Part 1: Sevens vs. doubles

**1a.** No fixed answer -- the point is to get a guess on record before counting. Many students
guess "sum of 7" is more likely (it's the most commonly known fact about two dice) or "doubles"
feels rarer (fewer ways to *name* a double casually). The actual answer (1b-1d) is that they're
exactly tied.

**1b.** $\lVert\Omega\rVert = 36$ (6 outcomes for the red die $\times$ 6 for the blue die,
ordered pairs).

**1c.** Outcomes summing to 7: $(1,6),(2,5),(3,4),(4,3),(5,2),(6,1)$ -- 6 outcomes.
$P(\text{sum}=7) = 6/36 = 1/6 \approx 0.167$.

**1d.** Doubles: $(1,1),(2,2),(3,3),(4,4),(5,5),(6,6)$ -- 6 outcomes.
$P(\text{doubles}) = 6/36 = 1/6 \approx 0.167$.

**1e.** They're **equally likely** -- both are exactly $6$ of the $36$ equally-likely outcomes.
Students who guessed "sum of 7" are more likely often reason from partial memory of dice-game
lore (7 is famous in craps) without actually enumerating; students who guessed "doubles" often
reason that a double feels like a narrower, more special category. Worth pointing out: this only
works out to a tie because 7 is the *most-repeated* sum on two dice (six ways to make it, the
maximum for any sum) -- any other sum (say, a sum of 8, with 5 ways) would actually lose to
doubles.

## Part 2: How many tickets?

$$\binom{30}{5} = \frac{30!}{5!\,25!} = \frac{30\times29\times28\times27\times26}{5\times4\times3\times2\times1}$$

Students only need the expression, not the integer, but for reference:

```{r ticket-count}
choose(30, 5)
```

$= 142{,}506$ tickets. **Common snag:** forgetting to divide by $5!$ (i.e. answering with the
*permutation* count $30\times29\times28\times27\times26$ instead of the *combination* count) --
that's the "why do you have to divide by something at the end" the hint is pointing at: without
dividing, every 5-number ticket gets counted once for each of the $5!=120$ orders you could have
written its numbers down in.

## Part 3: The birthday problem

**The guess itself has no fixed answer** (that's the point of a teaser), but the correct answer,
once computed, is **yes** -- with 25 people, a shared birthday is *more* likely than not. Most
students guess "no," reasoning (correctly, as far as it goes) that 25 is a small fraction of
365 -- the whole reason this is a famous "paradox" is that the intuitive comparison (25 vs. 365)
is the wrong comparison to make, as the derivation below shows.

### Setting up the calculation

Assume 365 equally likely birthdays (ignore Feb. 29 for simplicity) and that everyone's birthday
is independent of everyone else's. Directly computing $P(\text{at least one shared birthday})$
is awkward, because "at least one match" could happen in a huge number of different ways (exactly
one pair matches, two separate pairs match, three people share one birthday, etc.). The standard
trick: compute the complement instead.

$$P(\text{at least one match}) = 1 - P(\text{no matches at all})$$

"No matches at all" means all 25 birthdays are different -- a much easier event to count.

### Building "no matches" one person at a time

Line the 25 people up and assign birthdays one at a time, and ask: given everyone assigned so
far has a distinct birthday, what's the probability the *next* person also avoids all of them?

- Person 1: no constraint yet -- probability of "still all distinct" $=365/365$.
- Person 2: must avoid person 1's one taken birthday -- probability $=364/365$.
- Person 3: must avoid the 2 already taken -- probability $=363/365$.
- ...
- Person $k$: must avoid the $(k-1)$ already-taken birthdays -- probability $=(366-k)/365$.
- ...
- Person 25: must avoid the 24 already taken -- probability $=341/365$.

Because each assignment is independent given the ones before it, multiply all 25 probabilities
together (this is the same chain-of-conditional-probabilities logic Tuesday's Lab 3 Act 1 used
to build up the classifier PPV calculation, and that today's lecture formalizes as Bayes' rule):

$$P(\text{no matches}) = \prod_{k=1}^{25}\frac{366-k}{365}
= \frac{365}{365}\cdot\frac{364}{365}\cdot\frac{363}{365}\cdots\frac{341}{365}$$

### Computing it

```{r birthday-calc}
n <- 25
p_no_match <- prod((365 - 0:(n - 1)) / 365)
p_no_match
p_match <- 1 - p_no_match
p_match
```

$P(\text{no matches}) \approx 0.431$, so $P(\text{at least one match}) \approx 0.569$ --
**56.9%**, comfortably past the 50% line.

### Why this feels wrong (worth saying out loud in class)

The instinct to compare "25" against "365" is comparing the wrong quantities. The real driver is
the number of *pairs* of people, since any pair could be the one that matches:

```{r pairs}
choose(25, 2)
```

25 people make $\binom{25}{2}=300$ distinct pairs. Each individual pair only has a $1/365$
chance of sharing a birthday, but with 300 independent-ish chances to get a match, the odds of
*at least one* hit compound quickly -- this is the same "many small chances add up" logic as
Act 1's classifier false-positive rate, just running in the opposite direction (there, many
chances to falsely flag; here, many chances to coincidentally match).

### A table for context, if useful during the reveal

```{r birthday-table}
ns <- c(5, 10, 15, 20, 23, 25, 30, 40, 50, 70)
data.frame(
  n = ns,
  P_match = round(1 - sapply(ns, function(n) prod((365 - 0:(n - 1)) / 365)), 3)
)
```

The commonly-cited "tipping point" is $n=23$ (just past 50%); this WOD deliberately asks about
25 instead so students can't just recall that number from prior exposure to the classic version
of the puzzle.
