---
title: "Week 3, Lecture 2 WOD: Answer Key"
subtitle: "A Probability Puzzle"
author: "NRES 746, Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 6, fig.height = 4)
```

Each item repeats the handout's question, then gives the answer, a
short rationale, and mistakes to watch for. Questions with one correct
answer are answered directly. The guesses (1a, Part 3) have no right
answer at the time they're made. Part 3 is worked out in full at the
end, for whenever you reveal it in class. All numbers are computed in R
when this file is knitted.

## Part 1: Sevens vs. doubles

**1a.**

> Before doing any counting: which do you think is more likely, rolling
> a **sum of 7**, or rolling **doubles** (both dice show the same
> number)? Circle one: sum of 7 / doubles / equally likely

**Answer.** Any guess is fine. The correct answer (1b-1d) is
**equally likely**.

**Why.** The point is to get a guess on record before counting, so
students can see where their intuition was off.

**Watch for.** Students who skip ahead and count first. Ask them to
circle their gut answer anyway.

**1b.**

> How many equally-likely outcomes are there in the full sample space
> $\Omega$ (all ordered pairs of red/blue results)?
> $\lVert \Omega \rVert =$ ____

**Answer.** 36.

**Why.** 6 outcomes for the red die times 6 for the blue die. The dice
are distinguishable, so (2, 5) and (5, 2) are different outcomes.

**Watch for.** Answering 21, which counts unordered pairs. Those 21
outcomes aren't equally likely: (2, 5) can happen two ways, but (3, 3)
only one.

**1c.**

> How many of those outcomes sum to 7? List them if it helps.
> $P(\text{sum}=7) =$ ____

**Answer.** 6 outcomes: (1,6), (2,5), (3,4), (4,3), (5,2), (6,1).
$P(\text{sum}=7) = 6/36 = 1/6$.

**Why.** Every red value from 1 to 6 has exactly one blue value that
makes 7.

**Watch for.** Listing only 3 outcomes (treating (1,6) and (6,1) as
the same) and getting $3/36$.

**1d.**

> How many outcomes are doubles? $P(\text{doubles}) =$ ____

**Answer.** 6 outcomes: (1,1) through (6,6).
$P(\text{doubles}) = 6/36 = 1/6$.

**Why.** Each red value has exactly one matching blue value.

**Watch for.** Students who got $3/36$ in 1c but $6/36$ here are
counting the two events inconsistently.

**1e.**

> Was your guess in 1a correct? If not, what made doubles or sevens
> feel more/less likely than it actually is?

**Answer (discussion).** They're tied, at 6 of 36 each.

**Why.** Common reasons for a wrong guess: 7 is famous as the most
common sum (true), so it feels like it should beat everything;
doubles feel special, so they feel rare. Worth pointing out: the tie
happens only because 7 is the most common sum. Any other sum, such as
8 (5 ways), loses to doubles.

```{r sums-vs-doubles}
sums <- outer(1:6, 1:6, "+")
table(sums)
```

**Watch for.** "I guessed right" with no reason. Ask what they would
have guessed for a sum of 8.

## Part 2: How many tickets?

> A small regional lottery asks you to pick 5 distinct numbers from 1
> to 30 (order doesn't matter). How many different tickets are
> possible? Leave your answer as an expression (use factorials!).

**Answer.**
$$\binom{30}{5} = \frac{30!}{5!\,25!} = \frac{30\times29\times28\times27\times26}{5!}
= `r format(choose(30, 5), big.mark = "{,}")`$$
(The integer isn't required.)

**Why.** There are $30 \times 29 \times 28 \times 27 \times 26 = 30!/25!$
ways to write down 5 distinct numbers in order. Each ticket appears
$5! = 120$ times in that list (once per ordering), so divide by $5!$.

**Watch for.**

- Stopping at $30!/25!$ (the ordered count), which is 120 times too
  big. This is what the hint's "how do you account for order" is
  getting at.
- Writing $30^5$, which allows repeated numbers.

## Part 3: A teaser (guess only)

> Suppose there are 25 people in this room. Do you think it's more
> likely than not (better than a 50% chance) that **at least two of
> them share a birthday**? Circle your guess: yes / no

**Answer.** Any guess is fine at the time. The correct answer is
**yes**: about
`r round(100 * (1 - prod((365 - 0:24) / 365)), 1)`%.

**Why.** The full calculation is below. The short version: what
matters is the number of *pairs* of people (300), not the number of
people compared with 365 days.

**Watch for.** Most students guess "no," comparing 25 with 365. That's
the comparison the reveal should correct.

### Full worked solution, for the reveal

Assume 365 equally likely birthdays (ignore Feb. 29) and independence
between people.

**Use the complement.** "At least one match" can happen in many ways
(one pair, two pairs, three people on one day, ...). "No matches"
happens in only one way, so compute that and subtract:
$$P(\text{at least one match}) = 1 - P(\text{no matches})$$

**Build "no matches" one person at a time.** Given that everyone so far
has a different birthday, person $k$ must avoid the $k - 1$ days
already taken:

- Person 1: $365/365$
- Person 2: $364/365$
- Person 3: $363/365$
- ...
- Person 25: $341/365$

Multiply these conditional probabilities (the same chaining used for
the detection-classifier calculation in Lab 3, Act 1):
$$P(\text{no matches}) = \prod_{k=1}^{25}\frac{366-k}{365}
= \frac{365}{365}\cdot\frac{364}{365}\cdots\frac{341}{365}$$

```{r birthday-calc}
n <- 25
p_no_match <- prod((365 - 0:(n - 1)) / 365)
c(p_no_match = p_no_match, p_match = 1 - p_no_match)
```

$P(\text{no matches}) \approx `r round(p_no_match, 3)`$, so
$P(\text{at least one match}) \approx `r round(1 - p_no_match, 3)`$.

**Why it feels wrong.** Any pair of people could be the match, and 25
people make many pairs:

```{r pairs}
choose(25, 2)
```

Each pair has only a $1/365$ chance of sharing a birthday, but 300
chances add up quickly.

**Optional aside: computing in log space.** The same answer can be
written with factorials, $P(\text{no matches}) = \frac{365!}{340!\,365^{25}}$,
but $365!$ is far too large for a computer to store. Taking logs
avoids that (the same trick as the log-likelihood later in the
course):

```{r birthday-log}
exp(lfactorial(365) - lfactorial(365 - n) - n * log(365))
```

**For context during the reveal:**

```{r birthday-table}
ns <- c(10, 15, 20, 23, 25, 30, 40, 50)
data.frame(n = ns,
           P_match = round(1 - sapply(ns, function(m) prod((365 - 0:(m - 1)) / 365)), 3))
```

The commonly cited tipping point is $n = 23$. The WOD asks about 25 so
students can't just recall the classic number.
