---
title: "Week 8, Lecture 1 WOD: Answer Key"
subtitle: "How Much Did the Prior Matter?"
author: "NRES 746, Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6, fig.height = 2.6)
```

```{r compute, include=FALSE}
teams <- list(
  A = list(prior = c(2, 2), data = c(45, 15)),
  B = list(prior = c(80, 20), data = c(3, 2)),
  C = list(prior = c(16, 4), data = c(10, 10))
)
beta_mean <- function(ab) ab[1] / sum(ab)
beta_sd <- function(ab) sqrt(prod(ab) / (sum(ab)^2 * (sum(ab) + 1)))
tab <- t(sapply(teams, function(tm) {
  po <- tm$prior + tm$data
  c(prior_worth = sum(tm$prior), n = sum(tm$data),
    prior_mean = beta_mean(tm$prior), mle = tm$data[1] / sum(tm$data),
    post_mean = beta_mean(po), prior_sd = beta_sd(tm$prior),
    lik_sd = beta_sd(tm$data + 1), post_sd = beta_sd(po))
}))
r2 <- function(v) formatC(v, format = "f", digits = 2)
```

Each item repeats the handout's question, then gives the answer, a
short rationale, and mistakes to watch for. All three panels are
Beta-binomial (the setup from Week 7 Wednesday), so the numbers below
are exact. Students don't see any of them; the point is to read the
pictures.

| Team | Prior | Survived | Prior mean | Observed | Posterior mean |
|:------|:--------------|:----------|:------:|:------:|:------:|
| A | Beta(2, 2) | 45 of 60 | `r r2(tab["A", "prior_mean"])` | `r r2(tab["A", "mle"])` | `r r2(tab["A", "post_mean"])` |
| B | Beta(80, 20) | 3 of 5 | `r r2(tab["B", "prior_mean"])` | `r r2(tab["B", "mle"])` | `r r2(tab["B", "post_mean"])` |
| C | Beta(16, 4) | 10 of 20 | `r r2(tab["C", "prior_mean"])` | `r r2(tab["C", "mle"])` | `r r2(tab["C", "post_mean"])` |

A Beta($\alpha$, $\beta$) prior is "worth" about $\alpha + \beta$
animals: 4, 100, and 20 for Teams A, B, and C.

Use this WOD to lead straight into the germination example: the
supplier prior there (worth 200 seeds, against 40 seeds of data) puts
the germination trial between Teams B and C.

## Part 1: Read the panels

**1a.**

> Rank the three teams from most to least influenced by their prior.
> Give a one-line reason for each.

**Answer.** B (most), then C, then A (least).

- **B:** the posterior is almost on top of the prior; the data barely
  moved it.
- **C:** the posterior moved well away from the prior but didn't
  leave it behind; it sits between the prior and wherever the data
  point.
- **A:** the prior is nearly flat, so the posterior's location and
  width must come almost entirely from the data.

**Why.** The prior's influence shows in how far the posterior moved
from it and how much narrower it got. Both are tiny for B, moderate
for C, and large for A.

**Watch for.** Ranking by how tall or narrow the posterior is (B looks
"most certain" and some students read that as "most data"; see 2a).
Calling A's prior "no prior": it's weak, not absent.

**1b.**

> Sketch the shape of Team C's likelihood on its panel (the height
> doesn't matter). Where does it peak, and is it wider or narrower than
> the prior?

**Answer.** It peaks at about 0.5, on the far side of the posterior
from the prior, and is a little wider than the prior (SD
`r r2(tab["C", "lik_sd"])` vs. `r r2(tab["C", "prior_sd"])`).

```{r key-team-c}
s_seq <- seq(0.001, 0.999, length.out = 500)
par(mar = c(4, 4, 0.5, 0.5))
plot(s_seq, dbeta(s_seq, 26, 14), type = "l", lwd = 2.5, ylim = c(0, 6),
     xlab = "Annual survival probability", ylab = "Density")
lines(s_seq, dbeta(s_seq, 16, 4), lwd = 2.5, col = "grey60")
lines(s_seq, dbeta(s_seq, 11, 11), lwd = 2.5, lty = 2, col = "darkorange")
legend("topleft", c("prior", "likelihood (rescaled)", "posterior"),
       lwd = 2.5, lty = c(1, 2, 1), col = c("grey60", "darkorange", "black"),
       bty = "n", cex = 0.8)
```

**Why.** The posterior is a compromise between prior and likelihood,
so the likelihood has to sit on the opposite side of the posterior
from the prior. Here the prior and the data are each worth 20
tortoises, so the posterior mean (`r r2(tab["C", "post_mean"])`) lands
halfway between the prior mean (0.80) and the data (0.50). The
posterior is narrower than both because it combines two sources of
information.

**Watch for.** Drawing the likelihood between the prior and the
posterior (that would pull the posterior *toward* the prior's side).
Drawing it narrower than the posterior: the posterior should be the
narrowest curve in the panel. Exact position isn't the point; "below
0.65, around 0.5" is full credit.

## Part 2: How much data?

**2a.**

> Which team collared the most tortoises, and which collared the
> fewest? Team B has the narrowest posterior of the three. Does that
> mean Team B had the most data?

**Answer.** Team A had the most (60), Team B the fewest (5), Team C in
between (20). No: B's posterior is narrow because its prior was
narrow, not because of its data.

**Why.** How much the data contribute shows up in how much the
posterior differs from the prior. B's posterior is barely narrower than
its prior; A's is far narrower than its (nearly flat) prior. A
posterior's width reflects prior and data together, so a narrow
posterior alone can't tell you how much data went into it.

**Watch for.** "B is narrowest, so B had the most data." This is the
misconception the question is built to surface. Ask the room: if you
only saw Team B's posterior, could you tell whether it came from a
strong prior or a big study? (No; you need the prior too. This is why
priors must be reported.)
