---
title: "Week 4, Lecture 2 WOD -- Answer Key"
subtitle: "Expectation and the Sample Mean"
author: "NRES 746 -- Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 6, fig.height = 4)
```

This WOD is ungraded, so this key isn't a grading rubric -- it's a quick reference for
circulating during the 8-10 minute window. Questions with a single correct derivation or number
are answered directly; the one genuinely open-ended item (2b) is given a sample answer and
discussion points rather than an exact-match rubric.

## Part 1: A parameter vs. a statistic

**1.** $\mu$ is **not** a random variable -- it's a fixed (if unknown) property of the entire
population of trout, and it does not change no matter who does the sampling. $\bar{x}$ **is** a
random variable -- it's computed from whichever 5 trout happen to get caught, so a different
sample of 5 would give a different $\bar{x}$. This is the core distinction between a
**parameter** (a fixed number describing a population, e.g. $\mu$, $\sigma^2$) and a
**statistic** (a number computed from a sample, e.g. $\bar{x}$, which is itself a random
variable with its own distribution -- exactly what Parts 3-4 investigate).

## Part 2: Working with expectations

**2a.**
$$\mathbb{E}[2X+3] = 2\mathbb{E}[X]+3 = 2(22)+3 = 47 \qquad \text{Var}(2X+3) = 2^2\text{Var}(X) = 4(16) = 64$$

```{r part2a}
mu <- 22; sigma2 <- 16
2 * mu + 3        # E[2X+3]
2^2 * sigma2      # Var(2X+3)
```

**Common mistakes to watch for:** (1) forgetting to square the coefficient on the variance
(writing $\text{Var}(2X+3)=2(16)=32$ instead of $2^2(16)=64$); (2) trying to fold the $+3$ into
the variance calculation at all (e.g. $2(16)+3$) -- the additive constant shifts the mean but
has no effect on spread.

**2b.** The special condition is **independence** of $X$ and $Y$. Students aren't expected to
derive this algebraically yet (that requires covariance, which the course hasn't covered), but
"independent" is a reasonable guess since the problem explicitly describes $X$ and $Y$ as
independent measurements, and Part 3 relies on this same property.

*Instructor note, for context:* the *exact* necessary condition is a bit weaker than full
independence -- $\text{Var}(X+Y)=\text{Var}(X)+\text{Var}(Y)$ holds whenever $X$ and $Y$ are
merely **uncorrelated** ($\text{Cov}(X,Y)=0$), which independence guarantees but doesn't
require. Worth mentioning if a student pushes on this, but not worth deriving cold in an 8-10
minute WOD -- flag it as a preview for whenever covariance gets formally introduced.

## Part 3: Averaging two measurements

**3a.** $M = \frac12 X_1 + \frac12 X_2$, so $a=b=\frac12$.

$$\mathbb{E}[M] = \tfrac12\mathbb{E}[X_1] + \tfrac12\mathbb{E}[X_2] = \tfrac12\mu+\tfrac12\mu = \mu$$

For the variance, apply $\text{Var}(aX)=a^2\text{Var}(X)$ to each term, then add (valid because
$X_1, X_2$ are independent, per Part 2b):

$$\text{Var}(M) = \text{Var}\!\left(\tfrac12 X_1\right) + \text{Var}\!\left(\tfrac12 X_2\right)
= \tfrac14\text{Var}(X_1) + \tfrac14\text{Var}(X_2) = \tfrac14\sigma^2+\tfrac14\sigma^2 = \frac{\sigma^2}{2}$$

**Common mistake to watch for:** forgetting to square the $\frac12$ coefficients, which gives
$\text{Var}(M)=\frac12\sigma^2+\frac12\sigma^2=\sigma^2$ (no reduction at all) instead of the
correct $\sigma^2/2$ -- this is the exact error Part 3a's "name $a$ and $b$" scaffolding is
designed to catch.

**3b.**

```{r part3b}
sigma2 / 2
```

$\text{Var}(M) = 8$, compared to $\text{Var}(X_1)=16$ alone -- **half**. Averaging two
independent measurements cuts the variance in half (increases precision) relative to a single
measurement.

## Part 4: Generalizing to n measurements

**4a.** $\bar{X} = \frac1n\sum_{i=1}^n X_i = \frac1n X_1 + \frac1n X_2 + \cdots + \frac1n X_n$,
i.e. each term has coefficient $\frac1n$.

$$\mathbb{E}[\bar{X}] = \sum_{i=1}^n \frac1n\mathbb{E}[X_i] = \sum_{i=1}^n \frac1n\mu = n\cdot\frac{\mu}{n} = \mu$$

$$\text{Var}(\bar{X}) = \sum_{i=1}^n \text{Var}\!\left(\frac1n X_i\right) = \sum_{i=1}^n \frac{1}{n^2}\text{Var}(X_i)
= n\cdot\frac{\sigma^2}{n^2} = \frac{\sigma^2}{n}$$

(the sum-of-variances step again requires independence across all $n$ measurements.)

**4b.**

```{r part4b}
sigma2 / 5
```

$\text{Var}(\bar{X}) = 3.2$ at $n=5$. Checking against Part 3b: plugging $n=2$ into
$\sigma^2/n$ gives $16/2=8$, matching Part 3b's answer exactly -- a good internal consistency
check that the general formula is correct.

**4c.** As $n\to\infty$, $\text{Var}(\bar{X})=\sigma^2/n \to 0$. In plain language: with a very
large sample, the sample mean becomes an arbitrarily precise estimate of the true population
mean $\mu$ -- the more data you collect, the less your particular sample's average is likely to
stray from the true value.

**4d.**

$$\text{SD}(\bar{X}) = \sqrt{\text{Var}(\bar{X})} = \sqrt{\sigma^2/n} = \frac{\sigma}{\sqrt{n}}$$

```{r part4d}
sqrt(sigma2) / sqrt(5)
```

The standard deviation of $\bar{X}$ (the standard error) is about **1.79** at $n=5$, and it
scales as $1/\sqrt{n}$, not $1/n$ -- a genuinely different rate than the variance. **Common
mistake to watch for:** students conflating these two formulas (e.g. writing
$\text{SD}(\bar{X})=\sigma^2/n$, or claiming the standard error halves whenever $n$ doubles).
Worth spelling out explicitly: to cut the standard error in half, you need to **quadruple**
$n$, not double it -- precision improves more slowly than students' intuition from the variance
formula alone would suggest. This is the standard "law of diminishing returns" behind why
doubling a study's sample size rarely doubles its precision.
