---
title: "Week 6, Lecture 2 WOD: Answer Key"
subtitle: "Find the Bugs"
author: "NRES 746, Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)
```

Each item repeats the handout's question, then gives the answer, a
short rationale, and mistakes to watch for. Every buggy function is run
below, so the symptoms described are real R output. The corrected
functions match the Lab 4, Act 2 key, and they are the ones built live
in today's demo.

```{r fixed-functions, include=FALSE}
detect_prob <- function(k, pobs = 0.02) 1 - (1 - pobs)^k
simulate_counts <- function(t_survey, k, N0 = 1000, lambda = 0.75^(1/25)) {
  N_t <- N0 * lambda^t_survey
  rbinom(length(t_survey), size = round(N_t), prob = detect_prob(k))
}
detect_decline <- function(t_survey, counts, alpha = 0.05) {
  s <- summary(lm(log(counts) ~ t_survey))$coefficients[2, ]
  unname(s[1] < 0 && s[4] <= alpha)
}
t1 <- seq(1, 25, by = 1)
```

## Part 1: Find and fix

**1a.**

> `detect_prob()`: circle the bug, write the corrected line, and say
> what would go wrong if you ran the buggy version.

**Answer.** `1 - pobs^k` should be `1 - (1 - pobs)^k`.

**Why.** An animal is missed on one person-day with probability
$1 - p_{obs}$, and missed on all $k$ days with probability
$(1 - p_{obs})^k$. The buggy line uses the chance of being *detected*
every day instead. The result is a detection probability near 1, so
the simulated counts would be close to the true abundance:

```{r bug1}
detect_prob_bug <- function(k, pobs = 0.02) 1 - pobs^k
c(buggy = detect_prob_bug(3), fixed = detect_prob(3))
```

**Watch for.** Writing `(1 - pobs)^k` without the leading `1 -`. That
is the probability of being missed, not detected.

**1b.**

> `simulate_counts()`: circle the bug, write the corrected line, and
> say what would go wrong if you ran the buggy version.

**Answer.** `rbinom(1, ...)` should be `rbinom(length(t_survey), ...)`.

**Why.** The first argument of `rbinom()` is how many draws to make.
With 1, the function returns a single count (using only the first
year's abundance) instead of one count per survey year. `lm()` then
fails because `counts` and `t_survey` have different lengths:

```{r bug2}
simulate_counts_bug <- function(t_survey, k, N0 = 1000, lambda = 0.75^(1/25)) {
  N_t <- N0 * lambda^t_survey
  rbinom(1, size = round(N_t), prob = detect_prob(k))
}
set.seed(1)
simulate_counts_bug(t1, k = 3)
tryCatch(detect_decline(t1, simulate_counts_bug(t1, k = 3)),
         error = function(e) conditionMessage(e))
```

**Watch for.** Students who say "`size` should be `N0`." That would
remove the decline entirely. `size` must follow `N_t`.

**1c.**

> `detect_decline()`: circle the bug, write the corrected line, and
> say what would go wrong if you ran the buggy version.

**Answer.** `||` (or) should be `&&` (and):
`s[1] < 0 && s[4] <= alpha`.

**Why.** With "or," any negative slope counts as a detected decline,
significant or not, and so does any significant *increase*. Under a
stable population ($\lambda = 1$), about half of all slopes are
negative by chance, so the buggy rule "detects" a decline roughly half
the time instead of about 2.5%:

```{r bug3}
detect_decline_bug <- function(t_survey, counts, alpha = 0.05) {
  s <- summary(lm(log(counts) ~ t_survey))$coefficients[2, ]
  s[1] < 0 || s[4] <= alpha
}
set.seed(2)
null_sims <- replicate(1000, simulate_counts(t1, k = 3, lambda = 1), simplify = FALSE)
c(buggy = mean(sapply(null_sims, function(y) detect_decline_bug(t1, y))),
  fixed = mean(sapply(null_sims, function(y) detect_decline(t1, y))))
```

**Watch for.** Changing `<=` to `<`. That's a harmless difference, not
the bug.

**1d.**

> `estimate_power()`: circle the bug, write the corrected line, and say
> what would go wrong if you ran the buggy version.

**Answer.** `detected <- ...` should be `detected[i] <- ...`.

**Why.** Without the index, each pass through the loop overwrites the
whole vector with a single `TRUE` or `FALSE`. After the loop,
`detected` holds only the last replicate, so the "power" is always
exactly 0 or 1:

```{r bug4}
estimate_power_bug <- function(k, s, R = 1000) {
  t_survey <- seq(1, 25, by = s)
  detected <- logical(R)
  for (i in seq_len(R)) {
    counts <- simulate_counts(t_survey, k)
    detected <- detect_decline(t_survey, counts)
  }
  mean(detected)
}
set.seed(3)
replicate(8, estimate_power_bug(3, 1, R = 50))
```

**Watch for.** Students who fix it by moving `mean()` inside the loop.
That still keeps only one replicate's result.

## Part 2: Catch it next time

> Pick one of the bugs. Describe a quick test, one line of code and the
> answer you'd expect, that would have caught it before you trusted the
> function.

**Sample answers (any one is fine).**

- **1a:** `detect_prob(1)` should return `0.02` (one day of effort
  should equal `pobs`).
- **1b:** `length(simulate_counts(1:25, 3))` should return `25`.
- **1c:** run `detect_decline()` many times on data simulated with
  `lambda = 1`. It should return `TRUE` only about 2-3% of the time.
- **1d:** run `estimate_power(3, 1)` twice. Each should be a value
  between 0 and 1 (about 0.8 here), not exactly 0 or 1.

**Why.** Good tests use cases where you already know the answer: an
edge case (`k = 1`), a shape check (`length()`), or a scenario with
known truth (no decline). These are the same checks built into the Lab
4, Act 2 handout, and the same ones to demand from AI-written code in
today's demo.

**Watch for.** "Run it and see if the answer looks reasonable." Bug 1a
gives counts that look like plausible numbers unless you know what to
expect. A test needs a specific expected value.
