---
title: "Week 6, Lecture 2 WOD: Answer Key"
subtitle: "Find the Bugs"
author: "NRES 746, Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)
```

Each item repeats the handout's question, then gives the answer, a
short rationale, and mistakes to watch for. Both buggy functions are
run below, so the symptoms described are real R output. The corrected
functions match the Lab 4, Act 2 key, and they are the ones built live
in today's demo.

```{r fixed-functions, include=FALSE}
detect_prob <- function(k, pobs = 0.02) 1 - (1 - pobs)^k
simulate_counts <- function(t_survey, k, N0 = 1000, lambda = 0.75^(1/25)) {
  N_t <- N0 * lambda^t_survey
  rbinom(length(t_survey), size = round(N_t), prob = detect_prob(k))
}
detect_decline <- function(t_survey, counts, alpha = 0.05) {
  s <- summary(lm(log(counts) ~ t_survey))$coefficients[2, ]
  unname(s[1] < 0 && s[4] <= alpha)
}
```

## Part 1: Find and fix

**1a.**

> `detect_prob()`: circle the bug, write the corrected line, and say
> what would go wrong if you ran the buggy version.

**Answer.** `1 - pobs^k` should be `1 - (1 - pobs)^k`.

**Why.** An animal is missed on one person-day with probability
$1 - p_{obs}$, and missed on all $k$ days with probability
$(1 - p_{obs})^k$. The buggy line uses the chance of being *detected*
every day instead. The result is a detection probability near 1, so
the simulated counts would be close to the true abundance:

```{r bug1}
detect_prob_bug <- function(k, pobs = 0.02) 1 - pobs^k
c(buggy = detect_prob_bug(3), fixed = detect_prob(3))
```

**Watch for.** Writing `(1 - pobs)^k` without the leading `1 -`. That
is the probability of being missed, not detected.

**1b.**

> `estimate_power()`: circle the bug, write the corrected line, and say
> what would go wrong if you ran the buggy version.

**Answer.** `detected <- ...` should be `detected[i] <- ...`.

**Why.** Without the index, each pass through the loop overwrites the
whole vector with a single `TRUE` or `FALSE`. After the loop,
`detected` holds only the last replicate, so the "power" is always
exactly 0 or 1:

```{r bug2}
estimate_power_bug <- function(k, s, R = 1000) {
  t_survey <- seq(1, 25, by = s)
  detected <- logical(R)
  for (i in seq_len(R)) {
    counts <- simulate_counts(t_survey, k)
    detected <- detect_decline(t_survey, counts)
  }
  mean(detected)
}
set.seed(3)
replicate(8, estimate_power_bug(3, 1, R = 50))
```

**Watch for.** Students who fix it by moving `mean()` inside the loop.
That still keeps only one replicate's result.

## Part 2: Catch it next time

> For one of the bugs, what quick check (one line of code and the
> answer you'd expect) would have caught it?

**Sample answers (either is fine).**

- **1a:** `detect_prob(1)` should return `0.02` (one day of effort
  should equal `pobs`).
- **1b:** run `estimate_power(3, 1)` twice. Each should be a value
  between 0 and 1 (about 0.8 here), not exactly 0 or 1.

**Why.** Good checks use cases where you already know the answer: an
edge case (`k = 1`) or a range of values the output must fall in.
These are the same kinds of checks to demand from AI-written code in
today's demo.

**Watch for.** "Run it and see if the answer looks reasonable." Bug 1a
gives counts that look plausible unless you know what to expect. A
check needs a specific expected value.
