---
title: "Week 8, Lecture 1 WOD: Answer Key"
subtitle: "Newton's Method by Hand (take-home)"
author: "NRES 746, Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.width = 6, fig.height = 3)
```

```{r compute, include=FALSE}
nll <- function(l) 6 * l - 18 * log(l)
d1 <- function(l) 6 - 18 / l
d2 <- function(l) 18 / l^2
f <- round(nll(c(1.5, 2, 2.5)), 4)
fwd <- (f[3] - f[2]) / 0.5
ctr <- (f[3] - f[1]) / 1
sec <- (f[3] - 2 * f[2] + f[1]) / 0.25

steps <- matrix(NA, 3, 4, dimnames = list(NULL, c("guess", "slope", "curv", "next")))
l <- 2
for (i in seq_len(3)) {
  steps[i, ] <- c(l, d1(l), d2(l), l - d1(l) / d2(l))
  l <- steps[i, 4]
}
from7 <- 7 - d1(7) / d2(7)
fd_step <- 2 - ctr / sec

log_steps <- numeric(4)
th <- log(7)
for (i in seq_along(log_steps)) {
  th <- th - (6 * exp(th) - 18) / (6 * exp(th))
  log_steps[i] <- exp(th)
}
r4 <- function(v) formatC(v, format = "f", digits = 4)
r3 <- function(v) formatC(v, format = "f", digits = 3)
```

Each item repeats the handout's question, then gives the answer, a
short rationale, and mistakes to watch for. All numbers are computed
in R when this file is knitted.

Plan about 8 minutes at the start of Wednesday's class to go over this,
focusing on 3b and 3c. 3c leads straight into the lecture: the fixes
students propose there (smaller steps, better starting values, fitting
on the log scale) are what real optimizers do.

## Part 1: Slope and curvature

**1a.**

> Find $\text{NLL}'(\lambda)$ and $\text{NLL}''(\lambda)$ for
> $\text{NLL}(\lambda) = 6\lambda - 18\log\lambda$. Then set
> $\text{NLL}'(\lambda) = 0$ and solve for the MLE, $\hat\lambda$.

**Answer.** $\text{NLL}'(\lambda) = 6 - \dfrac{18}{\lambda}$,
$\text{NLL}''(\lambda) = \dfrac{18}{\lambda^2}$, and $\hat\lambda = 3$.

**Why.** $\frac{d}{d\lambda}\log\lambda = 1/\lambda$ and
$\frac{d}{d\lambda}\lambda^{-1} = -\lambda^{-2}$ (the two minus signs
cancel). $6 - 18/\lambda = 0$ gives $\lambda = 18/6 = 3$, the sample
mean.

**Watch for.** A negative $\text{NLL}''$ (sign slip); the NLL has a
minimum, so its curvature must be positive.

## Part 2: Slopes without calculus

**2a.**

> Estimate $\text{NLL}'(2)$ two ways: with the forward formula (using
> $\lambda = 2$ and 2.5), and with a centered version that uses
> $\lambda = 1.5$ and 2.5. Compare both with the exact value from 1a.
> Which is closer?

**Answer.** Forward: $(-1.4932 - (-0.4766))/0.5 = `r r4(fwd)`$.
Centered: $(-1.4932 - 1.7016)/1 = `r r4(ctr)`$. Exact: $-3$. The
centered estimate is much closer.

**Why.** The forward difference is the slope of a chord from 2 to
2.5, which matches the curve's slope near the chord's middle, 2.25
(exact $\text{NLL}'(2.25) = -2$), not at 2. The centered chord's middle
is at 2.

**Watch for.** Dividing the centered difference by $h$ (0.5) instead
of $2h$ (1). Sign errors from subtracting a negative number.

**2b.**

> Curvature is the slope of the slope. Use that idea to build a
> finite-difference approximation to $f''(x)$ that uses only
> $f(x - h)$, $f(x)$, and $f(x + h)$. Then use it to estimate
> $\text{NLL}''(2)$ and compare with the exact value.

**Answer.**
$$f''(x) \approx \frac{f(x + h) - 2f(x) + f(x - h)}{h^2}$$
Estimate: $(-1.4932 + 0.9532 + 1.7016)/0.25 = `r r4(sec)`$. Exact:
$18/4 = 4.5$.

**Why.** Two forward slopes,
$f'(x + h/2) \approx [f(x+h) - f(x)]/h$ and
$f'(x - h/2) \approx [f(x) - f(x-h)]/h$, are a distance $h$ apart.
Their difference divided by $h$ is the slope of the slope at $x$:
$$\frac{1}{h}\left[\frac{f(x+h) - f(x)}{h} - \frac{f(x) - f(x-h)}{h}\right]$$
which simplifies to the formula above.

**Watch for.** Dividing by $h$ instead of $h^2$ (gives 2.32). Using
slopes at $x$ and $x + h$, which needs $f(x + 2h)$ and estimates the
curvature at $x + h$. A student who gets the right formula by
"guessing the pattern" should still be able to show where $h^2$ comes
from.

## Part 3: Newton's method

**3a.**

> In words: what does dividing by the curvature do to the size of the
> step? When will Newton's method take big steps, and when small ones?

**Answer.** Strong curvature means small steps; weak curvature (a
nearly straight NLL) means big steps. The sign of the slope sets the
direction (downhill).

**Why.** A sharply curved parabola has its bottom close by; a nearly
flat one has its bottom far away. Newton's method assumes the parabola
is a good stand-in for the NLL and jumps all the way to its bottom.

**Watch for.** "Steep slope means a big step" is only half of it: the
same slope gives a small step where the curvature is large.

**3b.**

> Start at $\lambda = 2$ and take three Newton steps, using your exact
> derivatives from 1a.

**Answer.**

| Step | Guess | Slope, NLL$'$ | Curvature, NLL$''$ | Next guess |
|:---:|:---:|:---:|:---:|:---:|
| 1 | `r r4(steps[1, 1])` | `r r4(steps[1, 2])` | `r r4(steps[1, 3])` | `r r4(steps[1, 4])` |
| 2 | `r r4(steps[2, 1])` | `r r4(steps[2, 2])` | `r r4(steps[2, 3])` | `r r4(steps[2, 4])` |
| 3 | `r r4(steps[3, 1])` | `r r4(steps[3, 2])` | `r r4(steps[3, 3])` | `r r4(steps[3, 4])` |

**Why.** The distance to the MLE goes 1, 0.33, 0.037, 0.0005: each
error is roughly the square of the one before. That fast ("quadratic")
convergence near the answer is why derivative-based methods are
popular.

**Watch for.** Rounding $\lambda$ to 2.67 and carrying the rounding
error forward (still converges, just more slowly). Plugging the new
guess into the old curvature.

**3c.**

> Now start at $\lambda = 7$ and take one step. What went wrong? Sketch
> the NLL curve and the parabola Newton's method used at $\lambda = 7$
> to explain why, and suggest one way to fix it.

**Answer.** $\text{NLL}'(7) = `r r3(d1(7))`$, $\text{NLL}''(7) =
`r r3(d2(7))`$, so the next guess is
$7 - `r r3(d1(7))`/`r r3(d2(7))` = `r r3(from7)`$. A Poisson mean
can't be negative, and the NLL isn't even defined there.

```{r key-overshoot}
l_seq <- seq(0.3, 10, length.out = 400)
parab <- function(l, l0) nll(l0) + d1(l0) * (l - l0) + 0.5 * d2(l0) * (l - l0)^2
par(mar = c(4, 4, 0.5, 0.5))
plot(l_seq, nll(l_seq), type = "l", lwd = 2.5, xlim = c(-4, 10), ylim = c(-10, 16),
     xlab = expression(lambda), ylab = "NLL")
l_par <- seq(-4, 10, length.out = 400)
lines(l_par, parab(l_par, 7), lwd = 2, lty = 2, col = "darkorange")
abline(v = 0, col = "grey60")
points(c(7, from7), c(nll(7), parab(from7, 7)), pch = 19, col = c("black", "darkorange"))
legend("topright", c("NLL", "Newton's parabola at 7"), lwd = 2, lty = c(1, 2),
       col = c("black", "darkorange"), bty = "n", cex = 0.8)
```

**Why.** Far from the MLE the NLL is nearly a straight line (small
curvature), so the matching parabola is very wide and its bottom is far
to the left, past zero. In general the update here simplifies to
$\lambda_{\text{new}} = 2\lambda - \lambda^2/3$, which is negative for
any start above 6.

Acceptable fixes (any one):

- **Shorter steps.** If a step makes the NLL worse or leaves the
  allowed range, halve it and try again ("backtracking").
- **Better starting values.** Start near a sensible guess, such as the
  sample mean.
- **Constraints.** Use a method that respects bounds (e.g.
  `method = "L-BFGS-B"` with `lower = 0.001`).
- **Fit on the log scale.** Optimize over $\theta = \log\lambda$, so
  every value of $\theta$ gives a positive $\lambda$. Newton's method on
  $\theta$ from $\lambda = 7$ gives $\lambda$ =
  `r paste(r3(log_steps), collapse = ", ")`: no overshoot. (This one
  comes up in Wednesday's lecture.)

**Watch for.** Concluding Newton's method is "broken." It works well
near the answer and badly far from it, which is why starting values
matter for every optimizer.

**3d.**

> *(If you have time.)* Take one Newton step from $\lambda = 2$ using
> your finite-difference estimates from 2a (centered) and 2b instead.
> How close is it to your first step in 3b?

**Answer.** $2 - (`r r4(ctr)`)/`r r4(sec)` = `r r4(fd_step)`$, versus
`r r4(steps[1, 4])` with exact derivatives. Close, and it would still
converge.

**Why.** Errors in the slope and curvature only change where each step
lands, not where the method stops ($\text{NLL}' = 0$). `optim()` uses a
much smaller $h$ (0.001 by default), so its estimates are closer still.

**Watch for.** Using the forward slope from 2a, which gives
$2 + 2.033/4.646 = 2.44$; still in the right direction, but a worse
step.
