---
title: "Week 7, Lecture 2 WOD: Answer Key"
subtitle: "Curvature by Hand (take-home)"
author: "NRES 746, Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 6, fig.height = 3)
```

```{r compute, include=FALSE}
haypiles <- c(2, 5, 3, 0, 4, 4)
curv_pois <- function(y) sum(y) / mean(y)^2
curv_6 <- curv_pois(haypiles)
curv_24 <- curv_pois(rep(haypiles, 4))

curv_binom <- function(k, n) n / ((k / n) * (1 - k / n))
wald <- function(k, n) k / n + c(-1, 1) * qnorm(0.975) / sqrt(curv_binom(k, n))
se_3 <- 1 / sqrt(curv_binom(12, 30))
se_30 <- 1 / sqrt(curv_binom(120, 300))
wald_chytrid <- wald(2, 20)

nll_chytrid <- function(p) -dbinom(2, 20, p, log = TRUE)
f_cut <- function(p) nll_chytrid(p) - nll_chytrid(0.1) - 2
range_chytrid <- c(uniroot(f_cut, c(1e-6, 0.1))$root, uniroot(f_cut, c(0.1, 0.9))$root)
```

Each item repeats the handout's question, then gives the answer, a
short rationale, and mistakes to watch for. Questions with one correct
answer are answered directly. Open-ended items (the "have you seen
this before" part of 2c, the second half of 2d, and 3b) get discussion
points instead. All numbers are computed in R when this file is
knitted.

Plan about 5-10 minutes at the start of Monday's class to go over this,
focusing on 2c and 2d.

## Part 1: Pika haypiles (Poisson)

**1a.**

> Find $\text{NLL}'(\lambda)$ and $\text{NLL}''(\lambda)$, where
> $\text{NLL}(\lambda) = n\lambda - \left(\sum y_i\right)\log\lambda$.

**Answer.** $\text{NLL}'(\lambda) = n - \dfrac{\sum y_i}{\lambda}$ and
$\text{NLL}''(\lambda) = \dfrac{\sum y_i}{\lambda^2}$.

**Why.** $\frac{d}{d\lambda}\log\lambda = 1/\lambda$, and
$\frac{d}{d\lambda}\lambda^{-1} = -\lambda^{-2}$; the two minus signs
cancel. $\sum y_i$ is a constant (the data are fixed).

**Watch for.** Treating $\sum y_i$ as something to differentiate.
Sign errors: $\text{NLL}''$ must be positive (the NLL has a minimum).
Students who differentiate the log-likelihood instead get
$-\sum y_i/\lambda^2$, which is fine as long as they flip the sign
before taking the square root.

**1b.**

> Evaluate $\text{NLL}''(\hat\lambda)$ at $\hat\lambda = 3$ for the six
> plots, and again for 24 plots with the same mean. Give the SE for
> each.

**Answer.** 6 plots: $18/9 = `r curv_6`$, SE $= 1/\sqrt{2} \approx
`r round(1 / sqrt(curv_6), 2)`$. 24 plots: $72/9 = `r curv_24`$, SE
$= 1/\sqrt{8} \approx `r round(1 / sqrt(curv_24), 2)`$.

**Why.** Plug $\sum y_i$ (18 or 72) and $\lambda = 3$ into 1a. The
first matches the `optim()` Hessian shown in Monday's lecture.

**Watch for.** Using $\sum y_i = 18$ for the 24-plot case. Reporting
the curvature as the SE (forgetting $1/\sqrt{\cdot}$).

**1c.**

> Substitute $\hat\lambda = \bar y$ into your $\text{NLL}''$ and
> simplify to a general formula for $\text{SE}(\hat\lambda)$. What
> happens to the SE when $n$ quadruples?

**Answer.** $\text{NLL}''(\hat\lambda) = \dfrac{n\hat\lambda}{\hat\lambda^2}
= \dfrac{n}{\hat\lambda}$, so $\text{SE}(\hat\lambda) =
\sqrt{\hat\lambda / n}$. Quadrupling $n$ halves the SE.

**Why.** $\sum y_i = n\bar y = n\hat\lambda$. The SE scales as
$1/\sqrt{n}$. (Since a Poisson variance equals its mean, this is the
familiar $\sqrt{\text{variance}/n}$ standard error of a mean.)

**Watch for.** Saying the SE drops by a factor of 4. That's the
curvature (4 times steeper); the SE drops by $\sqrt 4 = 2$.

## Part 2: Frog detection (binomial)

**2a.**

> Find $\text{NLL}''(p)$ for $\text{NLL}(p) = -K\log p - (N-K)\log(1-p)$.

**Answer.**
$\text{NLL}'(p) = -\dfrac{K}{p} + \dfrac{N-K}{1-p}$ and
$\text{NLL}''(p) = \dfrac{K}{p^2} + \dfrac{N-K}{(1-p)^2}$.

**Why.** $\frac{d}{dp}\log(1-p) = -\frac{1}{1-p}$ (chain rule); then
$\frac{d}{dp}(1-p)^{-1} = (1-p)^{-2}$ (chain rule again, the two minus
signs cancel). Both terms of $\text{NLL}''$ are positive.

**Watch for.** Dropping the chain-rule minus sign, which leaves
$\frac{K}{p^2} - \frac{N-K}{(1-p)^2}$ (can go negative, a giveaway).
The first derivative should match Lab 5 Act 1, 2a (with the sign
flipped if they used the log-likelihood there).

**2b.**

> Evaluate $\text{NLL}''(\hat p)$ and the SE for $K = 12$, $N = 30$ and
> for $K = 120$, $N = 300$. Check: one of your curvatures should be 125.

**Answer.** 3 wetlands: $\text{NLL}'' = `r curv_binom(12, 30)`$,
SE $\approx `r round(se_3, 3)`$. 30 wetlands:
$\text{NLL}'' = `r curv_binom(120, 300)`$, SE $\approx
`r round(se_30, 4)`$.

**Why.** For 3 wetlands: $12/0.16 + 18/0.36 = 75 + 50 = 125$. Ten
times the data gives ten times the curvature and an SE smaller by
$\sqrt{10} \approx 3.2$, matching the narrower dashed curve in Lab 5
Act 1, Part 3.

**Watch for.** Arithmetic with $0.4^2$ and $0.6^2$. Students who got
125 for the first may still forget to scale $K$ and $N$ together for the
second.

**2c.**

> Substitute $\hat p = K/N$ and simplify to a general formula for
> $\text{SE}(\hat p)$. Have you seen this formula before?

**Answer.** $\text{NLL}''(\hat p) = \dfrac{N}{\hat p(1-\hat p)}$, so
$\text{SE}(\hat p) = \sqrt{\dfrac{\hat p(1-\hat p)}{N}}$. This is the
standard error of a proportion from intro stats.

**Why.** With $K = N\hat p$ and $N - K = N(1-\hat p)$:
$\frac{N\hat p}{\hat p^2} + \frac{N(1-\hat p)}{(1-\hat p)^2} =
\frac{N}{\hat p} + \frac{N}{1-\hat p} = \frac{N}{\hat p(1-\hat p)}$.

**Discussion point.** The "canned" formula from intro stats is the
quadratic approximation to this likelihood. That's the payoff of the
whole WOD: familiar formulas fall out of the curvature of a
likelihood, and the same recipe works for models with no textbook
formula.

**Watch for.** Getting stuck on the common denominator in the last
step; the algebra is short once $K$ and $N-K$ are rewritten in terms of
$\hat p$.

**2d.**

> For 2 chytrid-positive frogs out of 20, build $\hat p \pm 1.96\,\text{SE}$.
> What's wrong with it? Compare it with the range you read off the NLL
> curve on Monday.

**Answer.** SE $= \sqrt{0.1 \times 0.9 / 20} \approx
`r round(1 / sqrt(curv_binom(2, 20)), 3)`$, so the interval is
`r round(wald_chytrid[1], 3)` to `r round(wald_chytrid[2], 3)`. The
lower end is a **negative prevalence**, which is impossible.
Monday's 2-unit range was about `r round(range_chytrid[1], 2)` to
`r round(range_chytrid[2], 2)`: entirely positive, and lopsided (longer
on the high side).

**Why.** The quadratic approximation is symmetric by construction, but
this NLL curve isn't: $\hat p$ sits close to the boundary at 0, where
the curve rises very steeply. With few positives the parabola is a poor
stand-in for the real curve. The likelihood-based range follows the
curve and so respects the boundary.

**Watch for.** Students who report "0 to 0.23" without comment
(truncating hides the problem rather than explaining it). Good answers
connect this to Monday's lecture: the Wald and likelihood-ratio
intervals agree when the curve is close to a parabola (lots of data,
$\hat p$ away from 0 or 1) and disagree here.

## Part 3: Check it in R

**3a.**

> *(Recommended: no AI.)* Write an NLL function for $K = 12$, $N = 30$
> using `dbinom()`, and minimize it with `optim(..., method = "Brent",
> lower = 0.001, upper = 0.999, hessian = TRUE)`. Does the Hessian match
> 2b?

**Answer.** About 125 (for example `r round(optim(0.5, function(p) -dbinom(12, 30, p, log = TRUE), method = "Brent", lower = 0.001, upper = 0.999, hessian = TRUE)$hessian[1, 1], 3)`). Yes, it matches.

```{r key-3a}
nll_frog <- function(p) -dbinom(12, size = 30, prob = p, log = TRUE)
fit <- optim(0.5, nll_frog, method = "Brent", lower = 0.001, upper = 0.999,
             hessian = TRUE)
fit$par
fit$hessian
```

**Why.** `optim()` approximates the second derivative by finite
differences, so it agrees with the calculus up to small numerical error.
The binomial coefficient term doesn't depend on $p$, so including it
(via `dbinom()`) doesn't change the curvature.

**Watch for.** Forgetting the minus sign (maximizing instead of
minimizing gives the MLE at a boundary and a nonsense Hessian).
Omitting `method = "Brent"`, which triggers a warning for
one-parameter problems. A label/disclosure mismatch here is something
to talk about, not a penalty.

**3b.**

> Where did you use AI, if anywhere, and what did you have to check or
> fix? If you didn't use it, what was hardest?

**Answer (discussion points).** No single correct answer. Useful
responses name a specific step (the chain rule in 2a, the algebra in 2c,
setting up `optim()`). If several students say the algebra in 2c was
hardest, walk through it on the board Monday.
