---
title: "Week 5, Lecture 1 WOD -- Answer Key"
subtitle: "Choosing a Data-Generating Model"
author: "NRES 746 -- Instructor Answer Key (not for student distribution)"
date: ""
output:
  pdf_document: default
  # word_document: default
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 6.5, fig.height = 3)
```

This WOD is ungraded, so this key is a quick reference for circulating during the 8-10 minute
window, not a rubric. Nothing here has a single correct answer. Each data set was simulated
from a specific generating model (the "truth", below), but several other answers are defensible.
For each part the key gives the truth, the acceptable variants, and the red flags to listen
for. Whatever a student picks, look for four things:

1. The **function for the expected value** has a shape that matches the pattern in the figure
   (decline, saturation, S-shape) and respects the range of the response.
2. The **noise distribution** respects the type of data (discrete count, 0/1, positive
   continuous).
3. The **leftover noise parameter** is named ($\sigma$, $k$, a gamma shape, ...), or the student
   notices that the distribution has none (Bernoulli, Poisson, exponential).
4. The two lines are written as *expected value* plus *noise around it*, not as a model of the
   raw data.

## The generating models behind the figure

The chunk below reproduces the handout's simulated data exactly (same seed and order of draws).

```{r generate-data}
set.seed(746)
yr <- 0:19
lek_mu <- 55 * exp(-0.06 * yr)
lek <- rnbinom(20, mu = lek_mu, size = 8)

cover <- runif(80, 5, 45)
p_true <- plogis(-2.5 + 0.1 * cover)
nest <- rbinom(80, 1, p_true)

forb <- runif(60, 2, 30)
mass_mu <- 350 * forb / (5 + forb)
mass <- rgamma(60, shape = 25, rate = 25 / mass_mu)
```

```{r truth-table, echo=FALSE}
truth <- data.frame(
  Data = c("1. Lek counts", "2. Nest fate", "3. Chick mass", "4. Hen movement"),
  Expected = c("55 exp(-0.06 t), t = 0, ..., 19",
               "logistic(-2.5 + 0.1 cover)",
               "350 forb / (5 + forb)",
               "constant (no data plotted)"),
  Noise = c("Negative binomial, mean = expected value, size k = 8",
            "Bernoulli(p), p = expected value",
            "Gamma, mean = expected value, shape 25 (scale = mean / 25)",
            "Described only; gamma, shape 0.8, mean 250 m is consistent")
)
knitr::kable(truth, col.names = c("Data set", "Expected value", "Noise around it"))
```

```{r truth-figure, echo=FALSE, fig.height=2.6}
par(mfrow = c(1, 3), mar = c(4, 4, 2.2, 0.8), mgp = c(2.4, 0.7, 0), cex = 0.75)
plot(yr + 1, lek, pch = 16, xlab = "Year of monitoring", ylab = "Peak male count",
     main = "Data set 1", ylim = c(0, 75))
lines(yr + 1, lek_mu, col = "red", lwd = 2)
lines(yr + 1, qnbinom(0.05, mu = lek_mu, size = 8), col = "red", lty = 2)
lines(yr + 1, qnbinom(0.95, mu = lek_mu, size = 8), col = "red", lty = 2)
plot(cover, jitter(nest, amount = 0.04), pch = 16, yaxt = "n", xlab = "Sagebrush cover (%)",
     ylab = "", main = "Data set 2", ylim = c(-0.15, 1.15))
axis(2, at = c(0, 1), labels = c("failed", "hatched"))
o <- order(cover); lines(cover[o], p_true[o], col = "red", lwd = 2)
plot(forb, mass, pch = 16, xlab = "Forb cover (%)", ylab = "Chick mass at 3 weeks (g)",
     main = "Data set 3")
o <- order(forb); lines(forb[o], mass_mu[o], col = "red", lwd = 2)
```

Red solid lines are the true expected values. The dashed red lines in panel 1 are the 5th and
95th percentiles of the true negative binomial. The $x$-axis in panel 1 is year of monitoring
$= t + 1$, while the generating function uses $t = 0, \ldots, 19$.

## Part 1: Data set 1 -- lek counts

**Truth and best answer.**

Expected value: negative exponential, $\mu_t = N_0 e^{-rt}$. Noise distribution: negative
binomial with mean $\mu_t$ and size $k$.

$$\mu_t = N_0 e^{-rt} \qquad Y_t \sim \text{NegBin}(\text{mean} = \mu_t,\ \text{size} = k)$$

The prompt says the counts fall by a constant *percentage* per year, which is exactly what an
exponential function does ($\mu_{t+1}/\mu_t = e^{-r}$ for every $t$). The noise parameter is
$k$: the variance is $\mu_t + \mu_t^2/k$, so smaller $k$ means more spread.

**1b. Why a negative binomial rather than a Poisson.** Counts are discrete and non-negative,
so Poisson and negative binomial are the natural first candidates. The prompt (and the figure)
say the scatter is large. The Poisson variance equals the mean, so we can check how much
scatter a Poisson would allow:

```{r lek-dispersion}
# Poisson would give Pearson dispersion (mean squared standardized residual) = 1
mean((lek - lek_mu)^2 / lek_mu)
# fraction of counts outside mu +/- 2 sqrt(mu) (about 5% under a Poisson)
mean(abs(lek - lek_mu) > 2 * sqrt(lek_mu))
# noise SD at the start and end of the series: Poisson vs negative binomial
rbind(Poisson = sqrt(lek_mu[c(1, 20)]), NegBin = sqrt(lek_mu[c(1, 20)] + lek_mu[c(1, 20)]^2 / 8))
```

The dispersion is about 3.6 rather than 1, and 40% of the counts fall outside a range that
should hold about 95% of Poisson counts. At the start of the series the Poisson would allow an
SD of about 7 while the negative binomial (truth) has about 21.

**Acceptable variants.**

- *Poisson* is a reasonable first guess for counts. Treat it as a good answer if the
  student flags that the scatter looks too large for it (equidispersion, from Monday's lecture).
- *Normal, lognormal, or gamma* are defensible approximations for large counts (these run from
  6 to 61), if the student says so. A normal with constant $\sigma$ would not shrink its noise
  as the mean falls.
- Any decreasing function is fine as long as it can't go negative: a straight line
  $\mu_t = a - bt$ eventually crosses zero, which is the "doubt" for a linear function.

**Red flags.** Choosing Poisson without any comment on the spread; a Ricker, Michaelis-Menten
or logistic function for a monotone decline in time (they don't have the right shape); leaving
out $k$ (a negative binomial is not complete without it).

## Part 2: Data set 2 -- nest fate

**Truth and best answer.**

Expected value: a logistic function of cover. Noise distribution: Bernoulli (a binomial with
size 1).

$$p_i = \frac{1}{1 + e^{-(a + b\,\text{cover}_i)}} \qquad Y_i \sim \text{Bernoulli}(p_i)$$

There is no leftover noise parameter. For a 0/1 response the expected value *is* the
probability of success, and the variance $p_i(1 - p_i)$ is fixed by that mean.

**2b. Why Bernoulli.** The response takes only two values, so the noise distribution must put
all its probability on 0 and 1. A normal, gamma, or Poisson can't do that.

**Why a bounded function.** The prompt hints that a probability can't leave $[0, 1]$. A straight
line is fine over the observed range but not beyond it:

```{r nest-linear}
lp <- lm(nest ~ cover)
range(fitted(lp))                               # inside [0, 1] over the observed 5-45%
predict(lp, data.frame(cover = c(0, 60)))       # outside [0, 1] when extrapolated
```

```{r nest-linear-figure, echo=FALSE, fig.height=2.6, fig.width=4.5}
par(mar = c(4, 4, 1, 1))
cv <- seq(0, 60, length.out = 200)
plot(cover, nest, pch = 16, xlim = c(0, 60), ylim = c(-0.2, 1.3), xlab = "Sagebrush cover (%)",
     ylab = "Hatched (1) or failed (0)")
lines(cv, plogis(-2.5 + 0.1 * cv), col = "red", lwd = 2)
lines(cv, predict(lp, data.frame(cover = cv)), lty = 2, lwd = 2)
abline(h = c(0, 1), col = "grey60")
legend(0, 0.85, c("logistic (truth)", "straight line"), col = c("red", "black"),
       lty = c(1, 2), lwd = 2, bty = "n", cex = 0.8)
```

**Acceptable variants.**

- *Michaelis-Menten with an asymptote of at most 1*, $p = a\,\text{cover}/(b + \text{cover})$ with
  $a \le 1$, is bounded, increasing, and defensible. A maximum-likelihood fit of both to these
  data (not something the students can do yet) finds the logistic slightly better, with
  negative log-likelihood 47.5 against 49.0, so the data don't strongly separate them.
- *Binomial*, if the student says each nest is a single trial (size 1). A binomial with
  $N = 80$ is a red flag: nests are 80 separate Bernoulli observations, not one experiment
  with 80 trials.

**Red flags.** Normal noise for a 0/1 response; a straight-line expected value with no
comment on the bounds; applying the function to the 0/1 outcome itself instead of to the
probability of success; adding a "$\sigma$" to a Bernoulli model.

## Part 3: Data set 3 -- chick mass

**Truth and best answer.**

Expected value: Michaelis-Menten. Noise distribution: gamma with mean $\mu_i$ and a shape
parameter $\alpha$ as the noise parameter.

$$\mu_i = \frac{a\,\text{forb}_i}{b + \text{forb}_i} \qquad Y_i \sim \text{Gamma}(\text{mean} = \mu_i,\ \text{shape} = \alpha)$$

Here $a$ is the asymptotic mass and $b$ is the forb cover at which the expected mass reaches
half of $a$. Mass is positive, and its spread grows with its mean (below), which is what a gamma with a
fixed shape gives.

**3b. Why gamma.** The response is continuous and positive, which excludes Poisson and
Bernoulli and makes a normal only an approximation. The spread grows with the mean:

```{r mass-spread}
bins <- cut(forb, c(0, 8, 16, 30))
res <- mass - mass_mu
tapply(mass, bins, mean)          # mean mass in each forb bin
tapply(res, bins, sd)             # SD of the residuals rises with the mean
tapply(res / mass_mu, bins, sd)   # CV roughly constant; the truth is 1 / sqrt(25) = 0.2
```

**Acceptable variants.**

- *Lognormal*: also positive with spread proportional to the mean. If the student puts the
  function on the *log* mean, that models the median rather than the mean:
  $E[Y] = e^{m + s^2/2}$. To keep the function on the expected value, use
  $m = \ln \mu_i - s^2/2$. With $\mu = 200$ and CV 0.2 that is $m = 5.279$, $s = 0.198$.
- *Normal with constant $\sigma$*: defensible here. The residual SD only rises from about 48 to
  64 g across the range, and with mean masses of 190 to 280 g and a CV of 0.2, a normal rarely
  goes negative. Ask whether the spread grows with the mean, and whether they'd trust a normal
  outside the observed range.
- *Function*: any saturating shape in the grab bag is fine (Michaelis-Menten, a power law
  with exponent below 1, a logistic-type curve). A hump (Ricker) or a straight line doesn't
  match the description ("gains shrink at high cover").

**Worth raising.** A Michaelis-Menten function passes through the origin, so it predicts zero
mass at zero forb cover. Chicks are not born weighing nothing, so an intercept would be a
reason to doubt it if the data extended to low cover.

**3c (if time).** For a gamma with shape $\alpha$ and scale $s$, the mean is $\alpha s$. Setting
this equal to $\mu_i$ gives

$$s = \frac{\mu_i}{\alpha}, \qquad \text{Var}(Y_i) = \alpha s^2 = \frac{\mu_i^2}{\alpha}, \qquad \text{CV} = \frac{1}{\sqrt{\alpha}}.$$

The shape therefore sets the *relative* noise: a large shape means small relative scatter, and
the standard deviation grows in proportion to the mean. In R (rate $= 1/$scale):

```{r gamma-check}
alpha <- 25; mu <- 200
y <- rgamma(1e6, shape = alpha, scale = mu / alpha)   # equivalently rate = alpha / mu
c(mean = mean(y), CV = sd(y) / mean(y), one_over_sqrt_alpha = 1 / sqrt(alpha))
```

**Notation note.** The handout writes the gamma shape as $\alpha$ so that it doesn't collide
with the $a$ students typically use for a function's parameter (the intercept, or the
Michaelis-Menten asymptote). Bolker writes the gamma shape as $a$, so students who consult the
book will see the other symbol. Accept any consistent naming.

## Part 4: If time -- a fourth data set

**Best answer.** The expected value is a **constant**, $\mu$, because no predictor is
available. The model still has an expected value, it just doesn't vary. The noise must be
positive and right-skewed with a few long moves: an exponential, gamma, or lognormal
distribution with mean $\mu$.

$$\mu_i = \mu \qquad Y_i \sim \text{Gamma}(\text{mean} = \mu,\ \text{shape} = \alpha)$$

All three candidates reproduce the description ("a typical hen moves 100-200 m, a few days
more than a kilometer") when the mean is about 250 m:

```{r hen-candidates}
mu <- 250
cv <- 1.5; s_ln <- sqrt(log(1 + cv^2)); m_ln <- log(mu) - s_ln^2 / 2
data.frame(
  distribution = c("Exponential (CV = 1)", "Gamma, shape 0.8 (CV = 1.12)", "Lognormal, CV = 1.5"),
  median_m = c(qexp(0.5, 1 / mu), qgamma(0.5, 0.8, scale = mu / 0.8), exp(m_ln)),
  prob_over_1km = c(pexp(1000, 1 / mu, lower.tail = FALSE),
                    pgamma(1000, 0.8, scale = mu / 0.8, lower.tail = FALSE),
                    plnorm(1000, m_ln, s_ln, lower.tail = FALSE))
)
```

The exponential has no noise parameter (its CV is fixed at 1). The gamma adds a shape and can
make the tail heavier (CV above 1) or lighter. The lognormal has a log-scale SD as its noise
parameter.

**Red flags.** A normal distribution (symmetric around 250 m, puts probability below zero,
and cannot produce a few very long moves); a Poisson (distance is continuous); a uniform.
"There is no predictor so there is no model" misses the point: an intercept-only model is
still a model, with a constant expected value, and it is what you would fit before adding
covariates such as habitat or time of year.

## Discussion prompts for the neighbor comparison

Where students disagree, these are the productive questions.

- **Lek counts:** what would a Poisson predict for the scatter, and does the figure agree? (The
  answer is in 1b.)
- **Nest fate:** a straight line fits the observed range. What goes wrong if you use it to
  predict for a nest with 60% cover?
- **Chick mass:** gamma, lognormal, and normal are all defensible. What would you check in a
  residual plot to tell them apart?
- **Hen movement:** exponential, gamma, and lognormal all match the description. What extra
  data would separate them?
