---
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}
```

**NRES 746 -- Workout of the Day #2: Pseudocoding a Permutation Test**

&nbsp;

Name: ___________________________________________________

&nbsp;

**Instructions:** Work alone for about 5 minutes on Parts 1-2. Then get
together with your neighbor(s) and spend about 5 minutes on Part 3,
translating your pseudocode into actual R syntax **by hand, on paper**
(no laptops).

## Background: what is a permutation test?

In previous stats classes, you've seen tests for whether a sample's mean differed
from a known null distribution: e.g., a z-test, and a t-test.
These tests rely on knowing (or assuming) the shape
of the population distribution the data came from.

A **permutation test** drops that requirement. It's a *nonparametric*
(distribution-free) test: instead of simulating data from an assumed
distribution, we generate a null distribution for our test statistic by
**randomly reshuffling the labels** (e.g., treatment vs. control) on the
data we already have, over and over. If the group labels really don't
matter, shuffling them shouldn't change/reduce the expected
difference between group means very much. If the observed difference is
much larger than what typical reshufflings produce, that's evidence the
labels *do* actually matter.

## Part 1: The scenario

Here's the scenario: pygmy short-horned lizards were sampled from a
cheatgrass-removal treatment area (trtA) and a nearby untreated control
area (Control). We want to know whether body mass differs between the
two groups.

```
   trtA:      175  168  168  190  156  181  182  175  174  179
   Control:   185  169  173  173  188  186  175  174  179  180
```

**1a.** What is the observed test statistic here (the single number
we'll compare against the null distribution)?

```{r}
blank_space()
```

**1b.** Is the alternative hypothesis one-tailed or two-tailed in this
case? Briefly explain why.

```{r fig.height=0.3*2}
blank_lines(2)
```

## Part 2: Write the pseudocode

**2.** Working from the description above (don't just copy it back), write
out -- in plain English, step by step -- the algorithm you would need to
run a permutation test on this data. Think about: what gets shuffled, what
gets recomputed each time, how many times you repeat it, and how you turn
the result into a p-value.

```{r fig.height=0.3*8}
blank_lines(8)
```

## Part 3: Translate to R (group, pen and paper)

**3.** As a group, fill in the blanks below to turn your pseudocode into
real R syntax. You don't need working code -- just your best attempt at
the correct functions and syntax, written by hand.

```
permutation_test <- function(data, group, value, reps){

  observed_dif <- ___________________________________________

  null_difs <- ___________________________________________

  for(i in 1:reps){

    shuffled_labels <- ___________________________________________

    null_difs[i] <- ___________________________________________

  }

  p_value <- ___________________________________________

  return(p_value)
}
```

**3a.** Which R function did your group use to do the actual shuffling?

```{r}
blank_lines()
```

**3b.** What was the trickiest part of going from pseudocode to R syntax?
Flag it -- we'll come back to it!

```{r fig.height=0.3*2}
blank_lines(2)
```
