---
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}
```

**NRES 746 -- Workout of the Day #3: Spot the Bug**

&nbsp;

Name: ___________________________________________

&nbsp;

**Instructions:** Work alone for about 5 minutes on Part 1. Then get
together with your neighbor(s) and spend a few minutes on Part 2.

## Part 1: Find the bug

Each snippet below has one bug. None of them are typos in the strict
sense -- the R is syntactically valid enough to run (or at least to try
to run). Circle the bug and write your fix directly on the snippet.

**1a.** Argument order

```

ricker <- function(a, b, x) {
  a * x * exp(-b * x)
}

# Evaluate the curve at x = 10, using a = 2, b = 0.5
ricker(10, 2, 0.5)

```

**1b.** Plotting a curve

```

a <- 2; b <- 0.3
curve(a * x * e^(-b * x), from = 0, to = 15)

```

**1c.** Evaluating across several parameter values

```

avals <- c(0.5, 1, 1.5)
peak_vals <- sapply(avals, ricker(x = 5, b = 0.2))

```

## Part 2: Write it yourself (individual, then compare with a neighbor)

**2.** Assume `ricker <- function(x, a, b) a * x * exp(-b * x)` is
already defined. In one line of R, use `sapply()` to compute the
Ricker function's value at $x=5$ for three different values of $a$
(0.5, 1, and 2), holding $b=0.2$ fixed.

```{r fig.height=0.3*3}
blank_space()
```
