---
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}
```

**NRES 746 -- Workout of the Day #4: Limits and a Derivative**

&nbsp;

Name: ___________________________________________

&nbsp;

**Instructions:** Work alone for about 5-8 minutes. Then compare with a
neighbor.

## Part 1: What happens at the edges?

For each function below, use the "throw out the lower-order terms"
trick from Monday's lecture to state what happens as $x\to0$ and as
$x\to\infty$. You don't need to show work -- just the two limiting
behaviors for each.

**1a.** Michaelis-Menten: $f(x) = \dfrac{ax}{b+x}$

As $x\to0$: ________________________ &nbsp;&nbsp;&nbsp;&nbsp; As $x\to\infty$: ________________________

**1b.** Ricker: $f(x) = ax\,e^{-bx}$

As $x\to0$: ________________________ &nbsp;&nbsp;&nbsp;&nbsp; As $x\to\infty$: ________________________

**1c.** Logistic: $f(x) = \dfrac{1}{1+e^{-bx}}$

As $x\to0$: ________________________ &nbsp;&nbsp;&nbsp;&nbsp; As $x\to\infty$: ________________________

**1d.** Holling type III: $f(x) = \dfrac{ax^2}{b^2+x^2}$

As $x\to0$: ________________________ &nbsp;&nbsp;&nbsp;&nbsp; As $x\to\infty$: ________________________

## Part 2: The logistic's steepest point

The logistic function, $f(x) = \dfrac{1}{1+e^{-bx}}$, is the simplest
sigmoid in the bestiary -- it's symmetric around $x=0$, which is also
where it's rising fastest.

**2a.** Rewrite $f(x)$ as $(1+e^{-bx})^{-1}$, then find $f'(x)$ using
the chain rule (twice: once for the outer power, once for the inner
exponential).

*Reminders:* $\dfrac{d}{dx}\,u^n = n\,u^{n-1}\,u'$ &nbsp;&nbsp;&nbsp;&nbsp; $\dfrac{d}{dx}\,e^{g(x)} = g'(x)\,e^{g(x)}$

Show your work:

```{r fig.height=0.3*3}
blank_space()
```

**2b.** Evaluate $f'(x)$ at $x=0$.

$f'(0) =$ ________________________

**2c.** What does this value tell you about the logistic curve at its
midpoint? (Hint: think about what the number itself represents, not
just how you calculated it.)

```{r fig.height=0.3*1}
blank_lines(1)
```
