---
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}
```

**NRES 746, Week 6, Lecture 2: Find the Bugs**

&nbsp;

Name: ___________________________________________

&nbsp;

**Instructions:** Work alone for about 6 minutes, then compare with a
neighbor. Keep laptops closed.

## The code

A classmate wrote the four functions from last week's Lab 4, Act 2
(the toad monitoring power analysis), possibly with help from an AI
assistant. As a reminder, the intended model is:

- each animal is detected on one person-day with probability `pobs`,
  so a bout of `k` person-days detects it with probability
  $1 - (1 - p_{obs})^k$;
- abundance declines as $N_t = N_0 \lambda^t$;
- each survey's count is Binomial($N_t$, detection probability);
- a decline is detected if the slope of log(count) on year is negative
  **and** its $p$-value is at most `alpha`.

Each function below contains **exactly one bug**. The code runs
(or at least starts to run), but gives wrong answers.

```{r buggy-code, echo=TRUE, eval=FALSE}
detect_prob <- function(k, pobs = 0.02) {
  1 - pobs^k
}

simulate_counts <- function(t_survey, k, N0 = 1000, lambda = 0.75^(1/25)) {
  N_t <- N0 * lambda^t_survey
  rbinom(1, size = round(N_t), prob = detect_prob(k))
}

detect_decline <- function(t_survey, counts, alpha = 0.05) {
  s <- summary(lm(log(counts) ~ t_survey))$coefficients[2, ]
  s[1] < 0 || s[4] <= alpha      # s[1] = slope, s[4] = p-value
}

estimate_power <- function(k, s, R = 1000) {
  t_survey <- seq(1, 25, by = s)
  detected <- logical(R)
  for (i in seq_len(R)) {
    counts <- simulate_counts(t_survey, k)
    detected <- detect_decline(t_survey, counts)
  }
  mean(detected)
}
```

## Part 1: Find and fix (individual, ~6 min)

For each function, circle the bug on the code above, write the
corrected line, and say what would go wrong if you ran the buggy
version.

**1a.** `detect_prob()`

```{r fig.height=0.3*2}
blank_lines(2)
```

**1b.** `simulate_counts()`

```{r fig.height=0.3*2}
blank_lines(2)
```

**1c.** `detect_decline()`

```{r fig.height=0.3*2}
blank_lines(2)
```

**1d.** `estimate_power()`

```{r fig.height=0.3*2}
blank_lines(2)
```

## Part 2: Catch it next time (with a neighbor, ~3 min)

**2.** Pick one of the bugs. Describe a quick test, one line of code
and the answer you'd expect, that would have caught it before you
trusted the function.

```{r fig.height=0.3*3}
blank_lines(3)
```
