---
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}
```

**NRES 746, Week 6, Lecture 2: Find the Bugs**

Name: ___________________________________________

**Instructions:** Work alone for about 3 minutes, then compare with a
neighbor. Keep laptops closed.

## The code

Here are two short functions, possibly written with AI help. The first
comes from today's live-coding problem (finishing Lab 4's toad
monitoring design).

- `detect_prob(k)`: one person-day of searching detects an animal with
  probability `pobs`, so `k` person-days should detect it with
  probability $1 - (1 - p_{obs})^k$.
- `sim_detect(R)`: should simulate `R` surveys of a site where the frog
  from Lab 5 is present (3 visits per survey, detection probability
  0.4 per visit) and return the fraction of surveys with at least one
  detection.

Each function below contains **exactly one bug**. Both run without an
error, but give wrong answers.

```{r buggy-code, echo=TRUE, eval=FALSE}
detect_prob <- function(k, pobs = 0.02) {
  1 - pobs^k
}

sim_detect <- function(R = 1000, visits = 3, p = 0.4) {
  detected <- logical(R)
  for (i in seq_len(R)) {
    n_det <- rbinom(1, size = visits, prob = p)
    detected <- n_det > 0
  }
  mean(detected)
}
```

## Part 1: Find and fix (individual, ~3 min)

For each function, circle the bug on the code above, write the
corrected line, and say what would go wrong if you ran the buggy
version.

**1a.** `detect_prob()`

```{r fig.height=0.3*2}
blank_lines(2)
```

**1b.** `sim_detect()`

```{r fig.height=0.3*2}
blank_lines(2)
```

## Part 2: Catch it next time (with a neighbor, ~2 min)

**2.** For one of the bugs, what quick check (one line of code and the
answer you'd expect) would have caught it?

```{r fig.height=0.3*2}
blank_lines(2)
```
