---
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
header-includes:
  - \usepackage{needspace}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# keeps a prompt and its answer space on one PDF page: place `r nq(...)` on its
# own line just above the prompt. Emits nothing in non-LaTeX output. lines =
# lines of prompt text, rows = rows of blank space below it (0.3 in each),
# extra = any other height (in inches), e.g. a figure
nq <- function(lines, rows = 0, extra = 0) {
  if (!knitr::is_latex_output()) return("")
  knitr::asis_output(sprintf("\n\\needspace{%.2fin}\n", 0.23 * lines + 0.15 + 0.3 * rows + extra))
}
```

**NRES 746, Week 8, Lecture 1: How Much Did the Prior Matter?**

Name: ___________________________________________

**Instructions:** Work alone on Part 1 (about 4 minutes), then compare
with a neighbor for Part 2. Keep laptops closed.

## The studies

Three research teams estimate the annual survival probability of adult
desert tortoises at three different sites, using radio-collared
animals. Each team used a prior based on earlier work and a different
number of collared tortoises. Each panel shows one team's prior (grey)
and posterior (black). The likelihoods are not shown.

```{r prior-post-panels, fig.width=6.5, fig.height=5.4}
s_seq <- seq(0.001, 0.999, length.out = 500)
teams <- list(
  A = list(prior = c(2, 2), data = c(45, 15)),
  B = list(prior = c(80, 20), data = c(3, 2)),
  C = list(prior = c(16, 4), data = c(10, 10))
)
par(mfrow = c(3, 1), mar = c(3.2, 4, 1.4, 0.5), mgp = c(2, 0.6, 0))
for (tm in names(teams)) {
  pr <- teams[[tm]]$prior
  po <- pr + teams[[tm]]$data
  plot(NA, xlim = c(0, 1), ylim = c(0, 11), xaxs = "i", yaxs = "i",
       xlab = "Annual survival probability", ylab = "Density",
       main = paste("Team", tm), cex.main = 1, font.main = 1)
  abline(v = seq(0, 1, 0.1), col = "grey90")
  lines(s_seq, dbeta(s_seq, pr[1], pr[2]), lwd = 2.5, col = "grey60")
  lines(s_seq, dbeta(s_seq, po[1], po[2]), lwd = 2.5)
  if (tm == "A") {
    legend("topleft", c("prior", "posterior"), lwd = 2.5,
           col = c("grey60", "black"), bty = "n", cex = 0.9)
  }
  box()
}
```

`r nq(4, 0)`

## Part 1: Read the panels (individual)

**1a.** Rank the three teams from most to least influenced by their
prior. Give a one-line reason for each.

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(4, 0)`

**1b.** Sketch the shape of Team C's likelihood on its panel (the
height doesn't matter). Where does it peak, and is it wider or narrower
than the prior?

```{r fig.height=0.3*2}
blank_lines(2)
```

`r nq(4, 3)`

## Part 2: How much data? (with a neighbor)

**2a.** Which team collared the most tortoises, and which collared the
fewest? Team B has the narrowest posterior of the three. Does that mean
Team B had the most data?

```{r fig.height=0.3*3}
blank_lines(3)
```
