---
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
header-includes:
  - \usepackage{needspace}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}

# keeps a prompt and its answer space on one PDF page: place `r nq(...)` on its
# own line just above the prompt. Emits nothing in non-LaTeX output. lines =
# lines of prompt text, rows = rows of blank space below it (0.3 in each)
nq <- function(lines, rows = 0, extra = 0) {
  if (!knitr::is_latex_output()) return("")
  knitr::asis_output(sprintf("\n\\needspace{%.2fin}\n", 0.23 * lines + 0.15 + 0.3 * rows + extra))
}
```

**NRES 746 -- Week 5, Lecture 1: Choosing a Data-Generating Model**

&nbsp;

Name: ___________________________________________

&nbsp;

**Instructions:** Work alone for about 8-10 minutes, then compare with a
neighbor. Several of these have more than one defensible answer, so
argue for your choice where you and your neighbor disagree. You do not need to estimate any parameter values. 

## The study system

Suppose a state wildlife agency runs a long-term monitoring program for greater sage-grouse in a Great Basin study area. For each data set below, propose a **data-generating model**: a deterministic function for the *expected value* of the response (which can be as simple as a constant, if nothing predicts it) and a probability distribution for the noise around that expected value. Write each
model as a deterministic line and a stochastic line, for example
$\mu_i = a + b x_i$ and $Y_i \sim \text{Normal}(\mu_i, \sigma)$, where $\mu_i$ is the expected value of $Y_i$. Write the distribution with
its mean set to $\mu_i$ where you can, and name any other parameter it needs
to describe the spread (like $\sigma$ for the normal). If you are not sure how
to set the mean for your distribution, write "mean $= \mu_i$" anyway and note
which parameter controls the spread.

NOTE: the expected value is not necessarily a *parameter* of the stochastic distribution (although it is for the normal and Poisson distributions). Don't worry about that right now- just think of the deterministic function as modeling the expected value and the stochastic distribution as a way to describe the unpredictable variability (spread) in the response.  

Here's your grab bag (Bolker Ch. 3 and 4). Use any of these, in any combination:

- **Deterministic functions (for the expected value):** linear, quadratic, power law, negative
  exponential, Ricker, Michaelis-Menten, logistic
- **Distributions (for the noise):** Bernoulli/binomial, Poisson, negative binomial,
  uniform, normal, lognormal, gamma, exponential, beta

```{r data-figure, fig.width=6.5, fig.height=1.75}
set.seed(746)

yr <- 0:19
lek <- rnbinom(20, mu = 55 * exp(-0.06 * yr), size = 8)

cover <- runif(80, 5, 45)
nest <- rbinom(80, 1, plogis(-2.5 + 0.1 * cover))

forb <- runif(60, 2, 30)
mass <- rgamma(60, shape = 25, rate = 25 / (350 * forb / (5 + forb)))

par(mfrow = c(1, 3), mar = c(4, 4, 2.2, 0.8), mgp = c(2.4, 0.7, 0), cex = 0.75)
plot(yr + 1, lek, pch = 16, xlab = "Year of monitoring",
     ylab = "Peak male count", main = "Data set 1", ylim = c(0, 65))
plot(cover, jitter(nest, amount = 0.04), pch = 16, yaxt = "n",
     xlab = "Sagebrush cover (%)", ylab = "", main = "Data set 2",
     ylim = c(-0.15, 1.15))
axis(2, at = c(0, 1), labels = c("failed", "hatched"))
plot(forb, mass, pch = 16, xlab = "Forb cover (%)",
     ylab = "Chick mass at 3 weeks (g)", main = "Data set 3")
```

## Part 1: Data set 1 -- lek counts

Each spring, biologists count the males displaying at a lek (a
traditional breeding display ground). The plot shows the peak count at
one lek over 20 years. Counts trend downward at roughly a constant
percentage per year, and the scatter around that trend is large.

Expected value: ______________ &nbsp; Noise distribution: ______________

**1a.** Write your model.

```{r fig.height=0.3*2.6}
blank_space()
```

**1b.** Why this distribution? 

```{r fig.height=0.3*2}
blank_lines(2)
```

## Part 2: Data set 2 -- nest fate

Crews monitor 80 nests. Each nest either hatches at least one egg
(success) or fails. Crews also measure the percent sagebrush cover
around each nest. Nests in denser cover look more likely to succeed. (The expected
value of a 0/1 response is the probability of success, and a probability
can never go below 0 or above 1, however much cover a nest has.)

Expected value: ______________ &nbsp; Noise distribution: ______________

**2a.** Write your model.

```{r fig.height=0.3*2.6}
blank_space()
```

**2b.** Why this distribution? 

```{r fig.height=0.3*2}
blank_lines(2)
```

## Part 3: Data set 3 -- chick mass

Crews weigh chicks at 3 weeks of age and record the percent forb cover
at the brood-rearing site (forbs host the insects chicks eat). Mass
rises with forb cover, but the gains shrink at high cover. Mass is
always positive.

Expected value: ______________ &nbsp; Noise distribution: ______________

**3a.** Write your model.

```{r fig.height=0.3*2.6}
blank_space()
```

**3b.** Why this distribution? 

```{r fig.height=0.3*2}
blank_lines(2)
```

`r nq(4, 3)`

**3c. If time.** If your distribution's own parameters are not the mean, work
out how to express them using $\mu_i$. (For a gamma with shape $\alpha$ and
scale $s$, the mean is $\alpha s$. What scale gives a mean of $\mu_i$, and what
does the shape do to the noise?)

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(7, 3, extra = 0.5)`

## Part 4: If time -- a fourth data set

Telemetry crews record how far each radio-collared hen moves between
consecutive daily locations (meters). Most daily moves are short (a
typical hen moves 100-200 m), but a few days involve moves of more than
a kilometer. There is no predictor in this data set.

**4.** What would you use for the expected value, and which distribution
for the noise? Does the expected value need to depend on a predictor at
all? Explain.

```{r fig.height=0.3*3}
blank_lines(3)
```
