---
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
header-includes:
  - \usepackage{needspace}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}

# draws a table with given headers/body for handwritten answers ("" = blank
# cell); headers/values may be plotmath expressions for math; pair with
# fig.height = header_height + nrow(body) * row_height
blank_table <- function(headers, body, col_widths = NULL,
                         header_height = 0.35, row_height = 0.5,
                         cex = 0.95, bold_rows = integer(0)) {
  nc <- length(headers); nr <- length(body)
  if (is.null(col_widths)) col_widths <- rep(1, nc)
  col_widths <- col_widths / sum(col_widths)
  cx <- c(0, cumsum(col_widths))
  h_total <- header_height + nr * row_height
  par(mar = c(0, 0, 0, 0), family = "serif")
  plot(NA, xlim = c(0, 1), ylim = c(0, h_total), axes = FALSE, xlab = "", ylab = "",
       xaxs = "i", yaxs = "i")
  y <- h_total
  segments(0, y, 1, y, lwd = 1.5)
  y <- y - header_height
  segments(0, y, 1, y, lwd = 1.5)
  for (i in seq_len(nr)) {
    y <- y - row_height
    segments(0, y, 1, y, lwd = if (i == nr) 1.5 else 0.5,
             col = if (i == nr) "black" else "grey70")
  }
  for (v in cx) segments(v, 0, v, h_total, lwd = 1)
  yh <- h_total - header_height / 2
  for (j in seq_len(nc)) text((cx[j] + cx[j + 1]) / 2, yh, labels = headers[[j]], font = 2, cex = cex)
  y <- h_total - header_height
  for (i in seq_len(nr)) {
    yc <- y - row_height / 2
    fnt <- if (i %in% bold_rows) 2 else 1
    for (j in seq_len(nc)) {
      val <- body[[i]][[j]]
      if (!identical(val, "")) text((cx[j] + cx[j + 1]) / 2, yc, labels = val, font = fnt, cex = cex)
    }
    y <- y - row_height
  }
}

# keeps a prompt and its answer space on one PDF page: place `r nq(...)` on its
# own line just above the prompt. Emits nothing in non-LaTeX output. lines =
# lines of prompt text, rows = rows of blank space below it (0.3 in each),
# extra = any other height (in inches), e.g. a table or code block
nq <- function(lines, rows = 0, extra = 0) {
  if (!knitr::is_latex_output()) return("")
  knitr::asis_output(sprintf("\n\\needspace{%.2fin}\n", 0.23 * lines + 0.15 + 0.3 * rows + extra))
}

# chunk option tight_mar = TRUE trims the default plot margins
knitr::knit_hooks$set(tight_mar = function(before, options, envir) {
  if (before && isTRUE(options$tight_mar)) par(mar = c(4, 4.5, 1, 1))
})

nll_const <- function(p, y) -sum(dbinom(y, size = 1, prob = p, log = TRUE))
y <- c(1, 1, 0, 1)
```

**NRES 746 -- Week 5, Lecture 2: Writing the Likelihood**

&nbsp;

Name: ___________________________________________

&nbsp;

**Hand this in.** Like every WOD, this one is collected, so make sure your
name is at the top and turn in your own copy (not a group copy) next Monday.

**Instructions:** This workout takes the place of today's lecture. Plan
on about 30 minutes: work alone on Parts 1-3 (~15 min), check in with
your group (~3 min), then work through Parts 4-5 with your group (~10
min). Please keep laptops closed and do not use AI tools for this one.
Everything here is designed for pencil and paper. Calculators are fine
for arithmetic, and fractions are fine as final answers. If you get
stuck, note where and move on. Nobody will call time, so someone in
your group should keep an eye on the clock.

## Where we left off

On Monday you went *forward*: you chose a data-generating model and
imagined the data it would produce. Today we go *backward*: the data are
already in hand, and the question is what they tell us about the
parameters. The probability formulas are the ones you already know from
Monday and Lab 3. What changes is which quantity you treat as known and
which you treat as unknown.

We will stay with the sage-grouse nest-fate data from Monday. Here are
four nests, with outcome $y_i=1$ if at least one egg hatched and
$y_i=0$ if the nest failed:

```{r nest-data}
nests <- data.frame(
  Nest = 1:4,
  Cover = c(30, 20, 10, 30),
  Outcome = c(1, 1, 0, 1)
)
knitr::kable(nests, col.names = c("Nest", "Sagebrush cover (%)",
                                  "Outcome (1 = hatched, 0 = failed)"),
             align = "ccc")
```

`r nq(5, 0, extra = 0.5)`

## Part 1: One observation (individual, ~4 min)

For now, ignore cover and suppose every nest has the same unknown
probability $p$ of success.

**1a.** Write down the probability of each possible outcome for a
single nest.

$P(y=1 \mid p) =$ ________________ &nbsp;&nbsp;&nbsp;&nbsp; $P(y=0 \mid p) =$ ________________

`r nq(4, 3)`

**1b.** Write one formula that gives the probability of an observed
outcome $y$ (either 0 or 1) as a function of $y$ and $p$, with no "if"
statements or separate cases. (The binomial pmf from Lab 3, with a single
trial, is one place to start.)

```{r fig.height=0.3*3}
blank_space()
```

`r nq(7, 0)`

> **Definition.** The **likelihood** of a parameter value is the
> probability of the data you actually observed, if that parameter
> value were the truth. It is written as a function of the parameter,
> with the data held fixed: for a single observation,
> $\mathcal{L}(p \mid y) = P(y \mid p)$.

**1c.** Nest 1 hatched ($y_1=1$). Evaluate the likelihood of $p=0.75$
and of $p=0.25$ given this observation.

$\mathcal{L}(0.75 \mid y_1=1) =$ ________________ &nbsp;&nbsp;&nbsp;&nbsp; $\mathcal{L}(0.25 \mid y_1=1) =$ ________________

`r nq(4, 2)`

**1d.** In your formula from 1b, which symbol is the data and which is
the parameter? When you did probability calculations in Lab 3, which
one did you treat as known? In a likelihood, which one do you treat as
known?

```{r fig.height=0.3*2}
blank_lines(2)
```

`r nq(5, 7, extra = 0.5)`

## Part 2: Several observations (individual, ~8 min)

Now use all four nests. Assume every nest has the same success
probability $p$, and that the nests are independent of one another.

**2a.** Write the likelihood of all four observations,
$\mathcal{L}(p \mid y_1, y_2, y_3, y_4)$, in terms of your formula
from 1b. Use product notation ($\prod$).

```{r fig.height=0.3*7}
blank_space()
```

`r nq(2, 0, extra = 0.3)`

**2b.** Plug in the observed outcomes and simplify to the form
$p^{j}(1-p)^{k}$. What are $j$ and $k$?

$j =$ ________ &nbsp;&nbsp;&nbsp;&nbsp; $k =$ ________

`r nq(2, 0, extra = 1.85)`

**2c.** Evaluate your answer to 2b at the three values of $p$ below.
Fractions are fine.

```{r fill-lik-table, fig.height=0.35+3*0.5, fig.width=4}
blank_table(
  headers = list("p", "Likelihood"),
  body = list(
    list("0.25", ""),
    list("0.50", ""),
    list("0.75", "")
  ),
  col_widths = c(1, 1.6),
  cex = 0.95
)
```

`r nq(2, 3)`

**2d.** Add up your three likelihood values. Do they sum to 1? Should
they? What does your answer say about how to read a likelihood value?

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(3, 0, extra = 2.4)`

**2e.** Plot your three values on the axes below, then sketch the whole
curve $\mathcal{L}(p)$ for $0 \le p \le 1$. Mark where you think the
peak is. Is that where you would have guessed from the raw data alone?

```{r sketch-axes, fig.width=5, fig.height=2.4}
par(mar = c(4, 4.5, 1, 1))
plot(NA, xlim = c(0, 1), ylim = c(0, 0.12), xaxs = "i", yaxs = "i",
     xlab = "p", ylab = "Likelihood")
abline(v = seq(0, 1, 0.25), h = seq(0, 0.12, 0.03), col = "grey85")
box()
```

`r nq(5, 4, extra = 0.5)`

## Part 3: The log-likelihood (individual, ~3 min)

**3a.** Take the log of your product from 2a and use the log rules to
rewrite it as a sum. (Recall that $\log(uv) = \log u + \log v$ and
$\log(u^c) = c \log u$.) Write your result for a general set of $n$ nests
with outcomes $y_1, \ldots, y_n$.

$\ell(p) = \log \mathcal{L}(p \mid y_1,\ldots,y_n) =$

```{r fig.height=0.3*4}
blank_space()
```

`r nq(7, 4)`

**3b.** Suppose the monitoring program had 2,000 nests and each nest
contributed a likelihood of about 0.5. Roughly how big is the product
of all 2,000 terms? (Use $\log_{10}2 \approx 0.30$ and give your answer
as $10^{\text{something}}$.) Standard computer arithmetic cannot store
positive numbers much smaller than $10^{-308}$. Why might you want to work
with the sum from 3a instead? What is its value here (use
$\ln 2 \approx 0.69$)?

```{r fig.height=0.3*4}
blank_space()
```

`r nq(7, 3, extra = 0.5)`

## Group check-in (~3 min)

Compare your answers to Parts 1-3 with your group. Where you differ,
argue it out and revise. The sum-to-1 question (2d) and the location of
the peak (2e) are good places to start. Then answer together:

**Group question.** Multiplying the four nest likelihoods together
depended on an assumption. Name one biological reason it might fail for
real sage-grouse nests.

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(6, 0, extra = 1.1)`

## Part 4: Adding the covariate (with your group, ~6 min)

Monday's model had success probability rising with sagebrush cover. A
logistic function keeps it between 0 and 1:

$$p_i = \frac{1}{1 + e^{-(a + b x_i)}}$$

where $x_i$ is the cover around nest $i$. Every nest now has its own
$p_i$. Suppose a candidate pair of parameter values is $a=-2.2$ and
$b=0.11$. Here are the success probabilities it predicts (rounded) for
our four nests:

```{r pred-table}
nests$p_pred <- c(0.75, 0.50, 0.25, 0.75)
knitr::kable(nests, col.names = c("Nest", "Sagebrush cover (%)",
                                  "Outcome (1 = hatched, 0 = failed)",
                                  "Predicted p"),
             align = "cccc")
```

`r nq(1, 3)`

**4a.** Check the predicted $p$ for nest 2 (20% cover) by hand.

```{r fig.height=0.3*3}
blank_space()
```

`r nq(3, 3)`

**4b.** Assuming the nests are independent, what is the likelihood of
this candidate parameter pair, $\mathcal{L}(a=-2.2, b=0.11 \mid \text{data})$?
Careful with the nest that failed.

```{r fig.height=0.3*3}
blank_space()
```

`r nq(4, 3)`

**4c.** Compare this to the best of your three constant-$p$ likelihoods
from 2c. Which model do the data support more, and by what factor? Does
that show that cover really matters for nest success? Give at least one
reason for caution.

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(4, 5)`

**4d.** Write the likelihood for $n$ nests in general,
$\mathcal{L}(a, b \mid y_1,\ldots,y_n, x_1,\ldots,x_n)$, using
$p_i = 1/(1+e^{-(a+bx_i)})$. Then underline the quantities that are
fixed data and circle the quantities that the likelihood is a function of.

```{r fig.height=0.3*5}
blank_space()
```

`r nq(11, 0, extra = 1.3)`

## Part 5: From math to R code, on paper (with your group, ~5 min)

Later in the course you will fit models like this by writing the
likelihood as an R function and handing it to an optimizer. You will
get that practice in Lab 4 and beyond. For now, we only write and read
code, with no computer. The built-in `dbinom(x, size = 1, prob = p)`
gives the Bernoulli probability from Part 1: `x` is the observed outcome
(0 or 1), `size` is the number of trials per observation, and `prob` is
the success probability. Setting `log = TRUE` returns the log of the
probability.

**5a.** Fill in the blanks so this function returns the *negative*
log-likelihood of a vector of nest outcomes `y` for a single success
probability `p`.

```{r nll-const-skeleton, echo=TRUE, eval=FALSE}
nll_const <- function(p, y) {
  -sum(dbinom(x = ________, size = ________, prob = ________, log = TRUE))
}
```

`r nq(3, 3)`

**5b.** With `y <- c(1, 1, 0, 1)`, what does `exp(-nll_const(0.5, y))`
return? Work it out by hand. Which entry of your table in 2c should it
match?

```{r fig.height=0.3*3}
blank_space()
```

`r nq(4, 0, extra = 3.7)`

**5c.** A classmate ran the code below with a working `nll_const` and
got this plot. (`sapply()` evaluates a function at each value of a vector
in turn.)

```{r nll-grid-plot, echo=TRUE, tight_mar=TRUE, fig.width=4.5, fig.height=2.4}
p_grid <- seq(0.05, 0.95, by = 0.01)
nll_vals <- sapply(p_grid, function(p) nll_const(p, y))
plot(p_grid, nll_vals, type = "l", xlab = "p",
     ylab = "neg. log-likelihood", xaxt = "n")
axis(1, at = seq(0, 1, 0.25))
abline(v = seq(0, 1, 0.25), col = "grey85")
```

`r nq(4, 4)`

Which value of $p$ gives the smallest negative log-likelihood? Does it
agree with your sketch in 2e? Why do you think we work with the
*negative* log-likelihood rather than the log-likelihood itself?

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(3, 4)`

**What is still fuzzy?** (individual) What is the muddiest point about
likelihood so far? What would you ask if this had been a lecture?

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(2, 0, extra = 1.5)`

**5d. Optional challenge, if you finish early.** Write the covariate version, using your
formula from 4d.

```{r nll-cover-skeleton, echo=TRUE, eval=FALSE}
cover <- c(30, 20, 10, 30)

nll_cover <- function(par, y, cover) {
  p <- ______________________________
  -sum(________________________________)
}
```

`r nq(3, 3)`

If you ran `exp(-nll_cover(c(-2.2, 0.11), y, cover))`, what number
should you get, approximately, and why?

```{r fig.height=0.3*3}
blank_lines(3)
```
