---
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}
```

**NRES 746 -- Week 4, Lecture 2: Expectation and the Sample Mean**

&nbsp;

Name: ___________________________________________

&nbsp;

**Instructions:** Work alone for about 8-10 minutes. Then compare with a
neighbor.

## Part 1: A parameter vs. a statistic

Suppose the length of brook trout in a stream (X) follows some distribution ($P(X=x)=f(x)$)
with true mean $\mu = \mathbb{E}[X]$ and variance $\sigma^2 = \mathbb{E}[(X-\mu)^2]$ -- these describe the
*entire population* of trout in the stream, whether or not anyone ever
catches one.

You catch $n=5$ trout, measure them, and compute their average length
$\bar{x}$.

**1.** Is $\mu$ a random variable? Is $\bar{x}$ a random variable? Briefly explain your reasoning.

```{r fig.height=0.3*3}
blank_lines(3)
```

## Part 2: working with expectations

Here are some useful facts, true for any random variable $X$ and constants $a,b$:

$$\mathbb{E}[aX+b] = a\mathbb{E}[X]+b \qquad \text{Var}(aX) = a^2\text{Var}(X)$$
Also, for any two random variables X and Y:

$$\mathbb{E}[X + Y]= \mathbb{E}[X] + \mathbb{E}[Y]$$

One more thing: the variance of any constant is zero.

Suppose trout length $X$ has $\mathbb{E}[X]=\mu=22$ cm and $\text{Var}(X)=\sigma^2=16$
cm$^2$.

**2a.** What is $\mathbb{E}[2X+3]$? What is $\text{Var}(2X+3)$?

$\mathbb{E}[2X] =$ ________ &nbsp;&nbsp;&nbsp;&nbsp; $\text{Var}(2X) =$ ________

**2b.** A second, independent measurement $Y$ of a different trout has
the same mean and variance as $X$. It is always true that
$\mathbb{E}[X+Y]=\mathbb{E}[X] + \mathbb{E}[Y]$. It is also true in this case that
$\text{Var}(X+Y)=\text{Var}(X)+\text{Var}(Y)$. However, this second fact is not always true.
What do you think might be the special condition that makes it true? 

```{r fig.height=0.3*3}
blank_lines(3)
```

## Part 3: Averaging two measurements

Let $X_1$ and $X_2$ be two independent measurements of the same
quantity, each with mean $\mu$ and variance $\sigma^2$. Define their
average $M = \dfrac{X_1+X_2}{2}$.

**3a.** Rewrite $M$ as $aX_1 + bX_2$ (i.e. what are $a$ and $b$?), then
use the facts from Part 2 to derive $\mathbb{E}[M]$ and $\text{Var}(M)$ in terms
of $\mu$ and $\sigma^2$. Show every step.

```{r fig.height=0.3*9}
blank_space()
```

**3b.** Using $\mu=22$ and $\sigma^2=16$ (from Part 2), what is
$\text{Var}(M)$ as a number? Compare it to $\text{Var}(X_1)=16$ alone
-- how does the variance of an average of two measurements compare to the variance of each of its constituent parts?

$\text{Var}(M) =$ ________

## Part 4: Generalizing to n measurements

NOTE: We will do part 4 together on the whiteboard in a minute, but see if you can think this one through if you have time:

Let $X_1, X_2, \ldots, X_n$ be independent measurements,
each with mean $\mu$ and variance $\sigma^2$, and define the sample
mean $\bar{X} = \dfrac{1}{n}\sum_{i=1}^n X_i$.

**4a.** Following the same logic as Part 3 (now with $n$ terms instead
of 2), derive $\mathbb{E}[\bar{X}]$ and $\text{Var}(\bar{X})$ in terms of $\mu$,
$\sigma^2$, and $n$.

```{r fig.height=0.3*11}
blank_space()
```

**4b.** Check your formula against $n=5$, $\sigma^2=16$: what number
does $\text{Var}(\bar{X})$ come out to? How does it compare to Part 3b when you
plug in $n=2$?

$\text{Var}(\bar{X}) =$ ________

**4c.** As $n\to\infty$, what happens to $\text{Var}(\bar{X})$? What
does that mean, in plain language, about how precisely you can pin down
$\mu$ with a very large sample?

```{r fig.height=0.3*3}
blank_lines(3)
```

**4d.** One more thing to be careful about: $\text{Var}(\bar{X})$ and
the *standard deviation* of $\bar{X}$ (often called the standard
error) are not the same formula. If $\text{Var}(\bar{X})=\sigma^2/n$,
write down the standard deviation of $\bar{X}$ -- how does it scale
with $n$?

```{r fig.height=0.3*3}
blank_lines(3)
```
