---
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}
```

**NRES 746 -- Week 3, Lecture 2: A Probability Puzzle**

&nbsp;

Name: ___________________________________________

&nbsp;

**Instructions:** Work alone for about 5-8 minutes, guessing before you
compute anything. Then compare with a neighbor.

## Part 1: Sevens vs. doubles

Roll two fair six-sided dice (one red, one blue, so you can tell them
apart).

**1a.** Before doing any counting: which do you think is more likely,
rolling a **sum of 7**, or rolling **doubles** (both dice show the same
number)? Circle one:

> sum of 7 &nbsp;&nbsp;&nbsp;&nbsp; doubles &nbsp;&nbsp;&nbsp;&nbsp; equally likely

**1b.** Now check your guess. First: how many equally-likely outcomes are
there in the full sample space $\Omega$ (all ordered pairs of red/blue
results)?

$\lVert \Omega \rVert =$ ________

**1c.** How many of those outcomes sum to 7? List them if it helps.

$P(\text{sum}=7) =$ ________

**1d.** How many outcomes are doubles?

$P(\text{doubles}) =$ ________

**1e.** Was your guess in 1a correct? If not, what made doubles or
sevens feel more/less likely than it actually is?

```{r fig.height=0.3*2}
blank_lines(2)
```

## Part 2: How many tickets?

A small regional lottery asks you to pick 5 distinct numbers from 1 to
30 (order doesn't matter - {3, 8, 14, 20, 29} is the same ticket as
{29, 20, 14, 8, 3}). How many different tickets are possible?

Leave your answer as an expression (factorials, or $\binom{n}{k}$
notation if you already know it). No need to provide an actual integer value!

*Hint:* how many choices do you have for the first number you write
down? How many for the second, given the first is already used? ... and
then why do you have to divide by something at the end?

```{r fig.height=0.3*6}
blank_space()
```

## Part 3: A teaser (guess only -- no calculation)

Suppose there are 25 people in this room. Do you think it's more likely
than not (better than a 50% chance) that **at least two of them share a
birthday**? Circle your guess:

> yes &nbsp;&nbsp;&nbsp;&nbsp; no

We'll come back to this one once we have the tools to compute it
exactly.
