---
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}
```

**NRES 746 -- Workout of the Day #6: Tracing the Binomial**

&nbsp;

Name: ___________________________________________

&nbsp;

**Instructions:** Work alone for about 5-8 minutes. Then compare with a
neighbor.

## Part 1: Trace the algorithm by hand

`dbinom()` is not magic -- it's just the binomial pmf,
$$P(X=k) = \binom{N}{k}p^k(1-p)^{N-k},$$
evaluated one piece at a time. Trace through it by hand for $N=5$,
$k=2$, $p=0.4$, filling in each step **before** moving to the next.

**Step 1.** Compute $\binom{N}{k} = \binom{5}{2}$ (the number of ways to
choose which 2 of the 5 trials are "successes"):

$\binom{5}{2} =$ ________

**Step 2.** Compute $p^k = 0.4^2$:

$0.4^2 =$ ________

**Step 3.** Compute $(1-p)^{N-k} = 0.6^{3}$:

$0.6^3 =$ ________

**Step 4.** Multiply the three pieces from Steps 1-3 together:

$P(X=2) =$ ________

## Part 2: Find the bug

A student wrote the following R function to compute the binomial pmf
completely from scratch -- no `dbinom()` allowed. It has exactly **one**
bug. Circle it, and write the corrected line below.

```
binom_pmf <- function(N, k, p) {
  choose(N, k) * p^k * (1 - p)^(N - k - 1)
}
```

Corrected line:

```{r fig.height=0.3*2}
blank_space()
```

**2b.** If you ran the buggy version with `binom_pmf(5, 2, 0.4)`, would
it return a number bigger or smaller than your hand-traced answer from
Part 1? How can you tell without running it?

```{r fig.height=0.3*3}
blank_lines(3)
```
