---
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
header-includes:
  - \usepackage{needspace}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# keeps a prompt and its answer space on one PDF page: place `r nq(...)` on its
# own line just above the prompt. Emits nothing in non-LaTeX output. lines =
# lines of prompt text, rows = rows of blank space below it (0.3 in each),
# extra = any other height (in inches), e.g. a figure
nq <- function(lines, rows = 0, extra = 0) {
  if (!knitr::is_latex_output()) return("")
  knitr::asis_output(sprintf("\n\\needspace{%.2fin}\n", 0.23 * lines + 0.15 + 0.3 * rows + extra))
}
```

**NRES 746, Week 7, Lecture 1: Reading the Interval**

Name: ___________________________________________

**Instructions:** Work alone on Part 1 (about 4 minutes), then compare
with a neighbor for Part 2. Keep laptops closed.

## The data

You swab 20 frogs from a pond for chytrid fungus, and 2 test positive.
Model the number of positives as binomial with unknown prevalence $p$.
The figure shows the negative log-likelihood (NLL) for these data,
minus its minimum value (so the lowest point of the curve is at zero).

```{r chytrid-nll, fig.width=6.5, fig.height=3.4}
p_grid <- seq(0.001, 0.4, by = 0.001)
nll <- -dbinom(2, size = 20, prob = p_grid, log = TRUE)
par(mar = c(4, 4.5, 0.5, 0.5))
plot(NA, xlim = c(0, 0.4), ylim = c(0, 5), xaxs = "i", yaxs = "i",
     xlab = "Prevalence, p", ylab = "NLL minus its minimum", xaxt = "n")
axis(1, at = seq(0, 0.4, by = 0.1))
axis(1, at = seq(0, 0.4, by = 0.02), labels = FALSE, tcl = -0.25)
abline(v = seq(0, 0.4, by = 0.02), h = seq(0, 5, by = 0.5), col = "grey85")
lines(p_grid, nll - min(nll), lwd = 2)
box()
```

`r nq(3, 1)`

## Part 1: Read the curve (individual)

**1a.** What is the maximum likelihood estimate of $p$?

$\hat{p} \approx$ ________

`r nq(5, 1)`

**1b.** A common rule of thumb treats every value of $p$ whose NLL is
within 2 units of the minimum as plausible. Draw that cutoff on
the figure, then read off the approximate range of plausible values.

From ________ to ________

`r nq(3, 3)`

**1c.** Is this range *symmetric* around $\hat{p}$? Why might the curve
be lopsided for these data?

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(4, 3)`

## Part 2: More frogs (with a neighbor)

**2a.** A second survey swabs 80 frogs and finds 8 positive. Sketch its
curve on the same figure. Will its range of plausible values be wider
or narrower than yours, and does $\hat{p}$ change? Compare sketches
with your neighbor.

```{r fig.height=0.3*3}
blank_lines(3)
```
