---
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
header-includes:
  - \usepackage{needspace}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}

# fill-in table drawn with base graphics so row height is controllable ("" = blank
# cell); headers/values may be plotmath expressions for math; pair with
# fig.height = header_height + nrow(body) * row_height
blank_table <- function(headers, body, col_widths = NULL,
                         header_height = 0.35, row_height = 0.5,
                         cex = 0.95, bold_rows = integer(0)) {
  nc <- length(headers); nr <- length(body)
  if (is.null(col_widths)) col_widths <- rep(1, nc)
  col_widths <- col_widths / sum(col_widths)
  cx <- c(0, cumsum(col_widths))
  h_total <- header_height + nr * row_height
  par(mar = c(0, 0, 0, 0), family = "serif")
  plot(NA, xlim = c(0, 1), ylim = c(0, h_total), axes = FALSE, xlab = "", ylab = "",
       xaxs = "i", yaxs = "i")
  y <- h_total
  segments(0, y, 1, y, lwd = 1.5)
  y <- y - header_height
  segments(0, y, 1, y, lwd = 1.5)
  for (i in seq_len(nr)) {
    y <- y - row_height
    segments(0, y, 1, y, lwd = if (i == nr) 1.5 else 0.5,
             col = if (i == nr) "black" else "grey70")
  }
  for (v in cx) segments(v, 0, v, h_total, lwd = 1)
  yh <- h_total - header_height / 2
  for (j in seq_len(nc)) text((cx[j] + cx[j + 1]) / 2, yh, labels = headers[[j]], font = 2, cex = cex)
  y <- h_total - header_height
  for (i in seq_len(nr)) {
    yc <- y - row_height / 2
    fnt <- if (i %in% bold_rows) 2 else 1
    for (j in seq_len(nc)) {
      val <- body[[i]][[j]]
      if (!identical(val, "")) text((cx[j] + cx[j + 1]) / 2, yc, labels = val, font = fnt, cex = cex)
    }
    y <- y - row_height
  }
}

# keeps a prompt and its answer space on one PDF page: place `r nq(...)` on its
# own line just above the prompt. Emits nothing in non-LaTeX output. lines =
# lines of prompt text, rows = rows of blank space below it (0.3 in each),
# extra = any other height (in inches), e.g. a table
nq <- function(lines, rows = 0, extra = 0) {
  if (!knitr::is_latex_output()) return("")
  knitr::asis_output(sprintf("\n\\needspace{%.2fin}\n", 0.23 * lines + 0.15 + 0.3 * rows + extra))
}
```

**NRES 746, Week 8, Lecture 1: Newton's Method by Hand (take-home)**

Name: ___________________________________________

**Bring this to class on Wednesday.** We'll start by going over it, and
Wednesday's lecture picks up where it leaves off. Work on paper; a
calculator is all you need.

## Putting it in context

Since Week 6 we've found MLEs with `optim()` without asking how it
finds them. On Wednesday we open the black box. This WOD works through
the core of one method, Newton's method, on a problem small enough to
do by hand.

The data are the Week 7 pika haypile counts from six talus plots: 2, 5,
3, 0, 4, 4 (sum 18), modeled as Poisson with mean $\lambda$. Dropping
terms that don't involve $\lambda$,

$$\text{NLL}(\lambda) = 6\lambda - 18\log\lambda$$

`r nq(3, 12)`

## Part 1: Slope and curvature

**1a.** Find $\text{NLL}'(\lambda)$ and $\text{NLL}''(\lambda)$. Then
set $\text{NLL}'(\lambda) = 0$ and solve for the MLE, $\hat\lambda$.

```{r fig.height=0.3*12}
blank_space()
```

`r nq(10, 2)`

## Part 2: Slopes without calculus

`optim()` can't do calculus. It estimates slopes from NLL values at
nearby points, using a **finite difference**: for a small step $h$,

$$f'(x) \approx \frac{f(x + h) - f(x)}{h}$$

Here are NLL values at three points spaced $h = 0.5$ apart:

| $\lambda$ | 1.5 | 2.0 | 2.5 |
|:---:|:---:|:---:|:---:|
| $\text{NLL}(\lambda)$ | 1.7016 | $-0.4766$ | $-1.4932$ |

**2a.** Estimate $\text{NLL}'(2)$ two ways: with the formula above
(using $\lambda = 2$ and 2.5), and with a centered version that uses
$\lambda = 1.5$ and 2.5. Compare both with the exact value from 1a.
Which is closer?

```{r fig.height=0.3*5}
blank_space()
```

Forward: ________ \qquad Centered: ________ \qquad Exact: ________

`r nq(4, 14)`

**2b.** Curvature is the slope of the slope. Use that idea to build a
finite-difference approximation to $f''(x)$ that uses only
$f(x - h)$, $f(x)$, and $f(x + h)$. Then use it to estimate
$\text{NLL}''(2)$ and compare with the exact value.

```{r fig.height=0.3*14}
blank_space()
```

Estimate: ________ \qquad Exact: ________

`r nq(9, 3)`

## Part 3: Newton's method

Newton's method looks for the point where $\text{NLL}'(\lambda) = 0$.
At the current guess, it replaces the NLL with a parabola that has the
same slope and curvature, then jumps to the bottom of that parabola:

$$\lambda_{\text{new}} = \lambda - \frac{\text{NLL}'(\lambda)}{\text{NLL}''(\lambda)}$$

**3a.** In words: what does dividing by the curvature do to the size of
the step? When will Newton's method take big steps, and when small
ones?

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(3, 0, extra = 0.35 + 3 * 0.55 + 0.3)`

**3b.** Start at $\lambda = 2$ and take three Newton steps, using your
exact derivatives from 1a. Keep at least three decimal places.

```{r newton-table, fig.width=6.2, fig.height=0.35+3*0.55}
blank_table(
  headers = list("Step", "Guess", "Slope, NLL'", "Curvature, NLL''", "Next guess"),
  body = list(
    list("1", "2", "", "", ""),
    list("2", "", "", "", ""),
    list("3", "", "", "", "")
  ),
  col_widths = c(0.6, 1, 1.2, 1.4, 1.1),
  row_height = 0.55
)
```

```{r fig.height=0.3*4}
blank_space()
```

`r nq(5, 10)`

**3c.** Now start at $\lambda = 7$ and take one step. What went wrong?
Sketch the NLL curve and the parabola Newton's method used at
$\lambda = 7$ to explain why, and suggest one way to fix it.

```{r fig.height=0.3*10}
blank_space()
```

```{r fig.height=0.3*3}
blank_lines(3)
```

`r nq(5, 5)`

**3d.** *(If you have time.)* `optim()`'s derivative-based methods
don't know the exact derivatives. Take one Newton step from
$\lambda = 2$ using your finite-difference estimates from 2a (centered)
and 2b instead. How close is it to your first step in 3b?

```{r fig.height=0.3*5}
blank_space()
```
