---
output:
  pdf_document: default
  # word_document: default   # uncomment (and comment out pdf_document) to knit to Word instead
header-includes:
  - \usepackage{needspace}
knit: (function(input, encoding) { rmarkdown::render(input, encoding = encoding, output_dir = dirname(input)) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 6.5, fig.height = 0.3)

# draws n full-width blank lines (margin to margin) for handwritten answers,
# as a single plot so spacing is deterministic; pair with fig.height = 0.3 * n
blank_lines <- function(n = 1) {
  par(mar = c(0, 0, 0, 0))
  plot(NA, xlim = c(0, 1), ylim = c(0, n), axes = FALSE, xlab = "", ylab = "")
  for (i in seq_len(n)) {
    segments(0, n - i + 0.5, 1, n - i + 0.5, lwd = 1)
  }
}

# reserves blank work space (no ruled lines) for equation manipulation /
# show-your-work; pair with fig.height set to however much room is needed
blank_space <- function() {
  par(mar = c(0, 0, 0, 0))
  plot.new()
}

# keeps a prompt and its answer space on one PDF page: place `r nq(...)` on its
# own line just above the prompt. Emits nothing in non-LaTeX output. lines =
# lines of prompt text, rows = rows of blank space below it (0.3 in each),
# extra = any other height (in inches), e.g. a code block
nq <- function(lines, rows = 0, extra = 0) {
  if (!knitr::is_latex_output()) return("")
  knitr::asis_output(sprintf("\n\\needspace{%.2fin}\n", 0.23 * lines + 0.15 + 0.3 * rows + extra))
}
```

**NRES 746, Week 7, Lecture 2: Curvature by Hand (take-home)**

Name: ___________________________________________

**Due at the start of class on Monday.** Try Parts 1 and 2 on paper
before you open a laptop. Part 3 uses R.

## Putting it in context

On Monday we saw that the curvature of the negative log-likelihood
(NLL) at its minimum measures how much the data tell us about a
parameter. For a parameter $\theta$ with MLE $\hat\theta$:

$$\text{SE}(\hat\theta) \approx \frac{1}{\sqrt{\text{NLL}''(\hat\theta)}}
\qquad \text{approximate 95\% interval: } \hat\theta \pm 1.96\,\text{SE}$$

`optim(..., hessian = TRUE)` estimates $\text{NLL}''(\hat\theta)$
numerically. Here you find it with calculus, for two models you already
know.

`r nq(8, 10)`

## Part 1: Pika haypiles (Poisson)

Monday's pika data: active haypiles in six talus plots, 2, 5, 3, 0, 4,
4, modeled as Poisson with mean $\lambda$. Dropping terms that don't
involve $\lambda$,

$$\text{NLL}(\lambda) = n\lambda - \left(\sum_{i=1}^{n} y_i\right)\log\lambda$$

**1a.** Find $\text{NLL}'(\lambda)$ and $\text{NLL}''(\lambda)$.

```{r fig.height=0.3*10}
blank_space()
```

`r nq(4, 6)`

**1b.** Evaluate $\text{NLL}''(\hat\lambda)$ at $\hat\lambda = 3$ for
the six plots, and again for 24 plots with the same mean (each count
appearing four times). Give the SE for each.

```{r fig.height=0.3*5}
blank_space()
```

6 plots: $\text{NLL}''(\hat\lambda) =$ ______, SE = ______ \qquad
24 plots: $\text{NLL}''(\hat\lambda) =$ ______, SE = ______

`r nq(4, 8)`

**1c.** Substitute $\hat\lambda = \bar{y} = \sum y_i / n$ into your
$\text{NLL}''$ and simplify to get a general formula for
$\text{SE}(\hat\lambda)$ in terms of $\hat\lambda$ and $n$. What happens
to the SE when $n$ quadruples?

```{r fig.height=0.3*8}
blank_space()
```

`r nq(8, 10)`

## Part 2: Frog detection (binomial)

In Lab 5, Act 1, you pooled three wetlands and heard the frog on $K = 12$
of $N = 30$ visits, so $\hat p = 0.4$. Dropping the binomial
coefficients,

$$\text{NLL}(p) = -K\log p - (N - K)\log(1 - p)$$

**2a.** Find $\text{NLL}''(p)$. (Watch the chain rule on $\log(1-p)$.)

```{r fig.height=0.3*10}
blank_space()
```

`r nq(5, 6)`

**2b.** Evaluate $\text{NLL}''(\hat p)$ and the SE for your survey
($K = 12$, $N = 30$) and for your colleague's survey ($K = 120$,
$N = 300$). Check: one of your curvatures should be 125.

```{r fig.height=0.3*5}
blank_space()
```

3 wetlands: SE = ______ \qquad 30 wetlands: SE = ______

`r nq(4, 10)`

**2c.** Substitute $\hat p = K/N$ and simplify to get a general formula
for $\text{SE}(\hat p)$. Have you seen this formula before?

```{r fig.height=0.3*8}
blank_space()
```

```{r fig.height=0.3*2}
blank_lines(2)
```

`r nq(6, 5)`

**2d.** Monday's WOD had 2 chytrid-positive frogs out of 20. Use your
formula from 2c to build the approximate 95% interval,
$\hat p \pm 1.96\,\text{SE}$. What's wrong with it? Compare it with
the range you read off the NLL curve on Monday.

Interval: from ________ to ________

```{r fig.height=0.3*4}
blank_lines(4)
```

`r nq(8, 1)`

## Part 3: Check it in R

**Using AI:** the coding question below is labeled with a
recommendation. After it, circle how you actually used AI. Using AI
never costs points; only undisclosed use does.

**What to turn in:** this handout, with the value you got in 3a.

**3a.** *(Recommended: no AI.)* Write an NLL function for the frog data
in 2b ($K = 12$, $N = 30$) using `dbinom()`, and minimize it with
`optim(..., method = "Brent", lower = 0.001, upper = 0.999, hessian = TRUE)`.
Does the Hessian match your answer to 2b?

`optim()` Hessian: ________

AI use (circle one): no AI / AI helped (e.g., debugging) / AI wrote the code

`r nq(3, 4)`

**3b.** Where did you use AI, if anywhere, and what did you have to
check or fix? If you didn't use it, what was hardest?

```{r fig.height=0.3*4}
blank_lines(4)
```
